---
title: "Aging Gene Sets"
author: Anna L Tyler
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  html_document:
    code_folding: hide
    collapsed: no
    toc: yes
    toc_float: yes
---

## Introduction

The purpose of this workflow is to check whether 
specific gene sets identified in 1a.KEGG_and_Biodomains.Rmd
have differential expression across age groups.

```{r param}
rm(list = ls())
library(here)

args = commandArgs(trailingOnly=TRUE)
gene.set.name <- args[1]

if(is.na(gene.set.name)){
    gene.set.name <- "Glutamate_Genes"
    #gene.set.name <- "Parkinson_disease-Mitochondrial_Metabolism_Genes"
    #gene.set.name <- "Ribosome-Immune_Response_Genes"
    #gene.set.name <- "Ribosome-Synapse_Genes"
    #gene.set.name <- "Gastric_acid_secretion-Mitochondrial_Metabolism_Genes"
}
gene.set.text <- gsub("_", " ", gene.set.name)

results.dir <- here("Results", "age_batch_12_gene")
dot.col <- "#dd1c77"

```

```{r load_code}
all.fun <- list.files(here("Code"), full.names = TRUE)
for(i in 1:length(all.fun)){source(all.fun[i])}

general.data.dir <- here("Data", "general")
group.processed.data.dir <- here("Results", "all", "processed_data")
processed.data.dir <- here("Results", "all_gene", "processed_data")
```

```{r libraries, message = FALSE, warning = FALSE, error = FALSE}
needed.libraries <- c("pheatmap", "DT", "gprofiler2", "cluster", "vioplot")
load_libraries(needed.libraries)
```


```{r info}
#gene information tables
gene.info <- read.table(file.path(general.data.dir, "mouse_gene_info.txt"), sep = "\t", header = TRUE)
scaled.expr <- readRDS(file.path(processed.data.dir, "Final_Adjusted_Expr.RDS"))
covar.table <- read.csv(file.path(processed.data.dir, "mouse_info_all_geno.csv"))

gene.set <- read.delim(here("Results", "age_batch_12_gene", "validation_gene_sets", 
    paste0(gene.set.name, ".txt")))
gene.id <- gene.info[match(gene.set[,1], gene.info[,"external_gene_name"]),"ensembl_gene_id"]
names(gene.id) <- gene.set[,1]
```

## Individual Gene Distributions by Age

```{r zero_line}
#adds zero line to grouped box plot.
add_zero_line <- function(){
  plot.dim <- par("usr")
  segments(plot.dim[1], 0, plot.dim[2], 0)
}
```

## All expression

The boxplot below shows expression for all genes and 
all individuals at both ages.

```{r gene_set, fig.width = 5, fig.height = 5}
gene.idx <- match(gene.id, colnames(scaled.expr))
sub.expr <- scaled.expr[,gene.idx]
colnames(sub.expr) <- names(gene.id)

u_age <- c(4, 12)
age.id <- lapply(u_age, 
    function(x) rownames(covar.table)[which(covar.table[,"age_batch"] == x)])

binned.expr <- lapply(age.id, function(x) sub.expr[x,])
names(binned.expr) <- u_age

boxplot(binned.expr, ylim = c(min(sub.expr), max(sub.expr)+0.5),
    ylab = "Scaled Expression")
diff.test <- pairwise_test(binned.expr)
add_pairwise_results(binned.expr, diff.test, y.nudge = 0.02)
```

## Boxes by Gene

The boxplot below splits up expression by gene. 

```{r by_gene, fig.width = 12, fig.height = 6}
means.by.age <- sapply(binned.expr, colMeans)
mean.diff.order <- order(means.by.age[,2]-means.by.age[,1], decreasing = TRUE)
ordered.expr <- lapply(binned.expr, function(x) x[,mean.diff.order])
plot.grouped.boxes(ordered.expr, type = "matrix", plot.type = "vioplot")
add_zero_line()
```

## Median Expression by Age

The violin plot shows the median expression for each
gene at different ages.

```{r, mean_diff}
#mean.expr <- lapply(binned.expr, colMeans)
mean.expr <- lapply(binned.expr, function(x) apply(x, 2, median))
vioplot(mean.expr, col = "lightgray", ylim = c(min(unlist(mean.expr)), 
    max(unlist(mean.expr))+0.02),
    main = paste(gene.set.text, "expression with age"))
mtext("Mean Expression", side = 2, line = 2.5)
stripchart(mean.expr, vertical = TRUE, pch = 16, method = "jitter", 
    col = dot.col, add = TRUE)
abline(h = 0)

test.diff <- pairwise_test(mean.expr)
add_pairwise_results(mean.expr, test.diff, y.nudge = 0.02)
```

## Median Expression by Age with Individual Genes

The plot below shows the same as above but with 
the median trajectories of individual genes tracked.

```{r mean_detail, fig.width = 7, fig.height = 7}
#mean.table <- sapply(binned.expr, colMeans)
mean.table <- sapply(binned.expr, function(x) apply(x, 2, median))

plot.new()
plot.window(xlim = c(0.5,2.5), ylim = c(min(mean.table)-0.02, max(mean.table)+0.01))
vioplot(mean.expr, col = "lightgray", add = TRUE, axes = FALSE)
plot.dim <- par("usr")
draw.rectangle(plot.dim[1], plot.dim[2], plot.dim[3], plot.dim[4], 
    border.col = NA, fill = rgb(1,1,1, alpha = 0.9))
for(i in 1:length(gene.set[,1])){
    points(1:ncol(mean.table), mean.table[i,], type = "b", col = i, pch = 16,
        lwd = 2)
    text(x = ncol(mean.table)+0.02, y = mean.table[i,ncol(mean.table)],
        labels = rownames(mean.table)[i], adj = 0, col = i, font = 2)
    text(x = 1-0.02, y = mean.table[i,1],
        labels = rownames(mean.table)[i], adj = 1, col = i, font = 2)
}
axis(2)
abline(h = 0)
mtext("Mean Expression", side = 2, line = 2.5, cex = 1.5)
mtext("Individual Gene Trajectories", side = 3, line = 2.5, font = 2, cex = 1.2)
text(x = 1:ncol(mean.table), y = min(mean.table)-0.02, 
    labels = paste(colnames(mean.table), "months"), adj = 0.5, cex = 1.5)
add_pairwise_results(mean.expr, test.diff, y.nudge = 0.02)

```
