---
title: "Comparison to Kl KO"
author: Anna L Tyler
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  html_document:
    code_folding: hide
    collapsed: no
    toc: yes
    toc_float: yes
---

## Introduction

One of the reviewers of the Klotho manuscript pointed out a paper
we hadn't noticed before. This paper compared brain transcriptomics
between WT and Kl knockout mice (pmid: 38862813). 

They report their DEG results in the supplemental files.
I downloaded these results to compare to ours. They have 
some additional stats in this file about whether the genes
changed in the mouse AD proteom, changed in the human AD
proteome and changed in specific types of AD models.


```{r param}
rm(list = ls())
library(here)
```

```{r load_code}
all.fun <- list.files(here("Code"), full.names = TRUE)
for(i in 1:length(all.fun)){source(all.fun[i])}
```


```{r libraries, message = FALSE, warning = FALSE, error = FALSE}
needed.libraries <- c("pheatmap", "vioplot", "RColorBrewer", "igraph")
load_libraries(needed.libraries)
```

```{r load_data}
ko_deg <- read.csv(here("Data", "Mouse", "DEG_from_Kl_KO_pmid38862813.csv"))
var_deg <- read.delim(here("Results", "for_paper", "Fig_2B_Data.txt"), comment.char = "#")
```

First, we looked at whether the VS haplotype effect size was
correlated with the log2FC from the KO model. Nope.

```{r common_gene}
common.gene <- intersect(ko_deg[,1], var_deg[,"geneName"])
matched.ko <- ko_deg[match(common.gene, ko_deg[,1]),]
matched.var <- var_deg[match(common.gene, var_deg[,"geneName"]),]

plot(matched.ko[,"log2FoldChange"], matched.var[,"genotype_beta"], pch = 16,
  xlab = "KO log2 Fold Change", ylab = "Haplotype effect size", xlim = c(-0.6, 0.6))
abline(v = 0, h = 0, col = "red")

coord.mat <- cbind(matched.ko[,"log2FoldChange"], matched.var[,"genotype_beta"])
rownames(coord.mat) <- matched.ko[,1]
num.layers = 3
hull.genes <- vector(mode = "list")
for(l in 1:num.layers){
  hull <- convex_hull(coord.mat)
  hull.genes[[l]] <- rownames(coord.mat)[hull$resverts]
  text(hull$rescoords[,1], hull$rescoords[,2], labels = rownames(coord.mat)[hull$resverts], pos = 4, cex = 0.5, col = "red")
  coord.mat <- coord.mat[-hull$resverts,]
}

```