---
title: "Klotho Alamar"
author: Anna L Tyler
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  html_document:
    code_folding: hide
    collapsed: no
    toc: yes
    toc_float: yes
---

The purpose of this workflow is to investigate the Alamar
data we got recently looking at abundance of 120 proteins
related to neurological function and disease.

The mice in this group are all the same age (12 months), 
They have humanized Klotho haplotype in the following genotype
combinations: KL-FC, KL-VS/FC and KL-VS.
All mice have humanized APOE4. 

So, the question here, is whether mice with the APOE4 allele
and the VS allele have a less AD-like profile than the APOE4 mice
with just the FC allele. And maybe the VS alone look worse than the
homs?

We cannot yet compare to other age groups or to APOE carriers,
but these data will come later.

The proteins in these sets are annotated in groups by Alamar, and
we can use those groups to look at specific subsets of proteins.

```{r param}
rm(list = ls())
library(here)
```

```{r load_code}
all.fun <- list.files(here("Code"), full.names = TRUE)
for(i in 1:length(all.fun)){source(all.fun[i])}
```

```{r libraries, message = FALSE, warning = FALSE, error = FALSE}
needed.libraries <- c("pheatmap", "vioplot", "RColorBrewer",
  "gprofiler2", "readxl", "stringr")
load_libraries(needed.libraries)
```

```{r load_data}
data.dir <- here("Data", "Mouse", "Alamar", "20260406_Jax_mousePLA_NULISA_Analysis_Software_Report")
processed.data.dir <- here("Results", "Alamar", "processed_data")
if(!file.exists(processed.data.dir)){dir.create(processed.data.dir, recursive = TRUE)}

annot <- as.matrix(read_excel(here("Data", "Mouse", "Alamar", "Alamar Nulisa Mouse 120 annotations.xlsx")))
#names that do not have a 1-to-1 correspondence in the data and the annotations
tricky.key <- as.matrix(read.delim(here("Data", "Mouse", "Alamar", "tricky_names.txt")))
norm.counts <- as.matrix(read_excel(file.path(data.dir, "NPQ_Values.xlsx"), sheet = 2))

#subset to KL mice
kl.idx <- grep("Klotho", colnames(norm.counts))
kl.mat <- apply(norm.counts[,kl.idx], 2, as.numeric)

#there are gene names and protein (target) names. We need to make
#sure  that the names being used match between the annotation and
#the abudance matrix.
data.names <- gsub("-", "", str_to_lower(norm.counts[,1]))
annot.gene <- gsub("-", "", str_to_lower(annot[,"Gene name"]))
annot.target <- gsub("-", "", str_to_lower(annot[,"Target"]))

#these are only mostly aligned. It looks as if Mapt (total tau)
#is causing some problems. Everything between Mapt in data.names
#and "Mapt" in annot.gene is misalinged. Awesome. Let's rearrange
#the data to match the annotation. The Tnfs are also a problem...
#After going in lots of circles, I've decided to make my own key
#by hand that aligns the trickier protein and gene names. The vast
#majority are fine, but some are really tricky.
check.table <- cbind(data.names, annot.gene, annot.target)

data.gene.match <- sapply(data.names, function(x) which(annot.gene == x))
data.target.match <- sapply(data.names, function(x) which(annot.target == x))
match.mat <- cbind(sapply(data.gene.match, length), sapply(data.target.match, length))
tricky.match <- Reduce("intersect", apply(match.mat, 2, function(x) which(x != 1)))
#names with no unique mapping. These go in the table made by hand.
#data.names[tricky.match]

#rename all rows in the data matrix using the target column in the annotation table
usable.names <- rep(NA, length(data.names))
for(i in 1:length(data.names)){
  if(match.mat[i,2] == 1 || match.mat[i,1] == 1){
    gene.annot.idx <- data.gene.match[[i]]
    gene.target.idx <- data.target.match[[i]]
    gene.name.idx <- unique(union(gene.annot.idx, gene.target.idx))
    if(length(gene.name.idx) > 1){
      gene.name.idx <- data.target.match[[i]] #happens for Snca
    }
    if(length(gene.name.idx) != 1){stop()}
    usable.names[i] <- str_to_title(annot.target[gene.name.idx])
  }else{
    tricky.idx <- which(str_to_lower(tricky.key[,1]) == data.names[i])
    usable.names[i]  <- tricky.key[tricky.idx,"target"]
  }
}
#cbind(data.names, usable.names)
rownames(kl.mat) <- usable.names
saveRDS(kl.mat, file.path(processed.data.dir, "Alamar_Proteins.RDS"))


kl.split <- strsplit(colnames(kl.mat), " ")
mouse.info <- Reduce("rbind", kl.split)
colnames(mouse.info) <- c("genotype", "ID", "sex")
mouse.info[,"genotype"] <- gsub("Klotho", "", gsub(".APOE4.hAb", "", mouse.info[,"genotype"]))

geno.cols <- c("FC" = "#31a354", "FC/VS" = "gray", "VS" = "#2b8cbe")
geno.col <- sapply(mouse.info[,"genotype"], function(x) geno.cols[which(names(geno.cols) == x)])
sex.cols = c("M" = "#998ec3", "F" = "#d8b365")
sex.col <- sapply(mouse.info[,"sex"], function(x) sex.cols[which(names(sex.cols) == x)])
```

## Individuals in PC space

```{r decomp, fig.width = 8, fig.height = 4}
par(mfrow = c(1,2))
plot.decomp(t(kl.mat), cols = sex.col, main = "Colored by sex")
legend("bottomright", legend = names(sex.cols), col = sex.cols, pch = 16)

plot.decomp(t(kl.mat), cols = geno.col, main = "Colored by genotype")
legend("bottomright", legend = names(geno.cols), col = geno.cols, pch = 16)
```

Adjusted for sex:

```{r adjust, fig.width = 8, fig.height = 4}
adj.kl <- adjust(t(kl.mat), dummy_covar(mouse.info[,"sex",drop=FALSE]))

par(mfrow = c(1,2))
plot.decomp(adj.kl, cols = sex.col, main = "Colored by sex")
legend("bottomright", legend = names(sex.cols), col = sex.cols, pch = 16)

plot.decomp(adj.kl, cols = geno.col, main = "Colored by genotype")
legend("bottomright", legend = names(geno.cols), col = geno.cols, pch = 16)
```

## ANOVA results

We performed an ANOVA to test for effects of genotype in the
sex-adjusted proteins. The qq plot for the p values is below.

```{r model, fig.width = 5, fig.height = 5}
geno.test <- apply(adj.kl, 2, function(x) aov(x~as.factor(mouse.info[,"genotype"])))
geno.p <- sapply(geno.test, function(x) summary(x)[[1]][1,"Pr(>F)"])
qqunif.plot(pvalues = geno.p, label.above = 2.55, xlim = c(0, 5.5), label.cex = 0.8)
```

## Significant proteins

```{r sig}
fdr <- 0.1
adj.p <- p.adjust(geno.p, "fdr")
sig.idx <- which(adj.p <= fdr)
```

There are multiple expression patterns across
the genotypes. Proteins significant below fdr = `r fdr`
are shown below.


```{r plot_one_protein, fig.width = 8, fig.height = 4}
plot_one_protein <- function(pname, k.mat){
  p.idx <- which(colnames(k.mat) == pname)
  boxplot(k.mat[,p.idx]~mouse.info[,"genotype"], col = geno.cols, xlab = "Genotype",
    ylab = pname, main = pname)
}

par(mfrow = c(1,2))
for(i in 1:length(sig.idx)){
  plot_one_protein(names(sig.idx)[i], adj.kl)
}
```

## Median abundance by genotype

We calculated the (scaled) median expression of each protein across
genotypes. All medians are shown below.

```{r geno_means, fig.width = 3, fig.height = 13}

geno.idx <- lapply(names(geno.cols), function(x) which(mouse.info[,"genotype"] == x))
names(geno.idx) <- names(geno.cols)

scaled.kl <- apply(adj.kl, 2, scale)
rownames(scaled.kl) <- mouse.info[,"ID"]

geno.kl <- lapply(geno.idx, function(x) scaled.kl[x,])
mean.by.geno <- sapply(geno.kl, function(x) apply(x, 2, median))

pheatmap(mean.by.geno, cluster_cols = FALSE)
```

The heat map below shows protein abundance for individual mice
ordered by genotype.

```{r individuals, fig.width = 7, fig.height = 13}
geno.mat <- t(Reduce("rbind", geno.kl))
geno.annot <- data.frame("genotype" = unlist(lapply(1:length(geno.kl), function(x) rep(names(geno.cols)[x], nrow(geno.kl[[x]])))))
rownames(geno.annot) <- colnames(geno.mat)
pheatmap(geno.mat, cluster_cols = FALSE, annotation_col = geno.annot, annotation_colors = list("genotype" = geno.cols))

#geno.decomp <- plot.decomp(geno.mat)
#ab.order <- order(geno.decomp$u[,1])
#pheatmap(geno.mat[ab.order,], cluster_rows = FALSE, cluster_cols = FALSE, annotation_col = geno.annot, annotation_colors = list("genotype" = geno.cols))

#pheatmap(geno.mat[which(adj.p <= fdr),], cluster_cols = FALSE, annotation_col = geno.annot, annotation_colors = list("genotype" = geno.cols))
```

The plot below shows only those proteins with a genotype fdr
less than `r fdr`. 

```{r just_sig, fig.width = 3, fig.height = 4}
pheatmap(mean.by.geno[which(adj.p <= fdr),], cluster_cols = FALSE)
```

## Functional groups {.tabset .tabset-fade .tabset-pills}

```{r plot_protein_group}
plot_protein_group <- function(pnames, k.mat, plot.label = ""){
  layout.mat <- matrix(c(1,1,1,2,3,0), nrow = 2, byrow = TRUE)

  layout(layout.mat)

  p.idx <- match(str_to_lower(pnames), str_to_lower(colnames(k.mat)))
  na.idx <- which(is.na(p.idx))
  #print(pnames[na.idx])

  p.idx <- p.idx[which(!is.na(p.idx))]
  by.geno <- lapply(geno.idx, function(x) k.mat[x,p.idx])
  plot.grouped.boxes(by.geno, type = "matrix", group.cols = geno.cols)
  add_zero_line()
  mtext(plot.label, side = 3, outer = TRUE, line = -1.5)

  boxplot(by.geno, col = geno.cols, main = "All values",
    ylab = "scaled protein abundance")
  add_zero_line()

  group.mean <- sapply(by.geno, function(x) apply(x, 2, median))
  #pheatmap(group.mean)
  boxplot(group.mean, main = "Group Medians", col = geno.cols, ylab = "scaled protein abundance")
  add_zero_line()
}
```


```{r plot_groups, results = "asis", fig.width = 10, fig.height = 7}
pgroups <- unique(annot[,"Function/Classification"])
group.genes <- lapply(pgroups, function(x) annot[which(annot[,"Function/Classification"] == x),"Target"])
names(group.genes) <- pgroups

for(i in 1:length(pgroups)){
  cat("###", pgroups[i], "\n")
  plot_protein_group(pnames = group.genes[[i]], k.mat = scaled.kl, plot.label = pgroups[i])
  cat("\n\n")
}
```

## Conclusions

The humanized Klotho haplotypes do significantly affect
protein abundance in the mouse brain in the presence of the
humanized APOE4 allele.

Consistent with the hypothesis that the VS allele is protevctive
in the heterozygous state, but harmful in the homozygous state,
the haplotype effects are non-additive. 

Epo and Pgf have higher expression in the FC/VS hets than the other
two genotypes. This pattern could suggest that VS hets have
a better blood/oxygen supply in their brains than other genotypes.

The VS homozygotes also have significantly higher phosphorylated
alpha-synuclein than either of the other genotypes, suggesting 
one possible detrimental effect of the VS homozygous state.

Other haplotype effects, however are closer to additive. 
Flt3l and Mif both increase with additional VS alleles.
Cxcl2 decreases with additional VS alleles, although, the
effect is more than additive.