---
title: "Alamar - Frailty"
author: Anna L Tyler
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  html_document:
    code_folding: hide
    collapsed: no
    toc: yes
    toc_float: yes
---

The purpose of this workflow is to align the Alamar data
with frailty data measured at 12 months. We want to try
HDMA with discrete input variables, anyway, and this might
be a good test for it. We have three different genotypes,
and two sexes. Maybe here we can relate those features to
frailty measurements through the proteins measured by Alamar.

The mice in this group are all the same age (12 months), 
They have humanized Klotho haplotype in the following genotype
combinations: KL-FC, KL-VS/FC and KL-VS.
All mice have humanized APOE4. 


```{r param}
rm(list = ls())
library(here)
```

```{r load_code}
all.fun <- list.files(here("Code"), full.names = TRUE)
for(i in 1:length(all.fun)){source(all.fun[i])}
```

```{r libraries, message = FALSE, warning = FALSE, error = FALSE}
needed.libraries <- c("pheatmap", "vioplot", "RColorBrewer",
  "gprofiler2", "readxl", "stringr")
load_libraries(needed.libraries)
```

```{r load_data}
frailty.dir <- here("Data", "Mouse", "Alamar", "frailty")

#read in Alamar data matrix created in 2.Alamar.Rmd
kl.mat <- readRDS(here("Results", "Alamar", "processed_data", "Alamar_proteins.RDS"))


kl.split <- strsplit(colnames(kl.mat), " ")
mouse.info <- Reduce("rbind", kl.split)
colnames(mouse.info) <- c("genotype", "ID", "sex")
mouse.info[,"genotype"] <- gsub("Klotho", "", gsub(".APOE4.hAb", "", mouse.info[,"genotype"]))

geno.cols <- c("FC" = "#31a354", "FC/VS" = "gray", "VS" = "#2b8cbe")
geno.col <- sapply(mouse.info[,"genotype"], function(x) geno.cols[which(names(geno.cols) == x)])
sex.cols = c("M" = "#998ec3", "F" = "#d8b365")
sex.col <- sapply(mouse.info[,"sex"], function(x) sex.cols[which(names(sex.cols) == x)])
```

The frailty data take quite a bit of processing...

```{r frailty_metadata}
all.frailty.files <- list.files(file.path(frailty.dir), full.names = TRUE)
all.id <- lapply(all.frailty.files, read_excel)

#for each of the mice with Alamar data, identify the 
#file location and the blind id.
frailty.info.table <- matrix(NA, nrow = nrow(mouse.info), ncol = 5)
rownames(frailty.info.table) <- mouse.info[,"ID"]
for(i in 1:nrow(mouse.info)){
  mouse.id <- mouse.info[i,"ID"]
  check.sheets <- lapply(all.id, function(x) which(x$"Mouse ID" == mouse.id))
  sheet.idx <- which(sapply(check.sheets, length) > 0)
  if(length(sheet.idx) == 0){next()}
  mouse.idx <- check.sheets[[sheet.idx]]
  in_sheet <- all.frailty.files[sheet.idx]
  id.info <- all.id[[sheet.idx]][mouse.idx,c("Blind ID (1, 2, 3…)", "Mouse ID", "Sex", "Weight")]
  all.info <- unlist(c(id.info, "sheet" = sheet.idx))
  frailty.info.table[i,] <- all.info
}
colnames(frailty.info.table) <- names(all.info)

#cbind(frailty.info.table, mouse.info)
num.mice <- length(which(!is.na(frailty.info.table[,1])))
```

I could identify `r num.mice` out of `r nrow(mouse.info)` that
had frailty data.

```{r info_stats}
all.info.table <- cbind(mouse.info[,c(2,3,1)], frailty.info.table[,c(4,5,1)])
colnames(all.info.table)[6] <- "blind_ID"

#remove the mice with no frailty data
not.na <- which(!is.na(all.info.table[,"Weight"]))
mouse.info.table <- all.info.table[not.na,]

count.mat <- table(data.frame(mouse.info.table[,c("sex", "genotype")]))
```

The mice break into the following groups. Not too bad.

```{r counts}
print(count.mat)
```

Now we will use the sheet numbers and blind IDs to pull 
out frailty data for each mouse. It's going to be too
annoying to pull out the numeric key for each label,
but I will say here that they are all on a scale of
0, 0.5, 1 with 0 being normal, 0.5 having a modest
abnormality, and 1 being very abnormal. Details are
in the excel spread sheets.

```{r frailty_data}
#it is very difficult to parse these sheets. They have varying
#placement and grouping of rows.
all.frailty <- lapply(all.frailty.files, function(x) read_excel(x, sheet = 2))

line.check <- 1-apply(all.frailty[[1]][,1:4], 2, function(x) as.numeric(is.na(x)))
take.rows <- which(rowSums(line.check) >= 3)

pheno.label <- as.character(unlist(all.frailty[[1]][take.rows,1]))
#the NA is a place where the labels for one phenotype got put on different rows
pheno.label[which(is.na(pheno.label))] <- "vaginal/uterine/penile prolapse"
assessment <- as.character(unlist(all.frailty[[1]][take.rows,2]))


#cbind(pheno.label, assessment)

frailty.mat <- matrix(NA, nrow = nrow(mouse.info.table), ncol = length(pheno.label))
rownames(frailty.mat) <- mouse.info.table[,1]
colnames(frailty.mat) <- pheno.label

for(i in 1:nrow(mouse.info.table)){
  sheet.num <- mouse.info.table[i,"sheet"]
  if(is.na(sheet.num)){next()}
  mouse.id <- mouse.info.table[i,"blind_ID"]
  frailty.sheet <- all.frailty[[as.numeric(sheet.num)]]
  mouse.idx <- which(colnames(frailty.sheet) == mouse.id)
  data.col <- unlist(frailty.sheet[,mouse.idx])
  mouse.frailty.info <- data.col[take.rows]
  frailty.mat[i,] <- mouse.frailty.info
}

#separate out the weight and the temperature because these are
#on a different scale than everything else.
weight.temp.idx <- which(colnames(frailty.mat) %in% c("Weight", "Temperature"))
weight.temp <- frailty.mat[,weight.temp.idx]
just.frailty <- frailty.mat[,-weight.temp.idx]
```

The frailty summary is below. There isn't enough data here
to do anything with, sadly. A few of the mice had issues,
but not nearly enough to warrant any kind of analysis.

```{r summary, fig.width = 7, fig.height = 5}
annot.df <- data.frame("genotype" = as.factor(mouse.info.table[,"genotype"]),
  "sex" = as.factor(mouse.info.table[,"sex"]))
rownames(annot.df) <- rownames(just.frailty)

geno.order <- order(mouse.info.table[,"genotype"])
pheatmap(t(just.frailty[geno.order,]), annotation_col = annot.df, cluster_cols = FALSE)
```

The following bar plot shows the total frailty score
for each mouse. More than half have a score of 0.

```{r total_score, fig.width = 7, fig.height = 5}
total.score <- rowSums(just.frailty)
barplot(sort(total.score), las = 2)
```

Just out of a misguided sense of curiosity, I checked to
see if any of the measured proteins were significantly 
correlated with the total frailty score. There were a few
that were significant with an FDR <= 0.1. These are shown 
below.

```{r test, fig.width = 8, fig.height = 8, warning = FALSE}
mouse.idx <- sapply(rownames(just.frailty), function(x) which(all.info.table[,"ID"] == x))
aligned.alamar <- kl.mat[,mouse.idx]
test.cor <- apply(aligned.alamar, 1, function(x) cor.test(x, total.score, method = "spearman"))

r <- sapply(test.cor, function(x) x$estimate)
p <- sapply(test.cor, function(x) x$p.value)
fdr <- p.adjust(p, "fdr")
#min(fdr)

high.cor <- which(fdr <= 0.1)

par(mfrow = c(2,2))
for(i in 1:length(high.cor)){
  plot.with.model(aligned.alamar[names(high.cor)[i],], total.score,
    xlab = names(high.cor)[i], ylab = "Frailty Score")
}

stripchart(total.score~as.factor(mouse.info.table[,"genotype"]),
  vertical = TRUE, pch = 16, method = "jitter")
```

I kind of want to try the mediation just to set it up.
I know it will be non-interpretable. 