---
title: "GTEx Validation Multi-Tissue"
author: Anna L Tyler
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  html_document:
    code_folding: hide
    collapsed: no
    toc: yes
    toc_float: yes
---

This workflow is based on 3.GTEx_validation_one_tissue.Rmd,
but looks at multiple tissues at once.


```{r clear}
rm(list = ls())
library(here)
```

```{r param}
gene.set <- "Glutamate_Genes"
#gene.set <- "Parkinson_disease-Mitochondrial_Metabolism_Genes"
#gene.set <- "Ribosome-Immune_Response_Genes"

tissue.files <- list.files(here("Data", "Human", "GTEx"), pattern = ".gct", full.names = TRUE)

tissue.names <- gsub(".gct", "", gsub("gene_reads_v11_brain_", "", basename(tissue.files)))

dot.col <- "#dd1c77"
scale.expr <- TRUE #if TRUE, scales expression across individuals

results.dir <- here("Results", "GTEx")
if(!file.exists(results.dir)){dir.create(results.dir)}
```

## Effects on `r gene.set` genes

This workflow is looking at expression in all tissues.

Expression values are scaled: `r as.character(scale.expr)`.

```{r load_code}
all.fun <- list.files(here("Code"), full.names = TRUE)
for(i in 1:length(all.fun)){source(all.fun[i])}
```

```{r libraries, message = FALSE, warning = FALSE, error = FALSE}
needed.libraries <- c("vioplot", "lme4", "pheatmap", "HardyWeinberg", "DT")
load_libraries(needed.libraries)
```


```{r read_data}
geno.cols <- get_geno_cols(here("Data", "general", "geno_cols.csv"))
het.col <- "gray"
hap.cols <- c("FCFC" = as.vector(geno.cols["FC/FC"]), 
  "FCVS" = het.col, "VSVS" = as.vector(geno.cols["VS/VS"]))
#barplot(rep(1, 3), col = hap.cols)

data.dir <- here("Data", "Human", "GTEX")

pheno <- read.delim(file.path(data.dir, "GTEx_Analysis_2021-02-11_v9_Annotations_SubjectPhenotypesDS.txt"))
haps <- read.table(file.path(data.dir, "id.geno.klotho.apoe.ALL.txt"), sep = " ", row.names = 1)
known.haps <- haps[which(haps[,1] != "TT00"),1,drop=FALSE]
ind.id <- sapply(strsplit(rownames(known.haps), "-"), function(x) paste(x[c(1:2)], collapse = "-"))
rownames(known.haps) <- ind.id
#table(known.haps)

val.genes <- read.table(here("Results", "age_batch_12_gene", "validation_gene_sets", 
  paste0(gene.set, ".txt")))[,1]
ortho.file <- here("Results", "age_batch_12_gene", 
  "validation_gene_sets", paste0(gene.set, "_Orthologs.csv"))
gene.orthos <- get_orthologs(mouse.id = val.genes, file.name = ortho.file)

expr.meta <- as.matrix(read.delim(file.path(data.dir, "GTEx_Analysis_v11_Annotations_SampleAttributesDS.txt")))
```

```{r read_expr}
expr <- lapply(tissue.files, function(x) read.delim(x, skip = 2))
expr.gene.info <- lapply(expr, function(x) x[,1:2])

common.genes <- lapply(expr.gene.info, 
    function(x) intersect(gene.orthos[,"human_external_gene_name"], x[,2]))
gene.idx <- lapply(1:length(common.genes), 
    function(x) match(common.genes[[x]], expr.gene.info[[x]][,2]))

expr.num <- lapply(1:length(expr), function(x) expr[[x]][gene.idx[[x]],3:ncol(expr[[x]])])
all.expr.samp.id <- lapply(expr.num, colnames)

expr.meta.samp.id <- expr.meta[,1] #sample IDs for expression, includes subject ID
meta.idx <- lapply(all.expr.samp.id, function(x) match(x, gsub("-", ".", expr.meta.samp.id, fixed = TRUE)))
sample.meta <- lapply(meta.idx, function(x) expr.meta[x,])

```

```{r align_vars}
#line up all variables by subject ID.
pheno.subj.id <- pheno[,"SUBJID"]

hap.subj.id <- rownames(known.haps)
mod.hap.subj.id <- sapply(strsplit(hap.subj.id, "-"), function(x) paste(x[1], x[2], sep = "-"))
rownames(known.haps) <- mod.hap.subj.id

#expression sample and subject IDs
expr.samp.id <- lapply(expr.num, colnames)
#extract subject IDs from sample IDs
mod.expr.subj.id <- lapply(expr.samp.id, 
    function(y) sapply(strsplit(y, ".", fixed = TRUE), function(x) paste(x[1], x[2], sep = "-")))

#subjects for which we have phenotype data, haplotype data, and at least some expression data
common.id <- Reduce("intersect", 
  list(pheno.subj.id, mod.hap.subj.id, Reduce("union", mod.expr.subj.id)))
#length(common.id)

#get expression for any of the individuals for which we have expression data
has.expr <- lapply(mod.expr.subj.id, function(x) intersect(x, common.id))
#sapply(has.expr, length)
expr.idx <- lapply(1:length(has.expr), function(x) match(has.expr[[x]], mod.expr.subj.id[[x]]))
#sapply(expr.idx, length)

matched.expr <- lapply(1:length(expr.idx), function(x) expr.num[[x]][,expr.idx[[x]]])
long.ind.id <- unlist(has.expr)
tissue.label <- lapply(1:length(matched.expr), function(x) rep(tissue.names[x], ncol(matched.expr[[x]])))
long.samp.id <- unlist(lapply(matched.expr, colnames))
full.long.id <- paste(unlist(tissue.label), long.samp.id, sep = "_")

all.expr <- Reduce("cbind", matched.expr)
log.expr <- log2(all.expr+1)

big.meta <- Reduce("rbind", sample.meta)
matched.meta <- big.meta[match(long.samp.id, gsub("-", ".", big.meta[,"SAMPID"], fixed = TRUE)),]

long.pheno.idx <- match(long.ind.id, pheno[,1])
matched.pheno <- pheno[long.pheno.idx,]
long.hap.idx <- match(long.ind.id, rownames(known.haps))
matched.haps <- known.haps[long.hap.idx,]

sex <- matched.pheno[,"SEX"]
sex[which(sex == 1)] <- "male"
sex[which(sex == 2)] <- "female"

var.table <- data.frame("age" = scale(matched.pheno[,c("AGE")]), "sex" = as.factor(sex), 
  "KL" = ordered(matched.haps, levels = c("FCFC", "FCVS", "VSVS")), 
  "subjID" = as.factor(long.ind.id), 
  "rin" = scale(as.numeric(matched.meta[,"SMRIN"])), 
  "center" = as.factor(matched.meta[,"SMCENTER"]))

matched.gene.info <- expr.gene.info[[1]][gene.idx[[1]],]
rownames(log.expr) <- matched.gene.info[,"Description"]
```

## Factor effects by gene

Now we can start fitting some models.
The plot below shows the -log10 p value for each factor as
it affects each gene in our list. Age and sex are the big 
movers here, not surprisingly. Klotho and its interaction 
with age don't seem to do anything.

```{r models, fig.width = 7, fig.height = 4}
#boxplot(t(log.expr), las = 2)
models <- apply(log.expr, 1, function(x) lmer(as.numeric(x) ~ (1 | rin) + (1 | center) + (1 | subjID) + sex + age * KL, data = var.table, REML = TRUE))
#lapply(models, summary)
test <- lapply(models, function(x) car::Anova(x, type = 3))
p <- sapply(test, function(x) x[,3])
rownames(p) <- rownames(test[[1]])
pheatmap(-log10(p+1e-16)[2:nrow(p),], cluster_rows = FALSE)

```

## Factor effects on mean expression

The bar plot below shows the -log10 p values for the 
effect of each factor on mean expression of all genes.

```{r mean}
mean.expr <- scale(colMeans(log.expr))
mean.model <- lmer(mean.expr ~ (1 | rin) + (1 | center) + (1 | subjID) + sex + age * KL, data = var.table, REML = TRUE)
#summary(mean.model)
test <- car::Anova(mean.model, type = 3)
barplot(-log10(test[2:nrow(test),3]), names = rownames(test)[2:nrow(test)], ylab = "-log10(p)")
abline(h = -log10(0.05))
```

## Mean expression by age

The plot below shows the overall effect of age on 
mean expression of these genes.

```{r}
dummy.var <- dummy_covar(var.table[,c("sex", "center")])
adj.expr <- adjust(t(log.expr), cbind(var.table[,"rin"], dummy.var))
mean.adj <- rowMeans(adj.expr)

plot.with.model(matched.pheno[,"AGE"], mean.adj, xlab = "Age", 
  ylab = "Mean Expression",report = "cor.test")
```


## Mean expression by age and KL genotype

The plot below shows mean expression adjusted for all the 
covariates as a function of age and for each KL genotype 
separately.

```{r plots, fig.width = 9, fig.height = 3}

#plot models for individual Klotho genotypes
all.models <- vector(mode = "list", length = length(hap.cols))
names(all.models) <- names(hap.cols)
par(mfrow = c(1,3))
for(i in 1:length(hap.cols)){
  kl.idx <- which(var.table[,"KL"] == names(hap.cols)[i])
  all.models[[i]] <- lm(mean.adj[kl.idx]~var.table[kl.idx,"age"])
  plot.with.model(matched.pheno[kl.idx,"AGE"], mean.adj[kl.idx], 
  xlab = "Age", ylab = "Mean Expression",
  report = "cor.test", main = names(hap.cols)[i],
  col = hap.cols[i])
}
```

## Overlayed

The following plot shows the same information as above
but overlayed in one plot.

```{r int_plot, fig.width = 5, fig.height = 5}
plot(var.table[,"age"], mean.adj, xlab = "Age", 
  ylab = "Mean Expression", type = "n", axes = FALSE)
axis(2)

x.vals <- segment_region(min(var.table[,"age"]), max(var.table[,"age"]), 5, "ends")
x.labels <- round(segment_region(min(matched.pheno[,"AGE"]), max(matched.pheno[,"AGE"]), 5, "ends"))
axis(1, at = x.vals, labels = x.labels)

for(i in 1:length(hap.cols)){
  kl.idx <- which(var.table[,"KL"] == names(hap.cols)[i])
  points(var.table[kl.idx,"age"], mean.adj[kl.idx], 
    col = mix_colors(hap.cols[i], "white", mixing.parameter = 0.8), 
    pch = 16)
}

#add lines over the dots
for(i in 1:length(hap.cols)){
  kl.idx <- which(var.table[,"KL"] == names(hap.cols)[i])
  abline(all.models[[i]], col = hap.cols[i], lwd = 3)
}
```


## Matching populations

The VS homs are all male and are older on average
than the rest of the patients. I'm going to try to
match this population so we can minimize biases.


```{r age_filter}
info.table <- data.frame("age" = matched.pheno[,c("AGE")], "sex" = as.factor(sex), 
  "KL" = ordered(matched.haps), "subjID" = as.factor(long.ind.id), 
  "rin" = as.numeric(matched.meta[,"SMRIN"]), 
  "center" = as.factor(matched.meta[,"SMCENTER"]))

vs.idx <- which(info.table[,"KL"] == "VSVS")
vs.subj <- unique(info.table[vs.idx,"subjID"])
num.vs <- length(vs.subj)
subj.info <- info.table[match(vs.subj, info.table[,"subjID"]),]
datatable(subj.info)
```

I'm going to try to select only males matched for age in the
and for genotype in the rest of the population. I will select
the same number of individuals in each group.

```{r vs_num}
u_subj <- unique(info.table[,"subjID"])
u_geno <- info.table[match(u_subj, info.table[,"subjID"]),"KL"]
vs.num <- length(which(u_geno == "VSVS"))
```

The are`r vs.num` VSVS homozygotes in this population.
We will sample subjects from the rest of the genotypes
using this number. We will match all to have the same 
mean age.


```{r match_pop}

#a function to sample a specified number of patients with
#a specified mean age and sex
sample_pop <- function(stat.table, mean.age, sex, genotype, num, age.tol = 5, replace = FALSE){
  sex.idx <- which(stat.table[,"sex"] == sex)
  geno.idx <- which(stat.table[,"KL"] == genotype)
  sub.table <- stat.table[intersect(sex.idx, geno.idx),]
  u_subj <- unique(sub.table[,"subjID"])
  rnd.subj <- sample(u_subj, num, replace = replace)  
  rnd.age <- mean(sub.table[match(rnd.subj, sub.table[,"subjID"]),"age"])
  outside.mean <- abs(rnd.age - mean.age) > age.tol

  while(outside.mean){
    rnd.subj <- sample(u_subj, num)  
    rnd.age <- mean(sub.table[match(rnd.subj, sub.table[,"subjID"]),"age"])
    outside.mean <- abs(rnd.age - mean.age) > age.tol
    }

  return(rnd.subj)

}
```

We will sample the other males with the right genotype and fit a model
in the matched populations.

```{r sample_and_test_lme4, eval = FALSE}
mean.age <- mean(subj.info[,"age"])

boot.p <-boot.effect <- matrix(NA, nrow = 100, ncol = 3)
colnames(boot.p) <- c("age", "KL", "age:KL")
for(i in 1:100){
  rnd.fc <- sample_pop(info.table, mean.age, "male", "FCFC", vs.num)
  #rnd.het <- sample_pop(info.table, mean.age, "male", "FCVS", vs.num)
  #all.subj <- c(subj.info[,"subjID"], rnd.fc, rnd.het)
  all.subj <- c(subj.info[,"subjID"], rnd.fc) #leave out the hets for now
  subj.idx <- which(var.table[,"subjID"] %in% all.subj)

  sub.df <- info.table[subj.idx,]
  #rescale for this pop
  sub.df[,"age"] <- scale(sub.df[,"age"])
  sub.df[,"rin"] <- scale(sub.df[,"rin"])
  balanced.model <- lmer(scale(mean.expr[subj.idx]) ~ (1 | rin) + (1 | center) + (1 | subjID) + age * KL, data = sub.df, REML = TRUE)
  #balanced.model <- lmer(scale(mean.expr[subj.idx]) ~ (1 | rin) + (1 | center) + age * KL, data = sub.df, REML = TRUE)

  #regular and regress out covar? 
  #test <- lm(mean.expr[subj.idx] ~ age * KL, data = var.table[subj.idx,])
  test <- car::Anova(balanced.model, type = 3)
  test.p <- test[2:nrow(test),3]
  boot.p[i,] <- test.p
  #barplot(-log10(test[2:nrow(test),3]), names = rownames(test)[2:nrow(test)], ylab = "-log10(p)")
  #abline(h = -log10(0.05))
}

boxplot(-log10(boot.p), ylab = "-log10(p)")
abline(h = -log10(0.05))
```



```{r sample_and_test_lm}
mean.age <- mean(subj.info[,"age"])
vs.id <- subj.info[,"subjID"]

#n.samp <- vs.num
n.samp <- 10
n.boot <- 500
boot.p <-boot.effect <- matrix(NA, nrow = n.boot, ncol = 5)
age.mat <- matrix(NA, nrow = n.boot, ncol = n.samp*2)
for(i in 1:n.boot){
  rnd.fc <- sample_pop(info.table, mean.age, "male", "FCFC", n.samp, replace = TRUE)
  rnd.het <- sample_pop(info.table, mean.age, "male", "FCVS", n.samp, replace = TRUE)
  all.subj <- c(sample(vs.id, n.samp, replace = TRUE), rnd.fc, rnd.het)
  #all.subj <- c(sample(vs.id, n.samp, replace = TRUE), rnd.fc) #leave out the hets for now
  subj.idx <- unlist(lapply(all.subj, function(x) which(var.table[,"subjID"] == x)))
  #subj.idx <- which(var.table[,"subjID"] %in% all.subj)

  sub.df <- info.table[subj.idx,]
  #rescale for this pop
  sub.df[,"age"] <- scale(sub.df[,"age"])
  sub.df[,"rin"] <- scale(sub.df[,"rin"])
  
  dummy.var <- cbind(dummy_covar(sub.df[,"center",drop=FALSE]), sub.df[,"rin",drop=FALSE])
  adj.expr <- adjust(matrix(scale(mean.expr[subj.idx]), ncol = 1), as.matrix(dummy.var))

  balanced.model <- lm(adj.expr ~ age * KL, data = sub.df)
  #summary(balanced.model)

  test.result <- summary(balanced.model)$coefficients
  #test.p <- test.result[2:nrow(test.result),4]
  test.p <- test.result[c("age", "KL.L", "age:KL.L"),4]
  
  boot.p[i,] <- test.p
  #barplot(-log10(test[2:nrow(test),3]), names = rownames(test)[2:nrow(test)], ylab = "-log10(p)")
  #abline(h = -log10(0.05))
}
colnames(boot.p) <- rownames(test.result)[2:nrow(test.result)]

boxplot(-log10(boot.p), ylab = "-log10(p)")
abline(h = -log10(0.05))
```

## More exploratory?
Should we be opening this up for more exploration?
I think this population is extremely underpowered. 
Maybe we should try a different population. 
But what would that be? I think we'd want to 
have a healthy population, not an Alzheimer's
population... Ugh.
