---
title: "GTEx Validation One Tissue"
author: Anna L Tyler
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  html_document:
    code_folding: hide
    collapsed: no
    toc: yes
    toc_float: yes
---

## Introduction

The purpose of this workflow is to see if we can recapitulate key findings
from the mouse experiment in a human cohort. We are using GTEx gene expression
to see if individuals with the different Klotho haplotypes (with or without 
different APOE genotypes) have variation in glutamate receptor genes. 

We will bin the individuals by age and adjust for sex. 

Of course, we know that tons of environmental variables affect Klotho 
levels. I'll do the simplest analysis at first and then go further if
we need to.


```{r clear}
rm(list = ls())
library(here)
```

```{r param}
args = commandArgs(trailingOnly=TRUE)

expr.filename <- args[1]

if(is.na(expr.filename)){
  expr.filename <- "gene_reads_v11_brain_amygdala.gct"
  #expr.filename <- "gene_reads_v11_brain_anterior_cingulate_cortex_ba24.gct"
  #expr.filename <- "gene_reads_v11_brain_caudate_basal_ganglia.gct"
  #expr.filename <- "gene_reads_v11_brain_hippocampus.gct"
  #expr.filename <- "gene_reads_v11_brain_nucleus_accumbens_basal_ganglia.gct"
}

tissue.name <- gsub(".gct", "", gsub("gene_reads_v11_", "", expr.filename))

gene.set.file <- "Glutamate_Genes.txt"
#gene.set.file <- "Ribosome-Synapse_Genes.txt"
#gene.set.file <- "Ribosome-Immune_Response_Genes.txt"

dot.col <- "#dd1c77"
scale.expr <- TRUE #if TRUE, scales expression across individuals

results.dir <- here("Results", "GTEx")
if(!file.exists(results.dir)){dir.create(results.dir)}
```

This workflow is looking at expression in `r tissue.name`.

Expression values are scaled: `r as.character(scale.expr)`.

The validation set was taken from `r gene.set.file`.

```{r load_code}
all.fun <- list.files(here("Code"), full.names = TRUE)
for(i in 1:length(all.fun)){source(all.fun[i])}
```

```{r libraries, message = FALSE, warning = FALSE, error = FALSE}
needed.libraries <- c("pheatmap", "vioplot", "corto")
load_libraries(needed.libraries)
```


```{r read_data}
geno.cols <- get_geno_cols(here("Data", "general", "geno_cols.csv"))
het.col <- "gray"
hap.cols <- c("FCFC" = as.vector(geno.cols["FC/FC"]), 
  "FCVS" = het.col, "VSVS" = as.vector(geno.cols["VS/VS"]))
#barplot(rep(1, 3), col = hap.cols)

data.dir <- here("Data", "Human", "GTEX")

pheno <- read.delim(file.path(data.dir, "GTEx_Analysis_2021-02-11_v9_Annotations_SubjectPhenotypesDS.txt"))
haps <- read.table(file.path(data.dir, "id.geno.klotho.apoe.ALL.txt"), sep = " ", row.names = 1)
#table(haps)

val.genes <- read.table(here("Results", "age_batch_12_gene", 
  "validation_gene_sets", gene.set.file))[,1]


ortho.file <- here("Results", "age_batch_12_gene", "validation_gene_sets",
  paste0(gsub(".txt", "", gene.set.file), "_Orthologs.csv"))

gene.orthos <- get_orthologs(mouse.id = val.genes, file.name = ortho.file)
```

```{r read_expr}
expr <- read.delim(file.path(data.dir, expr.filename), skip = 2)
expr.gene.info <- expr[,1:2]

age <- pheno[,"AGE"]
sex <- pheno[,"SEX"]
names(age) <- names(sex) <- pheno[,"SUBJID"]
```

## Age Distribution

Age is distributed in the following way:

```{r age}
hist(age)
```

We binned the subjects into the following age bins.

```{r bin_age}
#automatic generation of bins
age.bins <- round(segment_region(min(age), max(age), 4, "ends"))

#there aren't many individuals with expression and VSVS haplotypes
#check ages to make bins
#vs.idx <- which(matched.haps[,1] == "VSVS")
#matched.age[vs.idx]
#age.bins <- c(20, 55, 70)

bin.mat <- consec_pairs(age.bins)

#adjust so we have non-overlapping bins. Otherwise patients get double-assigned
bin.mat[2:nrow(bin.mat),1] <- bin.mat[2:nrow(bin.mat),1]+1

#assign each person a bin based on age
assign_bin <- function(value, bin.mat){
  bin.which <- which(apply(bin.mat, 1, function(x) segments.overlap(value, value, x[1], x[2])))
  return(bin.which)
}

age.bin <- sapply(age, function(x) assign_bin(x, bin.mat))
bin.names <- apply(bin.mat, 1, function(x) paste(x, collapse = "-"))
barplot_with_num(table(age.bin), names = bin.names, ylab = "Count",
  main = "Patients in each age group")
```

## Distribution of sex by age

Sex and age are distributed as follows:

```{r sex, fig.width = 7, fig.height = 5}
age.sex <- table(data.frame("age bin" = age.bin, "sex" = sex))
rownames(age.sex) <- bin.names
colnames(age.sex) <- c("Female", "Male")

#quartz(width = 7, height = 5)
a <- barplot(t(age.sex), beside = TRUE, legend = TRUE, 
  args.legend = c("x" = "topleft"), ylim = c(0, max(age.sex)*1.1))
text(x = a[1,], y = age.sex[,1]+15, labels = age.sex[,1])
text(x = a[2,], y = age.sex[,2]+15, labels = age.sex[,2])
```

## Gene expression by age

```{r extract_genes}
common.genes <- intersect(gene.orthos[,"human_external_gene_name"], expr.gene.info[,2])
gene.idx <- match(common.genes, expr.gene.info[,2])

expr.num <- expr[gene.idx,3:ncol(expr)]
```

```{r align_vars}
#line up all variables by subject ID.
pheno.subj.id <- pheno[,"SUBJID"]

hap.subj.id <- rownames(haps)
mod.hap.subj.id <- sapply(strsplit(hap.subj.id, "-"), function(x) paste(x[1], x[2], sep = "-"))
rownames(haps) <- mod.hap.subj.id

expr.subj.id <- colnames(expr.num)
mod.expr.subj.id <- sapply(strsplit(expr.subj.id, ".", fixed = TRUE), function(x) paste(x[1], x[2], sep = "-"))
colnames(expr.num) <- mod.expr.subj.id

common.id <- Reduce("intersect", list(pheno.subj.id, mod.hap.subj.id, mod.expr.subj.id))

#match up covariates
pheno.id.idx <- sapply(common.id, function(x) which(pheno.subj.id == x))
matched.age <- age[pheno.id.idx]
matched.age.bin <- age.bin[pheno.id.idx]
matched.sex <- sex[pheno.id.idx]

expr.id.idx <- sapply(common.id, function(x) which(colnames(expr.num) == x))
matched.expr <- expr.num[,expr.id.idx]
#matched.expr[1:10, 1:10]

hap.idx <- sapply(common.id, function(x) which(rownames(haps) == x))
matched.haps <- haps[hap.idx,]

matched.gene.info <- expr.gene.info[gene.idx,]
```

```{r adjust_expr}
#adjust by sex
adj.expr <- log2(adjust(t(matched.expr), matrix(matched.sex, ncol = 1))+1)
#name the genes something useful.
colnames(adj.expr) <- matched.gene.info[,"Description"]

if(scale.expr){
  scaled.expr <- apply(adj.expr, 2, scale)
  dimnames(scaled.expr) <- dimnames(adj.expr)
  use.expr <- scaled.expr
  ylab = "Scaled Expression"
}else{
  use.expr <- adj.expr
  ylab = "log2(count)"
}

binned.expr <- lapply(1:max(age.bin), function(x) use.expr[which(matched.age.bin == x),])
names(binned.expr) <- bin.names
```

The genes in this module have the following correlation structure.

```{r gene_cor}
#pheatmap(use.expr)
pheatmap(cor(use.expr, use = "pairwise.complete.obs"))
```

The following plot shows gene expression for this group
of genes by age. 

```{r zero_line}
#adds zero line to grouped box plot.
add_zero_line <- function(){
  plot.dim <- par("usr")
  segments(plot.dim[1], 0, plot.dim[2], 0)
}
```

```{r plot_expr, fig.width = 12, fig.height = 5}
if(scale.expr){
  mean.order <- 1:ncol(adj.expr)
}else{
  mean.order <- order(colMeans(adj.expr))
}
ordered.expr <- lapply(binned.expr, function(x) x[,mean.order])
#quartz(width = 12, height = 5)
plot.grouped.boxes(ordered.expr, type = "matrix", print.vals = NA, ylab = ylab,
  main = "Expression by Age Bins")
add_zero_line()
```

## Mean expression by age

Mean expression for each gene in each age group is shown below. 

```{r heat, fig.width = 3, fig.height = 5}
mean.mat <- sapply(binned.expr, function(x) colMeans(x, na.rm = TRUE))
#quartz(width = 3, height = 5)
pheatmap(mean.mat, cluster_cols = FALSE)
write.table(mean.mat, 
  file.path(results.dir, paste0("Mean_Expr_by_Age_", tissue.name, ".csv")),
  quote = FALSE, sep = ",")
```

As a group, these genes do seem to decline in expression with age.

```{r overall}
mean.mat.tests <- pairwise_test(mean.mat)

vioplot(mean.mat, col = hap.cols, main = "Module Mean Expression",
  ylim = c(min(mean.mat), max(mean.mat)+0.2))
add_pairwise_results(mean.mat, mean.mat.tests, y.nudge = 0.15, text.nudge = 0.01)

mtext("Mean Scaled Expression", side = 2, line = 2.5)

```

## Differential expression by age

Does gene expression vary across age groups?
The line marks an unadjusted p value of 0.05.

```{r age_test, fig.width = 12, fig.height = 5}
expr.diff.pvals <- rep(NA, ncol(ordered.expr[[1]]))
names(expr.diff.pvals) <- colnames(ordered.expr[[1]])

for(i in 1:ncol(ordered.expr[[1]])){
  test.expr <- lapply(ordered.expr, function(x) x[,i])
  test.diff <- aov.by.list(test.expr)
  expr.diff.pvals[i] <- test.diff[1, "Pr(>F)"]
}

barplot(-log10(expr.diff.pvals), las = 2, cex.names = 1.2,
  ylab = "-log10(p)")
abline(h = -log10(0.05))
```

## Expression by Kl haplotype {.tabset .tabset-fade .tabset-pills}

Now it's time to bring in the haplotypes.
We want to see if these genes are differentially 
expressed by Klotho haplotype only, and whether that
expression difference increases with age.

```{r kl_hap, fig.width = 12, fig.height = 5, results = "asis"}
kl.haps <- c("FCFC", "FCVS", "VSVS")
kl.groups <- lapply(kl.haps, function(x) rownames(matched.haps)[which(matched.haps[,1] == x)])

for(i in 1:length(ordered.expr)){
  cat("###", bin.names[i], "\n")
  age.expr <- ordered.expr[[i]]
  #bin these individuals into haplotypes
  hap.expr <- lapply(kl.groups, function(x) age.expr[which(rownames(age.expr) %in% x),,drop=FALSE])
  names(hap.expr) <- kl.haps
  plot.grouped.boxes(hap.expr, type = "matrix", main = names(ordered.expr)[i],
    ylab = "Scaled Expression", print.vals = NA, group.cols = hap.cols)
  add_zero_line()
  cat("\n\n")
}
```

## Quantifying the effects of Kl haplotype {.tabset .tabset-fade .tabset-pills}

We want to see whether Klotho haplotype has a significant
effect on expression of genes individually, and on the group
as a whole.

```{r module_overall, results = "asis", fig.height = 5, fig.width = 7}
stop()

kl.expr.by.age <- vector(mode = "list", length = length(ordered.expr))
names(kl.expr.by.age) <- names(ordered.expr)
kl.effect.mat <- matrix(NA, nrow = length(ordered.expr), ncol = length(kl.haps))
rownames(kl.effect.mat) <- names(ordered.expr)
colnames(kl.effect.mat) <- kl.haps

for(i in 1:length(ordered.expr)){
  cat("###", names(ordered.expr)[i], "\n")
  
  age.expr <- ordered.expr[[i]]

  #bin these individuals into haplotypes
  hap.expr <- lapply(kl.groups, function(x) age.expr[which(rownames(age.expr) %in% x),,drop=FALSE])
  names(hap.expr) <- kl.haps
  
  #calculate mean expression over all genes in the group
  mean.expr <- lapply(hap.expr, colMeans)

  #remove any categories that are all NA
  plot.expr <- mean.expr[which(!is.na(sapply(mean.expr, function(x) mean(x, na.rm = TRUE))))]
  kl.expr.by.age[[i]] <- mean.expr
  mean.by.geno <- sapply(mean.expr, function(x) mean(x, na.rm = TRUE))
  kl.effect.mat[i,] <- mean.by.geno

  
  #need to make the pairwise test function work with lists
  age.mat.tests <- pairwise_test(age.table)

  pval <- threshold_p(diff.p[1,"Pr(>F)"], return.text = TRUE)

  vioplot(plot.expr, col = hap.cols[names(plot.expr)], main = "Group Mean Expression")
  mtext("Mean Scaled Expr", side = 2, line = 2.5)
  stripchart(plot.expr, vertical = TRUE, pch = 16, method = "jitter", col = dot.col,
    add = TRUE)
  mtext(pval, side = 3, line = -2)
  cat("\n\n")
}

saveRDS(kl.expr.by.age, 
  file.path(results.dir, paste0("Mean_Expr_per_gene_by_Age_", tissue.name, ".RDS")))

write.table(kl.effect.mat, 
  file.path(results.dir, paste0("Mean_KL_Expr_by_Age_", tissue.name, ".csv")),
  sep = ",", quote = FALSE)
```

## Summary of Klotho effects by age

The following plot shows the effect of Klotho haplotype in 
each age bin.

```{r kl_age_summ, fig.width = 10, fig.height = 5}
plot.grouped.boxes(group.list = kl.expr.by.age, plot.grouping = "inner",
  ylab = "Mean Scaled Expression", main = "Effect of Haplotype at each Age",
  group.cols = hap.cols)
add_zero_line()
```

Bar plot showing the same thing.

```{r kl_by_age_summary, fig.width = 7, fig.height = 5}
barplot(t(kl.effect.mat), beside = TRUE, legend = TRUE, col = hap.cols,
  ylab = "Mean Scaled Expression")
abline(h = 0)
```

the following plot shows the effect of age in each Klotho haplotype.

```{r, age_kl_summ, fig.width = 10, fig.height = 5}
plot.grouped.boxes(kl.expr.by.age, ylab = "Mean Scaled Expression",
  main = "Effect of Age in each Haplotype")
add_zero_line()
```

The bar plot below shows the same thing in bars.

```{r age_by_kl_summary, fig.width = 7, fig.height = 5}
barplot(kl.effect.mat, beside = TRUE, legend = TRUE, col = hap.cols,
  ylab = "Mean Scaled Expression")
abline(h = 0)
```
