---
title: "GTEx Validation Multi-Tissue"
author: Anna L Tyler
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  html_document:
    code_folding: hide
    collapsed: no
    toc: yes
    toc_float: yes
---

## Introduction

This workflow is based on 3.GTEx_validation_one_tissue.Rmd,
but looks at multiple tissues at once.


```{r clear}
rm(list = ls())
library(here)
```

```{r param}
tissue.files <- list.files(here("Data", "Human", "GTEx"), pattern = ".gct", full.names = TRUE)

tissue.names <- gsub(".gct", "", gsub("gene_reads_v11_brain_", "", basename(tissue.files)))

dot.col <- "#dd1c77"
scale.expr <- TRUE #if TRUE, scales expression across individuals

results.dir <- here("Results", "GTEx")
if(!file.exists(results.dir)){dir.create(results.dir)}
```

This workflow is looking at expression in all tissues.

Expression values are scaled: `r as.character(scale.expr)`.

```{r load_code}
all.fun <- list.files(here("Code"), full.names = TRUE)
for(i in 1:length(all.fun)){source(all.fun[i])}
```

```{r libraries, message = FALSE, warning = FALSE, error = FALSE}
needed.libraries <- c("pheatmap", "vioplot")
load_libraries(needed.libraries)
```


```{r read_data}
geno.cols <- get_geno_cols(here("Data", "general", "geno_cols.csv"))
het.col <- "gray"
hap.cols <- c("FCFC" = as.vector(geno.cols["FC/FC"]), 
  "FCVS" = het.col, "VSVS" = as.vector(geno.cols["VS/VS"]))
#barplot(rep(1, 3), col = hap.cols)

data.dir <- here("Data", "Human", "GTEX")

pheno <- read.delim(file.path(data.dir, "GTEx_Analysis_2021-02-11_v9_Annotations_SubjectPhenotypesDS.txt"))
haps <- read.table(file.path(data.dir, "id.geno.klotho.apoe.ALL.txt"), sep = " ", row.names = 1)
#table(haps)

val.genes <- read.table(here("Results", "age_batch_12_gene", "Glutamate_Genes.txt"))[,1]
ortho.file <- here("Results", "age_batch_12_gene", "Glutamate_Gene_Orthologs.csv")
gene.orthos <- get_orthologs(mouse.id = val.genes, file.name = ortho.file)
```

```{r read_expr}
expr <- lapply(tissue.files, function(x) read.delim(x, skip = 2))
expr.gene.info <- lapply(expr, function(x) x[,1:2])

common.genes <- lapply(expr.gene.info, 
    function(x) intersect(gene.orthos[,"human_external_gene_name"], x[,2]))
gene.idx <- lapply(1:length(common.genes), 
    function(x) match(common.genes[[x]], expr.gene.info[[x]][,2]))

expr.num <- lapply(1:length(expr), function(x) expr[[x]][gene.idx[[x]],3:ncol(expr[[x]])])
```

```{r align_vars}
#line up all variables by subject ID.
pheno.subj.id <- pheno[,"SUBJID"]

hap.subj.id <- rownames(haps)
mod.hap.subj.id <- sapply(strsplit(hap.subj.id, "-"), function(x) paste(x[1], x[2], sep = "-"))
rownames(haps) <- mod.hap.subj.id

expr.subj.id <- lapply(expr.num, colnames)
mod.expr.subj.id <- lapply(expr.subj.id, 
    function(y) sapply(strsplit(y, ".", fixed = TRUE), function(x) paste(x[1], x[2], sep = "-")))
for(i in 1:length(expr.num)){
  colnames(expr.num[[i]]) <- mod.expr.subj.id[[i]]
}


common.id <- Reduce("intersect", list(pheno.subj.id, mod.hap.subj.id, Reduce("intersect", mod.expr.subj.id)))

#match up covariates
pheno.id.idx <- sapply(common.id, function(x) which(pheno.subj.id == x))
matched.age <- age[pheno.id.idx]
matched.age.bin <- age.bin[pheno.id.idx]
matched.sex <- sex[pheno.id.idx]

expr.id.idx <- sapply(common.id, function(x) which(colnames(expr.num) == x))
matched.expr <- expr.num[,expr.id.idx]
#matched.expr[1:10, 1:10]

hap.idx <- sapply(common.id, function(x) which(rownames(haps) == x))
matched.haps <- haps[hap.idx,]

matched.gene.info <- expr.gene.info[gene.idx,]
```


```{r}
#get rid of the huge expr object
rm(expr)

#build a big concatenated expression matrix
all.ind <- lapply(expr.num, rownames)


tissue.label <- lapply(expr)
```

```{r pheno}
age <- pheno[,"AGE"]
sex <- pheno[,"SEX"]
names(age) <- names(sex) <- pheno[,"SUBJID"]
```

## Age Distribution

Age is distributed in the following way:

```{r age}
hist(age)
```

We binned the subjects into the following age bins.

```{r bin_age}
#automatic generation of bins
age.bins <- round(segment_region(min(age), max(age), 4, "ends"))

#there aren't many individuals with expression and VSVS haplotypes
#check ages to make bins
#vs.idx <- which(matched.haps[,1] == "VSVS")
#matched.age[vs.idx]
#age.bins <- c(20, 55, 70)

bin.mat <- consec_pairs(age.bins)

#adjust so we have non-overlapping bins. Otherwise patients get double-assigned
bin.mat[2:nrow(bin.mat),1] <- bin.mat[2:nrow(bin.mat),1]+1

#assign each person a bin based on age
assign_bin <- function(value, bin.mat){
  bin.which <- which(apply(bin.mat, 1, function(x) segments.overlap(value, value, x[1], x[2])))
  return(bin.which)
}

age.bin <- sapply(age, function(x) assign_bin(x, bin.mat))
bin.names <- apply(bin.mat, 1, function(x) paste(x, collapse = "-"))
barplot_with_num(table(age.bin), names = bin.names, ylab = "Count",
  main = "Patients in each age group")
```

## Distribution of sex by age

Sex and age are distributed as follows:

```{r sex, fig.width = 7, fig.height = 5}
age.sex <- table(data.frame("age bin" = age.bin, "sex" = sex))
rownames(age.sex) <- bin.names
colnames(age.sex) <- c("Female", "Male")

#quartz(width = 7, height = 5)
a <- barplot(t(age.sex), beside = TRUE, legend = TRUE, 
  args.legend = c("x" = "topleft"), ylim = c(0, max(age.sex)*1.1))
text(x = a[1,], y = age.sex[,1]+15, labels = age.sex[,1])
text(x = a[2,], y = age.sex[,2]+15, labels = age.sex[,2])
```

## Gene expression by age




```{r adjust_expr}
#adjust by sex
adj.expr <- log2(adjust(t(matched.expr), matrix(matched.sex, ncol = 1))+1)
#name the genes something useful.
colnames(adj.expr) <- matched.gene.info[,"Description"]

if(scale.expr){
  scaled.expr <- apply(adj.expr, 2, scale)
  dimnames(scaled.expr) <- dimnames(adj.expr)
  use.expr <- scaled.expr
  ylab = "Scaled Expression"
}else{
  use.expr <- adj.expr
  ylab = "log2(count)"
}

binned.expr <- lapply(1:max(age.bin), function(x) use.expr[which(matched.age.bin == x),])
names(binned.expr) <- bin.names
```

The following plot shows gene expression for this group
of genes by age. 

```{r zero_line}
#adds zero line to grouped box plot.
add_zero_line <- function(){
  plot.dim <- par("usr")
  segments(plot.dim[1], 0, plot.dim[2], 0)
}
```

```{r plot_expr, fig.width = 12, fig.height = 5}
if(scale.expr){
  mean.order <- 1:ncol(adj.expr)
}else{
  mean.order <- order(colMeans(adj.expr))
}
ordered.expr <- lapply(binned.expr, function(x) x[,mean.order])
#quartz(width = 12, height = 5)
plot.grouped.boxes(ordered.expr, type = "matrix", print.vals = NA, ylab = ylab,
  main = "Expression by Age Bins")
add_zero_line()
```

## Mean expression by age

Mean expression for each gene in each age group is shown below. 

```{r heat, fig.width = 3, fig.height = 5}
mean.mat <- sapply(binned.expr, function(x) colMeans(x, na.rm = TRUE))
#quartz(width = 3, height = 5)
pheatmap(mean.mat, cluster_cols = FALSE)
write.table(mean.mat, 
  file.path(results.dir, paste0("Mean_Expr_by_Age_", tissue.name, ".csv")),
  quote = FALSE, sep = ",")
```

As a group, these genes do seem to decline in expression with age.

```{r overall}
test.diff <- anova(aov.by.matrix(mean.mat))
pdiff <- threshold_p(test.diff[1,"Pr(>F)"], return.text = TRUE)
vioplot(mean.mat, col = hap.cols, main = "Module Mean Expression")
mtext(pdiff, side = 3, line = -2)
mtext("Mean Scaled Expression", side = 2, line = 2.5)
```

## Differential expression by age

Does gene expression vary across age groups?
The line marks an unadjusted p value of 0.05.

```{r age_test, fig.width = 12, fig.height = 5}
expr.diff.pvals <- rep(NA, ncol(ordered.expr[[1]]))
names(expr.diff.pvals) <- colnames(ordered.expr[[1]])

for(i in 1:ncol(ordered.expr[[1]])){
  test.expr <- lapply(ordered.expr, function(x) x[,i])
  test.diff <- aov.by.list(test.expr)
  expr.diff.pvals[i] <- test.diff[1, "Pr(>F)"]
}

barplot(-log10(expr.diff.pvals), las = 2, cex.names = 1.2,
  ylab = "-log10(p)")
abline(h = -log10(0.05))
```

## Expression by Kl haplotype {.tabset .tabset-fade .tabset-pills}

Now it's time to bring in the haplotypes.
We want to see if these genes are differentially 
expressed by Klotho haplotype only, and whether that
expression difference increases with age.

```{r kl_hap, fig.width = 12, fig.height = 5, results = "asis"}
kl.haps <- c("FCFC", "FCVS", "VSVS")
kl.groups <- lapply(kl.haps, function(x) rownames(matched.haps)[which(matched.haps[,1] == x)])

for(i in 1:length(ordered.expr)){
  cat("###", bin.names[i], "\n")
  age.expr <- ordered.expr[[i]]
  #bin these individuals into haplotypes
  hap.expr <- lapply(kl.groups, function(x) age.expr[which(rownames(age.expr) %in% x),,drop=FALSE])
  names(hap.expr) <- kl.haps
  plot.grouped.boxes(hap.expr, type = "matrix", main = names(ordered.expr)[i],
    ylab = "Scaled Expression", print.vals = NA, group.cols = hap.cols)
  add_zero_line()
  cat("\n\n")
}
```

## Quantifying the effects of Kl haplotype {.tabset .tabset-fade .tabset-pills}

We want to see whether Klotho haplotype has a significant
effect on expression of genes individually, and on the group
as a whole.

```{r module_overall, results = "asis", fig.height = 5, fig.width = 7}
kl.expr.by.age <- vector(mode = "list", length = length(ordered.expr))
names(kl.expr.by.age) <- names(ordered.expr)
kl.effect.mat <- matrix(NA, nrow = length(ordered.expr), ncol = length(kl.haps))
rownames(kl.effect.mat) <- names(ordered.expr)
colnames(kl.effect.mat) <- kl.haps

for(i in 1:length(ordered.expr)){
  cat("###", names(ordered.expr)[i], "\n")
  
  age.expr <- ordered.expr[[i]]

  #bin these individuals into haplotypes
  hap.expr <- lapply(kl.groups, function(x) age.expr[which(rownames(age.expr) %in% x),,drop=FALSE])
  names(hap.expr) <- kl.haps
  
  #calculate mean expression over all genes in the group
  mean.expr <- lapply(hap.expr, colMeans)

  #remove any categories that are all NA
  plot.expr <- mean.expr[which(!is.na(sapply(mean.expr, function(x) mean(x, na.rm = TRUE))))]
  kl.expr.by.age[[i]] <- mean.expr
  mean.by.geno <- sapply(mean.expr, function(x) mean(x, na.rm = TRUE))
  kl.effect.mat[i,] <- mean.by.geno

  diff.p <- aov.by.list(plot.expr)
  pval <- threshold_p(diff.p[1,"Pr(>F)"], return.text = TRUE)

  vioplot(plot.expr, col = hap.cols[names(plot.expr)], main = "Group Mean Expression")
  mtext("Mean Scaled Expr", side = 2, line = 2.5)
  stripchart(plot.expr, vertical = TRUE, pch = 16, method = "jitter", col = dot.col,
    add = TRUE)
  mtext(pval, side = 3, line = -2)
  cat("\n\n")
}

saveRDS(kl.expr.by.age, 
  file.path(results.dir, paste0("Mean_Expr_per_gene_by_Age_", tissue.name, ".RDS")))

write.table(kl.effect.mat, 
  file.path(results.dir, paste0("Mean_KL_Expr_by_Age_", tissue.name, ".csv")),
  sep = ",", quote = FALSE)
```

## Summary of Klotho effects by age

The following plot shows the effect of Klotho haplotype in 
each age bin.

```{r kl_age_summ, fig.width = 10, fig.height = 5}
plot.grouped.boxes(group.list = kl.expr.by.age, plot.grouping = "inner",
  ylab = "Mean Scaled Expression", main = "Effect of Haplotype at each Age",
  group.cols = hap.cols)
add_zero_line()
```

Bar plot showing the same thing.

```{r kl_by_age_summary, fig.width = 7, fig.height = 5}
barplot(t(kl.effect.mat), beside = TRUE, legend = TRUE, col = hap.cols,
  ylab = "Mean Scaled Expression")
abline(h = 0)
```

the following plot shows the effect of age in each Klotho haplotype.

```{r, age_kl_summ, fig.width = 10, fig.height = 5}
plot.grouped.boxes(kl.expr.by.age, ylab = "Mean Scaled Expression",
  main = "Effect of Age in each Haplotype")
add_zero_line()
```

The bar plot below shows the same thing in bars.

```{r age_by_kl_summary, fig.width = 7, fig.height = 5}
barplot(kl.effect.mat, beside = TRUE, legend = TRUE, col = hap.cols,
  ylab = "Mean Scaled Expression")
abline(h = 0)
```
