---
title: "GTEx Validation One Tissue Continuous"
author: Anna L Tyler
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  html_document:
    code_folding: hide
    collapsed: no
    toc: yes
    toc_float: yes
---

## Introduction

This workflow was copied from 3.GTEx_validation_one_tissue.Rmd,
but doesn't bin people by age. Rather we use a continuous approach.

We want to see if we can recapitulate key findings from the mouse 
experiment in a human cohort without introducing artifacts from binning. 
We are using GTEx gene expression to see if individuals with the 
different Klotho haplotypes (with or without different APOE genotypes) 
have variation in glutamate receptor genes. 

We will still adjust for sex. 

Of course, we know that tons of environmental variables affect Klotho 
levels. I'll do the simplest analysis at first and then go further if
we need to.


```{r clear}
rm(list = ls())
library(here)
```

```{r param}
args = commandArgs(trailingOnly=TRUE)

expr.filename <- args[1]
gene.set <- args[2]

if(is.na(expr.filename)){
  expr.filename <- "gene_reads_v11_brain_amygdala.gct"
  #expr.filename <- "gene_reads_v11_brain_anterior_cingulate_cortex_ba24.gct"
  #expr.filename <- "gene_reads_v11_brain_caudate_basal_ganglia.gct"
  #expr.filename <- "gene_reads_v11_brain_hippocampus.gct"
  #expr.filename <- "gene_reads_v11_brain_nucleus_accumbens_basal_ganglia.gct"

  gene.set.file <- "Glutamate_Genes.txt"
  #gene.set.file <- "Ribosome-Synapse_Genes.txt"
  #gene.set.file <- "Ribosome-Immune_Response_Genes.txt"
  #gene.set.file <- "Parkinson_disease-Mitochondrial_Metabolism_Genes.txt"
}else{
  gene.set.file <- paste0(gene.set, ".txt")
}

tissue.name <- gsub(".gct", "", gsub("gene_reads_v11_", "", expr.filename))
gene.set.label <- gsub(".txt", "", gene.set.file)

dot.col <- "#dd1c77"
scale.expr <- FALSE #if TRUE, scales expression across individuals

results.dir <- here("Results", "GTEx")
if(!file.exists(results.dir)){dir.create(results.dir)}

quick.run <- TRUE #this subsets out the genes we need before adusting so it
                  #goes faster. Otherwise all genes are adjusted and written 
                  #to file for use in other workflows
```

This workflow is looking at expression in `r tissue.name`.

Expression values are scaled: `r as.character(scale.expr)`.

The validation set was taken from `r gene.set.file`.

```{r load_code}
all.fun <- list.files(here("Code"), full.names = TRUE)
for(i in 1:length(all.fun)){source(all.fun[i])}
```

```{r libraries, message = FALSE, warning = FALSE, error = FALSE}
needed.libraries <- c("pheatmap", "vioplot", "corto")
load_libraries(needed.libraries)
```


```{r read_data}
geno.cols <- get_geno_cols(here("Data", "general", "geno_cols.csv"))
het.col <- "#dfc27d"
hap.cols <- c("FCFC" = as.vector(geno.cols["FC/FC"]), 
  "FCVS" = het.col, "VSVS" = as.vector(geno.cols["VS/VS"]))
#barplot(rep(1, 3), col = hap.cols)

data.dir <- here("Data", "Human", "GTEX")

pheno <- read.delim(file.path(data.dir, "GTEx_Analysis_2021-02-11_v9_Annotations_SubjectPhenotypesDS.txt"))
haps <- read.table(file.path(data.dir, "id.geno.klotho.apoe.ALL.txt"), sep = " ", row.names = 1)
#table(haps)

val.genes <- read.table(here("Results", "age_batch_12_gene", 
  "validation_gene_sets", gene.set.file))[,1]

ortho.file <- here("Results", "age_batch_12_gene", "validation_gene_sets",
  paste0(gsub(".txt", "", gene.set.file), "_Orthologs.csv"))

gene.orthos <- get_orthologs(mouse.id = val.genes, file.name = ortho.file)
```

```{r read_expr}
expr.meta <- as.matrix(read.delim(file.path(data.dir, "GTEx_Analysis_v11_Annotations_SampleAttributesDS.txt")))
expr <- read.delim(file.path(data.dir, expr.filename), skip = 2)
expr.gene.info <- expr[,1:2]
```

```{r align_vars}
#line up all variables by subject ID. Use the subject IDs 
#from pheno as the reference format. Modify the subject ID
#formats for all other data tables so we can pull the same 
#individuals out of each one. We need to line up phenotype
#data, haplotype data, and expression data.
pheno.subj.id <- pheno[,"SUBJID"]
hap.subj.id <- rownames(haps) #from Robert; WGS
expr.subj.id <- colnames(expr) #subject IDs for expression
expr.meta.samp.id <- expr.meta[,1] #sample IDs for expression, includes subject ID


#modify each subject ID format to look like pheno subject IDs
mod.hap.subj.id <- sapply(strsplit(hap.subj.id, "-"), 
  function(x) paste(x[1], x[2], sep = "-"))
rownames(haps) <- mod.hap.subj.id

#align metadata with the expression samples we have
alt.samp.id <- gsub(".", "-", colnames(expr), fixed = TRUE)
common.id <- intersect(alt.samp.id, expr.meta[,1])
meta.idx <- which(expr.meta[,1] %in% common.id)
sample.meta <- expr.meta[meta.idx,]
meta.aligned.expr <- expr[,match(common.id, gsub(".", "-", colnames(expr), fixed = TRUE))]
colnames(meta.aligned.expr) <- common.id

#extract the subject ID from the sample ID
#so we can align phenotype and haplotype data
mod.expr.subj.id <- sapply(strsplit(colnames(meta.aligned.expr), "-"), 
  function(x) paste(x[1:2], collapse = "-"))

#find the IDs common to phenotype, haplotype, and expression
common.id <- Reduce("intersect", list(pheno.subj.id, mod.hap.subj.id, mod.expr.subj.id))

#match up all matrices based on common IDs
pheno.id.idx <- sapply(common.id, function(x) which(pheno.subj.id == x))
expr.id.idx <- sapply(common.id, function(x) which(mod.expr.subj.id == x))
hap.id.idx <- sapply(common.id, function(x) which(rownames(haps) == x))

matched.pheno <- pheno[pheno.id.idx,]
matched.haps <- haps[hap.id.idx,]
matched.expr <- meta.aligned.expr[,expr.id.idx]
  colnames(matched.expr) <- common.id
matched.meta <- sample.meta[sapply(common.id, function(x) grep(x, sample.meta[,1])),]


#filter out genes with fewer than five counts on average
total.counts <- rowSums(matched.expr)
keep.genes <- which(total.counts >= 5*ncol(matched.expr))
filtered.expr <- matched.expr[keep.genes,]
filtered.gene.info <- expr.gene.info[keep.genes,]

#convert to TPM
millions.per.ind <- colSums(filtered.expr)/1e6
tpm.expr <- sapply(1:length(millions.per.ind), 
  function(x) filtered.expr[,x]/millions.per.ind[x])
dimnames(tpm.expr) <- dimnames(filtered.expr)

age <- matched.pheno[,"AGE"]
sex <- matched.pheno[,"SEX"]
sex[which(sex == 1)] <- "male"
sex[which(sex == 2)] <- "female"
names(age) <- names(sex) <- common.id
```

## Age Distribution

Age is distributed in the following way:

```{r age}
hist(age)
```


## Distribution of sex by age

Sex and age are distributed as follows:

```{r sex, fig.width = 10, fig.height = 5}
age.sex <- table(data.frame("age" = age, "sex" = sex))

#quartz(width = 7, height = 5)
a <- barplot(t(age.sex), beside = TRUE, legend = TRUE, 
  args.legend = c("x" = "topleft"), ylim = c(0, max(age.sex)*1.1))
text(x = a[1,], y = age.sex[,1]+0.5, labels = age.sex[,1])
text(x = a[2,], y = age.sex[,2]+0.5, labels = age.sex[,2])
```



## `r gene.set` gene expression by age

```{r extract_fun}
#extract a subset of genes we are interested in.
#genes should be in columns
extract_genes <- function(expr.mat, gene.info, gene.names){
  gene.idx <- match(gene.names, gene.info[,"Description"])
  not.na <- which(!is.na(gene.idx))
  expr.sub <- expr.mat[,gene.idx[not.na]]
  colnames(expr.sub) <- gene.names[not.na]
  sub.info <- gene.info[gene.idx[not.na],]
  result <- list("expression" = expr.sub, "info" = sub.info)
  return(result)
}

```

```{r adjust_expr}
#get names for human orthologs of the mouse genes we are interested in
genes.of.interest <- intersect(gene.orthos[,"human_external_gene_name"], 
  expr.gene.info[,2])

if(quick.run){
  sub.set <- extract_genes(expr.mat = t(tpm.expr), 
    gene.info = filtered.gene.info, gene.names = genes.of.interest)
  sub.expr <- sub.set$expression
  sub.info <- sub.set$info
}else{
  sub.expr <- t(tpm.expr) #make sure genes are in columns
  sub.info <- filtered.gene.info
}

#dim(sub.expr)
#dim(sub.info)

#log transform
log.expr <- log2(sub.expr+1)

#adjust all genes for sex, site of collection, and RIN
#we might want to use them later
adj.mat <- cbind(dummy_covar(cbind(matched.meta[,"SMCENTER",drop=FALSE], 
  "sex" = sex)), "RIN" = as.numeric(matched.meta[,"SMRIN"]))
adj.expr <- adjust(log.expr, adj.mat, retain.intercept = FALSE)

ensembl.id <- sapply(strsplit(sub.info[,1], ".", fixed = TRUE), 
  function(x) x[1])
colnames(adj.expr) <- ensembl.id

#boxplot(adj.expr[,1:10])
#boxplot(t(adj.expr[1:10,]), las = 2)
#hist(rowMeans(adj.expr, na.rm = TRUE)) #individuals have mean 0
#hist(colMeans(adj.expr, na.rm = TRUE)) #genes have mean 0
#hist(adj.expr)

if(scale.expr){
  scaled.expr <- apply(adj.expr, 2, scale)
  dimnames(scaled.expr) <- dimnames(adj.expr)
  use.expr <- scaled.expr
  ylab = "Scaled Expression"
}else{
  use.expr <- adj.expr
  ylab = "log2(count)"
}


#boxplot(use.expr[,1:10])
#boxplot(t(use.expr[1:10,]), las = 2)
#hist(rowMeans(use.expr, na.rm = TRUE)) #individuals have mean 0
#hist(colMeans(use.expr, na.rm = TRUE)) #genes have mean 0
#hist(use.expr)
if(!quick.run){ #only save if we haven't subset the genes yet
  saveRDS(use.expr, 
    here("Results", "GTEx", paste0("Adjusted_Expr_", tissue.name, ".RDS")))

  #then subset out the genes of interest for the rest of the analysis
  sub.set <- extract_genes(use.expr, sub.info, genes.of.interest)
  use.expr <- sub.set$expression
  sub.info <- sub.set$info
}

colnames(use.expr) <- genes.of.interest
```



```{r interaction_model}

#test whether age and genotype interact 
int.model <- apply(use.expr, 2, 
  function(x) lm(x~age*ordered(matched.haps[,1], 
  levels = c("FCFC", "FCVS", "VSVS"))))
int.p <- sapply(int.model, function(x) coef(summary(x))[5,4])

max.y <- max(c(-log10(0.05), -log10(int.p)))
barplot(-log10(int.p), las = 2, ylim = c(0, max.y+0.1), ylab = "-log10(p)",
  main = "Interaction p values for Klotho genotype and age")
abline(h = -log10(0.05))
```

The genes in this module have the following correlation structure.

```{r gene_cor}
#pheatmap(use.expr)
pheatmap(cor(use.expr, use = "pairwise.complete.obs"))
```

The following plot shows gene expression for this group
of genes by age.

```{r plot_expr, fig.width = 12, fig.height = 5}
mean.expr <- rowMeans(use.expr)
plot.with.model(age, mean.expr, report = "cor.test", ylab = "Mean Expression")
```

As a group, these genes do seem to decline in expression with age.


## Expression by Kl haplotype {.tabset .tabset-fade .tabset-pills}

Now it's time to bring in the haplotypes.
We want to see if these genes are differentially 
expressed by Klotho haplotype only, and whether that
expression difference increases with age.

```{r kl_hap, fig.width = 12, fig.height = 5, results = "asis"}
kl.haps <- c("FCFC", "FCVS", "VSVS")
kl.groups <- lapply(kl.haps, 
  function(x) rownames(matched.haps)[which(matched.haps[,1] == x)])

layout(get.layout.mat(ncol(use.expr)))
for(i in 1:ncol(use.expr)){
  plot.with.model(age, use.expr[,i], xlab = "age", report = "cor.test",
    ylab = "Expression", main = colnames(use.expr)[i])
}
```

## Quantifying the effects of Kl haplotype {.tabset .tabset-fade .tabset-pills}

We want to see whether Klotho haplotype has a significant
effect on expression of genes individually, and on the group
as a whole.

```{r mean_order}
test_order <- function(data.mat.or.list, test.results){
  comp.rows <- test.results[[1]]

  if(class(data.mat.or.list)[1] == "matrix"){
    maxes <- apply(comp.rows, 1, function(x) c(max(data.mat.or.list[,x[1]]), max(data.mat.or.list[,x[2]])))
  }else{
    maxes <- apply(comp.rows, 1, function(x) c(max(data.mat.or.list[[x[1]]]), max(data.mat.or.list[[x[2]]])))
  }
  max.order <- order(apply(maxes, 2, max))
  return(max.order)
}

```

```{r module_overall, results = "asis", fig.height = 6, fig.width = 7}

kl.expr.by.age <- vector(mode = "list", length = length(ordered.expr))
names(kl.expr.by.age) <- names(ordered.expr)
kl.effect.mat <- matrix(NA, nrow = length(ordered.expr), ncol = length(kl.haps))
rownames(kl.effect.mat) <- names(ordered.expr)
colnames(kl.effect.mat) <- kl.haps

for(i in 1:length(ordered.expr)){
  cat("###", names(ordered.expr)[i], "\n")
  
  age.expr <- ordered.expr[[i]]

  #bin these individuals into haplotypes
  hap.expr <- lapply(kl.groups, function(x) age.expr[which(rownames(age.expr) %in% x),,drop=FALSE])
  names(hap.expr) <- kl.haps
  
  #calculate mean expression over all genes in the group
  mean.expr <- lapply(hap.expr, colMeans)

  #remove any categories that are all NA
  plot.expr <- mean.expr[which(!is.na(sapply(mean.expr, function(x) mean(x, na.rm = TRUE))))]
  kl.expr.by.age[[i]] <- mean.expr
  mean.by.geno <- sapply(mean.expr, function(x) mean(x, na.rm = TRUE))
  kl.effect.mat[i,] <- mean.by.geno

  #need to make the pairwise test function work with lists
  effect.tests <- pairwise_test(plot.expr)

  vioplot(plot.expr, col = hap.cols[names(plot.expr)], main = "Group Mean Expression",
    ylim = c(min(unlist(plot.expr), na.rm = TRUE), 
    max(unlist(plot.expr), na.rm = TRUE)+0.5))
  mtext("Mean Scaled Expr", side = 2, line = 2.5)
  stripchart(plot.expr, vertical = TRUE, pch = 16, method = "jitter", col = dot.col,
    add = TRUE)

  
  #reorder test effects so comparisons with the highest values are plotted later

  
  add_pairwise_results(data.mat.or.list = plot.expr, 
    test.results = effect.tests, y.nudge = 0.05, text.nudge = 0.03,
    min.beam.dist = 0.07)
  
  cat("\n\n")
}

saveRDS(kl.expr.by.age, 
  file.path(results.dir, 
  paste0("Mean_Expr_per_gene_by_Age_", tissue.name, "_", gene.set.label, ".RDS")))

write.table(kl.effect.mat, 
  file.path(results.dir, 
  paste0("Mean_KL_Expr_by_Age_", tissue.name, "_", gene.set.label, ".csv")),
  sep = ",", quote = FALSE)
```

## Summary of Klotho effects by age

The following plot shows the effect of Klotho haplotype in 
each age bin.

```{r kl_age_summ, fig.width = 10, fig.height = 5}
plot.grouped.boxes(group.list = kl.expr.by.age, plot.grouping = "inner",
  ylab = "Mean Scaled Expression", main = "Effect of Haplotype at each Age",
  group.cols = hap.cols)
add_zero_line()
```

Bar plot showing the same thing.

```{r kl_by_age_summary, fig.width = 7, fig.height = 5}
barplot(t(kl.effect.mat), beside = TRUE, legend = TRUE, col = hap.cols,
  ylab = "Mean Scaled Expression")
abline(h = 0)
```

the following plot shows the effect of age in each Klotho haplotype.

```{r, age_kl_summ, fig.width = 10, fig.height = 5}
plot.grouped.boxes(kl.expr.by.age, ylab = "Mean Scaled Expression",
  main = "Effect of Age in each Haplotype")
add_zero_line()
```

The bar plot below shows the same thing in bars.

```{r age_by_kl_summary, fig.width = 7, fig.height = 5}
barplot(kl.effect.mat, beside = TRUE, legend = TRUE, col = hap.cols,
  ylab = "Mean Scaled Expression")
abline(h = 0)
```
