---
title: "Validation from organoids"
author: Anna L Tyler
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  html_document:
    code_folding: hide
    collapsed: no
    toc: yes
    toc_float: yes
---

## Introduction

This workflow is a grasp at some kind of human validation
of the mouse results. The human data sets we have are not
well powered. This organoid set does not look at the variants,
but could at least tell us if different gene sets are affected
by Klotho levels in brain organoids.

```{r param}
rm(list = ls())
library(here)
```


```{r load_code}
all.fun <- list.files(here("Code"), full.names = TRUE)
for(i in 1:length(all.fun)){source(all.fun[i])}
```

```{r libraries, message = FALSE, warning = FALSE, error = FALSE}
needed.libraries <- c("pheatmap", "vioplot", "stringr")
load_libraries(needed.libraries)
```

```{r Geo2R}
# Version info: R 4.2.2, Biobase 2.58.0, GEOquery 2.66.0, limma 3.54.0
################################################################
#   Differential expression analysis with DESeq2
library(DESeq2)

# load counts table from GEO
urld <- "https://www.ncbi.nlm.nih.gov/geo/download/?format=file&type=rnaseq_counts"
tbl.file <- here("Data", "Human", "Organoids", "GSE171719_raw_counts_GRCh38.p13_NCBI.RDS")
if(!file.exists(tbl.file)){
     path <- paste(urld, "acc=GSE171719", "file=GSE171719_raw_counts_GRCh38.p13_NCBI.tsv.gz", sep="&");
     tbl <- as.matrix(data.table::fread(path, header=T, colClasses="integer"), rownames="GeneID")
     saveRDS(tbl, tbl.file)
}else{
     tbl <- readRDS(tbl.file)
}

# load gene annotations 
annot.file <- here("Data", "Human", "Organoids", "Human.GRCh38.p13.annot.RDS")
if(!file.exists(annot.file)){
     apath <- paste(urld, "type=rnaseq_counts", "file=Human.GRCh38.p13.annot.tsv.gz", sep="&")
     annot <- data.table::fread(apath, header=T, quote="", stringsAsFactors=F, data.table=F)
     rownames(annot) <- annot$GeneID
     saveRDS(annot, annot.file)
}else{
     annot <- readRDS(annot.file)
}


# sample selection
gsms <- "000111"
sml <- strsplit(gsms, split="")[[1]]

# group membership for samples
gs <- factor(sml)
groups <- make.names(c("KL","Untreated"))
levels(gs) <- groups
sample_info <- data.frame(Group = gs, row.names = colnames(tbl))

# pre-filter low count genes
# keep genes with at least N counts > 10, where N = size of smallest group
keep <- rowSums( tbl >= 10 ) >= min(table(gs))
tbl <- tbl[keep, ]

ds <- DESeqDataSetFromMatrix(countData=tbl, colData=sample_info, design= ~Group)

ds <- DESeq(ds, test="Wald", sfType="poscount")

counts <- vst(tbl)

# extract results for top genes table
r <- results (ds, contrast=c("Group", groups[1], groups[2]), alpha=0.05, pAdjustMethod ="fdr")

tT <- r[order(r$padj),] 
tT <- merge(as.data.frame(tT), annot, by=0, sort=F)

tT <- subset(tT, select=c("GeneID","padj","pvalue","lfcSE","stat","log2FoldChange","baseMean","Symbol","Description"))
#write.table(tT, file=stdout(), row.names=F, sep="\t")
```

## Differential expression of gene sets of interest {.tabset .tabset-fade .tabset-pills}

```{r gene_sets, results = "asis", fig.width = 12, fig.height = 8}

#gene.sets <- c("Glutamate_Genes", "Ribosome-Synapse_Genes", "Parkinson_disease-Mitochondrial_Metabolism_Genes")
gene.set.files <- list.files("/Users/atyler/Documents/Projects/Klotho/klotho_analysis/Results/age_batch_12_gene/validation_gene_sets",
     full.names = TRUE, pattern = "Genes.txt")
gene.sets <- gsub(".txt", "", basename(gene.set.files))

layout(matrix(c(0,1,3,2,2,2), byrow = TRUE, nrow = 2))
for(g in 1:length(gene.sets)){
     cat("###", gene.sets[g], "\n")
     #validation genes
     val.genes <- read.table(here("Results", "age_batch_12_gene", "validation_gene_sets", 
          paste0(gene.sets[g], ".txt")))[,1]
     
     ortho.file <- here("Results", "age_batch_12_gene", 
          "validation_gene_sets", paste0(gene.sets[g], "_Orthologs.csv"))
     
     gene.orthos <- get_orthologs(mouse.id = val.genes, file.name = ortho.file)
     gene.names <- gene.orthos[,"human_external_gene_name"]
     #gene.names <- c("TNFAIP8L3", "IL10RA", "IL13RA2", "PDGFB", "NFIB", "NFIA", "PDGFC", "CX3CL1", "ATM", "NFIL3", "NFIC", "IL17RD", "ILF2", "TMEM30A", "ARRDC4", "SLC9A7", "ELMOD1")

     gene.idx <- match(gene.names, tT[,"Symbol"])
     gene.result <- tT[gene.idx,]
     
     xmin <- min(gene.result[,"log2FoldChange"], na.rm = TRUE)
     xmax <- max(gene.result[,"log2FoldChange"], na.rm = TRUE)
     plot.width <- xmax - xmin
     plot.max <- xmax + (plot.width*0.2)
     plot.min <- max(abs(c(xmin, xmax)))*-1

     plot(gene.result[,"log2FoldChange"], -log10(gene.result[,"padj"]), pch = 16,
          xlab = "log2FC", ylab = "-log10(p)", xlim = c(plot.min, plot.max),
          main = gene.sets[g])
     text(x = gene.result[,"log2FoldChange"], y = -log10(gene.result[,"padj"]), 
          labels = gene.result[,"Symbol"], pos = 4)
     plot.dim <- par("usr")
     segments(x0 = 0, y0 = plot.dim[3], y1 = plot.dim[4])

     gene.id <- annot[match(gene.names, annot[,"Symbol"]), "GeneID"]
     expr.idx <- match(gene.id, rownames(counts))
     set.expr <- t(apply(log2(counts[expr.idx,]), 1, scale))
     rownames(set.expr) <- gene.names
     colnames(set.expr) <- colnames(counts)
     
     #annot.df <- data.frame("Treatment" = gs)     
     #rownames(annot.df) <- colnames(set.expr)
     #set.mean <- colMeans(set.expr)
     #pheatmap(set.expr, annotation_col = annot.df, cluster_cols = FALSE, scale = "row", cluster_rows = FALSE)

     #plot.decomp(t(set.expr))
     
     by.treatment <- lapply(groups, 
          function(x) t(set.expr[,which(gs == x)]))
     names(by.treatment) <- groups
     untreated.order <- order(colMeans(by.treatment$Untreated))
     ordered.set <- lapply(by.treatment[2:1], function(x) x[,untreated.order])
     plot.grouped.boxes(ordered.set, type = "matrix", group.cols = c("#eff3ff", "#4292c6"))
     add_zero_line()  

     #boxplot(by.treatment[2:1], main = gene.sets[g])
     gene.means <- lapply(2:1, function(x) colMeans(by.treatment[[x]]))
     vioplot(gene.means, names = names(by.treatment)[2:1], col = "lightgray")
     stripchart(gene.means, vertical = TRUE, pch = 16, col = "#dd1c77", 
          method = "jitter", add = TRUE)
     mtext("Expression", side = 2, line = 2.5)
     add_zero_line()
     cat("\n\n")
}
```


```{r figures}
#just volcano plots for paper figure

pdf(here("Results", "for_paper", "Human_Validation_Organoids.pdf"), width = 6, height = 9)
par(mfrow = c(3,2))
for(g in 1:length(gene.sets)){
     #validation genes
     val.genes <- read.table(here("Results", "age_batch_12_gene", "validation_gene_sets", 
          paste0(gene.sets[g], ".txt")))[,1]
     
     ortho.file <- here("Results", "age_batch_12_gene", 
          "validation_gene_sets", paste0(gene.sets[g], "_Orthologs.csv"))
     
     gene.orthos <- get_orthologs(mouse.id = val.genes, file.name = ortho.file)
     gene.names <- gene.orthos[,"human_external_gene_name"]
     #gene.names <- c("TNFAIP8L3", "IL10RA", "IL13RA2", "PDGFB", "NFIB", "NFIA", "PDGFC", "CX3CL1", "ATM", "NFIL3", "NFIC", "IL17RD", "ILF2", "TMEM30A", "ARRDC4", "SLC9A7", "ELMOD1")

     gene.idx <- match(gene.names, tT[,"Symbol"])
     gene.result <- tT[gene.idx,]
     
     xmin <- min(gene.result[,"log2FoldChange"], na.rm = TRUE)
     xmax <- max(gene.result[,"log2FoldChange"], na.rm = TRUE)
     plot.width <- xmax - xmin
     plot.max <- xmax + (plot.width*0.4)
     plot.min <- max(abs(c(xmin, xmax)))*-1

     ymin <- min(-log10(gene.result[,"padj"]), na.rm = TRUE)
     ymax <- max(-log10(gene.result[,"padj"]), na.rm = TRUE)*1.05

     split.name <- strsplit(gene.sets[g], "-")
     k.name <- gsub("_", " ", split.name[[1]][1])
     bd.name <- gsub("_", " ", gsub("_Genes", "", split.name[[1]][2]))
     plot.label <- paste("KEGG:", k.name, "\nBiodomain:", bd.name)
     plot(gene.result[,"log2FoldChange"], -log10(gene.result[,"padj"]), pch = 16,
          xlab = "log2FC", ylab = "-log10(p)", xlim = c(plot.min, plot.max),
          main = "", ylim = c(ymin, ymax))
     mtext(plot.label, side = 3, line = 0.8)
     text(x = gene.result[,"log2FoldChange"], y = -log10(gene.result[,"padj"]), 
          labels = gene.result[,"Symbol"], pos = 4)
     plot.dim <- par("usr")
     segments(x0 = 0, y0 = plot.dim[3], y1 = plot.dim[4])

     gene.id <- annot[match(gene.names, annot[,"Symbol"]), "GeneID"]
     expr.idx <- match(gene.id, rownames(counts))
     set.expr <- t(apply(log2(counts[expr.idx,]), 1, scale))
     #set.expr <- log2(counts[expr.idx,])
     rownames(set.expr) <- gene.names
     colnames(set.expr) <- colnames(counts)

     by.treatment <- lapply(groups, 
          function(x) t(set.expr[,which(gs == x)]))
     names(by.treatment) <- groups
     #boxplot(by.treatment[2:1], cex.axis = 1.5, ylab = "Average Normalized Expression",
     #     cex.lab = 1.2)
     #mtext(plot.label, side = 3, line = 0.8)
     #abline(h = 0)

     gene.means <- lapply(2:1, function(x) colMeans(by.treatment[[x]]))
     vioplot(gene.means, names = c("", ""), 
          col = "lightgray", cex.axis = 1.5, cex.lab = 1.2)
     plot.dim <- par("usr")
     plot.height <- plot.dim[4] - plot.dim[3]
     par(xpd = NA)
     text(x = c(1,2), y = plot.dim[3]-(plot.height*0.15),
          labels = paste(names(by.treatment)[2:1], "\n(3)"), cex = 1.5)
     stripchart(gene.means, vertical = TRUE, pch = 16, col = "#dd1c77", 
          method = "jitter", add = TRUE, cex = 1.5)
     mtext(plot.label, side = 3, line = 0.8)
     mtext("Expression (A.U.)", side = 2, line = 2.5, cex = 1.2)
     add_zero_line()
}
dev.off()
```

```{r sandbox, eval = FALSE}
#just checking that the fold change direction matches that in the paper
gene.name <- "SPARCL1" #up with KL added
gene.name <- "FGL2" #up with KL
gene.name <- "CDH7" #down with KL added
gene.name <- "GAPDH"

gene.idx <- match(gene.name, tT[,"Symbol"])
tT[gene.idx,]

gene.id <- annot[match(gene.name, annot[,"Symbol"]), "GeneID"]
expr.idx <- match(gene.id, rownames(counts))
#boxplot(log(counts[expr.idx,])~gs)
barplot(counts[expr.idx,], col = as.numeric(gs), names = sample_info[,1], main = gene.name)
```

It would seem that glutamate receptor genes go down 
when Klotho is induced and ribosomal genes go up. 
The mitochondrial genes don't have a coherent direction. 

In Figure 5C of the paper, they show KEGG pathways 
that were enriched among the up and down-regulated
genes in the organoids. They have both nicotine addiction
and glutamatergic synapses going down, which matches our 
VS mice.

The glutamate receptor genes go down in VS animals,
and when paired with this data set, that would seem
to indicate more KL in the brain. Very interesting!

They don't talk about this specifically in the paper,
but they cite another paper about astrocyte senescence
promoting glutamate toxicity.


## More broadly

Maybe we can look at all of the KEGG-Biodomain intersections
to see how things compare more broadly.


```{r broad}
biadj <- readRDS("/Users/atyler/Documents/Projects/Klotho/klotho_analysis/Results/age_batch_12_gene/biadjacency_matrix.RDS")
bd.kegg <- readRDS("/Users/atyler/Documents/Projects/Klotho/klotho_analysis/Results/Processed_Data/Human_KEGG_Intersections_for_GSEA.RDS")

pairs.to.check <- which(biadj != 0, arr.ind = TRUE)
pairs.to.check[,1] <- rownames(biadj)[as.numeric(pairs.to.check[,1])]
pairs.to.check[,2] <- colnames(biadj)[as.numeric(pairs.to.check[,2])]

fc.mat <- mean.mat <- matrix(NA, nrow = nrow(biadj), ncol = ncol(biadj))
dimnames(fc.mat) <- dimnames(mean.mat) <- dimnames(biadj)

for(i in 1:nrow(pairs.to.check)){
     bd.name <- pairs.to.check[i,1]
     kegg.name <- pairs.to.check[i,2]
     bd.idx <- which(names(bd.kegg) == bd.name)
     k.idx <- which(names(bd.kegg[[bd.idx]]) == kegg.name)
     bd.k.genes <- bd.kegg[[bd.idx]][[k.idx]]
     gene.id <- annot[match(bd.k.genes, annot[,"EnsemblGeneID"]), "GeneID"]
     gene.id <- gene.id[which(!is.na(gene.id))]
     
     gene.idx <- match(gene.id, tT[,"GeneID"])
     gene.idx <- gene.idx[which(!is.na(gene.idx))]
     gene.result <- tT[gene.idx,]
     #plot(gene.result[,"log2FoldChange"], -log10(gene.result[,"padj"]), pch = 16);abline(v = 0)
     fc.mat[bd.name, kegg.name] <- mean(gene.result[,"log2FoldChange"], na.rm = TRUE)
     #fc <- gene.result[,"log2FoldChange"]; names(fc) <- gene.result[,"Symbol"]
     
     expr.idx <- match(gene.id, rownames(counts))
     set.expr <- t(apply(counts[expr.idx,], 1, scale))

     by.treatment <- lapply(groups, 
          function(x) t(set.expr[,which(gs == x)]))
     names(by.treatment) <- groups
     #plot.grouped.boxes(by.treatment[2:1], type = "matrix")
     by.gene <- t(sapply(1:nrow(set.expr), 
          function(x) c(mean(by.treatment$Untreated[,x], na.rm = TRUE), mean(by.treatment$KL[,x], na.rm = TRUE))))
     #mean.diff <- by.gene[,2] - by.gene[,1]; names(mean.diff) <- gene.result[,"Symbol"]
     #boxplot(by.treatment[2:1])
     #boxplot(by.gene)
     #kl.effect <- mean(by.gene[,2] - by.gene[,1], na.rm = TRUE)
     #kl.effect <- mean(by.treatment$KL, na.rm = TRUE) - mean(by.treatment$Untreated, na.rm = TRUE)
     kl.model <- lm.by.list(by.treatment[c(2:1)])
     kl.effect <- coef(summary(kl.model))[2,1]
     mean.mat[bd.name, kegg.name] <- kl.effect
}
```


```{r plot_mat}
heatmap.neg.col = "purple"
heatmap.pos.col = "brown"

png("~/Desktop/organoid_biadj1.png", width = 12, height = 5, units = "in", res = 300)
par(mar = c(12, 12, 0,0), bg = NA)
imageWithText(fc.mat, col.scale = list(heatmap.neg.col, heatmap.pos.col), 
    split.at.vals = TRUE, show.text = FALSE, row.text.cex = 0.8, col.text.cex = 0.5, 
    grad.dir = "ends", row.text.shift = 0.02, col.text.shift = 0.04, na.col = "white")
plot.dim <- par("usr")
draw.rectangle(min.x = 0.5, max.x = ncol(fc.mat)+0.5, 
     min.y = 0.5, max.y = nrow(fc.mat)+0.5)
dev.off()

png("~/Desktop/organoid_biadj2.png", width = 12, height = 5, units = "in", res = 300)
par(mar = c(12, 12, 0,0), bg = NA)
imageWithText(mean.mat, col.scale = list(heatmap.neg.col, heatmap.pos.col), 
    split.at.vals = TRUE, show.text = FALSE, row.text.cex = 0.8, col.text.cex = 0.5, 
    grad.dir = "ends", row.text.shift = 0.02, col.text.shift = 0.04, na.col = "white")
plot.dim <- par("usr")
draw.rectangle(min.x = 0.5, max.x = ncol(mean.mat)+0.5, 
     min.y = 0.5, max.y = nrow(mean.mat)+0.5)
dev.off()
```

```{r plotting_fun}

#testing
common_points <- function(arr.list){
     all.pts <- unique(Reduce("rbind", arr.list))
     common.pts <- NULL
     for(i in 1:nrow(all.pts)){
          inlist <- lapply(arr.list, function(x) intersect(which(x[,1] == all.pts[i,1]), which(x[,2] == all.pts[i,2])))
          inboth <- all(sapply(inlist, length) > 0)
          if(inboth){
               common.pts <- rbind(common.pts, all.pts[i,])
          }
     }
     return(common.pts)
}

spread_coord <- function(coord.vals, min.dist, nudge.dist){
     
     coord.dist <- as.matrix(dist(coord.vals))
     diag(coord.dist) <- NA
     coord.dist[upper.tri(coord.dist)] <- NA
     need.to.nudge <- as.logical(length(which(coord.dist < min.dist)))
     new.coord <- coord.vals

     while(need.to.nudge){
          too.close <- unique(as.vector(which(coord.dist < min.dist, arr.ind = TRUE)))
          old.coord <- coord.vals[too.close]
          
          #nudge the minimum one down, and the maximum one up
          new.min <- min(old.coord) - nudge.dist
          new.max <- max(old.coord) + nudge.dist
          #print(cbind(new.min, new.max))
          #spread out coordinates evenly
          adj.coord <- segment_region(new.min, new.max, length(old.coord), "ends")
          #put the new coordinates in the same order as the old
          old.order <- order(old.coord)
          
          new.coord[too.close[old.order]] <- adj.coord

          coord.dist <- as.matrix(dist(new.coord))
          diag(coord.dist) <- NA
          coord.dist[upper.tri(coord.dist)] <- NA
          need.to.nudge <- as.logical(length(which(coord.dist < coord.dist)))
     }
     
     return(new.coord)
}

get_coord <- function(arr.idx){
     arr.labels <- arr.idx
     arr.labels[,1] <- rownames(mean.mat)[arr.idx[,1]]
     arr.labels[,2] <- colnames(mean.mat)[arr.idx[,2]]

     label.names <- apply(arr.labels, 1, function(x) paste(x, collapse = " - "))

     x.coord <- apply(arr.idx, 1, function(x) biadj[x[1], x[2]])
     y.coord <- apply(arr.idx, 1, function(x) mean.mat[x[1], x[2]])
     coord.table <- cbind("x" = x.coord, "y" = y.coord)
     rownames(coord.table) <- label.names
     return(coord.table)
}


add_labels <- function(arr.idx, label.nudges = NULL,
     pos = 4, cex = 0.5, min.y.dist = 0.1, nudge.dist = 0.05,
     col = "black"){
     
     pt.coord <- get_coord(arr.idx)
     pt.x <- pt.coord[,"x"]
     pt.y <- pt.coord[,"y"]
     label.names <- rownames(pt.coord)
     label.y <- spread_coord(coord.vals = pt.y, min.dist = min.y.dist, nudge.dist)
     #plot(label.y, pt.y)

     if(!is.null(label.nudges)){
          for(i in 1:length(label.nudges)){
               label.idx <- which(names(label.y) == names(label.nudges)[i])
               label.y[label.idx] <- label.y[label.idx]+label.nudges[i]
          }
     }
     
     text(pt.x, label.y, labels = label.names, pos = pos, cex = cex, col = col)
     
     #draw lines from the labels to the dots
     plot.dim <- par("usr")
     plot.width <- plot.dim[2] - plot.dim[1]
     pt.diff <- which(label.y - pt.y != 0)
     label.x <- pt.x

     #add a bit to label x if in position 4
     #subtract a bit if in position 2
     if(pos == 4){
          label.x[pt.diff] <- label.x[pt.diff]+(plot.width*0.012)
     }
     if(pos == 2){
          label.x[pt.diff] <- label.x[pt.diff]-(plot.width*0.012)
     }
     segments(x0 = pt.x, x1 = label.x, y0 = pt.y, y1 = label.y)
     invisible(label.names)
}
```

The following plot shows the comparison between the effects
of the VS allele in the mice and the effects of overexpressed
Klotho in organoids. Some of the points with similar effects 
are labeled.

```{r labeled_comp, fig.width = 7, fig.height = 7}

#quartz(width = 7, height = 7)
plot(as.vector(biadj), as.vector(mean.mat), pch = 16, 
     xlab = "Effect of VS allele in mice", xlim = c(-0.7, 1.5),
     ylab = "Effect of KL induction in human organoids")
abline(h = 0, v = 0)


#abline(h = -0.8) #below this line for common negative effects
both.neg <- which(mean.mat < -0.7, arr.ind = TRUE)
add_labels(arr.idx = both.neg, min.y.dist = 0.08, nudge.dist = 0.01,
     label.nudges = c("Autophagy - Huntington disease" = -0.07))

each.pos <- list(which(mean.mat > 0, arr.ind = TRUE), which(biadj > 0, arr.ind = TRUE))
both.pos <- common_points(each.pos)
add_labels(both.pos, min.y.dist = 0.08, nudge.dist = 0.01)

pos.neg.list <- list(which(mean.mat < 0, arr.ind = TRUE), which(biadj > 0, arr.ind = TRUE))
pos.neg <- common_points(pos.neg.list)
#add_labels(pos.neg, min.y.dist = 0.08, nudge.dist = 0.01)

#just wondering what this dot was. It's far enough away to label clearly
neg.pos.list <- list(which(biadj < 0, arr.ind = TRUE), which(mean.mat > 0.7, arr.ind = TRUE))
neg.pos <- common_points(neg.pos.list) 
add_labels(neg.pos, min.y.dist = 0.08, nudge.dist = 0.01)

#add a label for the mitochondrial blob
draw.rectangle(0.22, 0.64, -0.72, -0.02)
text(x = 0.65, y = mean(c(-0.7, -0.05)), 
     labels = "Mitochondrial Metabolism\nand Oxidative Stress", cex = 0.5,
     adj = 0)

```

```{r for_fig}
#shorten some names
colnames(biadj) <- colnames(mean.mat) <- gsub("Coronavirus disease - ", "", colnames(mean.mat))

pdf(here("Results", "for_paper", "human_mouse_comparison.pdf"), width = 8, height = 7)
#quartz(width = 7, height = 7)
mouse.effects <- as.vector(biadj)
org.effects <- as.vector(mean.mat)

text.cex = 0.7
min.dist = 0.2

plot.new()
plot.window(xlim = c(-0.7, 2),
     ylim = c(min(org.effects, na.rm = TRUE), max(org.effects, na.rm = TRUE)))

abline(h = 0, v = 0, col = "gray")
#abline(v = seq(-1, 1.5, 0.5), col = "gray", lty = 2)
#abline(h = seq(-1.5, 1.5, 0.5), col = "gray", lty = 2)

#add border box
#draw.rectangle(-1.1, 1.5, -1.5, 1.5)

#add a label for the mitochondrial blob
draw.rectangle(0.22, 0.64, -0.72, -0.02)
text(x = 0.65, y = mean(c(-0.7, -0.05)), 
     labels = "Mitochondrial Metabolism\nand Oxidative Stress", cex = text.cex,
     adj = 0)

points(mouse.effects, org.effects, pch = 16)
axis(1)
axis(2)
mtext("Effect of VS allele in mice", side = 1, line = 2.5)
mtext("Effect of KL induction in human organoids", side = 2, line = 2.5)

#abline(h = -0.8) #below this line for common negative effects
both.neg <- which(mean.mat < -0.68, arr.ind = TRUE)
add_labels(both.neg, min.y.dist = min.dist, nudge.dist = 0.01, cex = text.cex,
     label.nudges = c("Autophagy - Huntington disease" = -0.07))

each.pos <- list(which(mean.mat > 0, arr.ind = TRUE), which(biadj > 0, arr.ind = TRUE))
both.pos <- common_points(each.pos)
add_labels(both.pos, min.y.dist = min.dist, nudge.dist = 0.01, cex = text.cex)

#lighten the biggest blob.
big.group <- list(which(biadj < 0, arr.ind = TRUE), which(mean.mat > -0.68, arr.ind = TRUE))
big.idx <- common_points(big.group)
big.coord <- get_coord(big.idx)
points(big.coord, col = "gray", pch = 16)
dev.off()


```


```{r for_fig_alt}
#a version with group labels instead of individual labels

pdf(here("Results", "for_paper", "human_mouse_comparison_alt.pdf"), width = 8, height = 7)
#quartz(width = 7, height = 7)
mouse.effects <- as.vector(biadj)
org.effects <- as.vector(mean.mat)

text.cex = 1.2
min.dist = 0.2

plot.new()
plot.window(xlim = c(-0.7, 1.5),
     ylim = c(min(org.effects, na.rm = TRUE), max(org.effects, na.rm = TRUE)))

abline(h = 0, v = 0, col = "gray")
#abline(v = seq(-1, 1.5, 0.5), col = "gray", lty = 2)
#abline(h = seq(-1.5, 1.5, 0.5), col = "gray", lty = 2)

#add a label for the mitochondrial blob
draw.rectangle(0.22, 0.64, -0.72, -0.02)
text(x = 0.66, y = mean(c(-0.7, -0.05)), 
     labels = "Mitochondrial Metabolism\nand Oxidative Stress", 
     cex = text.cex, adj = 0)

#add a label for APP-synapse blob
draw.rectangle(-0.67, -0.16, -1.45, -0.58)
#add a backing rectangle to cover up 0 line
draw.rectangle(-0.14, 0.5, -1.32, -1, fill = "white", border.col = NA)
text(x = -0.14, y = mean(c(-1.45, -0.87)), 
     labels = "APP Metabolism\nand Synapse", 
     cex = text.cex, adj = 0)

#add a label for ribosomal blob
draw.rectangle(0.3, 0.9, 0.25, 1.5)
text(x = 0.92, y = mean(c(0.25, 1.5)), 
     labels = "Ribosome and\nProteostasis", cex = text.cex, adj = 0)

points(mouse.effects, org.effects, pch = 16)
axis(1)
axis(2)
mtext("Effect of VS allele in mice", side = 1, line = 2.5, cex = 1.5)
mtext("Effect of KL induction in human organoids", side = 2, line = 2.5, cex = 1.5)

#lighten the biggest blob.
big.group <- list(which(biadj < 0, arr.ind = TRUE), which(mean.mat > -0.58, arr.ind = TRUE))
big.idx <- common_points(big.group)
big.coord <- get_coord(big.idx)
points(big.coord, col = "gray", pch = 16)
dev.off()
```


```{r for_fig_all_info}
#Greg thinks we can cram all the information on there! Let's try.

pdf(here("Results", "for_paper", "human_mouse_comparison_all.pdf"), width = 8, height = 7)
#quartz(width = 7, height = 7)
mouse.effects <- as.vector(biadj)
org.effects <- as.vector(mean.mat)

small.cex = 0.7
big.cex  <- 1.2
min.dist = 0.2
box.col <- "darkgray"
box.lwd <- 2
box.lty = 2
#cat.cols <- c("#fb9a99", "#fdbf6f", "#de77ae")
cat.cols <- rep("#de77ae", 3)

plot.new()
plot.window(xlim = c(-0.7, 1.5),
     ylim = c(min(org.effects, na.rm = TRUE), max(org.effects, na.rm = TRUE)))

abline(h = 0, v = 0, col = "gray")
#abline(v = seq(-1, 1.5, 0.5), col = "gray", lty = 2)
#abline(h = seq(-1.5, 1.5, 0.5), col = "gray", lty = 2)

#add a label for APP-synapse blob
draw.rectangle(-0.67, 0.69, -1.45, -0.68, border = cat.cols[1], lwd = box.lwd, 
     fill = "white", lty = box.lty)
text(x = 0.74, y = mean(c(-1.45, -0.87)), 
     labels = "APP Metabolism\nand Synapse", 
     cex = text.cex, adj = 0, col = cat.cols[1])

#add APP individual labels
both.neg <- which(mean.mat < -0.68, arr.ind = TRUE)
add_labels(both.neg, min.y.dist = min.dist, nudge.dist = 0.01, cex = small.cex,
     label.nudges = c("Autophagy - Huntington disease" = -0.07))


#add a label for the mitochondrial blob after APP blob and fill in white
draw.rectangle(0.22, 0.64, -0.72, -0.02, fill = "white", border = cat.cols[2], 
     lwd = box.lwd, lty = box.lty)
text(x = 0.66, y = mean(c(-0.7, -0.05)), 
     labels = "Mitochondrial Metabolism\nand Oxidative Stress", 
     cex = big.cex, adj = 0, col = cat.cols[2])

#add points
points(mouse.effects, org.effects, pch = 16)
axis(1)
axis(2)
mtext(bquote("Effect Size ("~beta~"), Mouse"), side = 1, line = 2.5, cex = 1.5)
mtext(bquote("Effect Size ("~beta~"), Organoid"), side = 2, line = 2.5, cex = 1.5)

#add labels for ribosomal blob
each.pos <- list(which(mean.mat > 0, arr.ind = TRUE), which(biadj > 0, arr.ind = TRUE))
both.pos <- common_points(each.pos)
add_labels(both.pos, min.y.dist = min.dist, nudge.dist = 0.01, cex = small.cex)

#add a label for ribosomal blob
draw.rectangle(0.35, 1.4, 0.25, 1.5, border = cat.cols[3], lwd = box.lwd, 
     lty = box.lty)
par(xpd = NA)
text(x = mean(c(0.35, 1.4)), y = 1.6, col = cat.cols[3],
     labels = "Ribosome and Proteostasis", cex = text.cex, adj = 0.5)

#lighten the biggest blob.
big.group <- list(which(biadj < 0, arr.ind = TRUE), which(mean.mat > -0.68, arr.ind = TRUE))
big.idx <- common_points(big.group)
big.coord <- get_coord(big.idx)
points(big.coord, col = "gray", pch = 16)

#add correlation statistics to top left
cor.stats <- cor.test(mouse.effects, org.effects, 
     use = "pairwise.complete.obs", method = "spearman")
r <- cor.stats$estimate
p <- cor.stats$p.value
text(-0.65, 1.45, labels = bquote(rho ==.(signif(r, 2))), adj = 0)
text(-0.65, 1.3, labels = bquote(italic(p) ==.(signif(p, 2))), adj = 0)

#bounding box for stats
draw.rectangle(-0.67, -0.39, 1.22, 1.52, border = box.col, lwd = box.lwd)
dev.off()

#make table for supplements
did.measure <- which(!is.na(biadj), arr.ind = TRUE)
measured.labels <- sapply(1:2, function(x) dimnames(biadj)[[x]][did.measure[,x]])
mouse.vals <- as.vector(biadj[which(!is.na(biadj))])
organoid.vals <- as.vector(mean.mat[which(!is.na(mean.mat))])
comp.table <- cbind(measured.labels, mouse.vals, organoid.vals)
colnames(comp.table) <- c("Biodomain", "KEGG", "Effect of KL haplotype in mice", "Effect of KL induction in organoids")
write_table_with_header(comp.table, 
     here("Results", "for_paper", "Supplemental_File_Organoid_Comparison.txt"),
     header = "Comparison of the effect of the KL haplotype in mice and 
     the effects of KL induction in human iPSC-derived brain organoids
     on mean gene expression of groups of genes defined as the intersection
     between a Biodomain and a KEGG pathway. Intersections shown in Figure
     6 are included. Effects are the beta coefficients from linear models
     as described in the methods.")
```


Here is just the mitochondrial blob labeled,
so we can see what's in there. There are so many points to label,
that the labels spread into the upper quadrant, but we 
are not actually labeling any of the dots in that quadrant.
Most of the points have something to do with mitochondria,
oxidative stress and metal binding. I'd like to summarize this
group of dots in the figure.

```{r mito_blob, fig.width = 9, fig.height = 7}

#quartz(width = 7, height = 7)
plot(as.vector(biadj), as.vector(mean.mat), pch = 16, 
     xlab = "Effect of VS allele in mice", xlim = c(-0.7, 1.5),
     ylab = "Effect of KL induction in human organoids")
abline(h = 0, v = 0)

pos.neg.list <- list(which(mean.mat < 0, arr.ind = TRUE), which(biadj > 0, arr.ind = TRUE))
pos.neg <- common_points(pos.neg.list)
add_labels(arr.idx = pos.neg, min.y.dist = 0.05, nudge.dist = 0.6, cex = 0.7)

#test more of neg-neg intersections
#neg.neg.list <- list(which(mean.mat < -0.5, arr.ind = TRUE), which(mean.mat > -1, arr.ind = TRUE), which(biadj < 0, arr.ind = TRUE))
#neg.neg <- common_points(neg.neg.list)
#add_labels(arr.idx = neg.neg, min.y.dist = 0.05, nudge.dist = 0.3, cex = 0.7)


```

```{r other_checks, eval = FALSE}
plot(as.vector(biadj), as.vector(mean.mat), pch = 16, 
     xlab = "Effect of VS allele in mice", xlim = c(-0.7, 1.5),
     ylab = "Effect of KL induction in human organoids")
abline(h = 0, v = 0)

#test non-movers in organoids - do these need more body systems?
neg.non.list <- list(which(mean.mat > -0.1, arr.ind = TRUE), which(mean.mat < 0.1, arr.ind = TRUE), which(biadj < 0, arr.ind = TRUE))
neg.non <- common_points(neg.non.list)
idx <- row_col_to_idx(neg.non[,1], neg.non[,2], nrow(mean.mat))
points(biadj[idx], mean.mat[idx], col = "red")
test <- add_labels(arr.idx = neg.non, min.y.dist = 0.1, nudge.dist = 0.6, cex = 0.7, 
     col = "red")


#what about positive in organoid and negative in mice
neg.pos.list <- list(which(mean.mat > 0.1, arr.ind = TRUE), which(biadj < 0, arr.ind = TRUE))
neg.pos <- common_points(neg.pos.list)
idx <- row_col_to_idx(neg.pos[,1], neg.pos[,2], nrow(mean.mat))
points(biadj[idx], mean.mat[idx], col = "red")
test <- add_labels(arr.idx = neg.pos, min.y.dist = 0.1, nudge.dist = 0.6, cex = 0.7, 
     col = "red")

neg.blob.list <- list(which(mean.mat < 0, arr.ind = TRUE), which(mean.mat > -0.75, arr.ind = TRUE), which(biadj < 0, arr.ind = TRUE))
neg.blob <- common_points(neg.blob.list)
idx <- row_col_to_idx(neg.blob[,1], neg.blob[,2], nrow(mean.mat))
points(biadj[idx], mean.mat[idx], col = "red")
test <- add_labels(arr.idx = neg.blob, min.y.dist = 0.1, nudge.dist = 0.6, cex = 0.7, 
     col = "red")


split.test <- strsplit(test, " - ")
test.bd <- sapply(split.test, function(x) x[1])
test.k <- sapply(split.test, function(x) x[2])

par(mar = c(4, 14, 4, 2))
barplot(sort(table(test.bd)), las = 2, horiz = TRUE)
barplot(sort(table(test.k)), las = 2, horiz = TRUE)
```

## Check significance?

It occurred to me that with so many gene groups, we could 
have identified differential expression by chance. Do we 
need to check the significance of this finding overall?
Given random sets of genes, how likely is it that we would
see this concordance by chance?

How do we even test this? It seems like a lot of 
calculations. We could draw some number of sets of 
10 genes at random from each data set and see how
often we get concordance of the effects. But if the
genes really do have the same effect, we will find
stuff even in random sets. In fact, if all genes 
were concordant, every random draw would be concordant...


## check all

I'm curious about whether the glutamate genes and the
ribosomal signals are really the largest. I suspect they 
are, but I wanted to look even more broadly.
This code looks across all intersections.

```{r all}
u_bd <- names(bd.kegg)
#remove tau homeostasis, it's too small
u_bd <- u_bd[which(u_bd != "Tau Homeostasis")]
u_k <- unique(unlist(lapply(bd.kegg, names)))
all.mat <- matrix(NA, nrow = length(u_bd), ncol = length(u_k))
rownames(all.mat) <- u_bd
colnames(all.mat) <- u_k

for(b in 1:length(bd.kegg)){
     bd.name <- names(bd.kegg)[b]
     print(bd.name)
     for(k in 1:length(bd.kegg[[b]])){
          kegg.name <- names(bd.kegg[[b]])[k]
          bd.idx <- which(names(bd.kegg) == bd.name)
          k.idx <- which(names(bd.kegg[[bd.idx]]) == kegg.name)
          bd.k.genes <- bd.kegg[[bd.idx]][[k.idx]]
          gene.id <- annot[match(bd.k.genes, annot[,"EnsemblGeneID"]), "GeneID"]
          gene.id <- gene.id[which(!is.na(gene.id))]
     
          gene.idx <- match(gene.id, tT[,"GeneID"])
          gene.idx <- gene.idx[which(!is.na(gene.idx))]
          if(length(gene.idx) < 10){
               next()
          }
          gene.result <- tT[gene.idx,]
     
          expr.idx <- match(gene.id, rownames(counts))
          set.expr <- t(apply(counts[expr.idx,], 1, scale))

          by.treatment <- lapply(groups, 
               function(x) t(set.expr[,which(gs == x)]))
          names(by.treatment) <- groups
     
          by.gene <- t(sapply(1:nrow(set.expr), 
               function(x) c(mean(by.treatment$Untreated[,x], na.rm = TRUE), mean(by.treatment$KL[,x], na.rm = TRUE))))
          kl.model <- lm.by.list(by.treatment[c(2:1)])
          kl.effect <- coef(summary(kl.model))[2,1]
          all.mat[bd.name, kegg.name] <- kl.effect
     }
}
#remove empty cols
has.vals <- which(apply(all.mat, 2, function(x) length(which(!is.na(x)))) > 1)
plot.mat <- all.mat[,has.vals]
row.order <- order(decomp_with_nas(plot.mat, label.points = TRUE, plot.results = FALSE)$u[,1])
col.order <- order(decomp_with_nas(t(plot.mat), label.points = TRUE, add.zeros = TRUE, plot.results = FALSE)$u[,1])
pheatmap(plot.mat[row.order, col.order], cluster_rows = FALSE, cluster_cols = FALSE)

high.test <- which(plot.mat > 1, arr.ind = TRUE)
high.test.labels <- sapply(1:2, function(x) dimnames(plot.mat)[[x]][high.test[,x]])

low.test <- which(plot.mat < -1, arr.ind = TRUE)
low.test.labels <- sapply(1:2, function(x) dimnames(plot.mat)[[x]][low.test[,x]])

```

```{r the_rest, eval = FALSE}
plotDispEsts(ds, main="GSE171719 Dispersion Estimates")

# create histogram plot of p-values
hist(r$padj, breaks=seq(0, 1, length = 21), col = "grey", border = "white", 
         xlab = "", ylab = "", main = "GSE171719 Frequencies of padj-values")

# volcano plot
old.pal <- palette(c("#00BFFF", "#FF3030")) # low-hi colors
par(mar=c(4,4,2,1), cex.main=1.5)
plot(r$log2FoldChange, -log10(r$padj), main=paste(groups[1], "vs", groups[2]),
     xlab="log2FC", ylab="-log10(Padj)", pch=20, cex=0.5)
with(subset(r, padj<0.05 & abs(log2FoldChange) >= 0),
     points(log2FoldChange, -log10(padj), pch=20, col=(sign(log2FoldChange) + 3)/2, cex=1))
legend("bottomleft", title=paste("Padj<", 0.05, sep=""), legend=c("down", "up"), pch=20,col=1:2)

# MD plot
par(mar=c(4,4,2,1), cex.main=1.5)
plot(log10(r$baseMean), r$log2FoldChange, main=paste(groups[1], "vs", groups[2]),
     xlab="log10(mean of normalized counts)", ylab="log2FoldChange", pch=20, cex=0.5)
with(subset(r, padj<0.05 & abs(log2FoldChange) >= 0),
     points(log10(baseMean), log2FoldChange, pch=20, col=(sign(log2FoldChange) + 3)/2, cex=1))
legend("bottomleft", title=paste("Padj<", 0.05, sep=""), legend=c("down", "up"), pch=20,col=1:2)
abline(h=0)
palette(old.pal) # restore palette

################################################################
#   General expression data visualization
dat <- log10(counts(ds, normalized = T) + 1) # extract normalized counts

# box-and-whisker plot
lbl <- "log10(raw counts + 1)"
ord <- order(gs)  # order samples by group
palette(c("#1B9E77", "#7570B3", "#E7298A", "#E6AB02", "#D95F02",
          "#66A61E", "#A6761D", "#B32424", "#B324B3", "#666666"))
par(mar=c(7,4,2,1))
boxplot(dat[,ord], boxwex=0.6, notch=T, main="GSE171719", ylab="lg(norm.counts)", outline=F, las=2, col=gs[ord])
legend("topleft", groups, fill=palette(), bty="n")

# UMAP plot (multi-dimensional scaling)
library(umap)
dat <- dat[!duplicated(dat), ] # first remove duplicates
par(mar=c(3,3,2,6), xpd=TRUE, cex.main=1.5)
ump <- umap(t(dat), n_neighbors = 3, random_state = 123)
plot(ump$layout, main="UMAP plot, nbrs=3", xlab="", ylab="", col=gs, pch=20, cex=1.5)
legend("topright", inset=c(-0.15,0), legend=groups, pch=20,
       col=1:length(groups), title="Group", pt.cex=1.5)

```