---
title: "Klotho proteomics"
author: Anna L Tyler
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  html_document:
    code_folding: hide
    collapsed: no
    toc: yes
    toc_float: yes
---

## Introduction

The purpose of this workflow is to start looking at the proteomics
data from the Klotho mice.

```{r parameters}

rm(list = ls())

library(here)

general.data.dir <- here("Data", "general")
proteomics.data.dir <-  here("Data", "Proteomics")

use.carrier.status <- TRUE #if true, fits models with the ordered factor FC > WT > VS
                          #if false, fits modesl with the ordered factor FC/FC > FC/WT > WT/WT > WT/VS > VS/VS

general.processed.data.dir <- here("Results", "Processed_Data")

results.dir <- here("Results", "Proteomics")
if(!file.exists(results.dir)){dir.create(results.dir)}

processed.data.dir <- file.path(results.dir, "processed_data")
if(!file.exists(processed.data.dir)){dir.create(processed.data.dir)}

gene.dump.dir <- file.path(results.dir, "removed_genes")
if(!file.exists(gene.dump.dir)){dir.create(gene.dump.dir)}

tx.processed.data.dir <- here("Results", "all_gene", "processed_data")
mouse.data.dir <- here("Data", "Mouse")

adjust.ide = TRUE

compare.age <- 12

if(compare.age == 12){
    compare.age.text <- "twelve"
}
if(compare.age == 4){
    compare.age.text <- "four"
}

offline = FALSE #set to TRUE if offline. Enrichment queries will not be run.
```

```{r load_code}
all.fun <- list.files(here("Code"), full.names = TRUE)
for(i in 1:length(all.fun)){source(all.fun[i])}
```

```{r geno_type}

geno.cols <- get_geno_cols(here("Data", "general", "geno_cols.csv"))
ordered.geno <- c("FC/FC", "WT/FC", "WT/WT", "WT/VS", "VS/VS")
if(use.carrier.status){
    geno.cols <- geno.cols[c("FC/FC", "WT/WT", "VS/VS")]
    names(geno.cols) <- c("FC", "WT", "VS")
    ordered.geno <- c("FC", "WT", "VS")
    carrier.text <- "carriers"
}else{
    carrier.text <- "all_geno"
}

factor.cols <- c(geno.cols, "F" = "#99d8c9", "M" = "#bcbddc", 
    "4" = "#fdae6b", "12" = "#e6550d")
#dot.col <- "gray50"
dot.col <- "#dd1c77"

heatmap.neg.col <- "purple"
heatmap.pos.col <- "brown"
```

```{r libraries, message = FALSE, warning = FALSE, error = FALSE}
needed.libraries <- c("pheatmap", "VennDiagram", "DT", "knitr", "impute", "gprofiler2",
    "vioplot", "igraph", "stringr", "easyPubMed", "tm", "readxl")
load_libraries(needed.libraries)
```

```{r read_data}
gene.info <- read.delim(here("Data", "general", "mouse_gene_info.txt"))
mouse.info <- read.csv(file.path(mouse.data.dir, "metadata_validated.csv"))
orthos <- read.delim(file.path(general.data.dir, "human.mouse.orthologs.txt"))
#mouse.genes <- read.delim(here("Data", "general", paste0("mouse_", expression.type, "_info.txt")))

#read in the appropriate covariate table
covar.mat <- read.table(file.path(tx.processed.data.dir, 
    paste0("covariates_", carrier.text, ".csv")), sep = ",", header = TRUE, row.names = 1)
    
#convert to a data frame with appropriate ordered factors.
covar.table <- data.frame("sequencingBatch" = as.factor(covar.mat[,"sequencingBatch"]),
    "sex_ge" = as.factor(covar.mat[,"sex_ge"]),
    "age_batch" = ordered(covar.mat[,"age_batch"], levels = c("4", "12")),
    "ordered_geno" = ordered(covar.mat[,"ordered_geno"], levels = ordered.geno))

rownames(covar.table) <- rownames(covar.mat)
#write.table(covar.table, file.path(processed.data.dir, "covar_table.txt"), sep = "\t", quote = FALSE)

test.factor <- covar.table$ordered_geno
```

```{r read_proteome}
pr.mouse <- read.delim(file.path(proteomics.data.dir, "observations.tsv"))
#gene.features <- read.delim(file.path(proteomics.data.dir, "gg.features.tsv"))
#gene.mat <- read.delim(file.path(proteomics.data.dir, "gg.expression.tsv"))

#try using unique features. Does this make us less likely to attribute an ambiguous peptide to the wrong gene?
gene.features <- read.delim(file.path(proteomics.data.dir, "ug.features.tsv"))
gene.mat <- read.delim(file.path(proteomics.data.dir, "ug.expression.tsv"))
```

```{r check, eval = FALSE}
tx.dir <- here("Results", paste0("age_batch_", compare.age, "_gene"))
all.tx <- readRDS(file.path(tx.dir, "processed_data", "Batch_Adjusted_Expr.RDS"))
tx.covar <- read.csv(file.path(tx.dir, "processed_data", "covariates_carriers.csv"))


gene.chr <- gene.info[match(rownames(gene.mat), gene.info[,"external_gene_name"]), "chromosome_name"]
x.genes <- rownames(gene.mat)[which(gene.chr == "X")]
x.id <- gene.info[match(x.genes, gene.info[,"external_gene_name"]), "ensembl_gene_id"]
y.genes <- rownames(gene.mat)[which(gene.chr == "Y")]
y.id <- gene.info[match(y.genes, gene.info[,"external_gene_name"]), "ensembl_gene_id"]


#checking sex-specific genes
gene.name <- "Ddx3y"
gene.name <- "Ddx3x"
#gene.name <- sample(x.genes, 1)

gene.idx <- which(rownames(gene.mat) == gene.name)
stripchart(log2(as.numeric(gene.mat[gene.idx,]))~as.factor(pr.mouse[,"sex"]), main = gene.name,
    ylab = "log2(abundance)", xlab = "", vertical = TRUE, method = "jitter", pch = 16)


#look at all x-linked genes
f.abund <- sapply(x.genes, function(x) as.numeric(gene.mat[x,which(pr.mouse[,"sex"] == "F")]))
m.abund <- sapply(x.genes, function(x) as.numeric(gene.mat[x,which(pr.mouse[,"sex"] == "M")]))
f.mean <- colMeans(f.abund, na.rm = TRUE)
m.mean <- colMeans(m.abund, na.rm = TRUE)
comp.mat <- t(apply(cbind(f.mean, m.mean), 1, scale))
pdf("~/Desktop/x_linked.pdf", width = 7, height = 40)
pheatmap(comp.mat, cluster_cols = FALSE)
dev.off()

#which of these genes are differentially expressed in the sexes?
x.gene.id <- gene.info[match(x.genes, gene.info[,"external_gene_name"]), "ensembl_gene_id"]
sex.test <- lapply(x.gene.id, function(x) lm(all.tx[,x]~as.factor(tx.covar[,"sex_ge"])))
sex.effect <- sapply(sex.test, function(x) coef(x)[2])
names(sex.effect) <- x.gene.id
high.effect <- which(abs(sex.effect) > 0.5)

high.effect.names <- gene.info[match(names(high.effect), gene.info[,"ensembl_gene_id"]), "external_gene_name"]

protein.means <- t(apply(cbind("F" = rowMeans(gene.mat[high.effect.names,which(pr.mouse[,"sex"] == "F")], na.rm = TRUE),
    "M" = rowMeans(gene.mat[high.effect.names,which(pr.mouse[,"sex"] == "M")], na.rm = TRUE)), 1, scale))
colnames(protein.means) <- c("Pr_F", "Pr_M")

tx.means <- t(apply(cbind("F" = colMeans(all.tx[which(tx.covar[,"sex_ge"] == "Female"),names(high.effect)]),
    "M" = colMeans(all.tx[which(tx.covar[,"sex_ge"] == "Male"),names(high.effect)])), 1, scale))
colnames(tx.means) <- c("Tx_F", "Tx_M")

check.mat <- cbind(protein.means, tx.means)
pheatmap(check.mat, cluster_cols = FALSE)

match.genes <- c(which(apply(check.mat, 1, function(x) all(sign(x) == c(1, -1, 1, -1)))),
    which(apply(check.mat, 1, function(x) all(sign(x) == c(-1, 1, -1, 1)))))
mismatch.genes <- check.mat[setdiff(1:nrow(check.mat), match.genes),]
pheatmap(mismatch.genes, cluster_cols = FALSE)
mismatch.gene.names <- rownames(mismatch.genes)
mismatch.id <- gene.info[match(mismatch.gene.names, gene.info[,"external_gene_name"]), "ensembl_gene_id"]

pdf("~/Desktop/sex_effect.pdf", width = 8, height = 4)
for(i in 1:length(x.gene.id)){
    par(mfrow = c(1,2))
    plot(all.tx[,x.gene.id[i]]~as.factor(tx.covar[,"sex_ge"]), main = "Transcript", xlab = "", ylab = "transcript abundance")
    plot(as.numeric(gene.mat[x.genes[i],])~as.factor(pr.mouse[,"sex"]), main = "Protein", xlab = "", ylab = "protein abundance")
    mtext(x.genes[i], side = 3, line = -1.5, font = 2, cex = 1.2, outer = TRUE)
}
dev.off()
```

## Align proteomics and transcriptomics

Match our transcriptomic mouse information with the proteomic mouse information.

```{r match_mats, fig.width = 5, fig.height = 5}
all.trans.meta <- read.csv(here("Data", "Mouse", "metadata_validated.csv"))
trans.ind <- all.trans.meta[,"animalName"]
prot.ind <- pr.mouse[,"animal_id"]
plotVenn(list("transcriptomic" = unique(all.trans.meta[,"animalName"]), "proteomic" = unique(prot.ind)))
```

There were animals in each data set that did not appear in the other.
The animals that were included in the proteomics data set but not the
transcriptomics data set appear below. They all appear to be from a 
Trem2 experiment. I didn't include these in the Klotho transcriptomic
analysis, so these aren't a problem.

```{r not_in_prot}
not.in.prot <- setdiff(trans.ind, prot.ind)
trans.info <- all.trans.meta[match(not.in.prot, all.trans.meta[,"animalName"]),]
datatable(trans.info)
```

The table below shows the animals that were in the proteomics data 
set, but not in the transcriptomics data set. 

```{r not_in_trans}
not.in.trans <- setdiff(prot.ind, trans.ind)
prot.info <- pr.mouse[match(not.in.trans, pr.mouse[,"animal_id"]),]
datatable(prot.info)
```

We will set up a mouse information table separately for the proteomics
data and compare to the information we have in the transcriptomics data
just to be sure everything matches. 

```{r parse_geno}
parse_geno <- function(allele.label, dip.label){
    kl.type <- gsub("Klotho", "", allele.label)
    
    if(use.carrier.status){
        if(dip.label == "WT"){
            return("WT")
        }else{
            return(kl.type)
        }
    }else{
        if(dip.label == "HOM"){
            kl.geno <- paste0(kl.type, "/", kl.type)
        }
        if(dip.label == "HET"){
            kl.geno <- paste0("WT/", kl.type)
        }
        if(dip.label == "WT"){
            kl.geno <- "WT/WT"
        }
        return(kl.geno)
    }
}
```

We generated a table for each observation in the proteomics data. 
There are multiple replicates per individual mouse.

```{r pr_covar}
pr.id <- pr.mouse[,"animal_id"]
pr.sex <- pr.mouse[,"sex"]

raw.age <- pr.mouse[,"age"]/30 #convert age in days to age in months
#hist(raw.age, xlab = "months") #just double check these are 4 and 12-month olds
pr.age.m <- bin.vector(raw.age, c(4, 12)) #bin ages into four and 12 months old
pr.geno <- apply(pr.mouse[,c("klotho", "geno")], 1, function(x) parse_geno(x[1], x[2]))

split.pr.info <- strsplit(pr.mouse[,"obs_id"], "_")
pr.rep <- sapply(split.pr.info, function(x) tail(x, 1))
pr.run <- sapply(split.pr.info, function(x) x[1])
pr.unsure <- sapply(split.pr.info, function(x) x[3]) 
pr.date <- sapply(split.pr.info, function(x) x[4])

pr.info.table <- cbind(pr.id, pr.sex, pr.age.m, pr.geno, pr.rep, pr.run, pr.unsure, pr.date)
colnames(pr.info.table) <- c("animalID", "sex", "age_months", "genotype", "replicate", "run", "not_sure", "date")
write_table_with_header(pr.info.table, 
    file.path(processed.data.dir, "split_obsevation.txt"),
    sep = "\t", row.names = FALSE, header = "Mouse information for each observation. There are multiple observations per mouse because of technical replicates.")
```

We first went through the replicates to make sure that each individual
mouse had consistent information across all replicates. They all checked
out.

```{r rep_info}
u_id <- unique(pr.info.table[,"animalID"])
split_ind <- lapply(u_id, function(x) unique(pr.info.table[which(pr.info.table[,"animalID"] == x),c("animalID", "sex", "age_months", "genotype")]))
check.info <- unique(sapply(split_ind, nrow))
#all animals have one set of ID, sex, age, and genotype associated. Good!
```

We then aligned every mouse id in both the transcriptomic and
proteomic data sets to make sure the information was the same
in both.

```{r extract_info_fun}
get.pr.info <- function(animalID){
    animal.idx <- which(pr.info.table[,"animalID"] == animalID)[1]
    if(is.na(animal.idx)){
        animal.info <- c("animalID" = animalID, "sex" = NA, "age_months" = NA, "genotype" = NA)
    }else{
        animal.info <- pr.info.table[animal.idx,c("animalID", "sex", "age_months", "genotype")]
    }
    return(animal.info)
}

get.tr.info <- function(animalID){
    #use the covariate matrix because this has already gone through a QC process
    animal.idx <- which(rownames(covar.mat) == animalID)
    if(length(animal.idx) == 0){
        animal.info <- c("animalID" = animalID, "sex" = NA, "age_months" = NA, "genotype" = NA)
    }else{
        animal.sex <- substr(covar.mat[animal.idx, "sex_ge"], 1, 1)
        animal.age <- covar.mat[animal.idx,"age_batch"]
        animal.geno <- covar.mat[animal.idx,"ordered_geno"]
        animal.info <- c("animalID" = animalID, "sex" = animal.sex, "age_months" = animal.age, "genotype" = animal.geno)
    }
return(animal.info)
}

compare.info <- function(one.tr.info, one.pr.info){

    both.present <- TRUE
    
    #if the animal is not in the transcript data
    #or not in the protein data, just say so
    if(all(is.na(one.tr.info[2:length(one.tr.info)]))){
        check.result <- "Not present in transcript data"
        both.present = FALSE
    }
    if(all(is.na(one.pr.info[2:length(one.pr.info)]))){
        check.result <- "Not present in proteomic data"
        both.present = FALSE
    }

    if(both.present){
        #otherwise, compare data bitwise 
        check.stats <- sapply(1:length(one.tr.info), function(x) one.tr.info[x] == one.pr.info[x])
        if(all(check.stats)){
            check.result <- "Pass"
        }else{
            check.result <- "Fail"
        }
    }
    return(check.result)
}

#get meta-data information from finalized table
get_final_info <- function(animalID){
    animal.idx <- which(all.pr.info[,"animalID"] == animalID)
    return(all.pr.info[animal.idx,])
}
```

### Non-matches

```{r align}
all.id <- unique(c(prot.ind, trans.ind))
#plotVenn(list("prot" = unique(prot.ind), "trans" = unique(trans.ind)))
all.tr.info <- t(sapply(all.id, get.tr.info))
all.pr.info <- t(sapply(all.id, get.pr.info))

check.match <- sapply(1:length(all.id), 
    function(x) compare.info(one.tr.info = all.tr.info[x,], one.pr.info = all.pr.info[x,]))
fail.idx <- which(check.match == "Fail")
```

There were `r length(fail.idx)` IDs for which information in the 
transcriptomic and proteomic data did not match. 

The information from the transcriptomics data is below:

```{r}
kable(all.tr.info[fail.idx,])
```

The information from the proteomics data is below:

```{r}
kable(all.pr.info[fail.idx,])
```

## Action taken for non-matches

We have seen a couple of these mice before. In the 
transcriptomics data we converted 83901 to male 
based on transcriptomic evidence. We will also convert 
it to male here.

78933 was listed as female in the climb data and male by the
transcriptomic data. We used the climb sex (verified by 
Dylan), and we think that it might be an XO female based
on transcriptomic evidence. It does not need to be adjusted
here.

The mouse 89055 had inconclusive genotyping, but it looked
as if it might have at least one FC allele. However, it is
listed as WT in the proteomics meta data. The form that 
Dylan submitted also listed this animal as WT. I think 
the inconclusive genotyping might have misled me to believe
this mouse was FC. I will convert this to WT in the transcriptomics
data set.

```{r fail_checking}
fail.id <- all.id[fail.idx]
fail.table <- cbind(all.tr.info[fail.idx,], all.pr.info[fail.idx,])

switch.id <- "83901"
switch.idx <- which(all.pr.info[,"animalID"] == switch.id)
all.pr.info[switch.idx,"sex"] <- "M"
```

## Missingness

Now we are ready to start looking into the protein data. There 
were three replicates done per animal, and there are some other 
numbers in the observation ID that I'm not yet sure what to do 
with. I'm in touch with Mitch and Tim about this. 

We'll want to look at missing data across replicates to start
thinking about what we want to do with that.

```{r read_pr}
split_id <- Reduce("rbind", strsplit(colnames(gene.mat), "_"))
colnames(split_id) <- c("run", "submitter", "number", "date", "strain", "allele", "sex", "diplotype", "L7_id", "rep")
rownames(split_id) <- colnames(gene.mat)
animal_id <- sapply(rownames(split_id), function(x) pr.mouse[which(pr.mouse[,"obs_id"] == x), "animal_id"])
all.obs.info <- t(sapply(animal_id, get_final_info))

#split observations up by unique animal ID
split_obs <- lapply(all.pr.info[,"animalID"], function(x) t(gene.mat[,which(all.obs.info[,"animalID"] == x)]))

missing.by.obs <- sapply(split_obs, function(x) apply(x, 2, function(y) length(which(!is.na(y)))))

#pheatmap(missing.by.obs, show_rownames = FALSE)

#find genes and mice with no data at all
genes.missing.per.mouse <- apply(missing.by.obs, 2, function(x) length(which(x == 0)))
mice.missing.per.gene <- apply(missing.by.obs, 1, function(x) length(which(x == 0)))
```

### Missing by mouse

The bar plot below shows how many observations out of 10,000 are missing
for each mouse. 

```{r num_missing}
barplot(sort(genes.missing.per.mouse), ylab = "Number of Genes", 
    main = "Number of missing genes per mouse", xlab = "mouse")
no.data <- which(genes.missing.per.mouse == nrow(missing.by.obs))
```

The table below shows information for the mice that had no data.
We had no metadata for them either.

```{r no_data}
kable(all.pr.info[no.data,])
```

### Missing by gene

The bar plot below shows the number of mice with missing
data per gene. There were `r ncol(missing.by.obs)` mice. 
The horizontal line shows the 50\% mark. 

```{r missing_genes}
half.mark <- ncol(missing.by.obs)/2
barplot(sort(mice.missing.per.gene), names = NA, 
    main = "Number of missing mice by gene", xlab = "gene",
    ylab = "Number of genes")
abline(h = half.mark)

lots.missing <- which(mice.missing.per.gene >= half.mark)
```

There were `r length(lots.missing)` genes out of 10,000 for which 
at least 50\% of the mice had missing data. 

We removed the mice with no data, and removed the genes with more
than 50\% missingness across the mice.

```{r filter}
#remove mice with no data and genes with more than 50% missingness
remove.mice <- all.pr.info[no.data,"animalID"]
remove.genes <- names(lots.missing)
write_table_with_header(remove.genes, 
    file.path(gene.dump.dir, "High_Missingness_Genes.txt"),
    header = "Genes removed due to overall missing data.", col.names = FALSE)

mice.to.keep <- setdiff(all.pr.info[,"animalID"], remove.mice)
genes.to.keep <- setdiff(rownames(gene.mat), remove.genes)

final.mouse.table <- all.pr.info[match(mice.to.keep, all.pr.info[,"animalID"]),]
filtered.gene.mat <- gene.mat[genes.to.keep, which(all.obs.info[,"animalID"] %in% mice.to.keep)]
final.obs.info <- all.obs.info[which(all.obs.info[,"animalID"] %in% mice.to.keep),]

#identical(rownames(final.obs.info), colnames(filtered.gene.mat))
```


## Variance across replicates

Before we summarize the counts across replicates, we want to
take a look at how stable the different replicates are. 

The plot below shows the mean expression of all proteins
vs. their variance. There are a lot of middle-range proteins
that have extremely high variance. 


```{r rep_var}
log.gene.mat <- log2(filtered.gene.mat+1)
#boxplot(log.gene.mat[,1:10])
#hist(as.numeric(log.gene.mat["Ide",])) #checking Ide bimodality

gene.var <- sapply(final.mouse.table[,"animalID"], function(x) apply(log.gene.mat[,which(final.obs.info[,"animalID"] == x)], 1, var))
gene.mean <- sapply(final.mouse.table[,"animalID"], function(x) rowMeans(log.gene.mat[,which(final.obs.info[,"animalID"] == x)], na.rm = TRUE))

plot(gene.mean, gene.var, xlab = "mean of log2 expression", ylab = "variance of log2 expression")
```

The plot below shows the distribution of just the variance across
replicates. 99.9\% of the measurements live below 1.28 (marked).
Greg wants to err on the side of throwing out anything that has
a whif of irregularity, so I'm just tossing proteins with 
inter-replicate log2 variation > 1.28.

```{r var_hist}
hist(gene.var, breaks = 100, main = "Variation across replicates")
high.thresh <- get.percentile(gene.var[which(!is.na(gene.var))], 99.9)
abline(v = high.thresh)
too.high <- unique(which(gene.var > high.thresh, arr.ind = TRUE)[,1])
#length(too.high)
#sort(rownames(gene.var)[too.high])
write_table_with_header(sort(rownames(gene.var)[too.high]), 
    file.path(gene.dump.dir, "High_variance.txt"), col.names = FALSE,
    header = "Genes with variance across replicates exceeding 99.9% of all genes")
```

There are `r length(too.high)` proteins with variance across replicates
that exceed this threshold. I guess we can just remove these.

```{r filter_var}
to.keep <- setdiff(1:nrow(gene.var), too.high)
low.var.rep <- log.gene.mat[to.keep,]
#hist(as.numeric(low.var.rep["Ide",])) #checking Ide bimodality
```

## Reducing to gene-level numbers

We then took the mean of the observations across the replicates.

```{r mean_reps}
#for each unique animal ID, find all replicates and 
#average across them.
mean.gene.mat <- t(sapply(final.mouse.table[,"animalID"], 
    function(x) rowMeans(low.var.rep[,which(final.obs.info[,"animalID"] == x)], 
    na.rm = TRUE)))
#hist(as.numeric(mean.gene.mat[,"Ide"]))
```

```{r remove_outlier}
#there is an outlier in the decomposition that is causing problems 
#in the tests and plots I'm removing it here.
imp.gene.mat <- impute.knn(mean.gene.mat)$data
imp.decomp <- plot.decomp(imp.gene.mat, plot.results = FALSE)
outlier.idx <- which(imp.decomp$u[,1] > 0.3)
outlier.id <- rownames(mean.gene.mat)[outlier.idx]
keep.id <- setdiff(rownames(mean.gene.mat), outlier.id)
keep.idx <- which(all.obs.info[,"animalID"] %in% keep.id)

final.mouse.table <- final.mouse.table[match(keep.id, final.mouse.table[,"animalID"]),]
filtered.gene.mat <- filtered.gene.mat[,keep.idx]
final.obs.info <- all.obs.info[keep.idx,]
mean.gene.mat <- mean.gene.mat[keep.id,]
#identical(rownames(final.obs.info), colnames(filtered.gene.mat))

```


```{r factors}
genotype <- ordered(final.mouse.table[,"genotype"], levels = c("FC", "WT", "VS"))
sex <- ordered(final.mouse.table[,"sex"], levels = c("F", "M"))
age <- ordered(final.mouse.table[,"age_months"], levels = c(4, 12))
names(genotype) <- names(sex) <- names(age) <- final.mouse.table[,"animalID"]
#stripchart(mean.gene.mat[,"Ide"]~genotype, vertical = TRUE, method = "jitter")
#boxplot(mean.gene.mat[,"Pf4"]~genotype)
write_table_with_header(final.mouse.table, 
    file.path(processed.data.dir, "covariates.txt"),
    header = "Final information for all mice included in the analysis.")
```

## Filter for even distribution

We also want to make sure that values are relatively evenly spread
across the groups, sex, age, and genotype.

Below we count the number of non NA values we have in each subset.
The bar plot below shows how many animals there are of each sex.

```{r filter_sex}
sex.idx <- lapply(levels(sex), function(x) which(sex == x))
names(sex.idx) <- levels(sex)
barplot_with_num(sapply(sex.idx, length), text.gap = 0.25, text.shift = 0.15,
     main = "Number of mice of each sex")
```

```{r rep_thresh}
rep.thresh = 0.2
```

The dot plot below shows how many animals of each sex have counts in 
each gene. We removed genes with representation below 
`r rep.thresh*100`\%

```{r gene_count_sex, fig.width = 5, fig.height = 5}
#for each gene get the proportion of animals in each sex
#with complete data
sex.count <- t(apply(mean.gene.mat, 2, function(x) sapply(levels(sex), 
    function(y) c(length(which(!is.na(x[sex.idx[[y]]]))))/length(which(sex == y)))))
colnames(sex.count) <- levels(sex)

plot(sex.count, main = "Percentage represented per sex", xlim = c(0, 1), ylim = c(0, 1))
abline(0,1)
abline(v = rep.thresh, h = rep.thresh, col = "red")
to.remove <- which(sex.count < rep.thresh, arr.ind = TRUE)

write_table_with_header(unique(rownames(to.remove)), col.names = FALSE,
    file.path(gene.dump.dir, "Unbalanced_Sex.txt"), 
    header = "Genes removed because of low representation in at least one sex. Less than 20% of animals represented.")

to.keep <- setdiff(1:ncol(mean.gene.mat), to.remove)
text(sex.count[to.remove,1], sex.count[to.remove,2], 
    labels = rownames(sex.count)[to.remove], pos = 4, cex = 0.5)

sex.filtered.genes <- mean.gene.mat[,to.keep]
#plot(sex.filtered.genes[,"App"]~genotype)
#stripchart(sex.filtered.genes[,"Ide"]~genotype, vertical = TRUE, method = "jitter")
```

The bar plot below shows how many animals are in each age group.

```{r age}
age.idx <- lapply(levels(age), function(x) which(age == x))
names(age.idx) <- levels(age)
barplot_with_num(sapply(age.idx, length), text.gap = 5, text.shift = 1, 
    main = "Mice represented in each age", ylab = "Number of mice")
```

The dot plot shows how many 4 and 12-month olds represent each 
gene. We did not remove any genes based on imbalanced representation
across the ages.

```{r gene_count_age, fig.width = 5, fig.height = 5}
age.count <- t(apply(sex.filtered.genes, 2, 
    function(x) sapply(levels(age), 
    function(y) c(length(which(!is.na(x[age.idx[[y]]])))/length(which(age == y))))))
colnames(age.count) <- levels(age)

plot(age.count, main = "Percentage represented per age", xlim = c(0,1), ylim = c(0,1))
abline(0,1)
abline(h = rep.thresh, v = rep.thresh, col = "red")
to.remove <- which(age.count < rep.thresh, arr.ind = TRUE)

write_table_with_header(unique(rownames(to.remove)), col.names = FALSE,
    file.path(gene.dump.dir, "Unbalanced_age.txt"), 
    header = "Genes removed because of low representation in at least one age group. Less than 20% of animals represented.")


to.keep <- setdiff(1:ncol(sex.filtered.genes), to.remove)
if(length(to.remove) > 0){
    text(age.count[to.remove,1], age.count[to.remove,2], 
        labels = rownames(age.count)[to.remove], pos = 4, cex = 0.5)
}

age.filtered.genes <- sex.filtered.genes[,to.keep]
#boxplot(age.filtered.genes[,1:10])
#plot(age.filtered.genes[,1]~genotype)
```

The bar plot shows how many animals are of each genotype.

```{r genotype_rep}
geno.idx <- lapply(levels(genotype), function(x) which(genotype == x))
names(geno.idx) <- levels(genotype)
barplot_with_num(sapply(geno.idx, length), text.gap = 3, text.shift = 0.5, 
    main = "Mice per genotype")
```

The dot plots compare for each gene how many animals in 
each genotype have values for that gene. They are pretty well 
balanced. We did not remove any genes based on imbalanced
representation across genotypes.

```{r geno_balance, fig.height = 7, fig.width = 7}
geno.count <- t(apply(age.filtered.genes, 2, 
    function(x) sapply(levels(genotype), 
    function(y) c(length(which(!is.na(x[geno.idx[[y]]]))))/length(which(genotype == y)))))
colnames(geno.count) <- levels(genotype)

pairs(geno.count, main = "Percentage represented per genotype", 
    xlim = c(0,1), ylim = c(0,1))

to.remove <- which(geno.count < rep.thresh, arr.ind = TRUE)
to.keep <- setdiff(1:ncol(age.filtered.genes), to.remove)

write_table_with_header(unique(rownames(to.remove)), col.names = FALSE,
    file.path(gene.dump.dir, "Unbalanced_Genotype.txt"), 
    header = "Genes removed because of low representation in at least one genotype. Less than 20% of animals represented.")


filtered.gene.mat <- age.filtered.genes[,to.keep]
#boxplot(filtered.gene.mat[,1:10])
#plot(filtered.gene.mat[1,]~genotype)
```


## Re-check missingness

Let's check missingness again.

```{r missingness2}
missing.mice <- apply(filtered.gene.mat, 1, 
    function(x) length(which(is.na(x)))/ncol(filtered.gene.mat))
missing.genes <- apply(filtered.gene.mat, 2, 
    function(x) length(which(is.na(x)))/nrow(filtered.gene.mat))
```

The plot below shows the percent missing across genes.
We successfully removed genes with greater than 50\%
missingness. This causes problems when I look for 
interactions between variables because of sparsity.
In the interest of analyzing as many genes as possible,
I will keep the 50\% cutoff and only check for main effects.

```{r plot_missing}
missing.thresh = 0.5
hist(missing.genes, xlim = c(0,1))
abline(v = missing.thresh)
genes.to.keep <- which(missing.genes <= missing.thresh)
to.remove <- which(missing.genes > missing.thresh)

write_table_with_header(unique(names(to.remove)), col.names = FALSE,
    file.path(gene.dump.dir, "Missing_Recheck.txt"), 
    header = "Genes removed because of low overall missingness on second time through.")

#scale and center so we can compare differences across genes
#and take group means, etc.
centered.gene.mat <- apply(filtered.gene.mat[,genes.to.keep], 2, scale)
dimnames(centered.gene.mat) <- dimnames(filtered.gene.mat)
#stripchart(centered.gene.mat[,"Ide"]~genotype, vertical = TRUE, method = "jitter")
#vioplot(centered.gene.mat[,"App"]~genotype)


#this is a weird idea, but the 7th protein PC is correlated with 
#the first transcript PC. I'm going to try to adjust out the first
#6 PCs from the protein matrix and run everything. This is silly, 
#and I shouldn't waste my time with such things, but I can't resist.

#imp.mat <- impute.knn(centered.gene.mat)$data
#imp.decomp <- plot.decomp(imp.mat, pc = 6, plot.results = FALSE)
#adj.cent <- adjust(centered.gene.mat, imp.decomp$u)
#centered.gene.mat <- adj.cent
```

This leaves us `r length(to.keep)` proteins.

The following plot shows how much data individual mice are missing.
All mice have well below our threshold.

```{r missing_mice}
hist(missing.mice, xlim = c(0,1), main = "Number of mice with varying degrees of missing data")
abline(v = missing.thresh)
mice.to.keep <- which(missing.mice <= missing.thresh)
centered.gene.mat <- centered.gene.mat[mice.to.keep,]
```


```{r plot_gene_fun}
#for debugging/re-coding
#return.text = FALSE; plot.results = FALSE; stat.x = 0.1; stat.y = 0.9; stat.y.spread = 0.15; cex.lab = 1; plot.label = ""; ylab = "Abundance (A.U.)"; autoflip.stat.y = FALSE; autoplace.text = TRUE; n.samples = 25; ylim = NULL; jitter.factor = 1; min.contig = 5

test_effect <- function(values, plot.factor, return.text = FALSE,
    plot.results = FALSE, stat.x = 0.1, stat.y = 0.9, stat.y.spread = 0.15, 
    cex.lab = 1, plot.label = "", ylab = "Abundance (A.U.)", autoflip.stat.y = FALSE,
    autoplace.text = FALSE, n.samples = 25, min.contig = 5, jitter.factor = 1, ylim = NULL){
    
    model <- lm(values~plot.factor)
    #summary(model)

    r2 <- summary(model)$adj.r.squared
    p <- anova(model)$"Pr(>F)"[1]
    text.p <- threshold_p(p, sig.dig = 2, return.text = TRUE)
    linear.effect.size <- coefficients(summary(model))["plot.factor.L","Estimate"]
    linear.effect.p <- coefficients(summary(model))["plot.factor.L","Pr(>|t|)"]
    
        if(plot.results){
            box.cols <- sapply(levels(plot.factor), function(x) factor.cols[which(names(factor.cols) == x)])
            vioplot(values~plot.factor, col = box.cols, xlab = "", main = plot.label, 
                ylab = "", cex.names = cex.lab, ylim = ylim)
            mtext(ylab, side = 2, cex = cex.lab, line = 2)
            stripchart(values~plot.factor, vertical = TRUE, pch = 16, method = "jitter",
                add = TRUE, col = dot.col)
            plot.dim <- par("usr")
            segments(x0 = plot.dim[1], x1 = plot.dim[2], y0 = 0)
            
            if(autoplace.text){
  
                possible.coord <- autoplace_text2(x = jitter(as.numeric(plot.factor), factor = jitter.factor), 
                    y = values, n.samples = n.samples, min.contig = min.contig, plot.result = FALSE) 

                #place in region relative to specified stat.x and stat.y
                text.x <- fractional_pos(min(possible.coord[,1]), max(possible.coord[,1]), stat.x)
                
                plot.dim <- par("usr")
                plot.height <- plot.dim[4] - plot.dim[3]
                min.visible <- plot.dim[3] + (plot.height*0.1) #the minimum visible is above the minimum
                #define the maximum y position based on stat.y and the box we found
                max.y <- fractional_pos(min(possible.coord[,2]), max(possible.coord[,2]), stat.y)
                #define min.y based on max.y and stat.y.spread
                min.y <- max.y - (plot.height*stat.y.spread)
                
                if(min.y < min.visible){ #don't let the text go off the bottom of the plot
                    #move the y position up the amount that it is going off the bottom
                    off.bottom <- min.visible - min.y
                    max.y <- max.y + off.bottom
                    min.y <- min.visible
                }

                #y positions are evenly distributed from max.y to min.y
                text.y <- segment_region(max.y, min.y, 3, "ends")
                r2.y <- text.y[1]
                p.y <- text.y[2]
                effect.y <- text.y[3]
            }else{
                text.x <- fractional_pos(plot.dim[1], plot.dim[2], stat.x)
                if(autoflip.stat.y){
                    if(linear.effect.size < 0){
                        stat.y <- 1.1-stat.y #automatically flip if the effect size is negative. not sure if this is a good idea
                    }
                }
                text.y <- segment_region(stat.y, stat.y - stat.y.spread, 3, "ends")
                r2.y <- fractional_pos(plot.dim[3], plot.dim[4], text.y[1])
                p.y <- fractional_pos(plot.dim[3], plot.dim[4], text.y[2])
                effect.y <- fractional_pos(plot.dim[3], plot.dim[4], text.y[3])
            }
            text(text.x, r2.y, bquote(italic(R)^2==.(signif(r2, 2))), adj = 0)
            text(text.x, p.y, labels = text.p, adj = 0)
            text(text.x, effect.y, labels = bquote(beta==.(signif(linear.effect.size, 2))), adj = 0)
        }

    result <- list("r2" = r2, "model.p" = p, 
        "linear.effect.size" = linear.effect.size, "linear.effect.p" = linear.effect.p)
    invisible(result)
}


plot_gene <- function(gene.name, pr.mat, plot.label = "", plot.results = TRUE, stat.x = 0.1, 
    stat.y = 0.9, stat.y.spread = 0.15, cex.lab = 1, return.text = TRUE,
    ylab = "Abundance (A.U.)", autoflip.stat.y = FALSE, autoplace.text = FALSE, 
    n.samples = 25, min.contig = 5, jitter.factor = 1, ylim = NULL, plot.label.cex = 1){

    common.gene <- intersect(gene.name, colnames(pr.mat))
    if(length(common.gene) == 0){
        if(plot.results){
            plot.text("No genes found")
        }
        return(NA)
    }

    gene.idx <- match(gene.name, colnames(pr.mat))
    if(length(gene.idx) == 1 && is.na(gene.idx)){
        if(plot.results){
            plot.text(paste(paste(gene.name, collapse = ", "), "not found"))
            return(NA)
        }
    }
    not.found <- gene.name[which(is.na(gene.idx))]
    gene.idx <- gene.idx[which(!is.na(gene.idx))]
    if(length(gene.idx) == 0){
        if(plot.results){
            plot.text(paste("Could not find", paste(gene.name, collapse = ", ")))
        }
        return(NA)
    }
    
    if(plot.results){
        layout.mat <- matrix(c(1,2,6,3,4,5), byrow = TRUE, nrow = 2)
        layout(layout.mat)
    }

    #if there is more than one gene, take the average of all values
    gene.mat <- pr.mat[,gene.idx,drop=FALSE]
    #hist(gene.mat)
    if(ncol(gene.mat) > 1){
        mean.gene  <- matrix(rowMeans(gene.mat, na.rm = TRUE), ncol = 1)
        gene.label <- "gene group"
    }else{
        mean.gene <- gene.mat
        gene.label <- gene.name
    }

    #plot one: main effect of sex
    adj.vals <- adjust(mean.gene, dummy_covar(cbind(age, genotype)), 
        retain.intercept = FALSE)
    sex.result <- test_effect(values = adj.vals, plot.factor = sex, 
        return.text = return.text,
        plot.results = plot.results, stat.x = stat.x, stat.y = stat.y, 
        stat.y.spread = stat.y.spread, cex.lab = cex.lab, plot.label = 
        paste("Effect of sex on", gene.label), 
        ylab = ylab, autoflip.stat.y = autoflip.stat.y, autoplace.text = autoplace.text,
        n.samples = n.samples, min.contig = min.contig, jitter.factor = jitter.factor, ylim = ylim)
    
    #plot two: main effect of age
    adj.vals <- adjust(mean.gene, dummy_covar(cbind(sex, genotype)), 
        retain.intercept = FALSE)
    age.result <- test_effect(adj.vals, age, return.text = return.text,
        plot.results = plot.results, stat.x = stat.x, stat.y = stat.y, 
        stat.y.spread = stat.y.spread, cex.lab = cex.lab, plot.label = 
        paste("Effect of age on", gene.label), 
        ylab = ylab, autoflip.stat.y = autoflip.stat.y, autoplace.text = autoplace.text,
        n.samples = n.samples, min.contig = min.contig, jitter.factor = jitter.factor, ylim = ylim)

    #plot three: main effect of genotype
    adj.vals <- adjust(mean.gene, dummy_covar(cbind(age, sex)), 
        retain.intercept = FALSE)
    geno.result <- test_effect(values = adj.vals, plot.factor = genotype, return.text = return.text,
        plot.results = plot.results, stat.x = stat.x, stat.y = stat.y, 
        stat.y.spread = stat.y.spread, cex.lab = cex.lab, plot.label = 
        paste("Effect of genotype on", gene.label), 
        ylab = ylab, autoflip.stat.y = autoflip.stat.y, autoplace.text = autoplace.text,
        n.samples = n.samples, min.contig = min.contig, jitter.factor = jitter.factor, ylim = ylim)

    #plot four to look at interaction effect: main effect of genotype in 4 month old mice
    four.idx <- which(age == "4")
    four.result <- test_effect(values = adj.vals[four.idx], plot.factor = genotype[four.idx], 
        return.text = return.text, plot.results = plot.results, stat.x = stat.x, 
        stat.y = stat.y, stat.y.spread = stat.y.spread, cex.lab = cex.lab, 
        plot.label = paste("Effect of genotype on", gene.label, "\nin 4-month old mice"), 
        ylab = ylab, autoflip.stat.y = autoflip.stat.y, 
        autoplace.text = autoplace.text, n.samples = n.samples, min.contig = min.contig, 
        jitter.factor = jitter.factor, ylim = ylim)

    #plot five to look at interaction effect: main effect of genotype in 12 month old mice
    twelve.idx <- which(age == "12")
    twelve.result <- test_effect(values = adj.vals[twelve.idx], plot.factor = genotype[twelve.idx], 
        return.text = return.text, plot.results = plot.results, stat.x = stat.x, 
        stat.y = stat.y, stat.y.spread = stat.y.spread, cex.lab = cex.lab, 
        plot.label = paste("Effect of genotype on", gene.label, "\nin 12-month old mice"), 
        ylab = ylab, autoflip.stat.y = autoflip.stat.y, 
        autoplace.text = autoplace.text, n.samples = n.samples, min.contig = min.contig, 
        jitter.factor = jitter.factor, ylim = ylim)
    
    if(plot.results){
        #add a label in the black spot
        plot.text(plot.label, cex = plot.label.cex)
    }

    stat.mat <- rbind("sex" = sex.result, "age" = age.result, 
        "genotype" = geno.result, "geno_four" = four.result,
        "geno_twelve" = twelve.result)

    result <- list("expr_mat" = gene.mat, "not_found" = not.found, "stats" = stat.mat)
    invisible(result)
}

#plot_gene("App", centered.gene.mat, stat.x = 0.5)
#plot_gene("Apoe", centered.gene.mat)
#plot_gene("Ide", centered.gene.mat, autoplace.text = TRUE, n.samples = 25, min.contig = 5)
```

## IDE adjustment

As with the transcriptomes, we adjust for IDE abundance in case that is
a problem. It does have bimodal protein abundance, as shown below.

```{r ide_abundance, fig.width = 9, fig.height = 6.5}
plot_gene("Ide", centered.gene.mat, autoplace.text = TRUE)
```

```{r ide}
ide.idx <- which(colnames(centered.gene.mat) == "Ide")
centered.gene.mat <- adjust(centered.gene.mat, centered.gene.mat[,"Ide", drop = FALSE], 
    retain.intercept = FALSE)
write.table(centered.gene.mat, file.path(processed.data.dir, "centered_gene_mat.txt"),
    sep = "\t", quote = FALSE)
```


## PC Plots {.tabset .tabset-fade .tabset-pills}

We then imputed missing data in order to perform a PC analysis of the 
matrix. The plots below show the first four principal components of the
imputed protein abundance matrix. The different table show the same
data but colored by the different experimental factors: gentype, sex,
and age. Age seems to be correlated with PC1, but the other factors
are not correlated with any PC.

```{r impute}
imp.gene.mat <- impute.knn(centered.gene.mat)$data

genotype.col <- sapply(final.mouse.table[,"genotype"], function(x) geno.cols[which(names(geno.cols) == x)])
sex.col <- as.numeric(sex)
age.col <- as.numeric(age)

mean.decomp <- plot.decomp(imp.gene.mat, plot.results = FALSE, pc = 4)
```

The following plots show the first four PCs of the matrix colored
by different factors. The first PC seems to be age, but none of the
other factors seem to have an effect. 

```{r plot_decomp}
mouse.pc <- mean.decomp$u
colnames(mouse.pc) <- paste0("PC", 1:ncol(mouse.pc))
```

### Genotype
```{r geno_col, fig.width = 9, fig.height = 9}
pairs(mouse.pc, col = genotype.col, pch = 16, main = "colored by genotype")
```

### Sex
```{r sex_col, fig.width = 9, fig.height = 9}
pairs(mouse.pc, col = sex.col, pch = 16, main = "colored by sex")
```

### Age
```{r age_col, fig.width = 9, fig.height = 9}
pairs(mouse.pc, col = age.col, pch = 16, main = "colored by age")
```

## Fit models

We fit linear models to protein abundance explaining each abundance
with age, sex, and genotype. We used the unimputed matrix for this.

The qq plots for the p values are shown below. All factors seem to 
have a large effect, but age has the largest.

```{r get_lin_idx}
get_lin_idx <- function(model, factor.name){
    coef.mat <- summary(model)$coefficients
    want.idx <- grep(factor.name, rownames(coef.mat))
    dont.want.idx <- grep("Q", rownames(coef.mat))
    final.want <- setdiff(want.idx, dont.want.idx)
    #rownames(coef.mat)[final.want]
    return(final.want)
}
```

```{r fit_models_limma, eval = FALSE}
library(limma)
library(edgeR)
protein_data <- t(adjust(mean.gene.mat, dummy_covar(data.frame("sex" = sex, "age" = age))))

design <- model.matrix(~genotype)
colnames(design) <- levels(group)


fit_trend <- lmFit(protein_data, design)
fit_trend <- eBayes(fit_trend)
trend_results <- topTable(fit_trend, number = nrow(protein_data))


fit <- lmFit(protein_data, design)
fit <- eBayes(fit)
p.mat <- fit$p.value
effect.mat <- coef(fit)
trend_results <- topTable(fit, adjust = "BH", number = Inf)


age.p <- p.mat["age.L"]
sex.p <- p.mat[,"sexF"]-p.mat[,"sexM"]
geno.p <- p.mat[,"genotype.L"]

trend_results <- topTable(fit, adjust = "BH", number = Inf)

```


```{r fit_models, fig.width = 9, fig.height = 4}
#with interactions
age.models <- lapply(1:ncol(centered.gene.mat), function(x) lm(centered.gene.mat[,x]~(genotype+sex+age)^2))
sex.models <- lapply(1:ncol(centered.gene.mat), function(x) lm(centered.gene.mat[,x]~(age+genotype+sex)^2))
genotype.models <- lapply(1:ncol(centered.gene.mat), function(x) lm(centered.gene.mat[,x]~(sex+age+genotype)^2))

age.idx <- get_lin_idx(age.models[[1]], "age")
age.p <- t(sapply(age.models, function(x) summary(x)$coefficients[age.idx,"Pr(>|t|)"]))
colnames(age.p) <- gsub(".L", "", colnames(age.p))
age.effect <- t(sapply(age.models, function(x) summary(x)$coefficients[age.idx,"Estimate"]))
colnames(age.effect) <- gsub(".L", "", colnames(age.effect))

sex.idx <- get_lin_idx(sex.models[[1]], "sex")
sex.p <- t(sapply(sex.models, function(x) summary(x)$coefficients[sex.idx,"Pr(>|t|)"]))
colnames(sex.p) <- gsub(".L", "", colnames(sex.p))
sex.effect <- t(sapply(sex.models, function(x) summary(x)$coefficients[sex.idx,"Estimate"]))
colnames(sex.effect) <- gsub(".L", "", colnames(sex.effect))

geno.idx <- get_lin_idx(genotype.models[[1]], "genotype")
geno.p <- t(sapply(genotype.models, function(x) summary(x)$coefficients[geno.idx,"Pr(>|t|)"]))
colnames(geno.p) <- gsub(".L", "", colnames(geno.p))
geno.effect <- t(sapply(genotype.models, function(x) summary(x)$coefficients[geno.idx,"Estimate"]))
colnames(geno.effect) <- gsub(".L", "", colnames(geno.effect))

rownames(age.p) <- rownames(age.effect) <- rownames(sex.p) <- rownames(sex.effect) <- rownames(geno.p) <- rownames(geno.effect) <- colnames(centered.gene.mat)
```

For each factor, we tested for main effects as well as interaction
effects with other factors. The plots below show qq plots for each
of these tests. There appear to be some strong interactions, particularly
between age and genotype, and to a lesser extent between sex and genotype.

```{r qq_age, fig.width = 9, fig.height = 9}
par(mfrow = c(3,3))
test <- sapply(1:ncol(age.p), function(x) qqunif.plot(age.p[,x], plot.label = colnames(age.p)[x]))
test <- sapply(1:ncol(sex.p), function(x) qqunif.plot(sex.p[,x], plot.label = colnames(sex.p)[x]))
test <- sapply(1:ncol(geno.p), function(x) qqunif.plot(geno.p[,x], plot.label = colnames(geno.p)[x]))

factor.effect.mat <- cbind(age.effect, sex.effect, geno.effect)
factor.p.mat <- cbind(age.p, sex.p, geno.p)
factor.fdr.mat <- apply(factor.p.mat, 2, function(x) p.adjust(x, method = "fdr"))

write_table_with_header(factor.effect.mat, file.path(results.dir, "Protein_Abundance_Tests_Effects.csv"), 
    sep = ",", header = "Linear effect sizes (betas from linear models) for each protein abundance as it is affected by each tested factor.")
write_table_with_header(factor.p.mat, file.path(results.dir, "Protein_Abundance_Tests_P.csv"), 
    sep = ",", header = "P values for linear models for each protein abundance as it is affected by each tested factor.")
write_table_with_header(factor.p.mat, file.path(results.dir, "Protein_Abundance_Tests_FDR.csv"), 
    sep = ",", header = "FDR for linear models for each protein abundance as it is affected by each tested factor.")
```

```{r dups}
#some of the columns are duplicates of each other. 
#age:genotype is the same as genotype:age. Remove
#the duplicates
all.col <- sapply(strsplit(colnames(factor.p.mat), ":"), function(x) paste(sort(x), collapse = ":"))
dup.col <- which(duplicated(all.col))
to.keep <- setdiff(1:ncol(factor.p.mat), dup.col)
factor.effect.mat <- factor.effect.mat[,to.keep]
factor.p.mat <- factor.p.mat[,to.keep]
factor.fdr.mat <- factor.fdr.mat[,to.keep]
```

## Volcano plots

The volcano plots are shown below all on the same scale. 
This highlights the huge effects of age relative to all 
other factors. 

```{r volcano, fig.width = 9, fig.height = 6}
ylim = c(0, max(-log10(factor.p.mat)))
xlim = c(min(factor.effect.mat), max(factor.effect.mat))

par(mfrow = c(2,3))
for(i in 1:ncol(factor.p.mat)){
    plot(factor.effect.mat[,i], -log10(factor.p.mat[,i]), 
        main = colnames(factor.p.mat)[i], xlab = "Effect", ylab = "-log10(p)",
        xlim = xlim, ylim = ylim)
}
```

```{r thresh}
fdr.thresh <- 0.1
```

Volcano plots on independend scales are shown below.
Colored dots are significant at an FDR of `r fdr.thresh`.
The different colors indicate positive and negative effect
sizes.

```{r volcano_diff, fig.width = 9, fig.height = 6}
sig.list <- vector(mode = "list", length = ncol(factor.fdr.mat))
names(sig.list) <- colnames(factor.fdr.mat)
par(mfrow = c(2,3))
for(i in 1:ncol(factor.p.mat)){
    pt.col = rep("black", nrow(factor.fdr.mat))
    sig.idx <- which(factor.fdr.mat[,i] <= fdr.thresh)
    down.idx <- which(factor.effect.mat[,i] < 0)
    down.sig <- intersect(sig.idx, down.idx)
    up.idx <- which(factor.effect.mat[,i] > 0)
    up.sig <- intersect(sig.idx, up.idx)
    factor.sig <- list("Up" = rownames(factor.fdr.mat)[up.sig], 
        "Down" = rownames(factor.fdr.mat)[down.sig])
    sig.list[[i]] <- factor.sig
    if(length(up.idx) > 0){pt.col[up.sig] <- "#8c510a"}
    if(length(down.idx) > 0){pt.col[down.sig] <- "#01665e"}
    plot(factor.effect.mat[,i], -log10(factor.p.mat[,i]), 
        col = pt.col, pch = 16,
        main = colnames(factor.p.mat)[i], xlab = "Effect", 
        ylab = "-log10(p)")    
}
```

```{r num_sig}
par(mar = c(4, 12, 4, 4), mfrow = c(1,1))
num.sig <- sapply(sig.list, function(x) sapply(x, length))
a <- barplot(num.sig, beside = TRUE, main = "Number of significant proteins",
    col = c("#8c510a", "#01665e"), las = 2, horiz = TRUE, 
    xlim = c(0, (max(num.sig)*1.1)), legend = TRUE)
test <- sapply(1:nrow(a), function(b) text(x = num.sig[b,]+150, y = a[b,], 
    labels = num.sig[b,]))


write.table(num.sig, here("Results", "for_paper", "num_dap_table.txt"),
quote = FALSE, sep = "\t")
```

## Enrichments

The enrichments for these significant genes are shown below. 
I'm adding some discussion based on the results. This may change
in future iterations of the workflow.

### Age

* Proteins more abundant with age are enriched for synapse, ER,
    subcellular localization, and anion binding.

* Proteins less abundant with age are enriched for mitochondrion, 
    and mRNA splicing

### Sex

* Proteins more abundant in males are enriched for neuropeptide
    activity and siRNA biogenesis

* Proteins more abundant in females are enriched for nuclear 
    membrane, nuclear migration along microtubule, and mRNA
    transport.

### Genotype
* Proteins for which genotype had a positive effect were enriched 
    for a couple transcription factor motifs, abnormal immunoglobulin 
    levels, abnormal IgA secretion.

* Proteins for which genotype had a negative effect were enriched
    for MAPK signaling, blood coagulation, and TLR regulation.

### Genotype:age interaction
* Positive interaction: more in FC in 4-month olds, more in VS in `r compare.age.text`-month olds

* Proteins for which genotype and age had a positive interaction
    were enriched for small molecule metabolism, mitochondrial 
    matrix, and organic acid metabolism.

* Negative interaction: more in VS in 4-month olds, more in FC in `r compare.age.text`-month olds

* There were only two proteins for which genotype and age had a 
    negative interaction: Cnn2 and Tmed9. Obviously, these didn't
    have any enrichments.

* Tmed9 is a membrane trafficking protein in the synapse
* Cnn2 seems to be related to macrophage migration and phagocytosis.


```{r enrich, fig.width = 7, fig.height = 12}
#this file doesn't differ based on which age we are looking at, 
#since we look at this before we split out the age groups.

all.sig <- unlist(sig.list, recursive = FALSE)

if(!offline){
    enrich.file <- file.path(results.dir = "group_enrichment.RDS") 
    if(!file.exists(enrich.file)){
        enrich <- lapply(all.sig, 
            function(x) if(length(x) > 0){gost(x, organism = "mmusculus", source = "GO")})
        saveRDS(enrich, enrich.file)
    }else{
        enrich <- readRDS(enrich.file)
    }
    #pdf("~/Desktop/protein_enrich.pdf", width = 7, height = 12)
    plot.enrichment.group(enrich, max.term.size = 3000, 
        transformation = function(x) log10(x+1), cluster_cols = FALSE)
    #dev.off()
}
```


## Top proteins {.tabset .tabset-fade .tabset-pills}

The following tables show stats for the top proteins across
factors. The maximum number of genes shown for each group is
10.

```{r top_hits, results = "asis"}

for(i in 1:length(sig.list)){
    has.vals <- sapply(sig.list[[i]], length)
    if(any(has.vals) > 0){
        cat("###", names(sig.list)[i], "{.tabset .tabset-fade .tabset-pills}\n")
    }
    for(j in 1:length(sig.list[[i]])){
        sig.set <- sig.list[[i]][[j]]
        if(length(sig.set) > 0){
            cat("####", names(sig.list[[i]])[j], "\n")

            max.num <- min(10, length(sig.set)) #show no more than 10 genes
            sig.effect <- factor.effect.mat[match(sig.set, rownames(factor.effect.mat)),i,drop=FALSE]
            if(names(sig.list[[i]])[j] == "Up"){
                effect.order <- order(sig.effect, decreasing = TRUE)[1:max.num]
            }
            if(names(sig.list[[i]])[j] == "Down"){
                effect.order <- order(sig.effect, decreasing = FALSE)[1:max.num]
            }
        sorted.effect <- sig.effect[effect.order,1,drop=FALSE]
        top.idx <- match(rownames(sorted.effect), rownames(factor.fdr.mat))
        top.fdr <- factor.fdr.mat[top.idx,i]
        stat.table <- cbind(signif(sorted.effect, 2), signif(-log10(top.fdr), 2))
        colnames(stat.table) <- c("Effect Size", "-log10(FDR)")
        print(kable(stat.table))
        cat("\n\n")
        }
    }
}
```

## Examples {.tabset .tabset-fade .tabset-pills}

Below we show examples of each type of effect.


### Up with age

```{r age_up, fig.width = 9, fig.height = 6.5}
#pdf("~/Desktop/example.pdf", width = 9, height = 6.5)
#age effect up
#a note about automatic text placement. It's pretty good, but
#not perfect. It may require fiddling with parameters, especially
#for violin plots because I don't have a good way of defining
#where the violin will be drawn. I only jitter the points,
#so their spread is not based on their density as it is in the
#violin plots. Also because jitter is random, replotting can lead
#to different text placement.
plot_gene(gene.name = "Hapln2", pr.mat = centered.gene.mat, autoplace.text = TRUE) 
#dev.off()
```

### Down with age

```{r age_down, fig.width = 9, fig.height = 6.5}
#age effect down
plot_gene("Prom1", centered.gene.mat, autoplace.text = TRUE) 
```

### Positive genotype-age interaction
```{r geno_age_up, fig.width = 9, fig.height = 6.5}
#genotype-age interaction positive
plot_gene("Tspan2", centered.gene.mat, autoplace.text = TRUE, stat.y = 0.2) 
```

### Negative genotype-age interaction
```{r geno_age_down, fig.width = 9, fig.height = 6.5}
#genotype-age interaction negative
plot_gene("Tmed9", centered.gene.mat, autoplace.text = TRUE) 
```

### Higher in males
```{r male_up, fig.width = 9, fig.height = 6.5}
#higher in males
plot_gene("Serpina3k", centered.gene.mat, autoplace.text = TRUE)
```

### Higher in females
```{r female_up, fig.width = 9, fig.height = 6.5}
#higher in females
plot_gene("Ighg", centered.gene.mat, autoplace.text = TRUE) 
```

### Positive main effect of genotype
```{r geno_up, fig.width = 9, fig.height = 6.5}
#overall positive effect of genotype
plot_gene("Hmbox1", centered.gene.mat, autoplace.text = TRUE, stat.x = 0.18) 
```

### Negative main effect of genotype
```{r geno_down, fig.width = 9, fig.height = 6.5}
#overall negative effect of genotype
plot_gene("Csnk2b", centered.gene.mat, autoplace.text = TRUE) 
```


```{r patterns, eval = FALSE}
par(mfrow = c(2,3))
for(i in 1:length(sig.list)){
    category <- names(sig.list)[i]
    for(j in 1:length(sig.list[[i]])){
        direction <- names(sig.list[[i]])[j]
        cat.genes <- sig.list[[i]][[j]]
        if(length(cat.genes) > 0 && length(grep(":", category, fixed = TRUE)) == 0){
            sig.mat <- mean.gene.mat[,cat.genes]
            pheatmap(sig.mat, scale = "column")
            decomp_with_nas(sig.mat, cols = genotype.col)
            mean.mat <- apply(sig.mat, 1, function(x) mean(x, na.rm = TRUE))
            boxplot(mean.mat~get(category), xlab = category, ylab = "Mean abundance (A.U.)",
                main = paste(category, direction))
        }
    }

}
```


## Comparison to transcriptomics

We want to make as direct a comparison as we can to the transcriptomic
data. So we want to compare the genotype effects in just the `r compare.age.text`-month
old animals. However, we also want to be able to look at the same stats
for the four-month old animals. The following sections look at a single
age that is defined in the parameters section at the top of the workflow.
For this run, the age is `r compare.age` months.

We fit linear models to protein abundance in just the animals of the 
specified age. We then compared p values and effect sizes 
across all common genes. We also noted which genes (if any) have significant
effects of genotype in both the proteome and the transcriptome in the 
`r compare.age`
month animals.

The volcano plot for these tests is below. We have reduced power 
to detect differences in this data set compared to the transcriptomics,
likely because of missing data. 

```{r age_thresh}
raised.fdr.thresh <- 0.1
```


```{r age_test}
adj.pr <- adjust(centered.gene.mat, dummy_covar(data.frame("sex" = sex, "age" = age)), retain.intercept = FALSE)

age.idx <- which(age == compare.age)
age.genotype.models <- lapply(1:ncol(adj.pr), 
    function(x) lm(adj.pr[age.idx,x]~genotype[age.idx]))
geno.idx <- get_lin_idx(age.genotype.models[[1]], "genotype")
age.geno.p <- t(sapply(age.genotype.models, 
    function(x) summary(x)$coefficients[geno.idx,"Pr(>|t|)"]))
colnames(age.geno.p) <- colnames(centered.gene.mat)
age.geno.effect <- t(sapply(age.genotype.models, 
    function(x) summary(x)$coefficients[geno.idx,"Estimate"]))
age.adj.p <- p.adjust(age.geno.p, method = "fdr")
names(age.geno.effect) <- names(age.adj.p) <- colnames(centered.gene.mat)
age.sig.idx <- which(age.adj.p < raised.fdr.thresh)
#length(age.sig.idx)
```

At the same FDR threshold as in the transcriptomics 
(`r raised.fdr.thresh`), we get `r length(age.sig.idx)` 
significantly different proteins by genotype in the 
`r compare.age`-month old animals. These are shown in red 
on the volcano plots below.

```{r age_months}
pr.col <- rep("black", length(age.adj.p))

if(length(age.sig.idx) > 0){
    pr.col[age.sig.idx] <- "red"
}
plot(age.geno.effect, -log10(age.geno.p), xlab = paste("Effect of genotype at", compare.age, "months"), ylab = "-log10(p value)", 
    col = pr.col, pch = 16)
if(length(age.sig.idx) > 0){
    text(age.geno.effect[age.sig.idx], -log10(age.geno.p[age.sig.idx]), 
    labels = colnames(centered.gene.mat)[age.sig.idx], pos = 4, cex = 0.5)
}

geno.result.table <- cbind(as.numeric(age.geno.effect), as.numeric(age.geno.p), as.numeric(age.adj.p))
rownames(geno.result.table) <- colnames(centered.gene.mat)
colnames(geno.result.table) <- c("Effect", "p", "fdr")

write.table(geno.result.table, 
    file.path(results.dir, paste0("Effect_of_Genotype_at_", compare.age, "_months.csv")),
    sep = ",", quote = FALSE)
```

At `r compare.age` months, the upregulated proteins are enriched 
for mitochondrial function. This tracks with the transcriptomic 
results in which mitochondrial metabolism genes overall were up in VS.

The downregulated proteins are enriched for protein 
localization and blood coagulation. This doesn't have 
a clear analog in the transcriptomic data set. Greg thinks
the "blood coagulation" might be a marker of something else,
like ECM, since we take all the blood out of the brains before
testing.

```{r test_sig_age, fig.width = 15, fig.height = 7}
neg.idx <- which(age.geno.effect < 0)
pos.idx <- which(age.geno.effect > 0)
neg.sig <- intersect(age.sig.idx, neg.idx)
pos.sig <- intersect(age.sig.idx, pos.idx)
neg.pr <- colnames(centered.gene.mat)[neg.sig]
pos.pr <- colnames(centered.gene.mat)[pos.sig]

enrich.file <- file.path(results.dir, paste0("pos_neg_enrich_at_", compare.age, "_months.RDS"))
if(file.exists(enrich.file)){
    enrich <- readRDS(enrich.file)    
}
if(!file.exists(enrich.file) && !offline){
    enrich <- lapply(list("Positive" = pos.pr, "Negative" = neg.pr), function(x) gost(x, organism = "mmusculus", source = "GO"))
    saveRDS(enrich, enrich.file)
}

#quartz(width = 15, height = 7)
par(mfrow = c(1,2))
for(i in 1:length(enrich)){
    plot.enrichment(enrich[[i]], max.term.size = 3000, num.terms = 25, plot.label = paste(names(enrich)[i], "effects of genotype"))
}
#dev.off()
```

The code below plots out gene-GO term plots all together
as well as split by GO compartment (biological process,
molecular function, and cellular compartment).

Looking at all GO classificatoins together, we see that
the proteins upregulated in VS divided into two large clusters:
one related to mitochondria, and the other related to small
molecule metabolism. These clusters are highly connected,
suggesting that this small molecule metabolism might
be related to mitochondrial activity.

The proteins downregulated in VS cluster aroung three major
and two minor collections of GO terms. The three major clusters
are response to wounding (blood coagulation), protein metabolism,
and protein transport. The two minor clusters are enzyme activity,
and cholesterol/triglyceride stuff (connected to Apo genes).

These are pretty cool, actually, but we can't put these network
plots in a paper.


```{r go_part}
max.pval = 1e-3 #make this low to reduce some of the noise
enrich.file <- file.path(results.dir, paste0("GO_part_enrich_at_", compare.age, "_months.RDS"))

if(file.exists(enrich.file)){
    enrich.list <- readRDS(enrich.file)
    up.enrich <- enrich.list$up
    down.enrich <- enrich.list$down
    all.enrich <- enrich.list$all
}
if(!file.exists(enrich.file) && !offline){
    up.enrich <- gost(pos.pr, organism = "mmusculus", evcodes = TRUE, sources = "GO")
    down.enrich <- gost(neg.pr, organism = "mmusculus", evcodes = TRUE, sources = "GO")
    all.enrich <- gost(c(pos.pr, neg.pr), organism = "mmusculus", evcodes = TRUE, sources = "GO")
    enrich.list <- list("up" = up.enrich, "down" = down.enrich, "all" = all.enrich)
    saveRDS(enrich.list, enrich.file)
}

all.var <- ls() #check to see if we have the enrichment data
got.enrich <- length(which(all.var == "all.enrich")) > 0

if(got.enrich){

    go.parts <- c("GO:CC", "GO:MF", "GO:BP")


    pdf(file.path(results.dir, paste0("GO_gene_net_at_", compare.age, "_months.pdf")), width = 30, height = 20)
        up.net <- go_gene_net(enrichment.result = up.enrich, max.term.size = 3000,
            plot.results = TRUE, min.vertex.size = 2, max.vertex.size = 8, 
            max.pval = max.pval)
        mtext("Up in VS", side = 3, outer = TRUE, line = -2.5)    
        
        #up.mat <- go_gene_net(up.enrich, max.term.size = 3000, max.pval = max.pval, plot.as.heatmap = TRUE)
        #up.decomp <- plot.decomp(up.mat$biadjacency_matrix, label.points = TRUE)
        #plot.decomp(t(up.mat$biadjacency_matrix), label.points = TRUE)
        #png("~/Desktop/up_in_vs_adj.png", width = 11, height = 4, units = "in", res = 300)
        #par(mar = c(4,15,4,2))
        #imageWithText(up.mat$biadjacency_matrix[order(up.decomp$u[,1]), order(up.decomp$v[,1])], row.text.shift = 0.03, col.text.shift = 0.05, show.text = FALSE, col.scale = "blue", main = "Up in VS")
        #dev.off()

        down.net <- go_gene_net(enrichment.result = down.enrich, plot.results = TRUE, 
            max.pval = max.pval, max.term.size = 3000, min.vertex.size = 2, max.vertex.size = 8)
        mtext("Down in VS", side = 3, outer = TRUE, line = -2.5) 

        #down.mat <- go_gene_net(down.enrich, max.term.size = 3000, max.pval = max.pval, plot.as.heatmap = TRUE)
        #down.decomp <- plot.decomp(down.mat$biadjacency_matrix, label.points = TRUE)
        #plot.decomp(t(down.mat$biadjacency_matrix), label.points = TRUE)
        #row.order <- hclust(dist(down.mat$biadjacency_matrix))$order
        #col.order <- hclust(dist(t(down.mat$biadjacency_matrix)))$order
        #png("~/Desktop/down_in_vs_adj.png", width = 16, height = 6, units = "in", res = 300)
        #par(mar = c(4,22,4,2))
        #imageWithText(down.mat$biadjacency_matrix[row.order, col.order], row.text.shift = 0.03, col.text.shift = 0.05, show.text = FALSE, col.scale = "blue", main = "Down in VS")
        #dev.off()
    dev.off()

    for(i in 1:length(go.parts)){
        pdf(file.path(results.dir, paste0(gsub(":", "_", go.parts[i]), "_", compare.age, "_months_net.pdf")), 
            width = 15, height = 10)
        up.net <- go_gene_net(enrichment.result = up.enrich, max.term.size = 3000, 
            filter.source = go.parts[i], max.pval = max.pval,
            plot.results = TRUE, min.vertex.size = 2, max.vertex.size = 8)
            mtext("Up in VS", side = 3, outer = TRUE, line = -2.5)    
        
        down.net <- go_gene_net(enrichment.result = down.enrich, plot.results = TRUE, 
            filter.source = go.parts[i], max.pval = max.pval,
            max.term.size = 3000, min.vertex.size = 2, max.vertex.size = 8)
            mtext("Down in VS", side = 3, outer = TRUE, line = -2.5) 
        dev.off()
    }
}
```

Below we plot the mean abundance per genotype as we did
for the transcripts with the up and down clusters labeled.

```{r mean_by_geno, fig.width = 10, fig.height = 2.5}
adj.pr <- adjust(centered.gene.mat[,age.sig.idx], dummy_covar(data.frame("age" = age, "sex" = sex)), retain.intercept = FALSE)
age.pr <- adj.pr[age.idx,]

geno.abund <- sapply(levels(genotype), function(x) colMeans(age.pr[which(genotype[age.idx] == x),], na.rm = TRUE))
#col.order <- order(plot.decomp(geno.abund, plot.results = FALSE)$u[,1])
col.order <- order(age.geno.effect[age.sig.idx])

#use automatic summaries so we don't mix up which is positive and negative
#we can do something prettier by hand later
enrich.text <- lapply(enrich, function(x) words_to_paragraph(summarize_enrichment(x, 
    remove.words = c(data(PubMed_stopwords), stopwords(), "process", "positive", "negative"),
    num.words = 4, num.terms = 30, max.term.size = 3000), line.len = 5))

#in-line figure
imageWithText(t(geno.abund[col.order,]), split.at.vals = TRUE, 
    col.scale = c(heatmap.neg.col, heatmap.pos.col),
    grad.dir = "ends", show.text = FALSE, col.text.shift = 0.3, row.text.shift = 0.01, 
    col.names = NULL)
plot.dim <- par("usr")
plot.width = plot.dim[2] - plot.dim[1]
box.center <- 3:1
nudge.y <- 0.3
for(i in 1:length(geno.cols)){
    draw.rectangle(min.x = plot.dim[1]-(plot.width*0.02), max.x = 0, 
        min.y = box.center[i]-nudge.y, 
        max.y = box.center[i]+nudge.y, border = geno.cols[i], lwd = 3) 
}
text(x = length(neg.pr)/2, y = 3.5, 
    labels = enrich.text$Negative, adj = 0.5, cex = 0.7)
text(x = length(neg.pr)+length(pos.pr)/2, y = 3.5, 
    labels = enrich.text$Positive, adj = 0.5, cex = 0.7)

#png for putting in paper
pos.text <- "Mitochondria"
neg.text <- "ECM, integrins, MAPK signaling"
png(here("Results", "for_paper", paste0("DAP_", compare.age, "_months.png")), width = 8, height = 3, units = "in", res = 300)
par(bg = NA)
imageWithText(t(geno.abund[col.order,]), split.at.vals = TRUE, 
    col.scale = c(heatmap.neg.col, heatmap.pos.col),
    grad.dir = "ends", show.text = FALSE, col.text.shift = 0.3, row.text.shift = 0.015, 
    col.names = NULL)
plot.dim <- par("usr")
plot.width = plot.dim[2] - plot.dim[1]
box.center <- 3:1
nudge.y <- 0.3
for(i in 1:length(geno.cols)){
    draw.rectangle(min.x = plot.dim[1]-(plot.width*0.02), max.x = 0, 
        min.y = box.center[i]-nudge.y, 
        max.y = box.center[i]+nudge.y, border = geno.cols[i], lwd = 3) 
}
text(x = length(neg.pr)/2, y = 3.5, 
    labels = neg.text, adj = 0.5, cex = 1.2)
text(x = length(neg.pr)+length(pos.pr)/2, y = 3.5, 
    labels = pos.text, adj = 0.5, cex = 1.2)
dev.off()

png(here("Results", "for_paper", "colorbar_vertical_protein.png"), height = 4, width = 1.5, units = "in", res = 300)
par(bg = NA, xpd = NA)
imageWithTextColorbar(signif(geno.abund, 2), split.at.vals = TRUE, 
    col.scale = list(heatmap.neg.col, heatmap.pos.col), grad.dir = "ends", 
    cex = 1, round.nearest = 1)
plot.dim <- par("usr")
plot.width <- plot.dim[2] - plot.dim[1]
text(x = plot.dim[2]+(plot.width*0.4), y = mean(plot.dim[3:4]), 
    labels = "Abundance (A.U.)", srt = 270, cex = 1.2)
dev.off()

png(here("Results", "for_paper", "colorbar_horiz_protein.png"), height = 2, width = 5, units = "in", res = 300)
par(bg = NA, xpd = NA)
imageWithTextColorbar(signif(geno.abund, 2), split.at.vals = TRUE, 
    col.scale = list(heatmap.neg.col, heatmap.pos.col), grad.dir = "ends", 
    cex = 1, round.nearest = 1, orientation = "h", axis.line = 1)
plot.dim <- par("usr")
plot.width <- plot.dim[2] - plot.dim[1]
text(x = 0, y = 1.4, labels = "Abundance (A.U.)", srt = 0, cex = 1.2)
dev.off()
```

```{r mean_effect, eval = FALSE}
#just a check 

geno.annot <- data.frame("genotype" = genotype[age.idx])
pos.mat <- adjust(centered.gene.mat[age.idx,pos.pr], dummy_covar(data.frame(sex[age.idx])), retain.intercept = FALSE)
#pheatmap(t(pos.mat), annotation_col = geno.annot, annotation_colors = list("genotype" = geno.cols))
pos.mean <- rowMeans(pos.mat, na.rm = TRUE)
boxplot(pos.mean~genotype[age.idx], xlab = "genotype", ylab = "Mean abundance", main = "Positive Effect Proteins")

neg.mat <- adjust(centered.gene.mat[age.idx,neg.pr], dummy_covar(data.frame(sex[age.idx])), retain.intercept = FALSE)
#pheatmap(t(neg.mat), annotation_col = geno.annot, annotation_colors = list("genotype" = geno.cols))
neg.mean <- rowMeans(neg.mat, na.rm = TRUE)
boxplot(neg.mean~genotype[age.idx], xlab = "genotype", ylab = "Mean abundance", main = "Negative Effect Proteins")
```

Did we see differential expression of any of these genes in the
transcriptomic data? We test this below.

```{r transcriptome_results}
#get gene information for the proteins
protein.info.file <- file.path(processed.data.dir, "protein_info.tsv")
if(!file.exists(protein.info.file)){
    protein.info <- gconvert(colnames(centered.gene.mat), organism = "mmusculus")
    write.table(protein.info[,2:6], protein.info.file, sep = "\t", quote = FALSE, row.names = FALSE)
}else{
    protein.info <- read.delim(protein.info.file)
}
sig.tx <- read.delim(here("Results", paste0("age_batch_", compare.age, "_gene"), "DEG_clusters_carriers.txt"))
```

```{r matching}
common.pos <- intersect(pos.pr, sig.tx[,"geneName"])
common.neg <- intersect(neg.pr, sig.tx[,"geneName"])

pos.pr.idx <- match(common.pos, rownames(geno.result.table))
neg.pr.idx <- match(common.neg, rownames(geno.result.table))

pos.tr.idx <- match(common.pos, sig.tx[,"geneName"])
neg.tr.idx <- match(common.neg, sig.tx[,"geneName"])

all.pos.table <- cbind(sig.tx[pos.tr.idx,1:5], geno.result.table[pos.pr.idx,,drop=FALSE])
all.neg.table <- cbind(sig.tx[neg.tr.idx,1:5], geno.result.table[neg.pr.idx,,drop=FALSE])
colnames(all.pos.table)[3:5] <- colnames(all.neg.table)[3:5] <- paste0("transcript_", colnames(all.pos.table)[3:5])
colnames(all.pos.table)[6:8] <- colnames(all.neg.table)[6:8] <- paste0("protein_", colnames(all.pos.table)[6:8])

all.common.table <- rbind(all.pos.table, all.neg.table)
write.table(all.common.table, file.path(results.dir, paste0("Common_Effects_", compare.age, "_months.txt")),
    quote = FALSE, sep = "\t", row.names = FALSE)
```

The scatter plot below shows that there is no relationship 
between the effect of genotype on a gene's transcript and its
protein abundance, even for significantly affected proteins. 

If we are looking at 4-month olds, there won't be many
dots. This is because Klotho genotype has basically no
effect on any transcripts in the 4-month olds, so it's
hard to compare. In any case, even with the few genes 
shown, there is no relationship between the effects
of genotype on transcripts and those on proteins. 

```{r compare, fig.width = 5, fig.height = 5}
plot(all.common.table[,"transcript_genotype_beta"], 
    all.common.table[,"protein_Effect"], xlab = "Transcript Beta",
    ylab = "Protein Beta")

text(all.common.table[,"transcript_genotype_beta"], 
    all.common.table[,"protein_Effect"], labels = all.common.table[,"geneName"],
    pos = 4, cex = 0.7)
#abline(0,1)
abline(v = 0, h = 0)
```

Below we expanded the plot to include all transcripts with 
significant effects of genotype. It looks very similar to the
above plot.

```{r common}
#use ensembl IDs
common.gene <- intersect(rownames(sig.tx), protein.info[,"target"])
```

There were `r length(common.gene)` positively identifiable common 
genes between the two data sets.

Below we've plotted the effect size of genotype on each common 
transcript vs. the effect size of gentype on each common protein.
They are uncorrelated to a hilarious degree. Greg thinks that this
is probably grounds to dismiss the data. He says they usually see
a correlation of 0.15 to 0.3 for genotype effects. Bummer.

```{r align_genes, fig.width = 5, fig.height = 5}
tx.idx <- match(common.gene, rownames(sig.tx))
pr.name <- protein.info[match(common.gene, protein.info[,"target"]), "input"]
pr.idx <- match(pr.name, rownames(geno.result.table))

tx.effect <- sig.tx[tx.idx,"genotype_beta"]

combined.result.table <- cbind("Name" = pr.name, "ID" = common.gene, 
    "Transcript Effect" = tx.effect, 
    "Protein Effect" = geno.result.table[pr.idx, "Effect"])

plot.with.model(as.numeric(combined.result.table[,"Transcript Effect"]), 
    as.numeric(combined.result.table[,"Protein Effect"]),
    xlab = "Effect of genotype on transcript", ylab = "Effect of genotype on protein",
    report = "cor.test")

#combined.result.table[which(as.numeric(combined.result.table[,"Protein Effect"]) > 0.4),]

```

```{r top_example, fig.width = 9, fig.height = 6.5, eval = FALSE}
plot_gene("Ppbp", centered.gene.mat, autoplace.text = TRUE)
plot_gene("Apoe", centered.gene.mat, autoplace.text = TRUE)
plot_gene("App", centered.gene.mat, autoplace.text = TRUE)


ad.genes <- read.delim(here("Results", paste0("age_batch_", compare.age, "_gene"), "AD_genes.txt"), comment.char = "#")
ad.pos.pr <- intersect(ad.genes[,1], pos.pr)
ad.neg.pr <- intersect(ad.genes[,1], neg.pr)

plot_gene(ad.neg.pr[2], centered.gene.mat, autoplace.text = TRUE)

pdf("~/Desktop/common_pos.pdf", width = 9, height = 6.5)
for(i in 1:length(common.pos)){
    print(i)
    plot_gene(common.pos[i], autoplace.text = TRUE)
}
dev.off()

pdf("~/Desktop/examples.pdf", width = 9, height = 6.5)
#stats placed for age-specific animals
#stat placement are relative and run from 0 to 1 for both x and 1
plot_gene("Apoe", centered.gene.mat, autoplace.text = FALSE, stat.x = 0.7, stat.y = 0.9, autoflip.stat.y = FALSE, cex.lab = 1.5)
plot_gene("App", centered.gene.mat, autoplace.text = FALSE, stat.x = 0.7, stat.y = 0.9, autoflip.stat.y = FALSE, cex.lab = 1.5)
plot_gene("Pf4", centered.gene.mat, autoplace.text = FALSE, stat.x = 0.7, stat.y = 0.9, autoflip.stat.y = FALSE, cex.lab = 1.5)
plot_gene("Grin2b", centered.gene.mat, autoplace.text = FALSE, stat.x = 0.7, stat.y = 0.23, autoflip.stat.y = FALSE, cex.lab = 1.5)
dev.off()

#platelet factors from Dubal paper
plot_gene("Pf4", centered.gene.mat) #negative effect, which is the opposite of what we would predict
plot_gene("Tln1", centered.gene.mat) #no effect
plot_gene("Fermt3", centered.gene.mat, stat.x = 0.7, stat.y = 0.9) #negative effect
plot_gene("Thbs1", centered.gene.mat, stat.y = 0.9, stat.x = 0.7) #negative effect
plot_gene(c("Pf4", "Tln1", "Fermt3", "Thbs1"), centered.gene.mat, stat.y = 0.9, stat.x = 0.7)

plot_gene("Ckm", centered.gene.mat) #creatinine metabolism, no effect
#plot_gene("Gadph", centered.gene.mat) #not captured

platelet.genes <- gene.features[grep("platelet", gene.features[,"gene_desc"], ignore.case = TRUE),"gene_name"]
plot_gene(platelet.genes, centered.gene.mat, plot.label = "platelet factors")

glutamate.genes <- gene.features[grep("glutamate", gene.features[,"gene_desc"], ignore.case = TRUE),"gene_name"]
plot_gene(glutamate.genes, centered.gene.mat, plot.label = "glutamate genes")

#proteins annotated to blood coagulation tend to be 
#less abundant in VS at specified age
coag.table <- read.table("~/Downloads/GO_MFeatures.tsv", sep = "\t")
coag.genes <- coag.table[,8]
coag.list <- plot_gene(coag.genes, centered.gene.mat, autoplace.text = TRUE, plot.label = "coagulation") 

#proteins annotatied to mitochondrion are not differentially
#abundant at specified age
mito.table <- read.table("~/Downloads/GO_MFeatures (1).tsv", sep = "\t", fill = TRUE)
mito.genes <- mito.table[,8]
mito.list <- plot_gene(mito.genes, centered.gene.mat, autoplace.text = TRUE, plot.label = "mitochondria")

oxphos.genes <- c("Cox7a2l", "Ndufa6", "Slc25a33", "Pink1", "Nupr1", "Atp6v0a1",
    "Atp6v1b2", "Ndufa11", "Iscu", "Tmem135")
plot_gene(oxphos.genes, centered.gene.mat, autoplace.text = TRUE, plot.label = "oxphos")
```

## Biodomains and KEGG pathways

We looked at biodomains and KEGG pathways just to round this out.
There were no overall effects on biodomain or KEGG lists

```{r plot_list}
plot_list <- function(gene.list, pr.mat, list.label = "BD", plot.results = TRUE){

    gene.stats <- vector(mode = "list", length = length(gene.list))
    names(gene.stats) <- names(gene.list)
    
    if(plot.results){
        pdf(file.path(results.dir, paste0(list.label, ".pdf")), width = 9, height = 6)
    }
    for(i in 1:length(gene.list)){
        gene.names <- gene.info[match(gene.list[[i]], gene.info[,"ensembl_gene_id"]),"external_gene_name"]
        gene.stats[[i]] <- plot_gene(gene.names, pr.mat, plot.label = names(gene.list)[i], 
            plot.label.cex = 2, plot.results = plot.results)
    }
    if(plot.results){
        dev.off()
    }

    return(gene.stats)
}

plot_nested_list <- function(nested.gene.list, pr.mat, list.label = "KEGG_BD", plot.results = TRUE){

    gene.stats <- vector(mode = "list", length = length(nested.gene.list))
    names(gene.stats) <- names(nested.gene.list)
    
    if(plot.results){
        pdf(file.path(results.dir, paste0(list.label, ".pdf")), width = 9, height = 6)
    }
    for(i in 1:length(nested.gene.list)){
        if(length(nested.gene.list[[i]]) == 0){next()}
        sub.stats <- vector(mode = "list", length = length(nested.gene.list[[i]]))
        names(sub.stats) <- names(nested.gene.list[[i]])
        for(j in 1:length(nested.gene.list[[i]])){
            plot.label <- paste(names(nested.gene.list)[i], "AND\n", names(nested.gene.list[[i]])[j])
            gene.names <- gene.info[match(nested.gene.list[[i]][[j]], gene.info[,"ensembl_gene_id"]),"external_gene_name"]
            group.result <- plot_gene(gene.names, pr.mat, plot.label = plot.label, plot.label.cex = 1, plot.results = plot.results)
            sub.stats[[j]] <- group.result
        }
        gene.stats[[i]] <- sub.stats
    }
    if(plot.results){
        dev.off()
    }

    return(gene.stats)
}

get_list_stat <- function(list.result, 
    stat.label = c("sex", "age", "genotype", "geno_four", paste0("geno_", compare.age.text)), 
    stat.type = c("r2", "model.p", "linear.effect.size", "linear.effect.p")){


    stat.p <- rep(NA, length(list.result))
    names(stat.p) <- names(list.result)
    for(i in 1:length(list.result)){
        stat.mat <- list.result[[i]]$stats
        if(length(stat.mat) > 1){
            stat.p[i] <- as.numeric(stat.mat[stat.label, stat.type])
        }
    }
    return(stat.p)
}

get_nested_stat <- function(nested.list.result, 
    stat.label = c("sex", "age", "genotype", "geno_four", paste0("geno_", compare.age.text)), 
    stat.type = c("r2", "model.p", "linear.effect.size", "linear.effect.p")){

    stat.label <- stat.label[1]
    stat.type <- stat.type[1]

    stat.p <- vector(mode = "list", length(nested.list.result))
    names(stat.p) <- names(nested.list.result)
    
    for(i in 1:length(nested.list.result)){
        sub.stat.p <- rep(NA, length(nested.list.result[[i]]))
        names(sub.stat.p) <- names(nested.list.result[[i]])
        for(j in 1:length(nested.list.result[[i]])){
            if(length(nested.list.result[[i]][[j]]) > 1){
                stat.mat <- nested.list.result[[i]][[j]]$stats
            }else{
                stat.mat <- NA
            }
            if(length(stat.mat) > 1){
                sub.stat.p[j] <- as.numeric(stat.mat[stat.label, stat.type])
            }
        }
        stat.p[[i]] <- sub.stat.p
    }
    return(stat.p)
}
```

```{r gene_lists}
bd.list <- readRDS(file.path(general.processed.data.dir, "Mouse_Biodomains_for_GSEA.RDS"))
sub.bd.list <- readRDS(file.path(general.processed.data.dir, "Mouse_Subdomains_for_GSEA.RDS"))
bd.go.list <- readRDS(file.path(general.processed.data.dir, "Mouse_Biodomains_sub_GO_for_GSEA.RDS"))
kegg.list <- readRDS(file.path(general.processed.data.dir, "Mouse_KEGG_for_GSEA.RDS"))
kegg.bd.list <- readRDS(file.path(general.processed.data.dir, "Mouse_KEGG_Intersections_for_GSEA.RDS"))
```

There were no significantly differentially abundant biodomains at `r compare.age.text` months.

```{r bd, fig.width = 5, fig.height = 5}
bd.stats <- plot_list(gene.list = bd.list, pr.mat = centered.gene.mat, list.label = paste0("Biodomains_", compare.age, "_months"), plot.results = FALSE)
n.bd.genes <- sapply(bd.stats, function(x) if(length(x) > 0){ncol(x$expr_mat)}else{NA})
age.bd.p <- get_list_stat(bd.stats, stat.label = paste0("geno_", compare.age.text), stat.type = "model.p")
qqunif.plot(age.bd.p,  plot.label = "Biodomain p values")
```

There were no significantly differentially abundant biodomains at `r compare.age.text` months.

```{r k, fig.width = 5, fig.height = 5}
kegg.stats <- plot_list(bd.list, centered.gene.mat, "KEGG", plot.results = FALSE)
n.kegg.genes <- sapply(kegg.stats, function(x) if(length(x) > 0){ncol(x$expr_mat)}else{NA})
age.kegg <- get_list_stat(kegg.stats, stat.label = paste0("geno_", compare.age.text), stat.type = "model.p")
qqunif.plot(age.kegg, plot.label = "KEGG p values")
```

There were no significantly differentially abundant subdomains at `r compare.age.text` months.

```{r sbd, fig.width = 5, fig.height = 5}
sub.bd.stats <- plot_nested_list(nested.gene.list = sub.bd.list, centered.gene.mat, "Subdomains", plot.results = FALSE)
n.sbd.genes <- lapply(sub.bd.stats, function(x) sapply(x, function(y) if(length(y) > 0){ncol(y$expr_mat)}else{NA}))
pass.thresh <- which(unlist(n.sbd.genes) >= 10)
age.sub.bd <- get_nested_stat(nested.list.result = sub.bd.stats, stat.label = paste0("geno_", compare.age.text), stat.type = "model.p")
qqunif.plot(unlist(age.sub.bd)[pass.thresh],  plot.label = "Subdomain p values")
```

There were no significantly differentially abundant 
KEGG-biodomain intersections.


```{r bdk}
bd.kegg.stats <- plot_nested_list(nested.gene.list = kegg.bd.list, 
    centered.gene.mat, list.label = "KEGG-Biodomains", plot.results = FALSE)
n.bdk.genes <- lapply(bd.kegg.stats, function(x) sapply(x, function(y) if(length(y) > 1){ncol(y$expr_mat)}else{NA}))
#hist(unlist(n.bdk.genes), breaks = 100)
pass.thresh <- unlist(lapply(n.bdk.genes, function(x) which(x >= 10)))
age.kegg.bd.p <- unlist(get_nested_stat(nested.list.result = bd.kegg.stats, 
    stat.label = paste0("geno_", compare.age.text), stat.type = "model.p"))
age.kegg.bd.effect <- unlist(get_nested_stat(nested.list.result = bd.kegg.stats, 
    stat.label = paste0("geno_", compare.age.text), stat.type = "linear.effect.size"))

qqunif.plot(age.kegg.bd.p[names(pass.thresh)],  plot.label = "KEGG-Biodomain p values")
```

The volcano plot is shown below. Significant intersections (FDR < 0.1)
are highlighted.

```{r bdk_volcano, fig.width = 5, fig.height = 5}
stat.table <- cbind(age.kegg.bd.effect[names(pass.thresh)], 
    age.kegg.bd.p[names(pass.thresh)])
colnames(stat.table) <- c("effect", "p")

adj.p <- p.adjust(stat.table[,"p"], method = "fdr")
#plot(-log10(stat.table[,"p"]), -log10(adj.p))
sig.idx <- which(adj.p <= 0.1)
#stat.table[sig.idx,]
sig.col <- rep("black", length(adj.p))
sig.col[sig.idx] <- "red"

plot(stat.table[,"effect"], -log10(stat.table[,"p"]), 
    col = sig.col, pch = 16, xlab = "Effect Size", 
    ylab = "-log10(p)", main = "KEGG-Biodomain Intersections")
```

## Biodomain-Kegg examples {.tabset .tabset-fade .tabset-pills}

If there are significant effects, we plot the significant ones below.
We show a maximum of 5.

```{r example, fig.width = 9, fig.height = 6, results = "asis"}
max.show <- 5
show.num <- min(c(max.show, length(sig.idx)))
if(show.num > 0){
    sig.int <- rownames(stat.table)[sig.idx]
    split.int <- strsplit(sig.int, ".", fixed = TRUE)
    sig.table <- Reduce("rbind", split.int)
    if(show.num == 1){
        sig.table <- matrix(sig.table, nrow = 1)
    }
    colnames(sig.table) <- c("Biodomain", "KEGG")

    #plot out the significant KEGG-Biodomain intersections
    #pdf(file.path(results.dir, "KEGG-Biodomain_intersections.pdf"), width = 9, height = 6)
    
    for(i in 1:show.num){
        cat("###", paste(sig.table[i,], collapse = " AND "), "\n")
        bd.idx <- which(names(kegg.bd.list) == sig.table[i,1])
        k.idx <- which(names(kegg.bd.list[[bd.idx]]) == sig.table[i,2])
        gene.names <- gene.info[match(kegg.bd.list[[bd.idx]][[k.idx]], gene.info[,"ensembl_gene_id"]), "external_gene_name"]
        plot_gene(gene.names, centered.gene.mat, autoplace.text = TRUE, 
        plot.label = paste(sig.table[i,], collapse = "\nAND\n"))
        cat("\n\n")
    }
    #dev.off()
}
```


Mike wants a figure comparing the genotype effects on 
the short list of AD genes that I had before from Christian 
Behl's book.

He thinks showing these side-by-side will make the paper fly 
high. I think it will look like a mess. Let's see, shall we?


```{r ad_short_list}
#read in transcript results
tx.ad.short <- as.matrix(read.delim(here("Results", paste0("age_batch_", compare.age, "_gene"), "AD_short_list_means.txt"), comment.char = "#"))

#find the common gene names and IDs between proteins and transcripts
common.ad <- intersect(rownames(tx.ad.short), colnames(centered.gene.mat))
common.id <- gene.info[match(common.ad, gene.info[,"external_gene_name"]), "ensembl_gene_id"]

#get the transcript stats for the age-specific animals
tx.beta <- tx.ad.short[,"effect"]
tx.p <- tx.ad.short[,"p"]

#find the significant genes with an FDR <= 0.1
sig.tx <- names(tx.p)[which(tx.p <= 0.01)]

#adjust the protein data for sex
ad.pr <- adjust(centered.gene.mat[,common.ad], dummy_covar(data.frame("sex" = sex, "age" = age)), retain.intercept = FALSE)
age.ad.pr <- ad.pr[age.idx,]
#calculate means across genotypes for the genes of interest in the age-spcific animals
ad.pr.mean <- sapply(levels(genotype), function(x) colMeans(age.ad.pr[which(genotype[age.idx] == x),], na.rm = TRUE))
#attach the effect sizes and p values from the linear models with the age-specific animals
pr.effect <-  age.geno.effect[common.ad]
pr.p <- age.geno.p[1,common.ad]
ad.pr.mean <- cbind(ad.pr.mean, "effect" = pr.effect, "p" = pr.p)


#align the transcript and protein means based on gene names
aligned.mats <- get.xz(tx.ad.short, ad.pr.mean)
#pull out the names of genes that made it through the alignment
final.names <- rownames(aligned.mats[[1]]) 
#also get the IDs
final.id <- gene.info[match(final.names, gene.info[,"external_gene_name"]), "ensembl_gene_id"]

effect.mat <- sapply(aligned.mats, function(x) x[,"effect"])
colnames(effect.mat) <- c("Transcript", "Protein")
p.mat <- sapply(aligned.mats, function(x) x[,"p"])

sig.idx <- apply(p.mat, 2, function(x) which(x <= 0.01))
sig.tx.names <- lapply(aligned.mats, rownames)
for(i in 1:length(sig.tx.names)){
    sig.flag = rep("", length(sig.tx.names[[i]]))
    sig.flag[sig.idx[[i]]] <- "**"
    sig.tx.names[[i]] <- paste0(sig.flag, sig.tx.names[[i]], sig.flag)
}

row.order <- order(plot.decomp(aligned.mats[[1]], plot.results = FALSE)$u[,1])
#row.order <- order(effect.mat[1,])
#row.order <- order(aligned.mats[[1]][,3] - aligned.mats[[2]][,1], decreasing = TRUE)
#row.order <- order(aligned.mats[[1]][,1])

global.min <- min(unlist(lapply(aligned.mats, function(x) x[,1:3])))
global.max <- max(unlist(lapply(aligned.mats, function(x) x[,1:3])))

png(file.path(results.dir, "AD_comparison.png"), width = 8, height = 6, units = "in", res = 300)
layout.mat <- matrix(1:4, nrow = 1)
layout(layout.mat, widths = c(0.5, 0.7,0.7,1))

par(mar = c(24, 4, 4, 4))
imageWithTextColorbar(Reduce("rbind", lapply(aligned.mats, function(x) x[,1:3])), 
    split.at.vals = TRUE, col.scale = list(heatmap.neg.col, heatmap.pos.col), 
    grad.dir = "ends", cex = 1, round.nearest = 0.1)
mtext("Mean Expression/Abundance (A.U.)", side = 2, cex = 0.7, line = 2.7)

par(mar = c(4,6,4,0))
imageWithText(aligned.mats[[1]][row.order,1:3], show.text = FALSE, split.at.vals = TRUE,
    col.scale = c(heatmap.neg.col, heatmap.pos.col), grad.dir = "ends",
    col.text.shift = 0.04, row.text.shift = 0.3, col.text.rotation = 0,
    col.text.adj = 0.5, global.color.scale = TRUE, global.min = global.min,
    global.max = global.max, row.names = sig.tx.names[[1]])
mtext("Transcript\nAbundance", side = 3, line = 0)

imageWithText(aligned.mats[[2]][row.order,1:3], show.text = FALSE, split.at.vals = TRUE,
    col.scale = c(heatmap.neg.col, heatmap.pos.col), grad.dir = "ends",
    col.text.shift = 0.05, row.text.shift = 0.3, col.text.rotation = 0, 
    col.text.adj = 0.5, global.color.scale = TRUE, global.min = global.min,
    global.max = global.max, row.names = sig.tx.names[[2]])
mtext("Protein\nAbundance", side = 3, line = 0)

par(mar = c(4,6,4,1))
barplot(t(effect.mat[rev(final.names),]), beside = TRUE, las = 2, horiz = TRUE, 
    main = "Effect size of genotype", col = c("gray20", "gray80"),
    xlab = bquote(beta))
plot.dim <- par("usr")
segments(x0 = 0, y0 = 0, y1 = plot.dim[4])
legend("bottomright", fill = c("gray20", "gray80"), legend = colnames(effect.mat), cex = 0.8)

dev.off()

#for the paper
global.min <- min(sapply(aligned.mats, function(x) min(x[,1:3])))
global.max <- max(sapply(aligned.mats, function(x) max(x[,1:3])))
row.order <- order(sapply(rownames(p.mat), function(x) which(rownames(tx.ad.short) == x)), decreasing = FALSE)
sig.text <- rep("", nrow(p.mat))
sig.text[which(p.mat[,2] <= 0.01)] <- "*"
png(here("Results", "for_paper", "AD-related_proteins.png"), width = 8, height = 3, units = "in", res = 300)
par(mar = c(6,4,4,2), bg = NA)
imageWithText(t(aligned.mats[[2]][row.order,1:3]), show.text = FALSE, split.at.vals = TRUE,
    col.scale = c(heatmap.neg.col, heatmap.pos.col), grad.dir = "ends",
    col.text.shift = 0.25, row.text.shift = 0.06, col.text.rotation = 90,
    col.text.adj = 1, row.text.adj = 0.5, global.color.scale = TRUE, 
    global.min = global.min, global.max = global.max, 
    col.names = str_to_upper(rownames(aligned.mats[[2]])[row.order]))
plot.dim <- par("usr")
for(i in 1:length(geno.cols)){
    draw.rectangle(min.x = plot.dim[1]-0.5, max.x = 0.4, min.y = length(geno.cols)-i+0.6, max.y = length(geno.cols)-i+1.4,
        border.col = geno.cols[i], lwd = 3)
}
#text(x = 1:length(short.ad.p), y = rep(plot.dim[4]+0.1, length(short.ad.p)), labels = sig.text[row.order])
dev.off()

#plot_gene("App", centered.gene.mat, autoplace.text = TRUE)
#plot.with.model(as.numeric(aligned.mats[[1]]), as.numeric(aligned.mats[[2]]))
```

## Overall correlations

We wanted to look at whether overall higher transcript 
levels led to higher protein levels.

The plots below show that overall there is no correlation
between transcripts and protein levels. 

Greg said that this is expected in the following statement: 
"I think mass spec is highly variable in how well it detects 
certain peptides, and while RNA seq is better it still has PCR 
biases in amplification that works better for some genes than 
others. In other data sets, we've sometimes seen correlation coeffs 
of 0.15-0.30 for genotype effects (or case vs control), but almost 
no correlation for overall abundance."

```{r cor_test, fig.width = 10, fig.height = 5}
#use tx data that has been adjusted for technical sources of variation, but not biological sources
tx.dir <- here("Results", paste0("age_batch_", compare.age, "_gene"))
all.tx <- readRDS(file.path(tx.dir, "processed_data", "Batch_Adjusted_Expr.RDS"))
tx.covar <- read.csv(file.path(tx.dir, "processed_data", "covariates_carriers.csv"))
#adjust out all known sources of variation. We just want to test whether the transcript
#level is associated with protein level, not whether sex or genotype affects 
#either one. We don't need to adjust for age, because we are only looking in
#one age group at a time.
tx.adj.mat <- cbind(dummy_covar(data.frame("sex" = as.factor(tx.covar[,"sex_ge"]), 
    "geno" = as.factor(tx.covar[,"ordered_geno"]))),
    as.numeric(tx.covar[,"scaled_Ide_expression"]))
#adj.tx <- adjust(all.tx, tx.adj.mat, retain.intercept = FALSE) #adjust and mean-center

#don't adjust
adj.tx <- all.tx
#boxplot(adj.tx[,1:10])

#also adjust proteins for everything. Ide abundance has already been regressed
#out of centered.gene.mat
pr.adj.mat <- dummy_covar(data.frame("sex" =  as.factor(final.mouse.table[,"sex"]),
    "geno" = as.factor(final.mouse.table[,"genotype"])))

#using age.idx doesn't really matter because the transcript data are for a single age
#adj.pr <- adjust(centered.gene.mat[age.idx,], pr.adj.mat[age.idx,], retain.intercept = FALSE)

#don't adjust
adj.pr <- centered.gene.mat
#adj.pr <- mean.gene.mat #unfiltered for sex-biased genes
#boxplot(adj.pr[,1:10])

pr.id <- gene.info[match(colnames(adj.pr), gene.info[,"external_gene_name"]),"ensembl_gene_id"]
colnames(adj.pr) <- pr.id

common.id <- intersect(pr.id, colnames(adj.tx))
common.ind <- intersect(rownames(adj.pr), rownames(adj.tx))
aligned.pr <- adj.pr[common.ind, common.id]
aligned.tx <- adj.tx[common.ind, common.id]

f.id <- intersect(rownames(tx.covar)[which(tx.covar[,"sex_ge"] == "Female")], common.ind)
m.id <- intersect(rownames(tx.covar)[which(tx.covar[,"sex_ge"] == "Male")], common.ind)
test.cor <- sapply(1:ncol(aligned.pr), function(x) cor(aligned.pr[,x], aligned.tx[,x], 
    use = "pairwise.complete.obs"))
f.test.cor <- sapply(1:ncol(aligned.pr), function(x) cor(aligned.pr[f.id,x], aligned.tx[f.id,x], 
    use = "pairwise.complete.obs"))
m.test.cor <- sapply(1:ncol(aligned.pr), function(x) cor(aligned.pr[m.id,x], aligned.tx[m.id,x], 
    use = "pairwise.complete.obs"))

par(mfrow = c(1,2))
plot(density(test.cor), xlab = "Correlations", main = "Transcript-Protein Correlations", lwd = 3)
points(density(f.test.cor), col = factor.cols["F"], type = "l", lwd = 3)
points(density(m.test.cor), col = factor.cols["M"], type = "l", lwd = 3)
legend("topright", legend = c("all", "M", "F"), lty = 1, lwd = 3, 
    col = c("black", factor.cols[c("M", "F")]))
#abline(v = mean(test.cor), col = "black", lwd = 3)
#abline(v = mean(f.test.cor), col = factor.cols["F"], lwd = 3)
#abline(v = mean(m.test.cor), col = factor.cols["M"], lwd = 3)

abline(v = 0, col = "red")
boxplot(list("all" = test.cor, "F" = f.test.cor, "M" = m.test.cor), 
    ylab = "Correlation", main = "Transcript-Protein Correlations")
abline(h = 0)
mtext(paste(compare.age, "months"), side = 3, outer = TRUE, line = -1.5, font = 2, cex = 1.5)
```

```{r more_comparison, eval = FALSE}
n.pc <- 10
tx.decomp <- plot.decomp(aligned.tx, pc = n.pc, plot.results = FALSE)
tx.pc <- tx.decomp$u
imp.pr <- impute.knn(aligned.pr)$data
pr.decomp <- plot.decomp(imp.pr, pc = n.pc, plot.results = FALSE)
pr.pc <- pr.decomp$u

pc.pairs <- pair.matrix(1:n.pc)
tx.to.pr <- pr.to.tx <- rep(NA, nrow(pc.pairs))
for(i in 1:nrow(pc.pairs)){
    pc1 <- pc.pairs[i,1]
    pc2 <- pc.pairs[i,2]
    tx.to.pr[i] <- cor(tx.pc[,pc1], pr.pc[,pc2])
    pr.to.tx[i] <- cor(pr.pc[,pc1], tx.pc[,pc2])
}
final.result <- cbind(pc.pairs, tx.to.pr, pr.to.tx)
hist(c(tx.to.pr, pr.to.tx), breaks = 15)
boxplot(list(tx.to.pr, pr.to.tx))
top1 <- which.max(abs(tx.to.pr))
top2 <- which.max(abs(pr.to.tx))

final.result[top1,]
final.result[top2,]

tx.pc1 <- final.result[top1,1]
pr.pc1 <- final.result[top1,2]
par(mfrow = c(1,2))
plot.with.model(tx.pc[,tx.pc1], pr.pc[,pr.pc1], report = "cor.test",
    col = as.numeric(covar.table[rownames(aligned.tx),"sex_ge"]),
    xlab = paste("Transcript PC", tx.pc1), ylab = paste("Protein PC", pr.pc1))
plot.with.model(tx.pc[,tx.pc1], pr.pc[,pr.pc1], report = "cor.test",
    col = as.numeric(covar.table[rownames(aligned.tx),"ordered_geno"]),
    xlab = paste("Transcript PC", tx.pc1), ylab = paste("Protein PC", pr.pc1))


tx.pc2 <- final.result[top2,2]
pr.pc2 <- final.result[top2,1]
par(mfrow = c(1,2))
plot.with.model(tx.pc[,tx.pc2], pr.pc[,pr.pc2], report = "cor.test",
    col = as.numeric(covar.table[rownames(aligned.tx),"sex_ge"]))
plot.with.model(tx.pc[,tx.pc2], pr.pc[,pr.pc2], report = "cor.test",
    col = as.numeric(covar.table[rownames(aligned.tx),"ordered_geno"]))

par(mfrow = c(1,2))
pr.models <- apply(pr.pc, 2, function(x) lm(x~covar.table[rownames(aligned.tx),"ordered_geno"]))
pr.effect <- sapply(pr.models, function(x) coef(x)[2])
barplot(pr.effect)

tx.models <- apply(tx.pc, 2, function(x) lm(x~covar.table[rownames(aligned.tx),"ordered_geno"]))
tx.effect <- sapply(tx.models, function(x) coef(x)[2])
barplot(tx.effect)

```

We looked at the enrichments of genes with strong positive
and strong negative correlations between transcript and protein.

The high-correlation pairs are enriched for organelle-associated
genes, and translation genes.

The low-correlation pairs are enriched for synaptic genes. 
This enrichment was predicted by pmid 30664243, citing the 
large distance between the cell body and synaptic regions
of neurons. 

The enrichment tables are shown below.

```{r enrich_cor, fig.widtdh = 7, fig.height = 7}
#check.cor <- m.test.cor
check.cor <- test.cor

high.thresh <- get.percentile(check.cor, 99)
low.thresh <- get.percentile(check.cor, 1)
hist(check.cor, breaks = 100)
abline(v = c(high.thresh, low.thresh))
high.cor <- colnames(aligned.pr)[which(check.cor >= high.thresh)]
low.cor <- colnames(aligned.pr)[which(check.cor <= low.thresh)]
high.enrich <- gost(high.cor, organism = "mmusculus", source = "GO")
low.enrich <- gost(low.cor, organism = "mmusculus", source = "GO")

plot.enrichment(high.enrich, plot.label = "High Transcript-Protein Correlations",
    num.terms = 25)
plot.enrichment(low.enrich, plot.label = "Low Transcript-Protein Correlations",
    num.terms = 25)

```

For genes that are very highly correlated with sex, we do see
positive correlations between transcript and protein. However,
once you get down below 40\% variance explained by sex, that
correlation disappears.

```{r high_cor_example, fig.width = 5, fig.height = 5}
sex.col <- as.numeric(as.factor(tx.covar[common.ind,"sex_ge"]))
gene.name <- names(sort(factor.effect.mat[,"sex"], decreasing = TRUE))[2]
#gene.name <- "Tppp"
#get_ortho(gene.name)
gene.id <- gene.info[which(gene.info[,"external_gene_name"] == gene.name), "ensembl_gene_id"]

#pdf("~/Desktop/sex_correlation.pdf", height = 5, width = 5)
plot.with.model(aligned.tx[common.ind,gene.id], aligned.pr[common.ind,gene.id], 
    report = "cor.test", main = gene.name, xlab = "Transript level", 
    ylab = "Protein level", col = sex.col)
#dev.off()
```

```{r sex_specific_QC, eval = FALSE}
#There are two Y chr genes expressed in the brain: Eif2s3y and Ddx3y
#here we check for sex-specific expression
gene.name <- "Eif2s3y"
gene.name <- "Eif2s3x"
gene.name = "Ddx3y"
gene.name = "Ddx3x"

gene.id <- gene.info[which(gene.info[,"external_gene_name"] == gene.name), "ensembl_gene_id"]

par(mfrow = c(1,2))
boxplot(aligned.tx[common.ind,gene.id]~covar.table[common.ind,"sex_ge"], 
    xlab = "", ylab = "Transcript Abundance", main = "Transcript")
boxplot(aligned.pr[common.ind,gene.id]~covar.table[common.ind,"sex_ge"], 
    xlab = "", ylab = "Protein Abundance", main = "Protein")
mtext(gene.name, side = 3, outer = TRUE, font = 2, cex = 1.2, line = -1.5)
```

## Correlation quality control

We are a little suspicious that our protein data might not
be very accurate, so we are conducting some tests. Greg said
that complement transcripts and proteins have been very highly
correlated in their past data sets. Here we look to see if this
is true here.


```{r check_individ, fig.width = 12, fig.height = 9}

complement.idx <- grep("complement ", gene.features[,"gene_desc"], ignore.case = TRUE)
check.name <- gene.features[complement.idx,"gene_name"]
comp.cor <- matrix(NA, nrow = length(check.name), ncol = 2)
colnames(comp.cor) <- c("r", "p")
rownames(comp.cor) <- check.name
#pdf("~/Desktop/complement_report.pdf", width = 12, height = 9)
par(mfrow = c(3,4))
for(i in 1:length(check.name)){
    gene.name <- check.name[i]
    gene.id <- gene.info[which(gene.info[,"external_gene_name"] == gene.name), "ensembl_gene_id"]
    if(is.na(gene.id) || length(gene.id) == 0){next()}
    in_data <- Reduce("intersect", list(gene.id, colnames(aligned.tx), colnames(aligned.pr)))
    if(length(in_data) == 0){next()}

    comp.cor[i,] <- plot.with.model(aligned.tx[,gene.id], aligned.pr[,gene.id], report = "cor.test", 
        main = gene.name, xlab = "Transript level", ylab = "Protein level")
    }
boxplot(comp.cor[,"r"], ylab = "Correlation (r)", main = "All correlatoins")
stripchart(comp.cor[,"r"], vertical = TRUE, pch = 16, method = "jitter", 
    add = TRUE, cex = 2)
abline(h = 0)

#dev.off()
```

## Comparison to Matrisome Module 42

Previous proteomic work in the lab has shown that the protein
signatures in the mouse models match the human condition better
than the transcriptomics. That is, proteins that are more abundant
in AD patients relative to controls are also more abundant in 
mouse models with introduced risk variants. 

Greg said that they see the largest signal in a module of genes
related to the "matrisome" module from pmid: 35115731. He sent
me the genes in this module along with some stats for each.

As a measure of quality control we wanted to check how the
Klotho proteomics stack up to that, assuming the VS allele is 
the protective allele. We have set up our models so that the 
order of the genotypes goes FC < WT < VS, so proteins with 
positive effects are up in the *protective* situation, not the risk 
situation. So we would like to see that the proteins in this module
are less abundant in the VS animals, i.e. have negative effects in
the linear model.

The figure below shows the overall effect for the whole module.
It is not different overall across genotypes in the 12-month old
animals.


```{r orthos, fig.width = 9, fig.height = 6.5}
get_ortho <- function(gene.name, target = "mouse"){
    gene.idx <- unlist(apply(orthos, 2, function(x) which(x == gene.name)))
    if(length(gene.idx) == 0){
        return(paste("Couldn't find", gene.name))
    }else{
        gene.line <- orthos[gene.idx,]
    }

    if(target == "mouse"){
        result <- gene.line[,c("Mouse.Ortholog.Name", "Mouse.Ortholog.Ensembl")]
    }else{
        result <- gene.line[,c("Human.Gene.Name", "Human.Ensembl")]
    }
    names(result) <- c("name", "id")
    return(return(result))
}

#get_ortho("PLOD3")

mod42 <- read_excel(here("Data", "general", "TreatAD_scores_M42_module.xlsx")) #matrisome module
mod42.genes <- mod42$GeneName
mod42.mouse <- t(sapply(mod42.genes, get_ortho))

#pdf("~/Desktop/Matrisome.pdf", width = 9, height = 6.5)
mod.result <- plot_gene(mod42.mouse[,"name"], centered.gene.mat, 
    autoplace.text = TRUE, plot.label = "Module 42", plot.label.cex = 2)
```

The plot below shows the results for the individual proteins in
the module. This is a volcano plot showing the effect size of 
genotype at 12 months on each protein against the -log10(p) of
the effect. Proteins with -log10(p) greater than 1 are labeled.

None of these effects are significant, however, the general
trend of the effect is in the right direction. All of these
proteins were more abundant in human AD patients. Here, they
are less abundant in VS animals, which is consistent with their
label of being protective. A few of the proteins, including Flt1
and Bdh2 were more highly abundant in the VS animals. This is 
not consistent with the human data.

```{r mod_ind, fig.width = 5, fig.height = 5}
included.proteins <- colnames(mod.result$expr_mat)
mod.gene.results <- vector(mode = "list", length = length(included.proteins))
for(i in 1:length(included.proteins)){
    mod.gene.results[[i]] <- plot_gene(included.proteins[i], centered.gene.mat, 
        plot.results = FALSE)
}
#dev.off()

twelve.month.effect <- sapply(mod.gene.results, function(x) as.numeric(x$stats["geno_twelve", "linear.effect.size"]))
twelve.month.p <- sapply(mod.gene.results, function(x) as.numeric(x$stats["geno_twelve", "linear.effect.p"]))
#pdf("~/Desktop/Module42_Effect.pdf")
plot(twelve.month.effect, -log10(twelve.month.p), xlim = c(-0.6, 0.6),
    main = "Twelve-month effects on Module 42 proteins")
pretty.big <- which(-log10(twelve.month.p) > 1)
text(twelve.month.effect[pretty.big], -log10(twelve.month.p)[pretty.big], 
    labels = included.proteins[pretty.big], pos = 4)
#dev.off()
```