---
title: "Klotho proteomics"
author: Anna L Tyler
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  html_document:
    code_folding: hide
    collapsed: no
    toc: yes
    toc_float: yes
---

## Introduction

The purpose of this workflow is to start looking at the proteomics
data from the Klotho mice.

```{r parameters}

rm(list = ls())

library(here)

general.data.dir <- here("Data", "general")
proteomics.data.dir <-  here("Data", "Proteomics")

use.carrier.status <- TRUE #if true, fits models with the ordered factor FC > WT > VS
                          #if false, fits modesl with the ordered factor FC/FC > FC/WT > WT/WT > WT/VS > VS/VS

general.processed.data.dir <- here("Results", "Processed_Data")

results.dir <- here("Results", "Proteomics")
if(!file.exists(results.dir)){dir.create(results.dir)}

processed.data.dir <- file.path(results.dir, "processed_data")
if(!file.exists(processed.data.dir)){dir.create(processed.data.dir)}

gene.dump.dir <- file.path(results.dir, "removed_genes")
if(!file.exists(gene.dump.dir)){dir.create(gene.dump.dir)}

tx.processed.data.dir <- here("Results", "all_gene", "processed_data")
mouse.data.dir <- here("Data", "Mouse")

adjust.ide = TRUE

compare.age <- 12

if(compare.age == 12){
    compare.age.text <- "twelve"
}
if(compare.age == 4){
    compare.age.text <- "four"
}

offline = FALSE #set to TRUE if offline. Enrichment queries will not be run.
                #peptide votes also require being online, although these are
                #not run by default
```

```{r load_code}
all.fun <- list.files(here("Code"), full.names = TRUE)
for(i in 1:length(all.fun)){source(all.fun[i])}
```

```{r geno_type}

geno.cols <- get_geno_cols(here("Data", "general", "geno_cols.csv"))
ordered.geno <- c("FC/FC", "WT/FC", "WT/WT", "WT/VS", "VS/VS")
if(use.carrier.status){
    geno.cols <- geno.cols[c("FC/FC", "WT/WT", "VS/VS")]
    names(geno.cols) <- c("FC", "WT", "VS")
    ordered.geno <- c("FC", "WT", "VS")
    carrier.text <- "carriers"
}else{
    carrier.text <- "all_geno"
}

factor.cols <- list("genotype" = geno.cols, "sex" = c("F" = "#99d8c9", "M" = "#bcbddc"), 
    "age" = c("4" = "#fdae6b", "12" = "#e6550d"))
#dot.col <- "gray50"
dot.col <- "#dd1c77"

heatmap.neg.col <- "purple"
heatmap.pos.col <- "brown"
```

```{r libraries, message = FALSE, warning = FALSE, error = FALSE}
needed.libraries <- c("pheatmap", "VennDiagram", "DT", "knitr", "impute", "gprofiler2",
    "vioplot", "igraph", "stringr", "easyPubMed", "tm", "readxl", "hexbin", "RColorBrewer",
    "httr2", "jsonlite", "dplyr", "Biostrings", "cluster", "mediation", "pkgcond", "bmediatR")

load_libraries(needed.libraries)
```

The meta data file looks weird. Under "line" all mice are listed as
either Klotho(VS) or Klotho(FC), but this seems to be completely 
independent of the other columns. Under "snp" there are Trem2 and
Apoe mice in addition to Klotho-FC and Klotho-VS, and these labels
do not correspond to the labels under "line." I am filtering to 
just the animals with actual Klotho genotypes.

```{r read_data}
gene.info <- read.delim(here("Data", "general", "mouse_gene_info.txt"))
#mouse.info <- read.csv(file.path(mouse.data.dir, "metadata_validated.csv"))
orthos <- read.delim(file.path(general.data.dir, "human.mouse.orthologs.txt"))

#read in the appropriate covariate table
covar.mat <- read.table(file.path(tx.processed.data.dir, 
    paste0("covariates_", carrier.text, ".csv")), sep = ",", header = TRUE, row.names = 1)
    
#convert to a data frame with appropriate ordered factors.
covar.table <- data.frame("sequencingBatch" = as.factor(covar.mat[,"sequencingBatch"]),
    "sex_ge" = as.factor(covar.mat[,"sex_ge"]),
    "age_batch" = ordered(covar.mat[,"age_batch"], levels = c("4", "12")),
    "ordered_geno" = ordered(covar.mat[,"ordered_geno"], levels = ordered.geno))

rownames(covar.table) <- rownames(covar.mat)
#write.table(covar.table, file.path(processed.data.dir, "covar_table.txt"), sep = "\t", quote = FALSE)

#fasta sequences for mouse proteins. Downloaded July 27, 2026 from UniProt website
fasta <- readAAStringSet(here("Data", "Mouse", "uniprotkb_proteome_UP000000589_2026_07_27.fasta"))

#our KEGG/biodomain nested list from 1a.KEGG_and_Biodomains.Rmd
kegg.bd.list <- readRDS(file.path(general.processed.data.dir, "Mouse_KEGG_Intersections_for_GSEA.RDS"))

rna <- read.delim(here("Data", "Mouse", "rsem.merged.gene_counts.tsv"))
```

```{r read_proteome}
pr.mouse <- read.delim(file.path(proteomics.data.dir, "observations.tsv"))

#choose one of the levels of analysis
#the file I got said they usually use the gene level or precursor level.

analysis.level <- "gene" #one abundance per gene name
#analysis.level <- "protein group" #possibly multiple abundances per gene name and multiple genes per gene group.
#analysis.level <- "precursor" #many abundances per gene; rawest form, more missing data.
#analysis.level <- "unique gene" #gene groups that map only to one gene

analysis.text <- gsub(" ", "_", analysis.level)

if(analysis.level == "gene"){
    gene.features <- read.delim(file.path(proteomics.data.dir, "gg.features.tsv"))
    gene.mat <- read.delim(file.path(proteomics.data.dir, "gg.expression.tsv"))
    gene.col <- "gene_name" #where gene names are in the feature matrix
    id.col = "gene_name" #where rownames of the abundance matrix are in the feature matrix
}

if(analysis.level == "protein group"){
    #protein groups
    gene.features <- read.delim(file.path(proteomics.data.dir, "pg.features.tsv"))
    gene.mat <- read.delim(file.path(proteomics.data.dir, "pg.expression.tsv"))
    gene.col <- "gene_name"
    id.col = "protein_group"
}

if(analysis.level == "precursor"){
    #precursor peptides
    gene.features <- read.delim(file.path(proteomics.data.dir, "pr.features.tsv"))
    gene.mat <- read.delim(file.path(proteomics.data.dir, "pr.expression.tsv"))
    gene.col = "Genes"
    id.col = "precursor_id"
}

if(analysis.level == "unique"){
    #unique gene features
    gene.features <- read.delim(file.path(proteomics.data.dir, "ug.features.tsv"))
    gene.mat <- read.delim(file.path(proteomics.data.dir, "ug.expression.tsv"))
    gene.col = "gene_name"
    id.col = "gene_name"
}

```

We checked for reproducibility across all samples by correlating their
measured values. There are a few samples that look funny. I will keep 
an eye on these as we go through the filtering steps.

```{r reprod}
gene.mat.info <- t(sapply(strsplit(colnames(gene.mat), "_"), function(x) x))

#look for reproducibility across samples
sample.cor <- cor(gene.mat, use = 'pairwise.complete.obs')
pheatmap(sample.cor, show_rownames = FALSE, show_colnames = FALSE)
#sample.decomp <- plot.decomp(sample.cor)
#hist(rowMeans(sample.cor), breaks = 50)
#pheatmap(sample.cor[order(sample.decomp$u[,1]),order(sample.decomp$u[,1])], cluster_rows = FALSE, cluster_cols = FALSE, show_rownames = FALSE, show_colnames = FALSE)
cor.means <- rowMeans(sample.cor)
low.cor <- which(cor.means < 0.94)

#There is nothing obviously different about these samples.
#pr.mouse[match(names(low.cor), pr.mouse[,1]),]
```


## Align mouse identities with transcriptomic identities

Match our transcriptomic mouse information with the proteomic mouse information.
There were a few changes we made in the transcriptomic data, so we want to make
sure everything matches.

```{r match_mats, fig.width = 5, fig.height = 5}
all.trans.meta <- read.csv(here("Data", "Mouse", "metadata_validated.csv"))
trans.ind <- all.trans.meta[,"animalName"]
prot.ind <- pr.mouse[,"animal_id"]
plotVenn(list("transcriptomic" = unique(all.trans.meta[,"animalName"]), "proteomic" = unique(prot.ind)))
```

There were animals in each data set that did not appear in the other.
The animals that were included in the proteomics data set but not the
transcriptomics data set appear below. They all appear to be from a 
Trem2 experiment. I didn't include these in the Klotho transcriptomic
analysis, so these aren't a problem.

```{r not_in_prot}
not.in.prot <- setdiff(trans.ind, prot.ind)
trans.info <- all.trans.meta[match(not.in.prot, all.trans.meta[,"animalName"]),]
datatable(trans.info)
```

The table below shows the animals that were in the proteomics data 
set, but not in the transcriptomics data set. 

```{r not_in_trans}
not.in.trans <- setdiff(prot.ind, trans.ind)
prot.info <- pr.mouse[match(not.in.trans, pr.mouse[,"animal_id"]),]
datatable(prot.info)
```

We will set up a mouse information table separately for the proteomics
data and compare to the information we have in the transcriptomics data
just to be sure everything matches. 

```{r parse_geno}
parse_geno <- function(allele.label, dip.label){
    kl.type <- gsub("Klotho", "", allele.label)
    
    if(use.carrier.status){
        if(dip.label == "WT"){
            return("WT")
        }else{
            return(kl.type)
        }
    }else{
        if(dip.label == "HOM"){
            kl.geno <- paste0(kl.type, "/", kl.type)
        }
        if(dip.label == "HET"){
            kl.geno <- paste0("WT/", kl.type)
        }
        if(dip.label == "WT"){
            kl.geno <- "WT/WT"
        }
        return(kl.geno)
    }
}
```

We generated a table for each observation in the proteomics data. 
There are multiple replicates per individual mouse.

```{r pr_covar}
pr.id <- pr.mouse[,"animal_id"]
pr.sex <- pr.mouse[,"sex"]

raw.age <- pr.mouse[,"age"]/30 #convert age in days to age in months
#hist(raw.age, xlab = "months") #just double check these are 4 and 12-month olds
pr.age.m <- bin.vector(raw.age, c(4, 12)) #bin ages into four and 12 months old
#plot(pr.age.m, raw.age)
pr.geno <- apply(pr.mouse[,c("klotho", "geno")], 1, function(x) parse_geno(x[1], x[2]))

split.pr.info <- strsplit(pr.mouse[,"obs_id"], "_")
pr.rep <- sapply(split.pr.info, function(x) tail(x, 1))
pr.run <- sapply(split.pr.info, function(x) x[1])
pr.unsure <- sapply(split.pr.info, function(x) x[3]) 
pr.date <- sapply(split.pr.info, function(x) x[4])

pr.info.table <- cbind(pr.id, pr.sex, pr.age.m, pr.geno, pr.rep, pr.run, pr.unsure, pr.date)
colnames(pr.info.table) <- c("animalID", "sex", "age_months", "genotype", "replicate", "run", "not_sure", "date")

write_table_with_header(pr.info.table, 
    file.path(processed.data.dir, "split_obsevation.txt"),
    sep = "\t", row.names = FALSE, header = "Mouse information for each observation. There are multiple observations per mouse because of technical replicates.")
```

We first went through the replicates to make sure that each individual
mouse had consistent information across all replicates. They all checked
out.

```{r rep_info}
u_id <- unique(pr.info.table[,"animalID"])
split_ind <- lapply(u_id, function(x) unique(pr.info.table[which(pr.info.table[,"animalID"] == x),c("animalID", "sex", "age_months", "genotype")]))
check.info <- unique(sapply(split_ind, nrow))
#all animals have one set of ID, sex, age, and genotype associated. Good!
```

We then aligned every mouse id in both the transcriptomic and
proteomic data sets to make sure the information was the same
in both.

```{r extract_info_fun}

get.pr.info <- function(animalID){
    animal.idx <- which(pr.info.table[,"animalID"] == animalID)[1]
    if(is.na(animal.idx)){
        animal.info <- c("animalID" = animalID, "sex" = NA, "age_months" = NA, "genotype" = NA)
    }else{
        animal.info <- pr.info.table[animal.idx,c("animalID", "sex", "age_months", "genotype")]
    }
    return(animal.info)
}

get.tr.info <- function(animalID){
    #use the covariate matrix because this has already gone through a QC process
    animal.idx <- which(rownames(covar.mat) == animalID)
    if(length(animal.idx) == 0){
        animal.info <- c("animalID" = animalID, "sex" = NA, "age_months" = NA, "genotype" = NA)
    }else{
        animal.sex <- substr(covar.mat[animal.idx, "sex_ge"], 1, 1)
        animal.age <- covar.mat[animal.idx,"age_batch"]
        animal.geno <- covar.mat[animal.idx,"ordered_geno"]
        animal.info <- c("animalID" = animalID, "sex" = animal.sex, "age_months" = animal.age, "genotype" = animal.geno)
    }
return(animal.info)
}

compare.info <- function(one.tr.info, one.pr.info){

    both.present <- TRUE
    
    #if the animal is not in the transcript data
    #or not in the protein data, just say so
    if(all(is.na(one.tr.info[2:length(one.tr.info)]))){
        check.result <- "Not present in transcript data"
        both.present = FALSE
    }
    if(all(is.na(one.pr.info[2:length(one.pr.info)]))){
        check.result <- "Not present in proteomic data"
        both.present = FALSE
    }

    if(both.present){
        #otherwise, compare data bitwise 
        check.stats <- sapply(1:length(one.tr.info), function(x) one.tr.info[x] == one.pr.info[x])
        if(all(check.stats)){
            check.result <- "Pass"
        }else{
            check.result <- "Fail"
        }
    }
    return(check.result)
}

#get meta-data information from finalized table
get_final_info <- function(animalID){
    animal.idx <- which(all.pr.info[,"animalID"] == animalID)
    return(all.pr.info[animal.idx,])
}

get_gene_names <- function(bd.name, kegg.name){
    bd.idx <- which(names(kegg.bd.list) == bd.name)
    k.idx <- which(names(kegg.bd.list[[bd.idx]]) == kegg.name)
    bd.k.id <- kegg.bd.list[[bd.idx]][[k.idx]]
    #cat(bd.k.id, sep = "\n")
    bd.k.names <- gene.info[which(gene.info[,"ensembl_gene_id"] %in% bd.k.id),"external_gene_name"]
    return(bd.k.names)
}
```

### Non-matches

```{r align}
all.id <- unique(c(prot.ind, trans.ind))
#plotVenn(list("prot" = unique(prot.ind), "trans" = unique(trans.ind)))
all.tr.info <- t(sapply(all.id, get.tr.info))
all.pr.info <- t(sapply(all.id, get.pr.info))

check.match <- sapply(1:length(all.id), 
    function(x) compare.info(one.tr.info = all.tr.info[x,], one.pr.info = all.pr.info[x,]))
fail.idx <- which(check.match == "Fail")
```

There were `r length(fail.idx)` IDs for which information in the 
transcriptomic and proteomic data did not match. 

The information from the transcriptomics data is below:

```{r}
kable(all.tr.info[fail.idx,])
```

The information from the proteomics data is below:

```{r}
kable(all.pr.info[fail.idx,])
```

## Action taken for non-matches

We have seen a couple of these mice before. In the 
transcriptomics data we converted 83901 to male 
based on transcriptomic evidence. We will also convert 
it to male here.

78933 was listed as female in the climb data and male by the
transcriptomic data. We used the climb sex (verified by 
Dylan), and we think that it might be an XO female based
on transcriptomic evidence. It does not need to be adjusted
here.

The mouse 89055 had inconclusive genotyping, but it looked
as if it might have at least one FC allele. However, it is
listed as WT in the proteomics meta data. The form that 
Dylan submitted also listed this animal as WT. I think 
the inconclusive genotyping might have misled me to believe
this mouse was FC. I will convert this to WT in the transcriptomics
data set.

```{r fail_checking}
fail.id <- all.id[fail.idx]
fail.table <- cbind(all.tr.info[fail.idx,], all.pr.info[fail.idx,])

switch.id <- "83901"
switch.idx <- which(all.pr.info[,"animalID"] == switch.id)
all.pr.info[switch.idx,"sex"] <- "M"
```

## Missingness

Now we are ready to start looking into the protein data. There 
were three replicates done per animal, and there are some other 
numbers in the observation ID that I'm not yet sure what to do 
with. I'm in touch with Mitch and Tim about this. 

We'll want to look at missing data across replicates to start
thinking about what we want to do with that.

```{r read_pr}
split_id <- Reduce("rbind", strsplit(colnames(gene.mat), "_"))
colnames(split_id) <- c("run", "submitter", "number", "date", "strain", "allele", "sex", "diplotype", "L7_id", "rep")
rownames(split_id) <- colnames(gene.mat)
animal_id <- sapply(rownames(split_id), function(x) pr.mouse[which(pr.mouse[,"obs_id"] == x), "animal_id"])
all.obs.info <- t(sapply(animal_id, get_final_info))

#split observations up by unique animal ID
split_obs <- lapply(all.pr.info[,"animalID"], function(x) t(gene.mat[,which(all.obs.info[,"animalID"] == x)]))

missing.by.obs <- sapply(split_obs, function(x) apply(x, 2, function(y) length(which(!is.na(y)))))

#pheatmap(missing.by.obs, show_rownames = FALSE)

#find peptides and mice with no data at all
genes.missing.per.mouse <- apply(missing.by.obs, 2, function(x) length(which(x == 0)))
mice.missing.per.gene <- apply(missing.by.obs, 1, function(x) length(which(x == 0)))
```

### Missing by mouse

The bar plot below shows how many observations out of 
`r nrow(missing.by.obs)` are missing for each mouse. 
There are a subset of mice with no data at all.

```{r num_missing}
barplot(sort(genes.missing.per.mouse), ylab = "Number of Peptides", 
    main = "Number of missing genes per mouse", xlab = "mouse")
no.data <- which(genes.missing.per.mouse == nrow(missing.by.obs))
```

The table below shows information for the mice that had no data.
We had no metadata for them either.

```{r no_data}
kable(all.pr.info[no.data,])
```

Build an object that contains all the data and metadata.
We will carry this through all operations. We removed the
mice with no data.

```{r remove_no_data}
#make an object to hold all data and metadata together
obs_data <- list("mouse_info" = all.pr.info, "sample_info" = all.obs.info, 
    "pr_info" = gene.features, "data" = gene.mat, "gene.name.col" = gene.col,
    "gene.id.col" = id.col)

#remove mice with no data and genes with more than 20% missingness
remove.mice <- all.pr.info[no.data,"animalID"]
obs_data <- remove_mice(sample_data = obs_data, mice.to.remove = remove.mice)
```

Log normalize the data.

```{r log_norm_and_average}
obs_data <- log_norm_data(obs_data)
```

### Missing by gene

The bar plot below shows the number of mice with missing
data per gene. There were `r ncol(missing.by.obs)` mice. 
The horizontal line shows the 50\% mark.

```{r missing_genes}
missing.thresh <- ncol(missing.by.obs)*0.5
barplot(sort(mice.missing.per.gene), names = NA, 
    main = "Number of missing mice by gene", xlab = "gene",
    ylab = "Number of genes")
abline(h = missing.thresh)

lots.missing <- which(mice.missing.per.gene >= missing.thresh)
```

There were `r length(lots.missing)` genes out of `r nrow(missing.by.obs)` 
for which at least 50\% of the mice had missing data. 

We removed the mice with no data, and removed the peptides with more
than 50\% missingness across the mice.


```{r filter}
remove.genes <- names(lots.missing)
write_table_with_header(remove.genes, 
    file.path(gene.dump.dir, "High_Missingness_Genes.txt"),
    header = "Peptides removed due to overall missing data.", col.names = FALSE)
obs_data <- remove_genes(obs_data, rownames(remove.genes))
```

## Variance across replicates

Before we summarize the counts across replicates, we want to
take a look at how stable the different replicates are. 

The histogram below shows the the coefficient of variance
for all genes. This is the standard deviation divided by
the mean. We will set our cutoff at 0.1.


```{r rep_var}
log.gene.mat <- obs_data$log_data
sample.mouse.id <- obs_data$sample_info[,"animalID"]
u_mouse.id <- unique(sample.mouse.id)

#The minimum correlation across replicates is 0.995. Pretty good!
gene.cor <- lapply(u_mouse.id, function(x) cor(log.gene.mat[,which(sample.mouse.id == x)], use = "pairwise.complete.obs"))
rep.cor <- t(sapply(gene.cor, function(x) x[upper.tri(x, diag = FALSE)]))
#min(rep.cor)

#This takes a long time, so we will save it.
pep.cv.file <- here("Results", "Proteomics", paste0("Peptide_Variation_", analysis.text, ".RDS"))
if(!file.exists(pep.cv.file)){
    pep.var <- sapply(u_mouse.id, function(x) apply(log.gene.mat[,which(sample.mouse.id == x)], 1, function(y) sd(y, na.rm = TRUE)))
    pep.mean <- sapply(u_mouse.id, function(x) rowMeans(log.gene.mat[,which(sample.mouse.id == x)], na.rm = TRUE))
    pep.cv <- pep.var/pep.mean
    saveRDS(pep.cv, pep.cv.file)
}else{
    pep.cv <- readRDS(pep.cv.file)
}


hist(pep.cv, main = "Coefficient of Variation", breaks = 50)
abline(v = 0.1)
```


Greg wants to err on the side of throwing out anything that has
a whif of irregularity, so I'm just tossing proteins with CV > 0.1.

```{r var_hist}
too.high <- unique(which(pep.cv > 0.1, arr.ind = TRUE)[,1])

#length(too.high)
#sort(rownames(pep.cv)[too.high])
write_table_with_header(sort(rownames(pep.cv)[too.high]), 
    file.path(gene.dump.dir, "High_variance.txt"), col.names = FALSE,
    header = "Genes with variance across replicates exceeding 99.9% of all genes")
```

There are `r length(too.high)` peptides with CVs that exceed this threshold
in any mouse. 


```{r filter_var}
remove.peps <- rownames(pep.cv)[too.high]
obs_data <- remove_genes(obs_data, genes.to.remove = remove.peps)
```

## IDE adjustment

As with the transcriptomes, we adjust for IDE abundance in case that is
a problem. It does have bimodal protein abundance, as shown below.

```{r ide_abundance, fig.width = 9, fig.height = 6.5}
plot_pr_abund(gene.name = "Ide", sample_data = obs_data, 
    autoplace.text = TRUE, data.type = "log")
```

```{r ide}
ide.abund <- peptide_vals("Ide", obs_data, data.type = "log")
log.abund <- get_data(obs_data, data.type = "log")

ide.adjusted <- t(adjust(t(log.abund), t(ide.abund), retain.intercept = TRUE))
obs_data$log_data <- ide.adjusted
```

## Mean abundance across replicates

We then took the mean of the observations across the replicates.
We only calculated the mean for groups with values measured in
at least two replicates.

```{r mean_reps}
#for each unique animal ID, find all replicates and 
#average across them.
obs_data <- mean_abundance(obs_data, min.rep = 2)
```


```{r peptide_fun}
#Before we do anything else, I want to look for genes that are expressed 
#in only one sex, or very different between sexes. We will do this by 
#replacing NAs with 0s

#Eif2s3y looks good! It is only expressed in males, except for 
#a single run in one female that registered some expression, but
#the other two runs for that female had 0 detected. So not a sample-swap.
#DDx3x I think also looks good. That should be higher in females, but
#not by a huge amount. I'm not sure what to make of A1bg. This is encoded
#on Chr 15 and has no information about it on mouseMine. I'm not sure
#why we'd see it only in females.

parse_geno <- function(info.row){
    geno <- gsub("Klotho", "", info.row[6])
    num <- info.row[8]
    if(num == "WT"){
        return("WT")
    }else{
        return(geno)
    }
}

```


```{r peptide_check, eval = FALSE}
#I used this code to check up on the peptide precursors assigned
#to genes, because we were seeing a strange pattern in Ddx3y. 
#Ddx3y did have some mis-assigned protein precursors, but other
#genes seem fine. I'm commenting out this code, but will keep 
#it in case we see other weird genes. It really only works
#with the precursor data.

gene.names <- c("Ddx3y", "Ddx3x", "Eif2s3y")

gene.name = "Csnk2"
#glut.syn <- c("Slc17a7", "Slc17a6", "Grin1", "Grin2b", "Glul", "Sn1") 
#gene.name <- glut.syn[2]
#u_genes <- unique(obs_data$pr_info[,gene.name.col])
#gene.name  <- sample(u_genes, 1)

pdf(paste("~/Desktop/spot_check_", gene.name, ".pdf"), width = 12, height = 9)
    result <- plot_peptide_vote(gene.name, obs_data, data.type = "log")

    if(length(result) == 1){
        plot.text(paste("No data for", gene.name))
        next()
    }

    all_cor <- result$cross_gene_correlations
    in_gene_cor <- result$in_gene_cor
    all_cor$query_gene <- result$in_gene_cor
    if(!is.null(all_cor)){
        boxplot(all_cor, main = paste("peptide correlations for", gene.name))
        abline(h = seq(-1, 1, 0.25), lty = 2, col = "gray")
        abline(h = 0)
    }

    if(length(unique(result$gene_vote)) > 1){
        all.mean.vals <- colMeans(result$peptide_abundance)
        match.mean.vals <- colMeans(result$peptide_abundance[which(result$gene_vote == gene.name),])
        plot.with.model(all.mean.vals, match.mean.vals, xlab = "Mean with all peptides", 
            ylab = "Mean with matching peptides", main = gene.name)
    }

    common.pep <- intersect(rownames(result$peptide_abundance), rownames(obs_data$mean_abundance))
    pep_means <- obs_data$mean_abundance[common.pep,,drop=FALSE]
    pep_test <- vector(mode = "list", length = nrow(pep_means))
    names(pep_test) <- rownames(pep_means)

    mouse_var <- get_factor_var(obs_data, data.type = "mean")
    for(i in 1:length(pep_test)){
        test.df <- data.frame("abund" = pep_means[i,], mouse_var)
        pep_test[[i]] <- lm(abund~sex+age+genotype, data = test.df)
    }
    all.p <- t(sapply(pep_test, function(x) coef(summary(x))[2:4,"Pr(>|t|)"]))
    if(nrow(all.p) > 1){
        pheatmap(-log10(all.p), main = paste(gene.name, "peptides"))
    }else{
        barplot(-log10(all.p), ylab = "-log10(p)")
    }

    pep.coord <- pep_coord(gene.name, obs_data, fasta, data.type = "log")

    
    par(mfrow = c(1,3))
    for(f in 1:ncol(all.p)){
        plot.new()
        plot.window(xlim = c(min(pep.coord), max(pep.coord)), ylim = c(0, max(-log10(all.p))))
        axis(1);axis(2)
        for(i in 1:nrow(pep.coord)){
            pep <- rownames(pep.coord)[i]
            pep.idx <- which(rownames(all.p) == pep)
            segments(x0 = pep.coord[i,1], x1 = pep.coord[i,2], y0 = -log10(all.p[pep.idx,f]), lwd = 3)
        }
        mtext(colnames(all.p)[f], side = 3)
        mtext("-log10(p value)", side = 2, line = 2.5)
        mtext("Peptide Position", side = 1, line = 2.5)
    }

dev.off()

dummy.var <- dummy_covar(mouse_var[,c("sex", "age")])

png(paste0("~/Desktop/peptide_cor_", gene.name, ".png"), width = 9, height = 6, units = "in", res = 300)
    geno.idx <- lapply(levels(mouse_var$genotype), function(x) which(obs_data$mouse_info[,"genotype"] == x))
    pep_geno <- lapply(geno.idx, function(x) pep_means[,x,drop=FALSE])
    adj_pep <- lapply(pep_geno, function(x) adjust(t(x), dummy.var))
    pep_cor <- lapply(adj_pep, function(x) cor(x, use = "pairwise.complete.obs"))

    if(nrow(pep_geno[[1]]) > 1){
        row.order <- hclust(dist(pep_cor[[2]]))$order
    

        par(mfrow = c(2,3))
        for(i in 1:length(adj_pep)){
            imageWithText(pep_cor[[i]][row.order, row.order], use.pheatmap.colors = TRUE, show.text = FALSE,
                main = levels(mouse_var$genotype)[i], col.text.cex = 0.5, row.text.cex = 0.5)
        }

        all.cor <- sapply(pep_cor, as.numeric)
        geno.pairs <- pair.matrix(c(1:3))
        for(i in 1:nrow(geno.pairs)){
            geno1 <- geno.pairs[i,1]
            geno2 <- geno.pairs[i,2]
            plot.with.model(all.cor[,geno1], all.cor[,geno2], xlab = levels(mouse_var$genotype)[geno1],
                ylab = levels(mouse_var$genotype)[geno2])
        }
    }
dev.off()

var.mat <- get_factor_var(obs_data, data.type = "mean")

glut.syn <- c("Slc17a7", "Slc17a6", "Grin1", "Grin2b", "Glul", "Sn1") 
gene.name <- glut.syn[3]


pep_abund <- peptide_vals(gene.name, obs_data, data.type = "mean")
pep.coord <- pep_coord(gene.name, obs_data, fasta)
test <- decomp_with_nas(pep_abund)
pheatmap(pep_abund[order(test$u[,1]),], show_colnames = FALSE, cluster_cols = FALSE, cluster_rows = FALSE)

test <- decomp_with_nas(t(pep_abund), label.points = TRUE, add.zeros = TRUE)
plot(test$u[,1], test$u[,2], pch = 16, col = as.numeric(var.mat[,"sex"]))
k.test <- test.pam.k(test$u, plot.results = FALSE)
k.means <- sapply(k.test$cl.width, mean)
best.k <- names(k.means)[which.max(k.means)]
plot(test$u[,1], test$u[,2], pch = 16, col = as.numeric(k.test$mem[,best.k]))

#clusters seem to be driven by peptide missingness
mem <- k.test$mem[,best.k]
mem.ind <- lapply(1:max(mem), function(x) colnames(pep_abund)[which(mem == x)])
mem.df <- data.frame(as.factor(mem))
rownames(mem.df) <- colnames(pep_abund)
pheatmap(pep_abund[rownames(pep.coord),order(mem)], show_colnames = FALSE, 
    cluster_rows = FALSE, cluster_cols = FALSE, annotation_col = mem.df)


pheatmap(cor(pep_abund, use = "pairwise.complete.obs"), show_colnames = FALSE, show_rownames = TRUE,
    main = "Correlation between individuals")

pheatmap(cor(t(pep_abund), use = "pairwise.complete.obs"), show_colnames = FALSE, show_rownames = TRUE,
    main = "Correlation between peptides")

plot.new()
plot.window(xlim = c(min(pep.coord, na.rm = TRUE), max(pep.coord, na.rm = TRUE)),
    ylim = c(floor(min(pep_abund, na.rm = TRUE)), ceiling(max(pep_abund, na.rm = TRUE))))
for(i in 1:nrow(pep_abund)){
    if(is.na(pep.coord[i,1])){next()}
    stripchart(t(pep_abund[rownames(pep.coord)[i],]), add = TRUE, at = pep.coord[i,1],
        vertical = TRUE, method = "jitter", pch = 16, cex = 0.5)
}
axis(1); axis(2)
mtext("Abundance", side = 2, line = 2.5)
mtext("Peptide Position", side = 1, line = 2.5)
mtext(gene.name, side = 3)

tx.id <- gene.info[which(gene.info[,"external_gene_name"] == gene.name), "ensembl_gene_id"]
tx.idx <- which(rna[,"gene_id"] == tx.id)
tx.abund <- rna[tx.idx,3:ncol(rna)]
#plot(log2(as.numeric(tx.abund)))
common.ind <- intersect(colnames(pep_abund), gsub("X", "", names(tx.abund)))
pep.ind.idx <- match(common.ind, colnames(pep_abund))
tx.ind.idx <- match(common.ind, gsub("X", "", names(tx.abund)))

tx.pep.cor <- apply(pep_abund[,pep.ind.idx], 1, 
    function(x) cor(log2(as.numeric(tx.abund[tx.ind.idx])), x, use = "pairwise.complete.obs"))
hist(tx.pep.cor)

best.cor <- which.max(tx.pep.cor)

plot.with.model(log2(as.numeric(tx.abund[tx.ind.idx])), pep_abund[names(best.cor),pep.ind.idx],
    xlab = "Transcript Abundance", ylab = "Peptide Abundance")

```


## Filter for even distribution

We want to make sure that values are relatively evenly spread
across the groups, sex, age, and genotype.

Below we count the number of non NA values we have in each subset.
The bar plot below shows how many animals there are of each sex.

```{r filter_sex}
mouse.factors <- get_factor_var(obs_data, data.type = "mean")

sex.idx <- lapply(levels(mouse.factors$sex), function(x) which(mouse.factors$sex == x))
names(sex.idx) <- levels(mouse.factors$sex)
barplot_with_num(sapply(sex.idx, length), text.gap = 0.25, text.shift = 0.15,
     main = "Number of mice of each sex")
```

```{r rep_thresh}
rep.thresh = 0.2
```

The dot plot below shows how many animals of each sex have counts in 
each gene. We removed genes with representation below 
`r rep.thresh*100`\%

```{r gene_count_sex, fig.width = 5, fig.height = 5}
#for each gene get the proportion of animals in each sex
#with complete data

sex.count <- t(apply(obs_data$mean_abundance, 1, function(x) sapply(levels(mouse.factors[,"sex"]), 
    function(y) c(length(which(!is.na(x[sex.idx[[y]]]))))/length(which(mouse.factors[,"sex"] == y)))))
colnames(sex.count) <- levels(mouse.factors$sex)

plot(sex.count, main = "Percentage represented per sex", xlim = c(0, 1), ylim = c(0, 1))
abline(0,1)
abline(v = rep.thresh, h = rep.thresh, col = "red")
to.remove <- which(sex.count < rep.thresh, arr.ind = TRUE)

if(nrow(to.remove) > 0){
    write_table_with_header(unique(rownames(to.remove)), col.names = FALSE,
        file.path(gene.dump.dir, "Unbalanced_Sex.txt"), 
        header = "Genes removed because of low representation in at least one sex. Less than 20% of animals represented.")

    text(sex.count[to.remove,1], sex.count[to.remove,2], 
        labels = rownames(sex.count)[to.remove], pos = 4, cex = 0.5)
}

obs_data <- remove_genes(obs_data, rownames(to.remove))
```

The bar plot below shows how many animals are in each age group.

```{r age}
age.idx <- lapply(levels(mouse.factors$age), function(x) which(mouse.factors$age == x))
names(age.idx) <- levels(mouse.factors$age)
barplot_with_num(sapply(age.idx, length), text.gap = 5, text.shift = 1, 
    main = "Mice represented in each age", ylab = "Number of mice")
```

The dot plot shows how many 4 and 12-month olds represent each 
gene. We did not remove any genes based on imbalanced representation
across the ages.

```{r gene_count_age, fig.width = 5, fig.height = 5}
age.count <- t(apply(obs_data$mean_abundance, 1, 
    function(x) sapply(levels(mouse.factors$age), 
    function(y) c(length(which(!is.na(x[age.idx[[y]]])))/length(which(mouse.factors$age == y))))))
colnames(age.count) <- levels(mouse.factors$age)

plot(age.count, main = "Percentage represented per age", xlim = c(0,1), ylim = c(0,1))
abline(0,1)
abline(h = rep.thresh, v = rep.thresh, col = "red")
to.remove.idx <- which(age.count < rep.thresh, arr.ind = TRUE)
to.remove <- unique(rownames(to.remove.idx))

if(nrow(to.remove.idx) > 0){
    write_table_with_header(unique(rownames(to.remove)), col.names = FALSE,
        file.path(gene.dump.dir, "Unbalanced_age.txt"), 
        header = "Genes removed because of low representation in at least one age group. Less than 20% of animals represented.")

   text(age.count[to.remove,1], age.count[to.remove,2], 
        labels = rownames(age.count)[to.remove], pos = 4, cex = 0.5)
}

obs_data <- remove_genes(obs_data, to.remove)
```

The bar plot shows how many animals are of each genotype.

```{r genotype_rep}
geno.idx <- lapply(levels(mouse.factors$genotype), function(x) which(mouse.factors$genotype == x))
names(geno.idx) <- levels(mouse.factors$genotype)
barplot_with_num(sapply(geno.idx, length), text.gap = 3, text.shift = 0.5, 
    main = "Mice per genotype")
```

The dot plots compare for each gene how many animals in 
each genotype have values for that gene. They are pretty well 
balanced. We did not remove any genes based on imbalanced
representation across genotypes.

```{r geno_balance, fig.height = 7, fig.width = 7}
geno.count <- t(apply(obs_data$mean_abundance, 1, 
    function(x) sapply(levels(mouse.factors$genotype), 
    function(y) c(length(which(!is.na(x[geno.idx[[y]]]))))/length(which(mouse.factors$genotype == y)))))
colnames(geno.count) <- levels(mouse.factors$genotype)

pairs(geno.count, main = "Percentage represented per genotype", 
    xlim = c(0,1), ylim = c(0,1))

to.remove <- which(geno.count < rep.thresh, arr.ind = TRUE)

if(nrow(to.remove)){
    write_table_with_header(unique(rownames(to.remove)), col.names = FALSE,
        file.path(gene.dump.dir, "Unbalanced_Genotype.txt"), 
        header = "Genes removed because of low representation in at least one genotype. Less than 20% of animals represented.")
}

obs_data <- remove_genes(obs_data, rownames(to.remove))
```

I think we should probably look at combinations of factors too.
We can't test factor interactions without sufficient samples
in each group combination.

```{r combo_count}
mouse.table <- get_factor_var(obs_data, data.type = "mean")
dat.mat <- get_data(obs_data, "mean")
min.rep <- 3 #there should be at least three individuals with data for any group
all.combos <- NULL
all.counts <- NULL
for(s in 1:length(levels(mouse.table[,"sex"]))){
    for(a in 1:length(levels(mouse.table[,"age"]))){
        for(g in 1:length(levels(mouse.table[,"genotype"]))){
            full.combo <- c(levels(mouse.table[,"sex"])[s], levels(mouse.table[,"age"])[a], levels(mouse.table[,"genotype"])[g])
            all.combos <- rbind(all.combos, full.combo)
            s.idx <- which(mouse.table[,"sex"] == levels(mouse.table[,"sex"])[s]) 
            a.idx <- which(mouse.table[,"age"] == levels(mouse.table[,"age"])[a]) 
            g.idx <- which(mouse.table[,"genotype"] == levels(mouse.table[,"genotype"])[g]) 
            full.idx <- Reduce("intersect", list(s.idx, a.idx, g.idx))
            sub.mat <- dat.mat[,full.idx]
            num.vals <- apply(sub.mat, 1, function(x) length(which(!is.na(x))))
            all.counts <- rbind(all.counts, num.vals)
            #hist(num.vals)
        }
    }
}

#look for any genes that have fewer than three representatives in any group
below.min <- which(all.counts < min.rep, arr.ind = TRUE)
low.rep.gene <- colnames(all.counts)[unique(below.min[,2])]

#remove these genes/peptides/precursors
obs_data <- remove_genes(obs_data, low.rep.gene)
```

Before centering we will look at the correlations between
samples again. There are still some funny looking samples.

```{r check_cor}
#check correlation before centering
cor.check <- cor(obs_data$mean_abundance, use = "pairwise.complete.obs")
pheatmap(cor.check, show_rownames = FALSE, show_colnames = FALSE)
#sample.decomp <- plot.decomp(cor.check)
#pheatmap(cor.check[order(sample.decomp$u[,1]),order(sample.decomp$u[,1])], cluster_rows = FALSE, cluster_cols = FALSE, show_rownames = FALSE, show_colnames = FALSE)
mean.cor <- rowMeans(cor.check)
#hist(mean.cor, breaks = 50)
low.cor <- which(mean.cor < 0.99)
```

They are the following: 

```{r low_cor_mice}
low.cor.mice <- get_mouse_info(obs_data, names(low.cor))
kable(data.frame(low.cor.mice))
```

The difference between a lower correlation and a higher one
doesn't seem huge.

```{r example_cor, fig.width = 8, fig.height = 4}
par(mfrow = c(1,2))
plot.with.model(obs_data$mean_abundance[,1], obs_data$mean_abundance[,3], main = "Low correlation",
    xlab = colnames(obs_data$mean_abundance)[1], ylab = colnames(obs_data$mean_abundance)[2])
plot.with.model(obs_data$mean_abundance[,1], obs_data$mean_abundance[,4], main = "High correlation",
    xlab = colnames(obs_data$mean_abundance)[1], ylab = colnames(obs_data$mean_abundance)[4])

```

Should we remove these low-correlation mice? The code below can do this
as a test. It is comment out now.

```{r remove_low_cor, eval = FALSE}
obs_data <- remove_mice(obs_data, low.cor.mice[,1])
```

```{r final_mouse_info}
final.mouse.table <- get_factor_var(obs_data, data.type = "mean")
write_table_with_header(final.mouse.table, 
    file.path(processed.data.dir, "covariates.txt"),
    header = "Final information for all mice included in the analysis.")
```

```{r scale_center}
obs_data <- scale_abundance(obs_data)
```

This leaves us `r nrow(obs_data$scaled_abundance)` genes, 
protein groups, or peptide precursors and 
`r ncol(obs_data$scaled_abundance)` mice.


## Factors and PCs

We then imputed missing data in order to perform a PC analysis of the 
matrix. We fit linear models explaining each of the first 10 PCs in the
protein matrix with age, sex, and genotype. The bar plot below shows the 
variance explained by each PC of the protein abundance matrix. 

```{r pc_exp, fig.width = 5, fig.height = 5}
n.pc <- 10

imp.gene.mat <- impute.knn(t(get_data(obs_data, "scaled")))$data
mouse.factors <- get_factor_var(obs_data, data.type = "scaled")
mean.decomp <- plot.decomp(imp.gene.mat, plot.results = FALSE, pc = n.pc)
mouse.pc <- mean.decomp$u
colnames(mouse.pc) <- paste0("PC", 1:ncol(mouse.pc))


barplot_with_num(signif(mean.decomp$var.exp[1:n.pc], 2)*100, ylab = "Percent Variance Explained", 
    ylim = c(0, 45), main = "Variance Explained by First 10 PCs")
abline(h = 0)
```

The bar plot below shows how much
variance is explained in each PC by each factor. Age explained about 20% 
of the variance in PC1 (which itself explained 31\% of the variance in the
protein matrix). The other factors did not have as much influence. So age 
explains about 6\% of the variance in the protein matrix overall.


```{r var_exp, fig.width = 7, fig.height = 5}
test.geno <- apply(mouse.pc, 2, function(x) lm(x~as.numeric(mouse.factors$genotype)))
geno.r2 <- sapply(test.geno, function(x) summary(x)$adj.r.squared)

test.sex <- apply(mouse.pc, 2, function(x) lm(x~as.numeric(mouse.factors$sex)))
sex.r2 <- sapply(test.sex, function(x) summary(x)$adj.r.squared)

test.age <- apply(mouse.pc, 2, function(x) lm(x~as.numeric(mouse.factors$age)))
age.r2 <- sapply(test.age, function(x) summary(x)$adj.r.squared)


all.r2 <- rbind(geno.r2, sex.r2, age.r2)
factor.colors = c("genotype" = "#7fc97f", "sex" = "#beaed4", "age" = "#fdc086")
barplot(all.r2, beside = TRUE, col = factor.colors, ylim = c(0, 0.25), ylab = c("R2"))
legend("topright", legend = names(factor.colors), fill = factor.colors)
abline(h = 0)
```

## Fit models

We fit linear models to protein abundance explaining each abundance
with age, sex, and genotype. We used the unimputed matrix for this.

The qq plots for the p values are shown below. All factors seem to 
have a large effect, but age has the largest as we saw abuve.

```{r get_lin_idx}
get_lin_idx <- function(model, factor.name){
    coef.mat <- summary(model)$coefficients
    want.idx <- grep(factor.name, rownames(coef.mat))
    dont.want.idx <- grep("Q", rownames(coef.mat))
    final.want <- setdiff(want.idx, dont.want.idx)
    #rownames(coef.mat)[final.want]
    return(final.want)
}

test_model <- function(sample_data, factor.name = "genotype", 
    data.type = c("raw", "log", "mean", "scaled")){

    if(length(data.type) > 1){data.type = "scaled"} #default to scaled

    all.factors <- c("sex", "age", "genotype")
    
    abund.mat <- get_data(sample_data, data.type)
    mouse.var <- get_factor_var(sample_data, data.type)
    has.var <- which(apply(mouse.var, 2, function(x) length(unique(x))) > 1)
    other.factors <- setdiff(names(has.var), factor.name)

    factor.text <- paste(c(other.factors, factor.name), collapse = "+")
    fmla <- paste("Expression ~", paste0("(", paste(c(other.factors, factor.name), collapse = " + "), ")^2"))
    full.mats <- lapply(1:nrow(abund.mat), function(x) data.frame("Expression" = abund.mat[x,], mouse.var))
    
    models <- lapply(full.mats, function(x) lm(as.formula(fmla), data = x))

    factor.idx <- get_lin_idx(models[[1]], factor.name)

    factor.p <- t(sapply(models, function(x) summary(x)$coefficients[factor.idx,"Pr(>|t|)"]))
    colnames(factor.p) <- gsub(".L", "", colnames(factor.p), fixed = TRUE)

    factor.effect <- t(sapply(models, function(x) summary(x)$coefficients[factor.idx,"Estimate"]))
    colnames(factor.effect) <- gsub(".L", "", colnames(factor.effect), fixed = TRUE)

    rownames(factor.p) <- rownames(factor.effect) <- rownames(abund.mat)

    result <- list("effect" = factor.effect, "p" = factor.p)

    return(result)
}
```

```{r fit_models, fig.width = 9, fig.height = 4}
#with interactions
age_effects <- test_model(obs_data, factor.name = "age", data.type = "scaled")
sex_effects <- test_model(obs_data, factor.name = "sex", data.type = "scaled")
genotype_effects <- test_model(obs_data, factor.name = "genotype", data.type = "scaled")
```

```{r get_old_young}
young.mice <- obs_data$mouse_info[which(obs_data$mouse_info[,"age_months"] == "4"),1]
old.mice <- obs_data$mouse_info[which(obs_data$mouse_info[,"age_months"] == "12"),1]
old_data <- remove_mice(obs_data, young.mice)
young_data <- remove_mice(obs_data, old.mice)
```

```{r test_old_young, eval = FALSE}
genotype_effects_young <- test_model(sample_data = young_data, factor.name = "genotype", data.type = "scaled")
genotype_effects_old <- test_model(sample_data = old_data, factor.name = "genotype", data.type = "scaled")

par(mfrow = c(1,2))
qqunif.plot(genotype_effects_young$p[,"gennotype"], plot.label = "Young")
qqunif.plot(genotype_effects_old$p[,"genotype"], plot.label = "Old")

sig.young <- which(p.adjust(genotype_effects_young$p[,"genotype"], "fdr") < 0.05)
sig.old <- which(p.adjust(genotype_effects_old$p[,"genotype"], "fdr") < 0.05)

sig.young.table <- obs_data$pr_info[match(names(sig.young), obs_data$pr_info[,id.col]),]
sig.old.table <- obs_data$pr_info[match(names(sig.old), obs_data$pr_info[,id.col]),]

sig.young.genes <- sig.young.table[,gene.col]
sig.old.genes <- sig.old.table[,gene.col]
plotVenn(list("young" = sig.young.genes, "old" = sig.old.genes))

if(!offline){
    intersect(sig.young.genes, sig.old.genes)
    young.enrich <- gost(sig.young.genes, organism = "mmusculus")
    old.enrich <- gost(sig.old.genes, organism = "mmusculus")
    plot.enrichment(young.enrich, num.terms = 25, max.term.size = 3000)
    plot.enrichment(old.enrich, num.terms = 25, max.term.size = 3000)
}
```

For each factor, we tested for main effects as well as interaction
effects with other factors. The plots below show qq plots for each
of these tests. There appear to be some strong interactions, particularly
between age and genotype, and to a lesser extent between sex and genotype.

```{r qq_age, fig.width = 9, fig.height = 9}
par(mfrow = c(3,3))
test <- sapply(1:ncol(age_effects$p), function(x) qqunif.plot(age_effects$p[,x], plot.label = colnames(age_effects$p)[x]))
test <- sapply(1:ncol(sex_effects$p), function(x) qqunif.plot(sex_effects$p[,x], plot.label = colnames(sex_effects$p)[x]))
test <- sapply(1:ncol(genotype_effects$p), function(x) qqunif.plot(genotype_effects$p[,x], plot.label = colnames(genotype_effects$p)[x]))
```

```{r dups}
factor.effect.mat <- cbind(age_effects$effect, sex_effects$effect, genotype_effects$effect)
factor.p.mat <- cbind(age_effects$p, sex_effects$p, genotype_effects$p)
factor.fdr.mat <- apply(factor.p.mat, 2, function(x) p.adjust(x, method = "fdr"))

#some of the columns are duplicates of each other. 
#age:genotype is the same as genotype:age. Remove
#the duplicates
all.col <- sapply(strsplit(colnames(factor.p.mat), ":"), function(x) paste(sort(x), collapse = ":"))
dup.col <- which(duplicated(all.col))
to.keep <- setdiff(1:ncol(factor.p.mat), dup.col)
factor.effect.mat <- factor.effect.mat[,to.keep]
factor.p.mat <- factor.p.mat[,to.keep]
factor.fdr.mat <- factor.fdr.mat[,to.keep]

write_table_with_header(factor.effect.mat, file.path(results.dir, "Protein_Abundance_Tests_Effects.csv"), 
    sep = ",", header = "Linear effect sizes (betas from linear models) for each protein abundance as it is affected by each tested factor.")
write_table_with_header(factor.p.mat, file.path(results.dir, "Protein_Abundance_Tests_P.csv"), 
    sep = ",", header = "P values for linear models for each protein abundance as it is affected by each tested factor.")
write_table_with_header(factor.p.mat, file.path(results.dir, "Protein_Abundance_Tests_FDR.csv"), 
    sep = ",", header = "FDR for linear models for each protein abundance as it is affected by each tested factor.")

```

## Volcano plots

The volcano plots are shown below all on the same scale. 
This highlights the huge effects of age relative to all 
other factors. 

```{r volcano, fig.width = 9, fig.height = 6}
ylim = c(0, max(-log10(factor.fdr.mat)))
xlim = c(min(factor.effect.mat)*1.05, max(factor.effect.mat)*1.2)

par(mfrow = c(2,3))
for(i in 1:ncol(factor.fdr.mat)){
    plot(factor.effect.mat[,i], -log10(factor.fdr.mat[,i]), 
        main = colnames(factor.p.mat)[i], xlab = "Effect", ylab = "-log10(p)",
        xlim = xlim, ylim = ylim)
    label.these <- which(-log10(factor.fdr.mat[,i]) > 10)
    label.names <- sapply(names(label.these), function(x) get_prg_info(x, obs_data)[gene.col])
    
    if(length(label.these) > 0){
        text(factor.effect.mat[label.these,i], -log10(factor.fdr.mat[label.these,i]), labels = label.names,
            pos = 4, cex = 0.7)
    }
}
```

```{r thresh}
fdr.thresh <- 0.05
```

Volcano plots on independend scales are shown below.
Colored dots are significant at an FDR of `r fdr.thresh`.
The different colors indicate positive and negative effect
sizes.

```{r volcano_diff, fig.width = 9, fig.height = 6}
sig.list <- vector(mode = "list", length = ncol(factor.fdr.mat))
names(sig.list) <- colnames(factor.fdr.mat)
par(mfrow = c(2,3))
for(i in 1:ncol(factor.p.mat)){
    pt.col = rep("black", nrow(factor.fdr.mat))
    sig.idx <- which(factor.fdr.mat[,i] <= fdr.thresh)
    down.idx <- which(factor.effect.mat[,i] < 0)
    down.sig <- intersect(sig.idx, down.idx)
    up.idx <- which(factor.effect.mat[,i] > 0)
    up.sig <- intersect(sig.idx, up.idx)
    factor.sig <- list("Up" = rownames(factor.fdr.mat)[up.sig], 
        "Down" = rownames(factor.fdr.mat)[down.sig])
    sig.list[[i]] <- factor.sig
    if(length(up.idx) > 0){pt.col[up.sig] <- "#8c510a"}
    if(length(down.idx) > 0){pt.col[down.sig] <- "#01665e"}
    plot(factor.effect.mat[,i], -log10(factor.p.mat[,i]), 
        col = pt.col, pch = 16,
        main = colnames(factor.p.mat)[i], xlab = "Effect", 
        ylab = "-log10(p)")    
}
```

```{r num_sig}
par(mar = c(4, 12, 4, 4), mfrow = c(1,1))
num.sig <- sapply(sig.list, function(x) sapply(x, length))
a <- barplot(num.sig, beside = TRUE, main = "Number of significant proteins",
    col = c("#8c510a", "#01665e"), las = 2, horiz = TRUE, 
    xlim = c(0, (max(num.sig)*1.1)), legend = TRUE)
test <- sapply(1:nrow(a), function(b) text(x = num.sig[b,]+150, y = a[b,], 
    labels = num.sig[b,]))

write.table(num.sig, here("Results", "for_paper", paste0("num_dap_table_", analysis.text, ".txt")),
quote = FALSE, sep = "\t")
```

## Enrichments

The enrichments for these significant genes are shown below. 
I'm adding some discussion based on the results. This may change
in future iterations of the workflow.

### Age

* Proteins more abundant with age are enriched for synapse, ER,
    subcellular localization, and anion binding.

* Proteins less abundant with age are enriched for mitochondrion, 
    and mRNA splicing

### Sex

* Proteins more abundant in males are enriched for neuropeptide
    activity and siRNA biogenesis

* Proteins more abundant in females are enriched for nuclear 
    membrane, nuclear migration along microtubule, and mRNA
    transport.

### Genotype
* Proteins for which genotype had a positive effect were enriched 
    for a couple transcription factor motifs, abnormal immunoglobulin 
    levels, abnormal IgA secretion.

* Proteins for which genotype had a negative effect were enriched
    for MAPK signaling, blood coagulation, and TLR regulation.

### Genotype:age interaction
* Positive interaction: more in FC in 4-month olds, more in VS in `r compare.age.text`-month olds

* Proteins for which genotype and age had a positive interaction
    were enriched for small molecule metabolism, mitochondrial 
    matrix, and organic acid metabolism.

* Negative interaction: more in VS in 4-month olds, more in FC in `r compare.age.text`-month olds

* There were only two proteins for which genotype and age had a 
    negative interaction: Cnn2 and Tmed9. Obviously, these didn't
    have any enrichments.

* Tmed9 is a membrane trafficking protein in the synapse
* Cnn2 seems to be related to macrophage migration and phagocytosis.


```{r enrich, fig.width = 7, fig.height = 12}
#this file doesn't differ based on which age we are looking at, 
#since we look at this before we split out the age groups.

all.sig <- unlist(sig.list, recursive = FALSE)
sig.id <- lapply(all.sig, function(x) obs_data$pr_info[match(x, obs_data$pr_info[,id.col]),gene.col])

enrich.file <- file.path(results.dir, paste0("group_enrichment_", analysis.text, ".RDS"))

if(!offline){
    if(!file.exists(enrich.file)){
        enrich <- lapply(sig.id, 
            function(x) if(length(x) > 0){gost(x, organism = "mmusculus", source = "GO")})
        saveRDS(enrich, enrich.file)
    }else{
        enrich <- readRDS(enrich.file)
    }
}else{
    enrich <- readRDS(enrich.file)
}

enrich.exists <- as.logical(length(which(ls() == "enrich")))

if(enrich.exists){
    #pdf("~/Desktop/protein_enrich.pdf", width = 7, height = 12)
    plot.enrichment.group(enrich, max.term.size = 3000, 
        transformation = function(x) log10(x+1), cluster_cols = FALSE)
    #dev.off()
}
```


## Top proteins {.tabset .tabset-fade .tabset-pills}

The following tables show stats for the top proteins across
factors. The maximum number of genes shown for each group is
10.

```{r top_hits, results = "asis"}

for(i in 1:length(sig.list)){
    has.vals <- sapply(sig.list[[i]], length)
    if(any(has.vals) > 0){
        cat("###", names(sig.list)[i], "{.tabset .tabset-fade .tabset-pills}\n")
    }
    for(j in 1:length(sig.list[[i]])){
        sig.set <- sig.list[[i]][[j]]
        if(length(sig.set) > 0){
            cat("####", names(sig.list[[i]])[j], "\n")

            max.num <- min(10, length(sig.set)) #show no more than 10 genes
            sig.effect <- factor.effect.mat[match(sig.set, rownames(factor.effect.mat)),i,drop=FALSE]
            if(names(sig.list[[i]])[j] == "Up"){
                effect.order <- order(sig.effect, decreasing = TRUE)[1:max.num]
            }
            if(names(sig.list[[i]])[j] == "Down"){
                effect.order <- order(sig.effect, decreasing = FALSE)[1:max.num]
            }
        sorted.effect <- sig.effect[effect.order,1,drop=FALSE]
        top.idx <- match(rownames(sorted.effect), rownames(factor.fdr.mat))
        top.fdr <- factor.fdr.mat[top.idx,i]
        stat.table <- cbind(signif(sorted.effect, 2), signif(-log10(top.fdr), 2))
        colnames(stat.table) <- c("Effect Size", "-log10(FDR)")
        rownames(stat.table) <- sapply(rownames(stat.table), function(x) as.character(get_prg_info(x, obs_data)[gene.col]))
        print(kable(stat.table))
        cat("\n\n")
        }
    }
}
```

## Examples {.tabset .tabset-fade .tabset-pills}

Below we show examples of each type of effect.


### Up with age

```{r age_up, fig.width = 9, fig.height = 6.5}
#pdf("~/Desktop/example.pdf", width = 9, height = 6.5)
#age effect up
#a note about automatic text placement. It's pretty good, but
#not perfect. It may require fiddling with parameters, especially
#for violin plots because I don't have a good way of defining
#where the violin will be drawn. I only jitter the points,
#so their spread is not based on their density as it is in the
#violin plots. Also because jitter is random, replotting can lead
#to different text placement.

plot_pr_abund(gene.name = "Hapln2", obs_data, data.type = "scaled", autoplace.text = TRUE) 
#dev.off()
```

### Down with age

```{r age_down, fig.width = 9, fig.height = 6.5}
#age effect down
plot_pr_abund(gene.name = "Prom1", obs_data, data.type = "scaled", 
    autoplace.text = TRUE) 
```

### Positive genotype-age interaction
```{r geno_age_up, fig.width = 9, fig.height = 6.5}
#genotype-age interaction positive
#genetic effect gets more positive with age
plot_pr_abund("Tspan2", obs_data, data.type = "scaled", 
    autoplace.text = TRUE, stat.y = 0.2) 
```

### Negative genotype-age interaction
```{r geno_age_down, fig.width = 9, fig.height = 6.5}
#genotype-age interaction negative
#genetic effect gets more negative with age
plot_pr_abund("Tmed9", obs_data, data.type = "scaled", autoplace.text = TRUE) 
```

### Higher in males
```{r male_up, fig.width = 9, fig.height = 6.5}
#higher in males
plot_pr_abund("Serpina3k", obs_data, data.type = "scaled", autoplace.text = TRUE)
```

### Higher in females
```{r female_up, fig.width = 9, fig.height = 6.5}
#higher in females
plot_pr_abund("Ciz1", obs_data, data.type = "scaled", autoplace.text = TRUE) 
```

### Positive main effect of genotype
```{r geno_up, fig.width = 9, fig.height = 6.5}
#overall positive effect of genotype
plot_pr_abund("Hmbox1", obs_data, data.type = "scaled", 
    autoplace.text = TRUE, stat.x = 0.18) 
```

### Negative main effect of genotype
```{r geno_down, fig.width = 9, fig.height = 6.5}
#overall negative effect of genotype
test <- plot_pr_abund("Csnk2b", obs_data, data.type = "scaled", 
    autoplace.text = TRUE, stat.y = 0.25) 
```


```{r specific_markers, eval = FALSE}
#based on reviewer comments, we want to look at some markers of
#specific things we can put into the paper.
#glutamatergic synapses
test.genes <- c("Slc17a7", "Slc17a6", "Grin1", "Grin2b", "Glul", "Sn1"); plot.label = "Glutamatergic Synapse Markers"
test.genes <- c("Casq2", "Ifit3", "Apod", "Mobp", "Baz2a", "Tubb6", "Htra1", "Stat1", "Mbp", "Gfap", "C1qa", "C1qb", "C4b"); plot.label = "Up with age in pmid 37875604"
test.genes <- c("Rras2", "Mt1", "Tnc", "Hist1hb", "Prom1", "Marcksl1", "Mbp", "Lppr4", "Tfrc"); plot.label = "Down with age in pmid 37875604"
test.genes <- c("Egf", "Egfr", "Grb2", "Sos", "Ras", "Raf", "Mek", "Erk"); plot.label <- "EGFR cascade"

plot_multi_gene(gene.names = test.genes, sample_data = old_data, 
    data.type = "scaled", plot.label = plot.label, stat.x = 0.75, 
    stat.y = 0.25, test_factor = "mean")


young.mice <- obs_data$mouse_info[which(obs_data$mouse_info[,"age_months"] == "4"),1]
old.mice <- obs_data$mouse_info[which(obs_data$mouse_info[,"age_months"] == "12"),1]
old_data <- remove_mice(obs_data, young.mice)
young_data <- remove_mice(obs_data, old.mice)

test.genes <- get_gene_names(bd.name = "APP Metabolism", kegg.name = "Nicotine addiction"); plot.label = "App/nicotine"
test.genes <- get_gene_names(bd.name = "Synapse", kegg.name = "Nicotine addiction");plot.label = "Synapse/nicotine"
test.genes <- get_gene_names(bd.name = "Mitochondrial Metabolism", kegg.name = "Parkinson disease");plot.label = "Mito/Park"


test <- plot_multi_gene(test.genes, old_data, data.type = "scaled", plot.label = plot.label,
    test_factor = "mean", stat.x = 0.7, stat.y = 0.2, mouse_order = "genotype")


test <- plot_multi_gene(test.genes, sample_data = old_data, order_by = "geno_twelve", 
    plot.label = paste(plot.label, "in 12 month olds"), data.type = "scaled",
    stat.x = 0.75, stat.y = 0.2)

test <- plot_multi_gene(test.genes, sample_data = young_data, order_by = "geno_four", 
    plot.label = paste(plot.label, "in 4 month old animals"), stat.y = 0.2, stat.x = 0.01,
    data.type = "scaled")


gene.name = "C4b"
plot_pr_abund(gene.name, obs_data, data.type = "scaled", stat.x = 0.75, stat.y = 0.25)
#plot_pr_abund(gene.name, old_data, data.type = "scaled", stat.x = 0.75, stat.y = 0.25)
#plot_pr_abund(gene.name, sample_data = young_data, data.type = "scaled", stat.x = 0.75, stat.y = 0.25)

```


## Gene groups based on gene descriptions

We can try looking at some gene groups based on their description,
for example mitochondrial genes, glutamate receptor genes, etc.
The code is turned off, but can be run by hand for curiosity. 


```{r gene_groups, eval = FALSE}
search.term <- c("claudin", "occludin", "F-actin", "junctional adhesion", "tight junction"); plot.label = "Tight Junction Genes"
#search.term <- plot.label <- "lipid"
#search.term <- plot.label <- "cadherin"
#search.term <- plot.label <- "adhesion"
#search.term <- plot.label <- "e2f"

term.list <- lapply(search.term, function(x) obs_data$pr_info[grep(x, obs_data$pr_info[,"gene_desc"], ignore.case = TRUE),])
#lapply(term.list, nrow)
term.table <- Reduce("rbind", term.list)
#datatable(term.table)
search.genes <- unique(term.table[,gene.col])

#or search by gene name pattern
#search.genes <- obs_data$pr_info[grep("Icam", obs_data$pr_info[,"gene_name"], ignore.case = TRUE),"gene_name"]


#overall effect
search.result <- plot_multi_gene(gene.names = search.genes, sample_data = obs_data, 
    pr_order_by = "genotype", show_r2 = "genotype", test_factor = "mean",
    plot.label = plot.label, data.type = "scaled", impute.with.min = FALSE)

#effect in younger animals
search.result <- plot_multi_gene(gene.names = search.genes, sample_data = young_data, 
    pr_order_by = "geno_four", show_r2 = "geno_four", test_factor = "mean", 
    plot.label = plot.label, data.type = "scaled", impute.with.min = FALSE)


#effect in older animals
search.result <- plot_multi_gene(gene.names = search.genes, sample_data = old_data, 
    pr_order_by = "geno_twelve", show_r2 = "geno_twelve", test_factor = "mean",
    plot.label = plot.label, data.type = "scaled", impute.with.min = FALSE)

high.effect <- which(search.result[,"genotype"] > 0.05)
search.genes <- unique(sapply(strsplit(names(high.effect), ":"), function(x) x[1]))

#plot_pr_abund("Jam3", obs_data)
```

## Comparison to transcriptomics

We want to make as direct a comparison as we can to the transcriptomic
data. We'll start with an overall comparison.

We used proteins that mapped unambiguously to a single gene.
We adjusted the RNA for batch. We further adjusted both 
types of data for sex age and genotype so that we weren't
correlating effects of those factors. 

```{r read_rna}
#read in rna
rna <- readRDS(here("Results", "all_gene", "processed_data", "Stabilized_Counts.RDS"))
rna.covar <- read.csv(here("Results", "all_gene", "processed_data", "covariates_carriers.csv"))
rna_data <- list("mouse_info" = rna.covar, "tx_info" = gene.info, "data" = rna)

all_aligned <- align_protein_rna(obs_data, rna_data, 
    data.type = "mean", gene.names = NULL, 
    adjust.for = c("sex", "age", "genotype"), align.genes = TRUE)

aligned_pr <- all_aligned$aligned_proteins
aligned_tx <- all_aligned$aligned_transcripts

#dim(aligned_pr)
#dim(aligned_tx)
```


We checked correlation across proteins/transcripts; i.e. whether highly 
abundant transcripts coincide with highly expressed proteins in a 
given mouse. The histogram of these correlations is shown below. 
These should be between 0.3 and 0.6, which they are.

```{r general_abund}
all_mouse_cor <- sapply(1:ncol(aligned_pr), 
    function(x) cor(as.numeric(aligned_pr[,x]), aligned_tx[,x], use = "pairwise.complete.obs", 
    method = "spearman"))
#hist(all_mouse_cor, xlim = c(0.15, 0.6), main = "RNA-Protein Correlations within Mouse")
```

We also looked at the same correlations in just the young,
and just the old mice. 

```{r cor_young_old}
young_aligned <- align_protein_rna(young_data, rna_data, 
    data.type = "mean", gene.names = NULL, 
    adjust.for = c("sex", "genotype"), align.genes = TRUE)

young_aligned_pr <- young_aligned$aligned_proteins
young_aligned_tx <- young_aligned$aligned_transcripts

young_mouse_cor <- sapply(1:ncol(young_aligned_pr), 
    function(x) cor(as.numeric(young_aligned_pr[,x]), young_aligned_tx[,x], use = "pairwise.complete.obs", 
    method = "spearman"))
#hist(young_mouse_cor, xlim = c(0.15, 0.6), main = "RNA-Protein Correlations within Mouse")


old_aligned <- align_protein_rna(old_data, rna_data,
    data.type = "mean", gene.names = NULL, 
    adjust.for = c("sex", "genotype"), align.genes = TRUE)

old_aligned_pr <- old_aligned$aligned_proteins
old_aligned_tx <- old_aligned$aligned_transcripts

old_mouse_cor <- sapply(1:ncol(old_aligned_pr), 
    function(x) cor(as.numeric(old_aligned_pr[,x]), old_aligned_tx[,x], use = "pairwise.complete.obs", 
    method = "spearman"))
#hist(old_mouse_cor, xlim = c(0.15, 0.6), main = "RNA-Protein Correlations within Mouse")
```

The young mice have a higher correlation between transcripts
and proteins. This matches what has been seen before. The 
relationship tends to break down with age.

```{r cor_comp}
vioplot(list("all" = all_mouse_cor, "old" = old_mouse_cor, "young" = young_mouse_cor))
```


## Correlations between glutamate transcripts and proteins

Which proteins are our important transcripts correlated with?
Below we align the transcript and protein data along individuals,
adjust for age and sex, and scale across individuals.

```{r prep_cor}

test.genes <- get_gene_names(bd.name = "APP Metabolism", kegg.name = "Nicotine addiction")
tx.id <- gene.info[match(test.genes, gene.info[,"external_gene_name"]), "ensembl_gene_id"]

test_aligned <- align_protein_rna(pr_data = old_data, rna_data = rna_data,
    data.type = "mean", gene.names = NULL, 
    adjust.for = c("sex"), align.genes = FALSE)

test_scaled <- test_aligned
for(i in 1:length(test_scaled)){
    test_scaled[[i]] <- t(apply(test_scaled[[i]], 1, scale))
    dimnames(test_scaled[[i]]) <- dimnames(test_aligned[[i]])
}

aligned_pr <- test_scaled$aligned_proteins
aligned_tx <- test_scaled$aligned_transcripts

test.tx <- aligned_tx[tx.id,]
```

We then calculated correlations between each transcript in the
group and all proteins.

```{r test_cor}

tx.cor.mat <- matrix(NA, nrow = nrow(test.tx), ncol = nrow(aligned_pr))
rownames(tx.cor.mat) <- rownames(test.tx)
colnames(tx.cor.mat) <- rownames(aligned_pr)
for(tx in 1:nrow(test.tx)){
    this.tx <- test.tx[tx,]
    tx.cor <- apply(aligned_pr, 1, function(x) cor(x, this.tx, use = "pairwise.complete.obs"))
    #hist(tx.cor)
    tx.cor.mat[tx,] <- tx.cor
}

#max(tx.cor.mat)
#min(tx.cor.mat)
```

We filtered the correlations to those with a Spearman
correlation rho of at least 0.5. We made a combined matrix
of the transcripts and correlated proteins. The correlation
matrix of this combined matrix is shown below. There is a
group of positively correlated proteins and a group of 
negatively correlated proteins (all transcripts are in
the larger group, because they are all positively 
correlated with each other).

```{r name_and_thresh}

pr_info <- t(sapply(rownames(aligned_pr), function(x) get_prg_info(x, obs_data)))
pr_names <- unlist(pr_info[,gene.col])
u_pr <- unique(pr_names)
pr_name_dups <- rep(NA, length(pr_names))
for(i in 1:length(u_pr)){
    u_pr_idx <- which(pr_names == u_pr[i])
    if(length(u_pr_idx) > 1){
        dupped_name <- paste(pr_names[u_pr_idx], names(u_pr_idx), sep = ":")
    }else{
        dupped_name <- pr_names[u_pr_idx]
    }
    pr_name_dups[u_pr_idx] <- dupped_name

}
tx_names <- gene.info[match(rownames(test.tx), gene.info[,"ensembl_gene_id"]), "external_gene_name"]

rownames(tx.cor.mat) <- tx_names
colnames(tx.cor.mat) <- pr_name_dups

#pheatmap(tx.cor.mat, show_colnames = FALSE)
#plot.decomp(tx.cor.mat, label.points = TRUE, xlim = c(-0.3, 0.9))

cor.thresh = 0.5
high.cor <- apply(tx.cor.mat, 1, function(x) which(x > cor.thresh))
low.cor <- apply(tx.cor.mat, 1, function(x) which(x < (cor.thresh*-1)))
any.cor <- apply(tx.cor.mat, 1, function(x) which(abs(x) > cor.thresh))

#get all unique proteins with at least a 0.5 correlation
u_cor_prot <- unique(unlist(lapply(any.cor, names)))

#build a matrix of the test transcripts and the highly 
#correlated proteins
prot.idx <- which(colnames(tx.cor.mat) %in% u_cor_prot)
pr_tx <- rbind(test.tx, aligned_pr[prot.idx,])
rownames(pr_tx) <- c(tx_names, pr_names[prot.idx])

pr_tx_cor <- cor(t(pr_tx), use = "pairwise.complete.obs")
#pheatmap(pr_tx_cor)
pr.tx.decomp <- plot.decomp(pr_tx_cor, plot.results = FALSE)
gr1 <- which(pr.tx.decomp$u[,1] < 0.05)
gr2 <- which(pr.tx.decomp$u[,1] > 0.05)
gr.col <- rep("blue", nrow(pr.tx.decomp$u))
gr.col[gr2] <- "red"
#plot.decomp(pr_tx_cor, cols = gr.col)
gr.df <- data.frame(as.factor(as.numeric(as.factor(gr.col))))
rownames(gr.df) <- rownames(pr_tx_cor)
colnames(gr.df) <- "group"
pheatmap(pr_tx_cor, annotation_row = gr.df, show_rownames = FALSE, show_colnames = FALSE)
#hist(pr_tx_cor)
```

We looked for enrichments of the proteins in each of these
groups. The enrichments for these proteins are shown below.
The positively correlated proteins are enriched for zinc
binding, which is important in glutamatergic synapses. 

```{r pr_group_enrich, fig.width = 7, fig.height = 7}
gr1.genes <- as.vector(sapply(strsplit(colnames(pr_tx_cor)[gr1], ":"), function(x) x[1]))
gr2.genes <- as.vector(sapply(strsplit(colnames(pr_tx_cor)[gr2], ":"), function(x) x[1]))
gr1.enrich <- gost(setdiff(gr1.genes, tx_names), organism = "mmusculus")
gr2.enrich <- gost(setdiff(gr2.genes, tx_names), organism = "mmusculus")

plot.enrichment(gr1.enrich, max.term.size = 3000, num.terms = 25, 
    plot.label = "Positively Correlated Proteins")
plot.enrichment(gr2.enrich, max.term.size = 3000, num.terms = 25, 
    plot.label = "Negatively Correlated Proteins")
```

## Mediation with bmediatR

Here we prepare the data for mediation analysis. The first step
is to look for proteins that are significantly affected by genotype.
I've used a relatively permissive fdr threshold of 0.1.

```{r med_prep}
old.var <- get_factor_var(old_data, data.type = "scaled")

#test the effect of genotype on each protein, these data are scaled and sex-adjusted
common.id <- intersect(rownames(old.var), colnames(aligned_pr))
#pr.alone.model <- apply(aligned_pr[,common.id], 1, function(x) lm(x~old.var[common.id,"genotype"]))
#pr.effect <- sapply(pr.alone.model, function(x) coef(summary(x))[2,1]) #effect of genotype on the protein
#pr.p <- sapply(pr.alone.model, function(x) coef(summary(x))[2,4]) #p value of genotype on the protein

unordered.geno <- as.factor(as.character(old.var[common.id,"genotype"]))
pr.alone.model <- apply(aligned_pr[,common.id], 1, function(x) lm(x~unordered.geno))
pr.r2 <- sapply(pr.alone.model, function(x) summary(x)$adj.r.squared) #variance explained of genotype on the protein from anova
pr.f <- t(sapply(pr.alone.model, function(x) summary(x)$fstatistic)) #f statistic
pr.p <- apply(pr.f, 1, function(x) pf(x[1], x[2], x[3], lower.tail = FALSE)) #p value
#qqunif.plot(pr.p)


sig.idx <- which(p.adjust(pr.p, "fdr") < 0.1) #filter to proteins with some genetic effect
#length(sig.idx)
#dzank1.id <- "Q8C008"
#pr.p[dzank1.id]
#check_id <- Reduce("rbind", lapply(names(sig.idx), function(x) get_prg_info(x, obs_data)))
#id.enrich <- gost(check_id[,"gene_name"], organism = "mmusculus")
#plot.enrichment(id.enrich, num.terms = 25, max.term.size = 3000)
```

The heat map below shows the expression patterns across
genotypes clusters nicely into 9 cluster patterns.
This is very different from the transcriptome, which 
had only two clusters, both of which had the FC and VS
expression opposing each other with WT in the middle. 
Here we get lots of different patterns.

```{r investigate_anova, fig.width = 3, fig.height = 7}
sig.pr <- aligned_pr[sig.idx,common.id]
#pheatmap(sig.pr[,order(unordered.geno)])
geno.idx <- lapply(levels(unordered.geno), function(x) which(unordered.geno == x))
sig.mean <- Reduce("cbind", lapply(geno.idx, function(x) rowMeans(sig.pr[,x], na.rm = TRUE)))
colnames(sig.mean) <- levels(unordered.geno)
#pheatmap(sig.mean, show_rownames = FALSE)
#pairs(sig.mean)

#sig.cor <- cor(t(sig.pr), use = "pairwise.complete.obs")
#pheatmap(sig.cor, show_rownames = FALSE, show_colnames = FALSE)
#plot.decomp(sig.cor)
#k = 10
#cl = pam(sig.cor, k = k)$clustering


#ktest <- test.pam.k(sig.mean, k = 2:20, plot.results = FALSE)
#boxplot(ktest[[1]])
#best.mean <- which.max(sapply(ktest[[1]], mean))
k = 9 #this gets good clusters witjout being too redundant
cluster.mem <- pam(sig.mean, k = k)$clustering
cl.df <- data.frame("cluster" = as.factor(cluster.mem))
rownames(cl.df) <- rownames(sig.mean)

split.blood.group <- FALSE #plist out the blood coagulation group
if(split.blood.group){
    cl.idx <- which(cluster.mem == 1)
    cl.cor <- cor(t(aligned_pr[names(cl.idx),]), use = "pairwise.complete.obs")
    test.cl <- pam(cl.cor, k = 2)
    cl1.df <- data.frame(as.factor(test.cl$clustering))
    #pheatmap(cl.cor, annotation_row = cl.df)
    blood.proteins = which(cl1.df[,1] == 1)
    k = k + 1
    cluster.mem[cl.idx[blood.proteins]] <- k
    #remake the cluster
    cl.df <- data.frame("cluster" = as.factor(cluster.mem))
}

pheatmap(sig.mean[order(cluster.mem),], annotation_row = cl.df, cluster_rows = FALSE, 
    show_rownames = FALSE, cluster_cols = FALSE)


```

### Cluster Enrichments {.tabset .tabset-fade .tabset-pills}

The enrichments of the clusters are shown below along with their 
expression patterns.

```{r clust_enrich, fig.width = 10, fig.height = 7, results = "asis"}
pr.clust <- lapply(1:k, function(x) get_prg_info(pr_group_name = rownames(sig.mean)[which(cluster.mem == x)], obs_data))

clust.enrich.file <- here("Results", "Proteomics", paste0("Cluster_Enrichment_", analysis.text, ".RDS"))
if(!file.exists(clust.enrich.file)){
    clust.enrich <- lapply(pr.clust, function(x) gost(x[,"gene_name"], organism = "mmusculus"))
    #plot.enrichment.group(clust.enrich, max.term.size = 3000)
    saveRDS(clust.enrich, clust.enrich.file)
}else{
    clust.enrich <- readRDS(clust.enrich.file)
}
#plot.enrichment.group(clust.enrich, max.term.size = 3000)

#pdf("~/Desktop/cluster_enrich.pdf", width = 10, height = 7)
enrich.labels <- rep(NA, k)
for(i in 1:k){
    cat("#### Cluster", i, "\n")
    k.idx <- which(cluster.mem == i)
    layout(matrix(c(1,2,0,2), byrow = TRUE, nrow = 2), widths = c(0.5, 1))
    #geno.vector <- ordered(unlist(lapply(1:3, function(x) rep(ordered.geno[x], length(k.idx)))), levels = ordered.geno)
    #val.vector <- as.vector(sig.mean[k.idx,ordered.geno])
    #test_effect(val.vector, unlist(geno.vector), plot.results = TRUE, test_type = "anova", autoplace.text = TRUE, plot.label = "Mean Expression Pattern")
    boxplot(sig.mean[k.idx,ordered.geno], main = "Expression Pattern", col = geno.cols)
    abline(h = 0)
    enrich.terms <- plot.enrichment(clust.enrich[[i]],
        max.term.size = 3000, plot.label = paste("Cluster", i), num.terms = 25)
    enrich.labels[i] <- paste(enrich.terms[1:2,1], collapse = "; ")
    #plot_multi_gene(pr.clust[[i]][,"gene_name"], old_data, test_factor = "mean", plot.label = enrich.terms[1,1])
    cat("\n\n")
}
#dev.off()
```

### Labeled Cluster Heatmap

Here's the cluster heatmap again with cluster labels.

```{r cluster_labels, fig.width = 10, fig.height = 7}
label.df <- matrix(NA, ncol = 1, nrow = nrow(cl.df))
for(i in 1:9){
    label.df[which(cl.df == i),1] <- enrich.labels[i]
}
label.df <- data.frame(label.df)
colnames(label.df) <- "Cluster Labels"
rownames(label.df) <- rownames(cl.df)
pheatmap(sig.mean[order(cluster.mem),ordered.geno], annotation_row = label.df, 
    cluster_rows = FALSE, show_rownames = FALSE, cluster_cols = FALSE)

```

The plot below is a more finalized version

```{r clust_fig, fig.width = 5, fig.height = 7}
cl.order <- order(cluster.mem)

cl.name.file <- here("Data", paste0("cluster_labels_", analysis.text, ".txt"))
if(file.exists(cl.name.file)){
    cl.names <- as.matrix(read.table(cl.name.file, sep = "\t"))
}else{
    cl.names <- as.matrix(enrich.labels)
}
cl.names[,1] <- gsub("; ", "\n", cl.names[,1])


par(mar = c(4, 16, 4, 2))
imageWithText(sig.mean[cl.order,ordered.geno], show.text = FALSE,
    split.at.vals = TRUE, col.scale = c("purple", "brown"), row.names = NULL,
    col.text.rotation = 0, col.text.adj = 0.5, col.text.shift = 0.03,
    grad.dir = "ends")

for(cl in 1:k){
    cl.idx <- which(cl.df[cl.order,1] == cl)
    min.y <- nrow(cl.df) - max(cl.idx) 
    max.y <- nrow(cl.df) - min(cl.idx) 
    draw.rectangle(min.x = 0.5, max.x = 3.5, min.y, max.y, lwd = 2)
    text(x = 0.45, y = mean(c(min.y, max.y)), labels = cl.names[cl,1], adj = 1, cex = 0.9)
}
```

### Detailed Cluster Plots {.tabset .tabset-fade .tabset-pills}

The cluster patterns are shown in detail below. Each cluster is labeled
with the top enrichment term from above. They are sorted by genotype

```{r cluster_detail, fig.width = 10, fig.height = 7, results = "asis"}

#pdf("~/Desktop/cluster_detail.pdf", width = 10, height = 7)
for(i in 1:k){
    cat("#### Cluster", i, "\n")
    k.idx <- which(cl == i)
    plot_multi_gene(gene.name = pr.clust[[i]][,"gene_name"], sample_data = old_data, 
        test_factor = "mean", test_type = "anova", plot.label = enrich.labels[i], 
        mouse_order = "genotype")
    cat("\n\n")
}
#dev.off()

```


## Mediation with bmediatR

I did a mediation just with module means between a single 
transcriptomic module of interest and all the proteomic
modules. I put the proteins as the mediators because
I can run multiple mediators at once.

The heat map below shows that the probabilities for each
mediation model with the nine protein clusters and the 
single transcriptomic module. I've run many of these and
they all look the same. In every case it looks as if the
platelet module is driving the transcriptomic signature...

Is the whole transcriptomic signature we see driven by
this blood module? What is going on?

```{r summarize_mediation}
summarize_mediation <- function(post.mat, min.prob = 0.51, interesting.models = c("complete", "partial")){
    interesting.idx <- unique(unlist(lapply(interesting.models, function(x) grep(x, colnames(post.mat)))))
    high.prob.idx <- which(post.mat[,interesting.idx] > min.prob, arr.ind = TRUE)
    interesting.result <- post.mat[unique(high.prob.idx[,1]),,drop=FALSE]
    return(interesting.result)
}

```

```{r mediate_modules}
gd <- 0:(length(ordered.geno)-1)
names(gd) <- ordered.geno
exp.var <- old.var[colnames(test.tx), "genotype"]
Gdosage <- sapply(exp.var, function(x) gd[which(names(gd) == x)])
geno.design <- model.matrix(~exp.var-1)
rownames(geno.design) <- colnames(test.tx)

#get mean abundance by genotype for all modules
clust.pep <- lapply(pr.clust, function(x) x[,id.col])
pr_clust <- lapply(clust.pep, function(x) aligned_pr[x,])
pr_clust_mean <- sapply(pr_clust, function(x) colMeans(x, na.rm = TRUE))

#tx.deg <- read.delim(here("Results", "age_batch_12_gene", "DEG_clusters_carriers.txt"))
#tx.deg.cl <- lapply(1:2, function(x) tx.deg[which(tx.deg[,"cluster"] == x),])
#test.id <- tx.deg.cl[[2]][,"geneID"]

biadj <- readRDS(here("Results", "age_batch_12_gene", "biadjacency_matrix.RDS"))
has.biadj <- which(biadj != 0, arr.ind = TRUE)

mediation.profile <- vector(mode = "list", length = nrow(has.biadj))

#get mediation results (posterior probability matrices) for 
#all transcriptomic modules in the biadjacency matrix and 
#all protein clusters. Loop through transcript modules and
#test against all protein clusters.
sink("temp")
for(b in 1:nrow(has.biadj)){
    bd.name <- rownames(biadj)[has.biadj[b,1]]
    kegg.name <- colnames(biadj)[has.biadj[b,2]]
    
    test.genes <- get_gene_names(bd.name, kegg.name)
    test.id <- rna_data$tx_info[match(test.genes, rna_data$tx_info[,"external_gene_name"]),"ensembl_gene_id"]
    common.tx <- intersect(test.id, rownames(aligned_tx))
    
    rna_clust <- aligned_tx[common.tx,]
    rna_clust_mean <- colMeans(rna_clust, na.rm = TRUE)

    #put the pr as mediators because we can run multiple at once
    med.test <- bmediatR(y = rna_clust_mean, M = pr_clust_mean, 
        X = geno.design, ln_prior_c = "reactive", 
        options_X = list(sum_to_zero = TRUE, 
        center = FALSE, scale = FALSE))

    post <- get_posterior(med.test)
    post.mat <- post[,2:ncol(post)]
    rownames(post.mat) <- colnames(pr_clust_mean) <- paste("Cluster", 1:nrow(post.mat), enrich.labels, sep = "_")
    #pheatmap(post.mat, cluster_rows = FALSE, cluster_cols = FALSE, display_numbers = TRUE)
    mediation.profile[[b]] <- post.mat
}
sink()
```

The bar plots below show the proportion of transcription modules
in each biodomain that were mediated by each protein module.

```{r plot_summ, fig.width = 10, fig.height = 6}

#find all pairs for which the posterior probability is >= 0.51
#we are looking at both complete and partial mediation
med.summary <- lapply(mediation.profile, function(x) summarize_mediation(x, min.prob = 0.51))
k.names <- unlist(lapply(1:length(med.summary), function(x) rep(colnames(biadj)[has.biadj[x,2]], nrow(med.summary[[x]]))))
b.names <- unlist(lapply(1:length(med.summary), function(x) rep(rownames(biadj)[has.biadj[x,1]], nrow(med.summary[[x]]))))
summary.table <- Reduce("rbind", med.summary)
pr_cl <- sapply(strsplit(rownames(summary.table), "; "), function(x) x[1])
mediating.pairs <- cbind("Bd" = b.names, "KEGG" = k.names, "PR" = pr_cl)    

#pheatmap(all.stats, cluster_cols = FALSE)
#for each protein module, tally the proportion of transcript modules mediated
num.sig.bd <- apply(biadj, 1, function(x) length(which(!is.na(x))))
num.sig.k <- apply(biadj, 2, function(x) length(which(!is.na(x))))

u_pm <- unique(pr_cl)
for(pm in 1:length(u_pm)){
    pm_idx <- which(pr_cl == u_pm[pm])

    u_paths <- lapply(1:2, function(x) table(mediating.pairs[pm_idx,x])) 
    
    common.bd <- intersect(names(u_paths[[1]]), rownames(biadj))
    bd.table <- cbind(u_paths[[1]][common.bd], num.sig.bd[common.bd])
    prop.med.bd <- bd.table[,1]/bd.table[,2]

    common.k <- intersect(names(u_paths[[2]]), colnames(biadj))
    k.table <- cbind(u_paths[[2]][common.k], num.sig.k[common.k])
    prop.med.k <- k.table[,1]/k.table[,2]

    par(mfrow = c(1,2))
    par(mar = c(4, 12, 4, 2))
    barplot(sort(prop.med.bd), las = 2, horiz = TRUE, 
        xlab = "Proportion Biodomain Modules Mediated",
        main = "Biodomains", cex.names = 0.8, xlim = c(0, 1))

    barplot(sort(prop.med.k), las = 2, horiz = TRUE, 
        xlab = "Proportion KEGG Modules Mediated",
        main = "KEGG", cex.names = 0.5, xlim = c(0, 1))
    mtext(u_pm[pm], side = 3, outer = TRUE, line = -1.5, font = 2, cex = 1.2)
}
```

The plot below compares the posterior probabilities on mediation
for any protein modules with significant mediation.

```{r patterns, fig.width = 5, fig.height = 5}
#This looks at patterns of modulation on the original biadjacency matrix
u_mod <- unique(mediating.pairs[,3])
mod.biadj.list <- vector(mode = "list", length = length(u_mod))
names(mod.biadj.list) <- u_mod

for(m in 1:length(u_mod)){
    mod.biadj <- matrix(NA, nrow = nrow(biadj), ncol = ncol(biadj))
    dimnames(mod.biadj) <- dimnames(biadj)
    
    mod.idx <- which(mediating.pairs[,3] == u_mod[m])
    #pheatmap(summary.table[mod.idx,,drop=FALSE])
    mod.edge <- mediating.pairs[mod.idx,1:2,drop=FALSE]
    max.idx <- apply(summary.table[mod.idx,,drop=FALSE], 1, which.max)
    mod.edge.weight <- sapply(1:length(max.idx), function(x) summary.table[mod.idx[x], max.idx[x]])
    med.type <- colnames(summary.table)[max.idx]
    med.table <- cbind(mod.edge, mod.edge.weight, med.type)
    rownames(med.table) <- NULL

    mod.biadj.list[[m]] <- med.table
}


#Strength of the mediation plotted against each other
par(mar = c(12, 4, 4, 2))
med.strength <- lapply(mod.biadj.list, function(x) as.numeric(x[,"mod.edge.weight"]))
boxplot(med.strength, las = 2, ylab = "Mediation Posterior Probability")
```

Below we check how much the effect of genotype on transcripts
is attenuated by the protein modules.

```{r attenuation}

mediation_effect <- function(bd.name, kegg.name, pr_module_mean, 
    sample_data, rna_data, aligned_data, plot.results = FALSE, 
    stat.y = 0.9, cex.lab = 1,
    pr_module_name = "", plot.label = ""){
    
    test.genes <- get_gene_names(bd.name, kegg.name)
    test.id <- rna_data$tx_info[match(test.genes, rna_data$tx_info[,"external_gene_name"]),"ensembl_gene_id"]
    common.tx <- intersect(test.id, rownames(aligned_data$aligned_transcripts))
    
    rna_clust <- aligned_data$aligned_transcripts[common.tx,]
    rna_clust_mean <- colMeans(rna_clust, na.rm = TRUE)

    mouse.var <- get_factor_var(sample_data, data.type = "scaled")

    if(plot.results){
        par(mfrow = c(2,2))
    }
    main.geno.tx <- test_effect(values = rna_clust_mean, 
        plot.factor = old.var[names(rna_clust_mean),"genotype"], 
        plot.results = plot.results, cex.lab = cex.lab,
        stat.y = stat.y, test_type = "lm", ylab = "RNA Cluster Mean", 
        plot.label = paste(bd.name, kegg.name, sep = " : "))

    main.geno.pr <- test_effect(pr_module_mean, old.var[names(rna_clust_mean),"genotype"], 
        plot.results = plot.results, cex.lab = cex.lab,
        stat.y = stat.y, test_type = "lm", ylab = "Protein Cluster Mean", 
        plot.label = pr_module_name)

    adj.expr <- adjust(as.matrix(rna_clust_mean), pr_module_mean)

    adj.geno.tx <- test_effect(adj.expr, old.var[names(rna_clust_mean),"genotype"], 
        plot.results = plot.results, cex.lab = cex.lab,
        stat.y = stat.y, test_type = "lm", ylab = "Expression After Adjustment", 
        plot.label = "RNA after adjusting for Protein")

    adj.pr <- adjust(pr_module_mean, as.matrix(rna_clust_mean))

    adj.geno.pr <- test_effect(adj.pr, old.var[names(rna_clust_mean),"genotype"], 
        plot.results = plot.results, cex.lab = cex.lab,
        stat.y = stat.y, test_type = "lm", ylab = "Protein After Adjustment", 
        plot.label = "Protein after adjusting for RNA")

    all.results <- rbind(main.geno.tx, adj.geno.tx, main.geno.pr, adj.geno.pr)
    invisible(all.results)
}


```

The bar plot below shows how much variance in each protein
module is explained by genotype
```{r protein_r2}
#these are just dummy bd and kegg. I'm only looking at the main
#effect of genotype on the protein module here, but it is
#easy to check this way.
bd.name = "APP Metabolism"; kegg.name = "Nicotine addiction"; cluster.num = 3

all.med.effect <- lapply(1:ncol(pr_clust_mean), 
    function(x) mediation_effect(bd.name, kegg.name, 
    pr_clust_mean[,x,drop=FALSE],
    old_data, rna_data, test_scaled, plot.results = FALSE))

all.main.effect <- sapply(all.med.effect, function(x) as.numeric(x["main.geno.pr", "R2"]))
names(all.main.effect) <- sapply(strsplit(colnames(pr_clust_mean), ";"), function(x) x[1])

par(mar = c(4, 16, 4, 2))
barplot(sort(all.main.effect), las = 2, horiz = TRUE, xlab = "R2", 
    main = "Variance explained by genotype")
```

The plots below show that this mediation model is correct.
The signal in the RNA is completely removed by adjusting
for the protein signal, but not vice versa. The protein
is driving the signal we see in the transcriptome! What?!

This is true for both the VS-up and VS-down modules. The
mitochondrial protein module also seems to be partially
mediating the transcriptomic signal, but there is maybe
a more bi-directional relationship there. 

Other modules all seem to be more or less independent of 
the transcriptomic modules.

```{r mediation_plot, fig.width = 8, fig.height = 8}
bd.name = "APP Metabolism"; kegg.name = "Nicotine addiction"; cluster.num = 3

#bd.name = "Synapse"; kegg.name = "Ribosome"

med.effect <- mediation_effect(bd.name, kegg.name, 
    pr_clust_mean[,cluster.num,drop=FALSE], cex.lab = 1.2,
    sample_data = old_data, rna_data, aligned_data = test_scaled, plot.results = TRUE, 
    pr_module_name = colnames(pr_clust_mean)[cluster.num],stat.y = 0.2)
```

The comparison of R2 statistics below shows that genotype has a stronger
effect on the protein module than on the transcript module. Adjusting 
for the protein completely removes the effect of genotype on the 
transcripts. But adjusting for the transcripts does not remove the
effect of genotype on the protein module.

```{r stat_comparison, fig.width = 6, fig.height = 5}
#par(mfrow = c(1,3))
med.cols <- c("#2ca25f", "#99d8c9", "#756bb1", "#bcbddc")
barplot(as.numeric(med.effect[,1]), names = rownames(med.effect), ylab = "R2", main = "R2", col = med.cols)
legend("topleft", fill = med.cols, legend = c("Transcript", "Transcript Mediated", "Protein", "Protein mediated"))
abline(h = 0)

```

We checked these stats for all the mediating pairs. You can see
that the platelet module is the primary mediator of the 
transcriptomic response.

```{r plot_mediator_effects}

plot_mediator_effects <- function(mediating.results.list, mediating.pair.matrix, 
    mediator.name = "", plot.label = ""){

    par(mfrow = c(1,3), mar = c(4,4,6,2))

    if(mediator.name != ""){
        mod.idx <- which(mediating.pair.matrix[,3] == mediator.name)
    }else{
        mod.idx <- 1:nrow(mediating.pair.matrix)
    }

    all_mod_effects <- t(sapply(mediating.results.list[mod.idx], function(x) as.numeric(x[,"beta"])))
    stripchart(lapply(1:ncol(all_mod_effects), function(x) all_mod_effects[,x]), 
        vertical = TRUE, method = "jitter", pch = 16, main = "",
        group.names = rownames(mediating.results.list[[1]]), ylab = "beta", ylim = c(-0.8, 1),
        col = med.cols)
    mtext("Effect Size", side = 3, line = 0)
    legend("topright", fill = med.cols, legend = c("Transcript", "Transcript Mediated", "Protein", "Protein mediated"))
    abline(h = 0)

    all_mod_r2 <- t(sapply(mediating.results.list[mod.idx], function(x) as.numeric(x[,"R2"])))
    stripchart(lapply(1:ncol(all_mod_r2), function(x) all_mod_r2[,x]), 
        vertical = TRUE, method = "jitter", pch = 16, main = "",
        group.names = rownames(mediating.results.list[[1]]), ylab = "R2", ylim = c(0, 1),
        col = med.cols)
    mtext("Variance Explained", side = 3, line = 0)
    legend("topleft", fill = med.cols, legend = c("Transcript", "Transcript Mediated", "Protein", "Protein mediated"))
    abline(h = 0)

    all_mod_p <- t(sapply(mediating.results.list[mod.idx], function(x) -log10(as.numeric(x[,"p"]))))
    stripchart(lapply(1:ncol(all_mod_p), function(x) all_mod_p[,x]), 
        vertical = TRUE, method = "jitter", pch = 16, main = "",
        group.names = rownames(mediating.results.list[[1]]), ylab = "-log10(p)",
        col = med.cols, ylim = c(0, 14))
    mtext("-log10(p)", side = 3, line = 0)
    legend("topleft", fill = med.cols, legend = c("Transcript", "Transcript Mediated", "Protein", "Protein mediated"))
    abline(h = 0)

    mtext(plot.label, side = 3, cex = 1.2, font = 2, line = -2.5, outer = TRUE)

    result <- list("effects" = all_mod_effects, "R2" = all_mod_r2,
        "p" = all_mod_p)
    invisible(result)
}
```


```{r all_med_stats, fig.width = 12, fig.height = 4}
mod_effects <- vector(mode = "list", length = nrow(mediating.pairs))

for(m in 1:nrow(mediating.pairs)){
    bd.name = mediating.pairs[m,1]
    kegg.name = mediating.pairs[m,2]
    cl.num <- as.numeric(strsplit(mediating.pairs[m,3], "_")[[1]][2])

    mod_effects[[m]] <- mediation_effect(bd.name, kegg.name, pr_clust_mean[,cl.num,drop=FALSE],
        old_data, rna_data, test_scaled, plot.results = FALSE)
}

for(i in 1:length(u_mod)){
    plot_mediator_effects(mod_effects, mediating.pairs, mediator.name = u_mod[i],
    plot.label = u_mod[i])
}
```


```{r all_pairs, eval = FALSE}
#I'm looking for a way to visualize all mediation strength
#for the paper
k = 9
#get abundance for peptides in each cluster
clust.abund <- lapply(pr.clust, function(x) peptide_vals(x[,gene.col], old_data, 
    data.type = "scaled", adjust.for = "sex"))
#collect mean cluster values for each mouse into a matrix
mean_abund <- sapply(clust.abund, function(x) colMeans(x, na.rm = TRUE))
mean_abund <- mean_abund[colnames(test_scaled[[1]]),]
colnames(mean_abund) <- paste0("Cluster", 1:k)

#boxplot(mean_abund[,1]~as.factor(old.var[,"genotype"]))

#use each mean value to mediate all biodomain-kegg intersections.
all_pairs <- which(!is.na(biadj), arr.ind = TRUE)
all_pairs[,1] <- rownames(biadj)[as.numeric(all_pairs[,1])]
all_pairs[,2] <- colnames(biadj)[as.numeric(all_pairs[,2])]


pg_summary <- vector(mode = "list", length = ncol(mean_abund))
names(pg_summary) <- colnames(mean_abund)
for(pg in 1:ncol(mean_abund)){
    print(pg)
    cl_effects <- vector(mode = "list", length = nrow(all_pairs))
    for(a in 1:nrow(all_pairs)){
        bd.name = all_pairs[a,1]
        kegg.name = all_pairs[a,2]
        med_vals <- mean_abund[,1]

        cl_effects[[a]] <- mediation_effect(bd.name, kegg.name, 
            pr_module_mean <- mean_abund[,pg,drop=FALSE], old_data, rna_data, test_scaled, 
            plot.results = FALSE)
    }
    all.r2 <- t(sapply(cl_effects, function(x) as.numeric(x[,"R2"])))
    colnames(all.r2) <- rownames(cl_effects[[a]])
    #boxplot(-log10(all.r2))
    pg_summary[[pg]] <- all.r2
}

genotype_var_expr_pr <- sapply(pg_summary, function(x) as.numeric(x[1,3])) #one for each protein cluster
genotype_var_expr_tx <- pg_summary[[1]][,1]#same across all protein clusters, one for each kegg-bd intersection
names(genotype_var_expr_tx) <- apply(all_pairs, 1, function(x) paste(x, collapse = " : "))

proportion_tx_mediated <- lapply(pg_summary, function(x) (x[,1]-x[,2])/x[,1])
proportion_pr_mediated <- lapply(pg_summary, function(x) (x[,3]-x[,4])/x[,3])

mediating_strength_order <- order(sapply(proportion_tx_mediated, mean))
boxplot(proportion_tx_mediated[mediating_strength_order], ylim = c(-0.1,1.2),
    ylab = "Proportion", main = "Proportion of genotype effect on transcription mediated")
add_zero_line(h = 0.5)

boxplot(proportion_pr_mediated[mediating_strength_order], ylim = c(-0.1,1),
    ylab = "Proportion", main = "Proportion of genotype effect on protein mediated")
add_zero_line(h = 0.5)


layout(matrix(c(1,2), ncol = 1), heights = c(0.6, 1))

par(mar = c(0, 4, 4, 4))
barplot_with_num(signif(genotype_var_expr_pr[mediating_strength_order], 2), 
    ylab = "Variance Explained", text.shift = 0.08,
    main = "Proportion Variance Explained by genotype for each protein cluster",
    ylim = c(0, 0.6))
add_zero_line(h = seq(0, 0.6, 0.2), col = "darkgray", lty = 2)
add_zero_line(h = mean(genotype_var_expr_tx))

par(mar = c(4, 4, 0, 2))
plot.grouped.boxes(list("protein > transcript" = proportion_tx_mediated[mediating_strength_order],
    "transcript > protein" = proportion_pr_mediated[mediating_strength_order]),
    legend.y = 1, legend.x = 0, grid.y = seq(0, 1, 0.2), grid.col = "gray", 
    grid.lty = 2, ylab = "Proportion Mediated", axis.at = seq(0, 1, 0.2))


```

## Deeper dive

Let's look at the modules in a bit more detail.
The platelet cluster has a cluster of very highly correlated 
proteins. These are blood proteins, including the fibrinogen 
genes, some Apo genes, etc.

```{r cluster_cor, fig.width = 7, fig.height = 7}
max.cl <- as.numeric(strsplit(names(which.max(table(mediating.pairs[,3]))), "_")[[1]][2])
clust_cor <- cor(t(pr_clust[[max.cl]]), use = "pairwise.complete.obs")
#pheatmap(clust_cor)

k = 2 #I have tried looking at 3 clusters, and the results are not different.
mem <- pam(clust_cor, k = k)$clustering
mem.df <- data.frame("cluster" = as.factor(mem))

clust_id <- lapply(1:k, function(x) rownames(mem.df)[which(mem.df[,1] == x)])
clust_info <- lapply(clust_id, function(x) pr.clust[[max.cl]][match(x, pr.clust[[max.cl]][,id.col]),])
clust_names <- lapply(clust_info, function(x) x[,gene.col])

pr.labels <- Reduce("rbind", lapply(clust_info, function(x) x[,c(id.col, gene.col)]))
pr.names <- apply(pr.labels, 1, function(x) paste(x, collapse = "; "))
name.idx <- match(pr.labels[,1], rownames(clust_cor))
rownames(clust_cor)[name.idx] <- colnames(clust_cor)[name.idx] <- rownames(mem.df)[name.idx] <- pr.names
pheatmap(clust_cor, annotation_row = mem.df, fontsize = 5)
```


```{r part_enrich, fig.width = 8, fig.height = 7}
clust.enrich <- lapply(clust_names, function(x) gost(x, organism = "mmusculus"))
names(clust.enrich) <- paste("Cluster", 1:k)
subclust.mat <- plot.enrichment.group(clust.enrich, max.term.size = 3000, 
    cluster_cols = FALSE, transformation = "sqrt")

sub.clust.labels <- apply(subclust.mat, 2, function(x) names(x)[which.max(x)])
names(clust_names) <- sub.clust.labels
```

Let's look at the effects of these separately. 

```{r module_geno_effect, fig.width = 8, fig.height = 4}
cl_abund <- lapply(clust_id, function(x) pr_clust[[max.cl]][x,])
cl_mean <- lapply(cl_abund, function(x) colMeans(x, na.rm = TRUE))

par(mfrow = c(1,k))
for(i in 1:k){
test_effect(cl_mean[[i]], old.var[names(cl_mean[[i]]),"genotype"], test_type = "lm", 
    ylab = "Mean Abundance", plot.results = TRUE, stat.y = 0.9, 
    plot.label = sub.clust.labels[i], stat.x = 0.7)
}
```

The detailed plots are below.

```{r sub_detail}
for(i in 1:k){
plot_multi_gene(gene.name = clust_names[[i]], sample_data = obs_data, 
    test_factor = "mean", test_type = "anova", plot.label = sub.clust.labels[i], 
    mouse_order = "genotype")
}
```

I used bmediatR to see if one module mediated the other.
bmediatR says they are co-local, so the interpretation 
is that genotype affects them each independently.

```{r mean_cl, fig.width = 7, fig.height = 5}

sink("temp")
#Does one part of the module mediate the other
med.test <- bmediatR(y = as.matrix(cl_mean[[1]]), M = as.matrix(cl_mean[[2]]), 
        X = geno.design, ln_prior_c = "reactive", 
        options_X = list(sum_to_zero = TRUE, 
        center = FALSE, scale = FALSE))
sink()

post <- get_posterior(med.test)
post.mat <- post[,2:ncol(post)]

par(mar = c(10, 4, 4, 2))
barplot(as.numeric(post.mat), names = colnames(post.mat), las = 2)
abline(h = 0.5, lty = 2, col = "darkgray")
```

Now I want to check to see whether each piece of this module
mediates the significant transcript modules. We use the same
procedure as above when we checked mediation between the protein
modules and the transcript modules.

```{r overall_mediation}

cl.mediators <- Reduce("cbind", cl_mean)
colnames(cl.mediators) <- sub.clust.labels

cl.mediation.profile <- vector(mode = "list", length = nrow(has.biadj))

sink("temp")
for(b in 1:nrow(has.biadj)){
    bd.name <- rownames(biadj)[has.biadj[b,1]]
    kegg.name <- colnames(biadj)[has.biadj[b,2]]
    
    test.genes <- get_gene_names(bd.name, kegg.name)
    test.id <- rna_data$tx_info[match(test.genes, rna_data$tx_info[,"external_gene_name"]),"ensembl_gene_id"]
    common.tx <- intersect(test.id, rownames(aligned_tx))
    
    rna_clust <- aligned_tx[common.tx,]
    rna_clust_mean <- colMeans(rna_clust, na.rm = TRUE)

    #put the pr as mediators because we can run multiple at once
    med.test <- bmediatR(y = rna_clust_mean, M = cl.mediators, 
        X = geno.design, ln_prior_c = "reactive", 
        options_X = list(sum_to_zero = TRUE, 
        center = FALSE, scale = FALSE))

    post <- get_posterior(med.test)
    post.mat <- post[,2:ncol(post)]
    rownames(post.mat) <- colnames(cl.mediators)
    #pheatmap(post.mat, cluster_rows = FALSE, cluster_cols = FALSE, display_numbers = TRUE)
    mediation.profile[[b]] <- post.mat
}
sink()

med.summary <- lapply(mediation.profile, function(x) summarize_mediation(x, min.prob = 0.51))
k.names <- unlist(lapply(1:length(med.summary), function(x) rep(colnames(biadj)[has.biadj[x,2]], nrow(med.summary[[x]]))))
b.names <- unlist(lapply(1:length(med.summary), function(x) rep(rownames(biadj)[has.biadj[x,1]], nrow(med.summary[[x]]))))
summary.table <- Reduce("rbind", med.summary)
pr_cl <- sapply(strsplit(rownames(summary.table), "; "), function(x) x[1])
cl.mediating.pairs <- cbind("Bd" = b.names, "KEGG" = k.names, "PR" = substr(pr_cl, 1, nchar(pr_cl[1])))    
```

```{r bmediatr_summary}
bmed.summ.table <- summary.table[,c("complete med", "complete med (react)", 
    "partial med", "partial med (react)", "co-local")]
par(mar = c(10, 4, 4, 2))
boxplot(bmed.summ.table, las = 2, ylab = "Posterior probability")

report.table <- cbind(cl.mediating.pairs, bmed.summ.table)
write_table_with_header(report.table, here("Results", "for_paper", 
    "Supplemental_File_BmediatR.txt"), header = "Posterior probabilites from bmediatR for all mediations between protein modules and transcript modules for which one model had a posterior probability of at least 0.51")
```

The `r pr_cl[1]` portion of the cluster mediated the effect of genotype
on `r nrow(mediating.pairs)` transcriptional modules in all 
biodomains. The plot below shows the proportion of significant
modules in each biodomain that were significantly mediated by the
`r pr_cl[1]` protein module. It mediated the vast majority.


```{r med_paths}
#pheatmap(all.stats, cluster_cols = FALSE)
u_paths <- apply(cl.mediating.pairs, 2, table)

num.sig.bd <- apply(biadj, 1, function(x) length(which(!is.na(x))))
test.table <- cbind(u_paths[[1]][rownames(biadj)], num.sig.bd[rownames(biadj)])
prop.med <- test.table[,1]/test.table[,2]

par(mar = c(4, 14, 4, 2))
barplot(sort(prop.med), las = 2, horiz = TRUE) #biodomains
#barplot(sort(u_paths[[2]]), las = 2, horiz = TRUE)
#barplot(sort(u_paths[[3]]), las = 2, horiz = TRUE)
```

The plot below shows the strength of the mediations.

```{r all_cl_med_stats, fig.width = 12, fig.height = 4}
strong_mod_effect <- vector(mode = "list", length = nrow(cl.mediating.pairs))
strong_mediator <- cl.mediating.pairs[1,3]

for(m in 1:nrow(cl.mediating.pairs)){
    bd.name = cl.mediating.pairs[m,1]
    kegg.name = cl.mediating.pairs[m,2]
    
    strong_mod_effect[[m]] <- mediation_effect(bd.name, kegg.name, 
        cl.mediators[,strong_mediator,drop=FALSE], old_data, rna_data, test_scaled, 
        plot.results = FALSE)
}

plot_mediator_effects(strong_mod_effect, cl.mediating.pairs, 
    mediator.name = strong_mediator, plot.label = strong_mediator)
```

This is in comparison to the other part of the module with no 
significant mediation. The effects of genotyope on the transcript
are not really changed by adjusting for the protein module, and
vice versa.

The R2 and -log(p) do not diminish much after adjusting for the 
protein module.

```{r comparison}
weak_mod_effect <- vector(mode = "list", length = nrow(cl.mediating.pairs))
weak_mediator <- colnames(cl.mediators)[2]

for(m in 1:nrow(cl.mediating.pairs)){
    bd.name = cl.mediating.pairs[m,1]
    kegg.name = cl.mediating.pairs[m,2]
    
    weak_mod_effect[[m]] <- mediation_effect(bd.name, kegg.name, 
        cl.mediators[,weak_mediator,drop=FALSE], old_data, rna_data, test_scaled, 
        plot.results = FALSE)
}

plot_mediator_effects(weak_mod_effect, cl.mediating.pairs, plot.label = weak_mediator)
```

## Discussion

So here we have a single group of proteins, enriched for subcellular
localization, that appears to mediate a widespread transcriptomic
response influencing mitochondria, ribosomes, and synapses. 

Within this group of proteins we have proteins involved in:

* ER export and cargo sorting:
    - AP1S1, ARFGAP1, ARFIP2, SEC24C, GOLPH3, GLG1
    In neurons, SEC24 and ARFGAP1 are directly implicated in 
    trafficking and axonal targeting of synaptic proteins

* endolysosomal trafficking
    TMEM106B, ATG13, STXBP2, ZFYVE1, DYNC1LI1, DYNC2LI1, GPRASP2
    TMEM106B is especially interesting. It regulates lysosomal 
    localization, lysosomal transport, and dendritic lysosome trafficking in neurons
    and when altered affects lysosomal trafficking and synaptic signaling
    ATG13 is upstream of autophagy initiation, including mitophagy.

* Mitochondrial transport proteins
    - MTX2 (Metaxin 2) Part of the mitochondrial outer-membrane protein import machinery.
    - MRPL54, Mitochondrial ribosomal protein, 

* neuron-ribosome connection
    - CPEB3/CPEB4, These are regulators of dendritic mRNA translation
    CPEB3 controls translation of synaptic transcripts including AMPA 
    receptor-associated mRNAs and is a well-known mediator of local protein 
    synthesis in neurons.


* Synaptic trafficking
     - SEC24C, ARFGAP1, STXBP2, TMEM106B, PTPRF, SEZ6L, THSD7A, VIP, CCK, CPEB3, CPEB4
    - SEC24/ARFGAP-dependent trafficking has been demonstrated for neuronally polarized cargo and proper axonal targeting
    - TMEM106B regulates lysosomal trafficking within dendrites and influences synaptic function.
    - PEB3/CPEB4 regulate activity-dependent local translation of synaptic proteins

* Long-range vesicle transport
    - DYNC1LI1, DYNC2LI1, STXBP2

* Endolysosomal-autophagy control
    - TMEM106B, ATG13, ZFYVE1, GPRASP2

* local translational remodeling
    - CPEB3, CPEB4, ZC3H14, TNPO1

Maybe good proteins to talk about.
    SEC24C, ARFGAP1, AP1S1, GOLPH3, TMEM106B, ATG13, DYNC1LI1, CPEB3, and CPEB4

So trafficking and local proteostatic control of
synaptic membrane proteins and synaptic organelles
are the primary effects, and mitochondrial and ribosomal
signatures would then arise as downstream adaptive reactions.

Klotho deficiency downregulates ATG13, impairing ULK1 complex 
formation and inhibiting autophagy initiation during ER- and 
Golgi-stress responses. The authors concluded that Klotho 
regulates autophagy partly through control of ATG13 expression
(PMID: 31732843).


```{r pubmed_search}
stop()
library(easyPubMed)

search_genes <- sort(clust_names[[1]])
additional.strings <- c("AND brain", "AND Klotho", "AND neurodegeneration", "AND ribosome", "AND mitochondria", "AND synapse", "AND neuron")

count_mat <- matrix(NA, nrow = length(search_genes), ncol = length(additional.strings))
rownames(count_mat) <- search_genes
colnames(count_mat) <- additional.strings

for(s in 1:length(additional.strings)){
    search_string <- paste(clust_names[[1]], additional.strings[s])
    all.pubmed <- lapply(search_string, epm_query)
    num.articles <- sapply(all.pubmed, function(x) get_epm_meta(x)$exp_count)
    count_mat[,s] <- num.articles
}

pheatmap(log10(count_mat+1))

test.gene <- "Cck"
plot_rna_abund(test.gene, rna_data)
plot_pr_abund(test.gene, obs_data)

plot_multi_gene(unlist(clust_names), obs_data, mouse_order = "genotype")
```
