---
title: "Klotho in ROSMAP"
author: Anna L Tyler
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  html_document:
    code_folding: hide
    collapsed: no
    toc: yes
    toc_float: yes
---

The purpose of this workflow is to look at whether 
the genes we identified in the mouse experiment 
are differentially expressed by KL status in ROSMAP
data. 

Initially, though, we will just look at the demographics
of KL haplotype. In GTEx we were severely underpowered,
and the VS homozygotes were all male and older on average
than the rest of the subjects. We want to check all these
things in the ROSMAP data.

For gene expression, we used the harmonized counts filtered
and adjusted for age at death (syn26967456).

```{r param}
rm(list = ls())
library(here)
```

```{r params}
gene.set <- "Glutamate_Genes"
#gene.set <- "Ribosome-Synapse_Genes"
#gene.set <- "Parkinson_disease-Mitochondrial_Metabolism_Genes"

gene.group.label = gsub("_", " ", gene.set)

```

```{r load_code}
all.fun <- list.files(here("Code"), full.names = TRUE)
for(i in 1:length(all.fun)){source(all.fun[i])}
```

```{r libraries, message = FALSE, warning = FALSE, error = FALSE}
needed.libraries <- c("pheatmap")
load_libraries(needed.libraries)
```


```{r load_data}
kl.geno <- read.delim(here("Data", "Human", "rosmap.klotho.haplotypes.txt"), header = FALSE)
demo <- read.delim(here("Data",  "Human", "dataset_1175_basic_06-20-2022.txt"))
#demo <- read.csv(here("Data", "Human", "ROSMAP_clinical.csv"))
id.key <- read.csv(here("Data", "Human", "ROSMAP_IDkey.csv"))
expr.meta <- read.csv(here("Data", "Human", "ROSMAP_assay_rnaSeq_metadata.csv"))
comp.expr.file <- here("Data", "Human", "ROSMAP_Residualized_counts_(age-death).RDS")
if(!file.exists(comp.expr.file)){
  expr <- read.delim(here("Data", "Human", "ROSMAP_Residualized_counts_(age-death).tsv"))
  saveRDS(expr, comp.expr.file)
}else{
  expr <- readRDS(comp.expr.file)
}

#validation genes
val.genes <- read.table(here("Results", "age_batch_12_gene", "validation_gene_sets", 
  paste0(gene.set, ".txt")))[,1]
ortho.file <- here("Results", "age_batch_12_gene", 
  "validation_gene_sets", paste0(gene.set, "_Orthologs.csv"))
gene.orthos <- get_orthologs(mouse.id = val.genes, file.name = ortho.file)
```

We aligned all the data to include only individuals that had 
data in all modalities: Klotho haplotype, gene expression,
and were controls. We don't want AD or another neurodegenerative
disease to be affecting gene expression.

```{r align}
all.id <- list("KL" = kl.geno[,1], 
  "demographic" = demo[,"projid"],
  "expression" = gsub("X", "", colnames(expr)),
  "controls" = demo[which(demo[,"cogdx"] == 1),"projid"]
  )

id.match <- lapply(all.id, function(y) apply(id.key, 2, function(x) length(intersect(y, x))))
#sapply(id.match, max)
id.col <- sapply(id.match, which.max)

common.id <- lapply(1:length(all.id), function(x) intersect(all.id[[x]], id.key[,id.col[x]]))
common.idx <- lapply(1:length(common.id), function(x) match(common.id[[x]], id.key[,id.col[[x]]]))
merged.idx <- Reduce("intersect", common.idx)
#length(merged.idx)

#subset key to only patients with 
aligned.key <- id.key[merged.idx,]

id_idx <- function(sub.key, id.vals){
  key.match <- apply(sub.key, 2, function(x) length(intersect(id.vals, x)))
  #barplot(key.match)
  match.col <- which.max(key.match)  
  match.idx <- match(sub.key[,match.col], id.vals)
  return(match.idx)
}


#expression ID alignment
expr.idx <- id_idx(aligned.key, all.id$expression)
kl.idx <- id_idx(aligned.key, all.id$KL)
demo.idx <- id_idx(aligned.key, all.id$demographic)

#pull out genes of interest
hum.genes <- unique(gene.orthos[,"human_ensembl_gene_id"])
gene.idx <- sapply(hum.genes, function(x) grep(x, expr[,1]))

if(class(gene.idx) == "list"){ #if we didn't find all the genes
  gene.idx <- unlist(gene.idx)
  #reassign to just the genes we found
  hum.genes <- names(gene.idx) 
  #subset to just the genes we found
  gene.orthos <- gene.orthos[match(hum.genes, gene.orthos[,"human_ensembl_gene_id"]),]
}

group.expr <- expr[gene.idx,expr.idx]
#group.expr <- t(apply(expr[gene.idx,expr.idx], 1, scale))
rownames(group.expr) <- unique(gene.orthos[,"human_external_gene_name"])
#pheatmap(group.expr)

demo.kl <- cbind(kl.geno[kl.idx,], demo[demo.idx,], t(group.expr))
colnames(demo.kl)[1:2] <- c("KL_ID", "KL_hap")
#demo.kl[1:3, 1:5]

kl.hap <- unique(demo.kl[,"KL_hap"])
```

## Counts of KL haplotypes

After aligning the KL and demographic data, we want to look 
at the demographics of individuals in different KL haplotype
groups. First, let's just look at the counts of the haplotypes.

The bar plot below shows the number of individuals in each 
genotype group. Not great.

```{r kl_demo}
kl.counts <- table(demo.kl[,"KL_hap"])
barplot_with_num(kl.counts)
```

## Haplotype counts by sex

```{r kl_by_sex}
kl.sex <- table(demo.kl[,c("KL_hap", "msex")])
sex.cols <- c("female" = "#7fbf7b", "male" = "#af8dc3")
a <- barplot(t(kl.sex), beside = TRUE, col = sex.cols, 
  ylab = "Count", ylim = c(0, 100))
for(i in 1:nrow(a)){
  text(x = a[i,], y = kl.sex[,i]+2, labels = kl.sex[,i])
}
legend("topright", fill = sex.cols, legend = names(sex.cols))
```

## Correlation of haplotype and other factors

We looked for correlation of KL haplotype with various 
factors. There were none.


```{r kl_effect}
kl_effect <- function(factor.name, factor.label = "", ylim = NULL, plot.results = TRUE){
  kl.edu <- lapply(kl.hap, function(x) as.numeric(demo.kl[which(demo.kl[,"KL_hap"] == x), factor.name]))
  names(kl.edu) <- kl.hap
  if(plot.results){
    edu.test <- pairwise_test(kl.edu)
    if(is.null(ylim)){
      ymin <- min(as.numeric(demo.kl[,factor.name]), na.rm = TRUE)
      ymax <- max(as.numeric(demo.kl[,factor.name]), na.rm = TRUE)
      plot.height <- ymax - ymin
      plot.max <- ymax + (plot.height*0.2)
      ylim <- c(ymin, plot.max)
    }
    boxplot(kl.edu, xlab = "KL haplotype", ylab = factor.label, ylim = ylim, 
      main = factor.label)
    add_pairwise_results(kl.edu, edu.test)
  }
  invisible(kl.edu)
}
```

```{r other, fig.width = 12, fig.height = 8}
#other.things <- c("Age at Death" = "age_death", "Age at diagnosis" = "age_first_ad_dx", "Post-Mortem Interval" = "pmi", "Braak Score" = "braaksc", "Cerad Score" = "ceradsc")
other.things <- c("Education" = "educ", "Age at Death" = "age_death", "Tangles" = "tangles", "Braak Score" = "braaksc", "Amyloid" = "amyloid")

layout(get.layout.mat(length(other.things)))
for(i in 1:length(other.things)){
  kl_effect(factor.name = other.things[i], factor.label = names(other.things)[i])
}
```

## KL haplotype and expression of `r gene.group.label`

Is expression affected by KL haplotype?

```{r expr_kl, fig.width = 12, fig.height = 9}
mean.mat <- matrix(NA, nrow = nrow(group.expr), ncol = length(kl.hap))
rownames(mean.mat) <- rownames(group.expr)
colnames(mean.mat) <- kl.hap


if(nrow(group.expr) > 12){
  plot.results <- FALSE
}else{
  plot.results <- TRUE
  layout(get.layout.mat(nrow(group.expr)))
}

for(i in 1:nrow(group.expr)){
  hap.expr <- kl_effect(factor.name = rownames(group.expr)[i], 
    factor.label = rownames(group.expr)[i], plot.results = plot.results)
  mean.mat[i,] <- sapply(hap.expr, mean)
}
```

The average expression per genotype for each gene is shown
below.

```{r mean_mat}
pheatmap(mean.mat, cluster_cols = FALSE)
```

## Average of gene expression

Average expression across all genes.

```{r avg_expr}
avg.expr <- colMeans(group.expr)
demo.kl <- cbind(demo.kl, avg.expr)
kl_effect(factor.name = "avg.expr", 
  factor.label = "Average Expression")
```
