---
title: "GTEx Results Summary"
author: Anna L Tyler
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  html_document:
    code_folding: hide
    collapsed: no
    toc: yes
    toc_float: yes
---

## Introduction

This workflow reads in the results generated by 
3.GTEx_validation.Rmd and summarizes them across
all the brain regions tested.


```{r clear}
rm(list = ls())
library(here)
```

```{r load_code}
all.fun <- list.files(here("Code"), full.names = TRUE)
for(i in 1:length(all.fun)){source(all.fun[i])}
```

```{r libraries, message = FALSE, warning = FALSE, error = FALSE}
needed.libraries <- c("pheatmap", "vioplot", "RColorBrewer")
load_libraries(needed.libraries)
```

```{r params}
tissue.cols <- brewer.pal(12, "Paired")
```

## Effects of Age

```{r age_effect, fig.width = 12, fig.height = 7}
all.age.files <- list.files(here("Results", "GTEx"), pattern = "Mean_Expr_by_Age",
    full.names = TRUE)
tissues <- gsub("_", " ", gsub(".csv", "", gsub("Mean_Expr_by_Age_brain_", "", basename(all.age.files))))
all.age <- lapply(all.age.files, read.csv)
age.bins <- gsub(".", "-", gsub("X", "", colnames(all.age[[1]])), fixed = TRUE)
for(i in 1:length(all.age)){
    colnames(all.age[[i]]) <- age.bins
}
names(all.age) <- tissues
plot.grouped.boxes(all.age, type = "matrix", legend.x = 15, legend.y = 1,
    ylab = "Expression Relative to Tissue Mean")

```


```{r age_line, fig.width = 7, fig.height = 7}
age.table <- t(sapply(all.age, colMeans))

#quartz(width = 7, height = 7)
ymin <- min(age.table)-0.1
ymax <- max(age.table)+0.05
plot.new()
plot.window(xlim = c(1,ncol(age.table)),ylim = c(ymin, ymax))
for(i in 1:nrow(age.table)){
    points(1:ncol(age.table), age.table[i,], type = "b", col = tissue.cols[i], 
    pch = 16)
}
axis(2)
mtext("Expression Relative to Tissue Mean", side = 2, line = 2.5)
mtext("Glutamate Receptor Expression Declines with Age\nAcross all Brain Tissues", 
    side = 3, line = 1.5)
text(1:length(age.bins), y = ymin, labels = age.bins)
legend("topright", pch = 16, col = tissue.cols[1:nrow(age.table)], 
    legend = rownames(age.table))
```

```{r age_viol, fig.width = 7, fig.height = 6}
age.mat.tests <- pairwise_test(age.table)

vioplot(age.table, main = "Expression Across all Tissues with Age",
    ylim = c(min(age.table), max(age.table)+0.1))
add_pairwise_results(data.mat = age.table, 
    test.results = age.mat.tests, y.nudge = 0.02, text.nudge = 0.02)

mtext("Expression Relative to Tissue Mean", side = 2, line = 2.5)
```

## Effects of Klotho on Glutamate Receptor Gene Expression {.tabset .tabset-fade .tabset-pills}

Klotho haplotype seems to have varied effects
across regions.

```{r kl_files}
kl.files <- list.files(here("Results", "GTEx"), pattern = "Mean_KL_Expr_by_Age", full.name = TRUE)
kl.expr <- lapply(kl.files, read.csv)
names(kl.expr) <- tissues
```

```{r kl_line, results = "asis", fig.width = 12, fig.height = 5}
ymin <- min(unlist(kl.expr), na.rm = TRUE)-0.1
ymax <- max(unlist(kl.expr), na.rm = TRUE)+0.1

kl.by.age <- lapply(1:length(age.bins), function(x) 
    t(sapply(kl.expr, function(y) y[x,])))
names(kl.by.age) <- age.bins


for(i in 1:length(kl.by.age)){
    cat("###", names(kl.by.age)[i], "\n")
    ymin = min(as.numeric(kl.by.age[[i]]), na.rm = TRUE)-0.1
    ymax = max(as.numeric(kl.by.age[[i]]), na.rm = TRUE)

    layout(matrix(c(1,2), nrow = 1), widths = c(1,0.6))
    par(mar = c(4,4,4,0))
    plot.new()
    plot.window(xlim = c(1,3), ylim = c(ymin, ymax))
    for(bt in 1:nrow(kl.by.age[[i]])){
        points(1:3, kl.by.age[[i]][bt,1:3], pch = 16, 
            col = tissue.cols[bt], type = "b")
    }
    mtext(names(kl.by.age)[i], side = 3, line = 2.5)
    text(x = 1:3, y = ymin, labels = colnames(kl.by.age[[i]]))
    axis(2)
    mtext("Expression Relative to Tissue Mean", side = 2, line = 2.5)
    abline(h = 0)
    
    par(mar = c(4,0,4,2))
    plot.new()
    plot.window(xlim = c(0,1), ylim = c(0,1))
    legend(x = 0, y = 1, legend = rownames(kl.by.age[[i]]), 
        col = tissue.cols[1:nrow(kl.by.age[[i]])], pch = 16, lty = 1)
   cat("\n\n") 
}
```

## Glutamate gene expression by KL and in boxes {.tabset .tabset-fade .tabset-pills}


```{r boxes, results = "asis"}

for(i in 1:length(kl.by.age)){
    cat("###", names(kl.by.age)[i], "\n")
    
    age.batch <- apply(kl.by.age[[i]], 2, as.numeric)
    test.diff <- pairwise_test(age.batch)
    
    boxplot(age.batch, ylab = "Expression Relative to Tissue Mean",
        main = names(kl.by.age)[i], ylim = c(min(age.batch, na.rm = TRUE), 
        max(age.batch, na.rm = TRUE)+0.5))

    stripchart(lapply(1:ncol(age.batch), function(x) age.batch[,x]), 
        vertical = TRUE, method = "jitter", col = "#dd1c77",
        add = TRUE, pch = 16)
    abline(h = 0)
    
    add_pairwise_results(data.mat = age.batch, test.results = test.diff)
    
    cat("\n\n")
}
```


## Comparison to mouse data

We really only have the FC/FC and the VS/VS genotypes in the mice,
so only those are a direct comparison. 


In young mice, Klotho allele has no effect on glutamate receptor 
genes. We don't have any young VS/VS individuals. The youngest is
52, so I don't think we can really evaluate this similarity. 

In older mice, the VS/VS genotype reduces glutamate receptor 
expression. 

We have middle aged humans, and older humans (no older mice yet).
In middle aged humans, the VS/VS haplotype reduces glutamate
receptor gene expression in some brain areas, but seems to 
increase it in others.

It appears to do the same thing in humans in multiple brain 
tissues, which is consistent with previous observations that
the VS/VS genotype has potentially deleterious effects.

