---
title: "GTEx Results Summary"
author: Anna L Tyler
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  html_document:
    code_folding: hide
    collapsed: no
    toc: yes
    toc_float: yes
---

## Introduction

This workflow reads in the results generated by 
3.GTEx_validation.Rmd and summarizes them across
all the brain regions tested.


```{r clear}
rm(list = ls())
library(here)
```

```{r load_code}
all.fun <- list.files(here("Code"), full.names = TRUE)
for(i in 1:length(all.fun)){source(all.fun[i])}
```

```{r libraries, message = FALSE, warning = FALSE, error = FALSE}
needed.libraries <- c("pheatmap", "vioplot", "RColorBrewer")
load_libraries(needed.libraries)
```

```{r params}
args = commandArgs(trailingOnly=TRUE)

gene.set <- args[1]

if(is.na(gene.set)){
    gene.set <- "Glutamate_Genes"
    #gene.set <- "Ribosome-Synapse_Genes"
    #gene.set <- "Parkinson_disease-Mitochondrial_Metabolism_Genes"
}
gene.group.label = gsub("_", " ", gene.set)

tissue.cols <- brewer.pal(12, "Paired")

geno.cols <- get_geno_cols(here("Data", "general", "geno_cols.csv"))
het.col <- "#dfc27d"
hap.cols <- c("FCFC" = as.vector(geno.cols["FC/FC"]), 
  "FCVS" = het.col, "VSVS" = as.vector(geno.cols["VS/VS"]))

```

## Group Counts

```{r switch_points}
switch_points <- function(labelV){
    adj.labels <- consec_pairs(labelV)
    switch.pts <- c(0, which(!apply(adj.labels, 1, function(x) x[1] == x[2])), 
        length(labelV))
    label.segs <- matrix(NA, ncol = 3, nrow = (length(switch.pts)-1))
    colnames(label.segs) <- c("start", "stop", "label")
    for(i in 1:(length(switch.pts)-1)){
        label.start <- (switch.pts[i]+1)
        label.end <- switch.pts[(i+1)]
        label.seg <- unique(labelV[label.start:label.end])
        label.segs[i,] <- c(label.start, label.end, label.seg)
    }
    return(label.segs)
}

add_seg_label <- function(seg.table, ymin, ymax, x.shift = -0.5, cex = 1){
    for(i in 1:nrow(seg.table)){
        rect.start <- as.numeric(seg.table[i,1])+x.shift
        rect.end <- (as.numeric(seg.table[i,2])-x.shift)
        draw.rectangle(min.x = rect.start, max.x = rect.end,
            min.y = ymin, max.y = ymax)
        label.x <- mean(c(rect.start, rect.end))
        text(x = label.x, y = mean(c(ymin, ymax)), labels = seg.table[i,3], cex = cex)
    }
}

```


```{r group_counts}
all.count.files <- get.files(here("Results", "GTEx"), want = c("Expr_group_num",
    gene.set), full.name = TRUE)
all.counts <- lapply(all.count.files, readRDS)
count.tissue <- gsub(".RDS", "", gsub(paste0("_", gene.set), "", gsub("Expr_group_num_brain_", "", basename(all.count.files))))

#make a table of subject counts for each age group, then sex, then KL haplotype
count.table <- round(t(sapply(all.counts, function(x) x[1,])))
rownames(count.table) <- count.tissue
#pheatmap(count.table, display_numbers = TRUE, cluster_cols = FALSE, cluster_rows = FALSE)

known.geno <- setdiff(1:ncol(count.table), grep("unknown", colnames(count.table)))
known.table <- count.table[,known.geno]
#pheatmap(known.table, display_numbers = TRUE, cluster_cols = FALSE, cluster_rows = FALSE)

split.label <- strsplit(colnames(known.table), " ")
age.label <- sapply(split.label, function(x) x[1])
sex.label <- sapply(split.label, function(x) x[2])
kl.label <- sapply(split.label, function(x) x[3])
kl.label[which(kl.label == "FCFC")] <- "FC"
kl.label[which(kl.label == "VSVS")] <- "VS"
kl.label[which(kl.label == "FCVS")] <- "het"

png(here("Results", "for_paper", "Human_Expr_Count_Table.png"), 
    width = 10, height = 6, units = "in", res = 300)
par(mar = c(4, 14, 8, 4))
imageWithText(known.table, col.scale = "blue", row.text.shift = 0.05, 
    col.names = NULL, cex = 0.7)

age.segs <- switch_points(age.label)
sex.segs <- switch_points(sex.label)
kl.segs <- switch_points(kl.label)

add_seg_label(age.segs, ymin = 14-0.5, ymax = 15-0.5, x.shift = -0.5)
text(x = 0.25, y = 14, labels = "Age", adj = 1)
add_seg_label(sex.segs, ymin = 13-0.5, ymax = 14-0.5, x.shift = -0.5)
text(x = 0.25, y = 13, labels = "Sex", adj = 1)
add_seg_label(kl.segs, ymin = 12-0.5, ymax = 13-0.5, x.shift = -0.5, cex = 0.7)
text(x = 0.25, y = 12, labels = "KL", adj = 1)

#add delimiters for the ages and sexes
for(i in 1:nrow(sex.segs)){
    draw.rectangle(as.numeric(sex.segs[i,1])-0.5, as.numeric(sex.segs[i,2])+0.5,
        min.y = 0.5, max.y = 11.5, lwd = 1, border = "gray20")
}

for(i in 1:nrow(age.segs)){
    draw.rectangle(as.numeric(age.segs[i,1])-0.5, as.numeric(age.segs[i,2])+0.5,
        min.y = 0.5, max.y = 11.5, lwd = 2)
}
dev.off()
```

```{r group_mean_expr, eval = FALSE}

#make a table of subject counts for each age group, then sex, then KL haplotype
mean.expr.table <- t(sapply(all.counts, function(x) x[2,]))
rownames(mean.expr.table) <- count.tissue

known.geno <- setdiff(1:ncol(count.table), grep("unknown", colnames(count.table)))
known.expr <- mean.expr.table[,known.geno]
#pheatmap(known.expr, display_numbers = TRUE, cluster_cols = FALSE, cluster_rows = FALSE)

png(here("Results", "for_paper", "Human_Expr_Table.png"), 
    width = 10, height = 6, units = "in", res = 300)
par(mar = c(4, 14, 8, 4))
imageWithText(known.expr, show.text = FALSE, use.pheatmap.colors = TRUE, 
    row.text.shift = 0.05, col.names = NULL, cex = 0.7)

add_seg_label(age.segs, ymin = 14-0.5, ymax = 15-0.5, x.shift = -0.5)
add_seg_label(sex.segs, ymin = 13-0.5, ymax = 14-0.5, x.shift = -0.5)
add_seg_label(kl.segs, ymin = 12-0.5, ymax = 13-0.5, x.shift = -0.5, cex = 0.7)

#add delimiters for the ages and sexes
for(i in 1:nrow(sex.segs)){
    draw.rectangle(as.numeric(sex.segs[i,1])-0.5, as.numeric(sex.segs[i,2])+0.5,
        min.y = 0.5, max.y = 11.5, lwd = 1, border = "gray20")
}

for(i in 1:nrow(age.segs)){
    draw.rectangle(as.numeric(age.segs[i,1])-0.5, as.numeric(age.segs[i,2])+0.5,
        min.y = 0.5, max.y = 11.5, lwd = 2)
}

add_color_bar(known.expr, position = "topright", use.pheatmap.colors = TRUE,
    x.min.shift = -0.05, x.max.shift = -0.05, y.min.shift = 0.05, y.max.shift = 0.05)

dev.off()
```

## Effects of Age

```{r age_effect, fig.width = 12, fig.height = 7}
all.age.files <- get.files(here("Results", "GTEx"), 
    want = c(gene.set, "Mean_Expr_by_Age"),
    full.names = TRUE)
tissues <- gsub("_", " ", gsub(".csv", "", gsub("Mean_Expr_by_Age_brain_", "", basename(all.age.files))))
all.age <- lapply(all.age.files, read.csv)
age.bins <- gsub(".", "-", gsub("X", "", colnames(all.age[[1]])), fixed = TRUE)
for(i in 1:length(all.age)){
    colnames(all.age[[i]]) <- age.bins
}
names(all.age) <- gsub(paste("", gene.group.label), "", tissues)
plot.grouped.boxes(all.age, type = "matrix", legend.x = 15, legend.y = 1,
    ylab = "Expression Relative to Tissue Mean")

```


```{r age_line, fig.width = 7, fig.height = 7}
age.table <- t(sapply(all.age, colMeans))

#quartz(width = 7, height = 7)
ymin <- min(age.table)-0.1
ymax <- max(age.table)+0.05
plot.new()
plot.window(xlim = c(1,ncol(age.table)),ylim = c(ymin, ymax))
for(i in 1:nrow(age.table)){
    points(1:ncol(age.table), age.table[i,], type = "b", col = tissue.cols[i], 
    pch = 16)
}
axis(2)
mtext("Expression Relative to Tissue Mean", side = 2, line = 2.5)
mtext(paste(gene.group.label, "Gene Expression Trends with Age\nAcross all Brain Tissues"), 
    side = 3, line = 1.5)
text(1:length(age.bins), y = ymin, labels = age.bins)
legend("topright", pch = 16, col = tissue.cols[1:nrow(age.table)], 
    legend = rownames(age.table))

```

```{r age_viol, fig.width = 7, fig.height = 6}

age.mat.tests <- pairwise_test(age.table)

vioplot(age.table, main = "Expression Across all Tissues with Age",
    ylim = c(min(age.table), max(age.table)+0.1))
add_pairwise_results(data.mat.or.list = age.table, 
    test.results = age.mat.tests, y.nudge = 0.02, text.nudge = 0.02)

mtext("Expression Relative to Tissue Mean", side = 2, line = 2.5)
```

## Effects of Klotho on `r gene.group.label` Gene Expression {.tabset .tabset-fade .tabset-pills}

Klotho haplotype seems to have varied effects
across regions.

```{r kl_files}
kl.files <- get.files(here("Results", "GTEx"), 
    want = c("Mean_KL_Expr_by_Age", gene.set), full.name = TRUE)
kl.expr <- lapply(kl.files, read.csv)
names(kl.expr) <- tissues
```

```{r kl_line, results = "asis", fig.width = 12, fig.height = 5}
ymin <- min(unlist(kl.expr), na.rm = TRUE)-0.1
ymax <- max(unlist(kl.expr), na.rm = TRUE)+0.1

kl.by.age <- lapply(1:length(age.bins), function(x) 
    t(sapply(kl.expr, function(y) as.numeric(y[x,]))))
names(kl.by.age) <- age.bins
for(i in 1:length(kl.by.age)){
    colnames(kl.by.age[[i]]) <- colnames(kl.expr[[1]])
}

for(i in 1:length(kl.by.age)){
    cat("###", names(kl.by.age)[i], "\n")
    ymin = min(as.numeric(kl.by.age[[i]]), na.rm = TRUE)-0.1
    ymax = max(as.numeric(kl.by.age[[i]]), na.rm = TRUE)

    layout(matrix(c(1,2), nrow = 1), widths = c(1,0.6))
    par(mar = c(4,4,4,0))
    plot.new()
    plot.window(xlim = c(1,3), ylim = c(ymin, ymax))
    for(bt in 1:nrow(kl.by.age[[i]])){
        points(1:3, kl.by.age[[i]][bt,1:3], pch = 16, 
            col = tissue.cols[bt], type = "b")
    }
    mtext(names(kl.by.age)[i], side = 3, line = 2.5)
    text(x = 1:3, y = ymin, labels = colnames(kl.by.age[[i]]))
    axis(2)
    mtext("Expression Relative to Tissue Mean", side = 2, line = 2.5)
    abline(h = 0)
    
    par(mar = c(4,0,4,2))
    plot.new()
    plot.window(xlim = c(0,1), ylim = c(0,1))
    legend(x = 0, y = 1, legend = gsub(paste("", gene.group.label), "", rownames(kl.by.age[[i]])), 
        col = tissue.cols[1:nrow(kl.by.age[[i]])], pch = 16, lty = 1)
   cat("\n\n") 
}

```

## `r gene.group.label` Gene expression by KL and age {.tabset .tabset-fade .tabset-pills}


```{r boxes, results = "asis"}

for(i in 1:length(kl.by.age)){
    cat("###", names(kl.by.age)[i], "\n")
    
    age.batch <- apply(kl.by.age[[i]], 2, as.numeric)
    test.diff <- pairwise_test(age.batch)
    
    #remove columns of NAs, vioplot can't handle them.
    not.na.col <- which(apply(age.batch, 2, function(x) !all(is.na(x))))
    no.na <- age.batch[,not.na.col]

    vioplot(no.na,main = names(kl.by.age)[i], 
        ylim = c(min(age.batch, na.rm = TRUE), 
        max(age.batch, na.rm = TRUE)+0.55), col = hap.cols[colnames(no.na)])
    mtext("Expression Relative to Tissue Mean", side = 2, line = 2.5)

    stripchart(lapply(1:ncol(age.batch), function(x) age.batch[,x]), 
        vertical = TRUE, method = "jitter", col = "#dd1c77",
        add = TRUE, pch = 16)
    abline(h = 0)
    
    add_pairwise_results(data.mat.or.list = no.na, test.results = test.diff,
        y.nudge = 0.05, text.nudge = 0.04, min.beam.dist = 0.1)
    
    cat("\n\n")
}
```

```{r old_kl_effect}
pdf(here("Results", "for_paper", paste0("Human_Validation_vioplot_", gene.set, ".pdf")), 
    width = 5, height = 5)   
    #quartz(width = 5, height = 5)
    i = length(kl.by.age)
    par(lwd = 1)
    age.batch <- apply(kl.by.age[[i]], 2, as.numeric)
    test.diff <- pairwise_test(age.batch)

    #remove columns of NAs, vioplot can't handle them.
    not.na.col <- which(apply(age.batch, 2, function(x) !all(is.na(x))))
    no.na <- age.batch[,not.na.col]

    ylim <- c(min(age.batch, na.rm = TRUE), max(age.batch, na.rm = TRUE)+0.2)
    #ylim = c(-0.5, 0.45)
    vioplot(no.na, ylim = ylim, col = hap.cols[colnames(no.na)],
        cex.lab = 1.5, cex.axis = 1.5,)
    mtext("Median Expression", side = 2, line = 2.5, cex = 1.5)

    stripchart(lapply(1:ncol(age.batch), function(x) age.batch[,x]), 
        vertical = TRUE, method = "jitter", col = "#dd1c77",
        add = TRUE, pch = 16)
    abline(h = 0)

   reordered.result <- test.diff
    if(gene.set == "Glutamate_Genes" || gene.set == "Parkinson_disease-Mitochondrial_Metabolism_Genes"){
        #change order of beams
        new.order <- c(1,3,2)
        reordered.result[[1]] <- test.diff[[1]][new.order,]
        reordered.result[[2]] <- test.diff[[2]][new.order]
    }
    if(gene.set == "Parkinson_disease-Mitochondrial_Metabolism_Genes"){
        y.nudge = 0.07; min.beam.dist = 0.25; text.nudge = 0.06 
    }else{
        y.nudge = 0.05; min.beam.dist = 0.15; text.nudge = 0.04
    }

    add_pairwise_results(data.mat.or.list = no.na, test.results = reordered.result,
        y.nudge = y.nudge, text.nudge = text.nudge, min.beam.dist = min.beam.dist, 
        force.higher = FALSE)
    
dev.off()

#get stats for reporting higher in hets
if(gene.set == "Glutamate_Genes"){
    vs.fc.test <- test.diff[[2]][[2]] #VS vs. FC
    vs.fc.stats <- signif(c(vs.fc.test$parameter, 
        "conf.int" = vs.fc.test$conf.int[c(1,2)],
        vs.fc.test$statistic,
        "p" = vs.fc.test$p.value), 2)    
    saveRDS(vs.fc.stats, here("Results", "for_paper", "paper_stats_Human_VS_v_FC.RDS"))


    vs.het.test <- test.diff[[2]][[3]] #hom VS vs het
    vs.v.het.stats <- signif(c(vs.het.test$parameter, 
        "conf.int" = vs.het.test$conf.int[c(1,2)],
        vs.het.test$statistic,
        "p" = vs.het.test$p.value), 2)    
    saveRDS(vs.v.het.stats, here("Results", "for_paper", "paper_stats_Human_VS_v_Het_Stats.RDS"))
}
```

## Side-by-side boxplots

```{r geno_on_x, fig.width = 7, fig.height = 5}
plot.grouped.boxes(kl.by.age, type = "matrix", main = "Effects of age in each genotype",
    ylab = "Expression relative to tissue mean")
add_zero_line()
```


```{r age_on_x, fig.width = 7, fig.height = 5}
#rearrange to get the effects of age on the x axis
age.by.kl <- lapply(1:ncol(kl.by.age[[1]]), function(x) sapply(kl.by.age, function(y) y[,x]))
names(age.by.kl) <- colnames(kl.by.age[[1]])
plot.grouped.boxes(age.by.kl, type = "matrix", main = "Effects of genotype in each age",
    group.cols = hap.cols)
add_zero_line()
```

## Interaction Model

The plot below shows how `r gene.group.label` median expression
changes with age for each genotype. The p value is for the
interaction between age and genotype.

The error bars show the 95% confidence intervals for the 
median values.

```{r interaction_model, fig.width = 6, fig.height = 6}
age.factor <- ordered(unlist(lapply(1:length(kl.by.age), function(x) rep(names(kl.by.age)[x], length(kl.by.age[[x]])))))
expr.v <- unlist(lapply(kl.by.age, unlist))
geno.v <- ordered(unlist(lapply(kl.by.age, function(x) rep(colnames(x), each = nrow(x)))))
model <- lm(expr.v~age.factor*geno.v)
model.p <- threshold_p(coef(summary(model))["age.factor.L:geno.v.L","Pr(>|t|)"], return.text = TRUE)

#interaction.plot(geno.v, age.factor, expr.v, 
#    fun = function(x) mean(x, na.rm = TRUE), lwd = 3, xlab = "Genotype",
#    ylab = paste(gene.group.label, "Expression"), trace.label = "Age Group")

par(lwd = 3)
interaction.plot(age.factor, geno.v, expr.v, 
    fun = function(x) median(x, na.rm = TRUE), lwd = 3,
    ylab = paste(gene.group.label, "Median Expression"), 
    trace.label = "Genotype", legend = FALSE, lty = 1,
    xlab = "Age Group", col = hap.cols, main = model.p,
    ylim = c(-0.8, 1.2))

#add 95% confidence intervals to the median
bounds <- add_interaction_error(age.factor, geno.v, expr.v,
    fun = function(x) median(x, na.rm = TRUE), lwd = 3,
    cols = hap.cols, 
    error.fun = function(x) conf_int(x, 95, 1000, function(y) median(y, na.rm = TRUE)),
    horiz.width = 0.02)

par(lwd = 1)
legend("topright", legend = levels(geno.v), lty = 1, col = hap.cols, lwd = 3)

abline(h = 0)
```

```{r int_plot}

min.error <- min(bounds$lower_bounds, na.rm = TRUE)
max.error <- max(bounds$upper_bounds, na.rm = TRUE)
plot.height <- max.error - min.error
ymin <- min.error - (plot.height*0.05)
ymax <- max.error + (plot.height*0.05)

if(gene.set == "Ribosome-Synapse_Genes"){
    ymax <- max.error + (plot.height*0.12)
}

pdf(here("Results", "for_paper", paste0("Human_Validation_Interaction_", gene.set, ".pdf")), 
    width = 5, height = 5)
#quartz(width = 5, height = 5)
par(lwd = 3, bg = NA, mar = c(5,5,4,2))
interaction.plot(age.factor, geno.v, expr.v, 
    fun = function(x) median(x, na.rm = TRUE), lwd = 3,
    ylab = "Median Expression", 
    trace.label = "Genotype", legend = FALSE, lty = 1,
    xlab = "Age", col = hap.cols, main = "",
    ylim = c(ymin, ymax), cex.lab = 1.5, cex.axis = 1.5)

text(x = 1.5, y = ymax*0.9, labels = model.p, cex = 1.5)

#add 95% confidence intervals to the median
add_interaction_error(age.factor, geno.v, expr.v,
    fun = function(x) median(x, na.rm = TRUE), lwd = 3,
    cols = hap.cols, 
    error.fun = function(x) conf_int(x, 95, 1000, function(y) median(y, na.rm = TRUE)),
    horiz.width = 0.02)

par(lwd = 1)
legend("topright", legend = levels(geno.v), lty = 1, col = hap.cols, lwd = 3)

abline(h = 0)
dev.off()
```


## Comparison to mouse data

We really only have the FC/FC and the VS/VS genotypes in the mice,
so only those are a direct comparison. 

In young mice, Klotho allele has no effect on `r gene.group.label` 
genes. We don't have any young VS/VS individuals. The youngest is
52, so I don't think we can really evaluate this similarity. 

In older mice, the VS/VS genotype reduces `r gene.group.label` 
expression. 

We have middle aged humans, and older humans (no older mice yet).
In middle aged humans, the VS/VS haplotype reduces `r gene.group.label`
gene expression in some brain areas, but seems to 
increase it in others.

It appears to do the same thing in humans in multiple brain 
tissues, which is consistent with previous observations that
the VS/VS genotype has potentially deleterious effects.

