---
title: "GTEx Aging Signatures"
author: Anna L Tyler
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  html_document:
    code_folding: hide
    collapsed: no
    toc: yes
    toc_float: yes
---

I am thinking of ways to help Yi Juin with aging signatures
in the GTEx data. I want to play around with our KEGG-
Biodomain intersections a bit before I suggest anything.

```{r clear}
rm(list = ls())
library(here)
```

```{r load_code}
all.fun <- list.files(here("Code"), full.names = TRUE)
for(i in 1:length(all.fun)){source(all.fun[i])}
```

```{r libraries, message = FALSE, warning = FALSE, error = FALSE}
needed.libraries <- c("pheatmap", "vioplot", "corto")
load_libraries(needed.libraries)
```

```{r expr_files}
data.dir <- here("Data", "Human", "GTEX")
pheno <- read.delim(file.path(data.dir, "GTEx_Analysis_2021-02-11_v9_Annotations_SubjectPhenotypesDS.txt"))
age <- pheno[,"AGE"]
names(age) <- pheno[,"SUBJID"]

expr.files <- list.files(here("Results", "GTEx"), 
    pattern = "Adjusted_Expr", full.names = TRUE)

tissue.names <- gsub(".RDS", "", gsub("Adjusted_Expr_", "", basename(expr.files)))
file.pairs <- pair.matrix(expr.files)
tissue.pairs <- pair.matrix(tissue.names)
```



```{r tissue_pairs, eval = FALSE}
#correlations of expression in different brain tissues

cor.dist <- vector(mode = "list", length = nrow(file.pairs))
names(cor.dist) <- apply(tissue.pairs, 1, function(x) paste(x, collapse = "-"))

for(i in 1:nrow(file.pairs)){
    print(i)
    file1 <- file.pairs[i,1]
    file2 <- file.pairs[i,2]

    expr1 <- readRDS(file1)
    expr2 <- readRDS(file2)

    common.gene <- intersect(colnames(expr1), colnames(expr2))
    common.ind <- intersect(rownames(expr1), rownames(expr2))

    gene1.idx <- match(common.gene, colnames(expr1))
    gene2.idx <- match(common.gene, colnames(expr2))

    ind1.idx <- match(common.ind, rownames(expr1))
    ind2.idx <- match(common.ind, rownames(expr2))

    aligned1 <- expr1[ind1.idx, gene1.idx]
    aligned2 <- expr2[ind2.idx, gene2.idx]

    tissue.cor <- sapply(1:nrow(aligned1), 
        function(x) cor(aligned1[x,], aligned2[x,], 
        use = "pairwise.complete.obs", method = "spearman"))
    names(tissue.cor) <- rownames(aligned1)

    cor.dist[[i]] <- tissue.cor
}

#cor.order <- order(sapply(cor.dist, mean))
#par(mar = c(4,28, 4, 2))
#boxplot(cor.dist[cor.order], las = 2, horizontal = TRUE)
#abline(v = 0)
```

```{r sim_mat, eval = FALSE}
sim.mat <- matrix(NA, nrow = length(tissue.names), ncol = length(tissue.names))
rownames(sim.mat) <- colnames(sim.mat) <- tissue.names
for(i in 1:nrow(tissue.pairs)){
    tissue1 <- tissue.pairs[i,1]
    tissue2 <- tissue.pairs[i,2]
    sim.mat[tissue1, tissue2] <- mean(cor.dist[[i]])
    sim.mat[tissue2, tissue1] <- mean(cor.dist[[i]])
}
pheatmap(sim.mat, display_numbers = TRUE)
```

I thought it might be interesting to see how the kegg-biodomain
intersections look at different ages.

```{r kegg-biodomain}
kg.bd <- readRDS(here("Results", "Processed_Data", 
    "Human_KEGG_Intersections_for_GSEA.RDS"))
```

```{r}
full.cor.list <- tissue.ind.means <- vector(mode = "list", length = length(tissue.names))
names(full.cor.list) <- names(tissue.ind.means) <- tissue.names

for(tx in 1:length(tissue.names)){
    print(tx)
    tx.expr <- readRDS(expr.files[tx])
    all.age.cor <- all.ind.means <- vector(mode = "list", length = length(kg.bd))
    names(all.age.cor) <- names(all.ind.means) <- names(kg.bd)
    all.mean <- rowMeans(tx.expr, na.rm = TRUE)
    #overall expression changes with age, so we are
    #going to put everyone on the same level by regressing
    #out mean expression. I'm not sure why this is. I'm
    #already using TPM. Shouldn't that fix that kind of thing?
    #plot.with.model(age[names(all.mean)], all.mean)
    adj.tx <- adjust(matX = tx.expr, adj.mat = matrix(all.mean, ncol = 1), retain.intercept = FALSE)
    #adj.tx <- tx.expr
    
    for(bd in 1:length(kg.bd)){
        k.age.cor <- rep(NA, length(kg.bd[[bd]]))
        names(k.age.cor) <- names(kg.bd[[bd]])
        ind.means <- vector(mode = "list", length = length(kg.bd[[bd]]))
        names(ind.means) <- names(kg.bd[[bd]])

        for(k in 1:length(kg.bd[[bd]])){    
            int.id <- kg.bd[[bd]][[k]]
            common.genes <- intersect(colnames(adj.tx), int.id)
            if(length(common.genes) < 10){ #only look at intersections with at least 10 genes
                k.age.cor[k] <- NA
                next()
                }
            int.expr <- adj.tx[,common.genes]
            #plot.decomp(int.expr, cols = colors.from.values(age))
            #plot.decomp(t(int.expr))
            #pheatmap(int.expr, scale = "row")
            #pheatmap(int.expr)
            #boxplot(t(int.expr))
            #for each individual, calculate mean expression for the group of genes
            mean.expr <- apply(int.expr, 1, mean)
            ind.means[[k]] <- mean.expr
            #plot.with.model(age[names(mean.expr)], mean.expr, report = "cor.test")
            age.cor <- cor(mean.expr, age[rownames(int.expr)], method = "spearman")
            k.age.cor[k] <- age.cor
        }
        all.age.cor[[bd]] <- k.age.cor
        all.ind.means[[bd]] <- ind.means
    }
    full.cor.list[[tx]] <- all.age.cor
    tissue.ind.means[[tx]] <- all.ind.means

    #pdf(paste0("~/Desktop/", tissue.names[tx], ".pdf"), width = 10, height = 7)
    #cor.order <- order(sapply(all.age.cor, function(x) mean(x, na.rm = TRUE)))
    #par(mar = c(4,16, 4, 2))
    #boxplot(all.age.cor[cor.order], las = 2, horizontal = TRUE, main = tissue.names[tx])
    #abline(v = 0)
    #dev.off()
}

sim_mat <- function(cor.list){
    bd <- names(cor.list)
    all.k <- unique(unlist(lapply(cor.list, names)))
    sim.mat <- matrix(NA, nrow = length(bd), ncol = length(all.k))
    rownames(sim.mat) <- bd
    colnames(sim.mat) <- all.k
    for(i in 1:length(cor.list)){
        sim.mat[i,names(cor.list[[i]])] <- cor.list[[i]]
    }
    return(sim.mat)
}

all.mat <- lapply(full.cor.list, sim_mat)
for(tx in 1:length(full.cor.list)){
    pdf(paste0("~/Desktop/heat_", tissue.names[tx], ".pdf"), height = 40, width = 7)
    test.decomp <- decomp_with_nas(all.mat[[tx]], add.zeros = TRUE, plot.results = FALSE)
    #row.order <- order(test.decomp$u[,1])
    #col.order <- order(test.decomp$v[,1])
    row.order <- order(rowMeans(all.mat[[tx]], na.rm = TRUE))
    col.order <- order(colMeans(all.mat[[tx]], na.rm = TRUE))
    par(mar = c(4,4, 4, 2))
    pheatmap(t(all.mat[[tx]][row.order, col.order]), 
        cluster_rows = FALSE, cluster_cols = FALSE,
        main = tissue.names[tx])
    par(mar = c(4,24, 4, 2))
    boxplot(all.mat[[tx]][,col.order], las = 2, horizontal = TRUE)
    dev.off()

}

col.vals <- sapply(all.mat, function(x) colMeans(x, na.rm = TRUE))
row.vals <- sapply(all.mat, function(x) rowMeans(x, na.rm = TRUE))

pdf("~/Desktop/kegg_mean_cor.pdf", width = 8,  height = 45)
row.order <- order(rowMeans(col.vals, na.rm = TRUE))
col.order <- order(colMeans(col.vals, na.rm = TRUE))
pheatmap(col.vals[row.order, col.order], cluster_rows = FALSE, cluster_cols = FALSE)
dev.off()


pdf("~/Desktop/bd_mean_cor.pdf", width = 7, height = 7)
row.order <- order(rowMeans(row.vals, na.rm = TRUE))
col.order <- order(colMeans(row.vals, na.rm = TRUE))
pheatmap(row.vals[row.order, col.order], cluster_rows = FALSE, cluster_cols = FALSE)
dev.off()

row.order <- order(rowMeans(col.vals, na.rm = TRUE))
boxplot(t(col.vals[row.order,]), las = 2)

pairs(col.vals)
```

## Networks

```{r net}
all.thresh <- vector(mode = "list", length = length(tissue.names))
names(all.thresh) <- tissue.names

pdf("~/Desktop/nets.pdf")
for(tx in 1:length(tissue.names)){
    thresh.mat <- all.mat[[tx]]
    all.vals <- as.vector(thresh.mat[which(!is.na(thresh.mat))])
    high.thresh <- get.percentile(all.vals, 99)
    low.thresh <- get.percentile(all.vals*-1, 99)

    keep.vals <- union(which(thresh.mat > high.thresh), which(thresh.mat <= -low.thresh))
    zero.out <- setdiff(1:length(thresh.mat), keep.vals)
    thresh.mat[zero.out] <- NA
    keep.rows <- setdiff(1:nrow(thresh.mat), which(apply(thresh.mat, 1, function(x) all(is.na(x)))))
    keep.cols <- setdiff(1:ncol(thresh.mat), which(apply(thresh.mat, 2, function(x) all(is.na(x)))))
    sub.mat <- thresh.mat[keep.rows, keep.cols]
    all.thresh[[tx]] <- sub.mat
    #pheatmap(sub.mat, cluster_rows = FALSE, cluster_cols = FALSE)
    edge.list <- which(!is.na(sub.mat), arr.ind = TRUE)
    edge.weights <- sub.mat[which(!is.na(sub.mat))]
    edge.list[,1] <- rownames(sub.mat)[edge.list[,1]]
    edge.list[,2] <- colnames(sub.mat)[as.numeric(edge.list[,2])]

    sub.net <- graph_from_edgelist(edge.list, directed = FALSE)
    text.size <- scale.between.vals(degree(sub.net)/max(degree(sub.net)), 0.3, 1.2)
    plot(sub.net, vertex.size = degree(sub.net), layout = layout_with_kk,
        main = tissue.names[tx], edge.color = colors.from.values(edge.weights, 
        use.pheatmap.colors = TRUE), edge.width = abs(edge.weights)*10,
        vertex.label.cex = text.size)
    add_color_bar(edge.weights, use.pheatmap.colors = TRUE, position = "topright")
}
dev.off()

all.rownames <- unique(unlist(lapply(all.thresh, rownames)))
all.colnames <- unique(unlist(lapply(all.thresh, colnames)))

all.thresh.mat <- matrix(NA, nrow = length(all.rownames), ncol = length(all.colnames))
rownames(all.thresh.mat) <- all.rownames
colnames(all.thresh.mat) <- all.colnames
for(tx in 1:length(all.thresh)){
    all.thresh.mat[rownames(all.thresh[[tx]]), colnames(all.thresh[[tx]])] <- all.thresh[[tx]]
}

row.order <- order(rowMeans(all.thresh.mat, na.rm = TRUE))
col.order <- order(colMeans(all.thresh.mat, na.rm = TRUE))
pheatmap(all.thresh.mat[row.order, col.order], cluster_rows = FALSE, cluster_cols = FALSE)
```

How can we summarize this result and then look for aging subtypes?
Each individual needs a score in multiple categories and multiple
brain tissues. But we also don't have longitudinal data, only 
snapshots, so do we only look in older people? Do we fit some
kind of model to serve as a reference, and then people deviate 
from that reference?

People could be scored based on their deviation from the mean
of other people their age in each grouping.

```{r ind_score}
#tissue.ind.means has mean expression values for every
#intersection in every tissue.

#tissue.ind.means

for(tx in 1:length(tissue.names)){
    tx.means <- tissue.ind.means[[tx]]
    for(bd in 1:length(tx.means)){
        all.k.means <- Reduce("rbind", tx.means[[bd]])
        rownames(all.k.means) <- names(tx.means[[bd]])[which(sapply(tx.means[[bd]], length) > 0)]
        test <- apply(all.k.means, 1, function(x) cor(x, age[colnames(all.k.means)]))
        pheatmap(all.k.means)
        plot.decomp(t(all.k.means), cols = colors.from.values(age[colnames(all.k.means)]), main = names(tissue.ind.means[[tx]])[bd])
        plot.decomp(all.k.means)
        boxplot(all.k.means[,order(colMeans(all.k.means))])
    }
}
```