## Batches Pre-Normalization {.tabset .tabset-fade .tabset-pills}


```{r pc_fun}
plot_pc_factors <- function(expr.decomp){
  for(i in 2:ncol(info)){
    cat("###", colnames(info)[i], "\n")
    pc1.var.exp <- signif(expr.decomp$var.exp[1]*100, 2)
    pc2.var.exp <- signif(expr.decomp$var.exp[2]*100, 2)
    
    layout(matrix(c(1,2), nrow = 1), widths = c(1, 0.4))
    par(mar = c(6,6,4,0))
    if(is.factor(info[,i])){
      plot(expr.decomp$u, pch = 16, xlab = paste0("PC1 (", pc1.var.exp, "%)"),
        ylab = paste0("PC2 (", pc2.var.exp, "%)"), col = as.numeric(info[,i]),
        main = colnames(info)[i])
    }else{
      age.cols = brewer.pal(9, "RdPu")
      u_age <- sort(unique(info[,i]))
      age.col <- sapply(info[,i], function(x) age.cols[which(u_age == x)])
      plot(expr.decomp$u, pch = 16, xlab = paste0("PC1 (", pc1.var.exp, "%)"),
        ylab = paste0("PC2 (", pc2.var.exp, "%)"), col = age.col, main = colnames(info)[i])
    }
    par(mar = c(6,0,4,2))
    plot.new()
    plot.window(xlim = c(0,1), ylim = c(0,1))
    if(is.factor(info[,i])){
      legend(0, 1, legend = levels(info[,i]), pch = 16, col = unique(as.numeric(info[,1])))
    }else{
      legend(0, 1, legend = u_age, pch = 16, col = age.cols[1:length(u_age)])
    }
    cat("\n\n")
  }
}
```

The following plots show the relationships between the first
two PCs of the un-normalized gene expression matrix and relevant
variables.

```{r pre_norm_pc, fig.width = 6, fig.height = 5, results = "asis"}
#pdf("~/Desktop/test.pdf", width = 6, height = 5)
plot_pc_factors(pre.norm.decomp)
#dev.off()
```

## Parsing biodomains


```{r parse_bd}
#make a list of the biodomains. Each biodomain has a subdomain
#with a list of human genes.

annotated_bd <- readRDS(file.path(biodomain.dir, "annotated_biodomains_Oct23.rds"))
u_bd <- unique(annotated_bd$Biodomain)
u_sbd <- unique(annotated_bd$Subdomain)

bd_list <- vector(mode = "list", length = length(u_bd))
names(bd_list) <- u_bd

for(bd in 1:length(u_bd)){
  bd.idx <- which(annotated_bd$Biodomain == u_bd[bd])
  u_sub.bd <- unique(annotated_bd$Subdomain[bd.idx])
  u_sub.bd[which(is.na(u_sub.bd))] <- u_bd[bd]
  sub.list <- vector(mode = "list", length = length(u_sub.bd))
  names(sub.list) <- u_sub.bd
  for(sbd in 1:length(u_sub.bd)){
    if(u_sub.bd[sbd] == u_bd[bd]){
      sbd.idx <- which(is.na(annotated_bd$Subdomain[bd.idx]))
    }else{
      sbd.idx <- which(annotated_bd$Subdomain[bd.idx] == u_sub.bd[sbd])
    }
    sbd.genes <- Reduce("union", annotated_bd$ensembl_id[bd.idx[sbd.idx]])
  }

}
```

From 1.Klotho_Initial_Data_Visualization.Rmd after biodomain
expression section:

## Cluster Enrichment {.tabset .tabset-fade .tabset-pills}

I thought it might be interesting to look at the enrichment
of the obvious clusters in the plots above, but the enrichments
are really redundant, so I'm taking out the plots. It will 
probably be more informative to look at intersections with 
GO terms and KEGG pathways.


```{r cluster_enrich, results = "asis", fig.height = 7, fig.width = 7}
stop()
for(i in 1:length(gene.enrich)){
  cat("###", names(mean.bd.vals)[i], "\n")
  plot.enrichment.group(gene.enrich[[i]], max.term.size = 3000, 
    n.terms = 30, cluster_cols = FALSE, plot.label = names(mean.bd.vals)[i])
  cat("\n\n")
}
```


## from KEGG and Biodomain Intersections


To calculate a significance threshold on the largest contributing
pathways, we need to perform permutations and calculate a null 
distribution for something. What we are interested in is the 
expression's ability to discriminate the genotypes, so we need
to calculate a statistic about that. We shuffled the mouse labels,
which breaks the correlation between genotype and expression,
but not the correlation structure of gene expression.

We need a statistic that quantifies the ability of the gene 
expression to distinguish genotypes along the first principal
component.

```{r perm_fun}
#This function permutes the output of merge_mean_mats()
#This is a matrix with pathway intersections in rows and individuals
#in columns. It shuffles the individual labels and then recalculates
#pathway means for each genotype based on the shuffled individuals.
#it returns the first two principal components of the shuffled matrices.

perm_mean_mat <- function(ind.mean.mat, n.perm = 1000){
    perm.genotype.score <- vector(mode = "list", length = n.perm)

    for(p in 1:n.perm){
        perm.order <- sample(1:ncol(ind.mean.mat)) #shuffle individuals
        
        #shuffle the individual labels on the mean gene expression
        perm.ind.mean.mat <- ind.mean.mat[,perm.order]
        colnames(perm.ind.mean.mat) <- colnames(sub.vals)

        perm.geno.mats <- lapply(geno.idx, function(x) perm.ind.mean.mat[,x])
        perm.geno.means <- sapply(perm.geno.mats, rowMeans)
        
        #plot.decomp(t(perm.geno.means), label.points = TRUE, cols = geno.cols, cex = 1.5)
        mean.perm.decomp <- plot.decomp(t(perm.geno.means), plot.results = FALSE) 
        perm.genotype.score[[p]] <- mean.perm.decomp$u
    }

    return(perm.genotype.score)
}

```

```{r perm}

true.decomp <- plot.decomp(t(sub.vals), plot.results = FALSE)

nperm = 1000
perm.geno.score <- perm_mean_mat(sub.vals, nperm)
```

The following plot shows the score for each genotype along
PC1 against the permuted result for PC1. The VS genotypes
do tend to be on the low end and the FC genotypes tend to
be on the high end, but they are still well within the 
random distributions.

```{r plot_perm_pc1}
perm1 <- t(sapply(perm.geno.score, function(x) x[,1]))
vioplot(perm1, col = geno.cols)
points(x = 1:length(geno.cols), y = mean.decomp$u[,1], pch = "*", cex = 3, col = "#c51b8a")
abline(h = 0)
```

The following plot is the same as above but for PC2

```{r plot_perm_pc2}
perm2 <- t(sapply(perm.geno.score, function(x) x[,2]))
vioplot(perm2, col = geno.cols)
points(x = 1:length(geno.cols), y = mean.decomp$u[,2], pch = "*", cex = 3, col = "#c51b8a")
abline(h = 0)
```

The following plot shows the calculated PC values for each genotype
against random distrbutions of PC1 and PC2. These are not particularly
inspiring. The VS genotypes are at the edge of their distributions, but 
still within the random point cloud.


```{r both_pc, fig.width = 9, fig.height = 6}
par(mfrow = c(2,3))
for(i in 1:ncol(perm1)){
    xmin <- min(c(perm1[,i], mean.decomp$u[i,1]))
    xmax <- max(c(perm1[,i], mean.decomp$u[i,1]))
    ymin <- min(c(perm2[,i], mean.decomp$u[i,2]))
    ymax <- max(c(perm2[,i], mean.decomp$u[i,2]))

    plot(perm1[,i], perm2[,i], xlim = c(xmin, xmax), ylim = c(ymin, ymax),
        xlab = "PC1", ylab = "PC2", main = names(geno.cols)[i],)
    points(mean.decomp$u[i,1], mean.decomp$u[i,2], col = geno.cols[i], pch = 16, cex = 3)
}

```

Maybe the individual scores don't matter? Maybe it's about how well
they are separated or clustered? The VS genotype should group together,
the FC genotypes should group together, and they should be separated 
from each other.

Or maybe we should test individual pathways, not overall?

```{r sandbox, eval = FALSE}
path.name <- "Ribosome"
path.name <- "Tryptophan metabolism"


path.idx <- which(names(mean.bd.k.vals) == path.name)
path.decomp <- plot_mean_bd_k(mean.bd.k.vals[[path.idx]], covar.table)
path.sub <- merge_mean_mats(mean.bd.k.vals[path.idx])

geno.path <- lapply(geno.idx, function(x) path.sub[,x])
geno.path.means <- sapply(geno.path, rowMeans)
mean.path.decomp <- plot.decomp(t(geno.path.means), cols = geno.cols, label.points = TRUE)

perm.path <- perm_mean_mat(path.sub, 1000)

perm1 <- t(sapply(perm.path, function(x) x[,1]))
vioplot(perm1, col = geno.cols, main = path.name)
points(x = 1:length(geno.cols), y = mean.path.decomp$u[,1], pch = "*", cex = 3, col = "#c51b8a")
abline(h = 0)


```


## ANOVA

What if we just do an ANOVA on each intersection.
We've done ANOVA's on individual genes, this will
look at the means of biodomain-kegg intersections.


```{r anova}

test_pathways <- function(nested.gene.list, expr.mat){

    all.aov <- max.mean.dist <- vector(mode = "list", length = length(nested.gene.list))
    names(all.aov) <- names(max.mean.dist) <- names(nested.gene.list)
    
    for(bd in 1:length(nested.gene.list)){
        bd.aov <- max.bd.dist <- vector(mode = "list", length = length(nested.gene.list[[bd]]))
        names(bd.aov) <- names(max.bd.dist) <- names(nested.gene.list[[bd]])
        if(length(nested.gene.list[[bd]]) == 0){next()}
        for(k in 1:length(nested.gene.list[[bd]])){
            gene.ids <- nested.gene.list[[bd]][[k]]
            common.genes <- intersect(gene.ids, colnames(expr.mat))
            gene.mat <- expr.mat[,common.genes]
            geno.mats <- lapply(geno.idx, function(x) t(gene.mat)[,x])
            geno.means <- sapply(geno.mats, rowMeans)
            #boxplot(geno.means)
            bd.aov[[k]] <- aov.by.matrix(geno.means)
            max.bd.dist[[k]] <- as.vector(dist(t(geno.means)))
        }
        all.aov[[bd]] <- bd.aov
        max.mean.dist[[bd]] <- max.bd.dist
    }
    return(list("aov" = all.aov, "dist" = max.mean.dist))
}

get_nested_stats <- function(anova.list, dist.list){
    bd.p <- bd.dist <- matrix(NA, ncol = length(kegg.bd.list), nrow = length(kegg.list))
    colnames(bd.p) <- colnames(bd.dist) <- names(kegg.bd.list)
    rownames(bd.p) <- rownames(bd.dist) <- names(kegg.list)
    for(bd in 1:length(kegg.bd.list)){
        sub.p <- sapply(anova.list[[bd]], function(x) summary(x)[[1]][,"Pr(>F)"][1])
        sub.dist <- sapply(dist.list[[bd]], max)
        if(length(sub.p) > 0){
            bd.p[names(sub.p),bd] <- sub.p
            bd.dist[names(sub.dist),bd] <- sub.dist
        }
    }

    result <- list("p" = bd.p, "dist" = bd.dist)
}


bd.kegg.test <- test_pathways(kegg.bd.list, adj_expr)
bd.kegg.aov <- bd.kegg.test$aov
bd.kegg.dist <- bd.kegg.test$dist

all.dist <- unlist(bd.kegg.dist)
hist(all.dist)

nperm = 10
perm.p <- perm.dist <- vector(mode = "list", length = nperm)
for(p in 1:nperm){
    print(p)
    rnd.sample <- sample(1:nrow(adj_expr))
    sampled.expr <- adj_expr[rnd.sample,]
    rownames(sampled.expr) <- rownames(adj_expr)
    bd.kegg.perm <- test_pathways(kegg.bd.list, sampled.expr)
    
    bd.kegg.aov.perm <- bd.kegg.perm$aov
    bd.kegg.dist.perm <- bd.kegg.perm$dist

    aov.stats <- get_nested_stats(anova.list = bd.kegg.aov.perm, dist.list =  bd.kegg.dist.perm)

    perm.p[[p]] <- aov.stats$p
    perm.dist[[p]] <- aov.stats$dist
}

qqunif.plot(as.vector(unlist(perm.p)))
hist(as.vector(unlist(perm.dist)))

pv <- as.vector(bd.p)
pv <- pv[which(!is.na(pv))]
adj.p <- p.adjust(pv, "fdr")
#length(which(adj.p < 0.05))
p.order <- order(pv, decreasing = FALSE)
test <- cbind("p" = pv[p.order], "adjusted_p" = adj.p[p.order])
thresh.p <- max(test[which(test[,"adjusted_p"] <= 0.00011),"p"])

sig.mat <- bd.p
sig.mat[which(sig.mat > thresh.p)] <- NA
row.vals <- which(apply(sig.mat, 1, function(x) !all(is.na(x))))
col.vals <- which(apply(sig.mat, 2, function(x) !all(is.na(x))))

sig.p <- -log10(bd.p[row.vals, col.vals])
sig.dist <- bd.dist[row.vals, col.vals]


pdf("~/Desktop/sig_intersections.pdf", width = 7, height = 17)
pheatmap(sig.p, cluster_rows = FALSE, cluster_cols = FALSE)
pheatmap(sig.dist, cluster_rows = FALSE, cluster_cols = FALSE)
dev.off()
 
pass.thresh <- which(!is.na(sig.p), arr.ind = TRUE)
diff.names <- apply(pass.thresh, 1, function(x) paste(rownames(sig.p)[x[1]], colnames(sig.p)[x[2]], sep = " - "))
diff.idx <- match(diff.names, rownames(sub.vals))

sub.means <- sapply(geno.idx, function(x) rowMeans(sub.vals[,x]))
pheatmap(sub.means[diff.idx,])

plot_mean_bd_k(sub.vals[diff.idx,], covar.table)
```

Is is interesting that the WT are not intermediate, but span 
the full range? That is that Klotho nudges you to one side of
normal or another side, but doesn't push you outside of the 
normal range.

The WT only seem intermediate because their mean is zero.



## AD gene enrichment stats

Originally, I had a wanted to show that the DEG in the Klotho
mice were enriched for "AD genes." I did this by getting a 
list of AD genes and doing a hypergeometric test to see if
this list were enriched in the DEG. The list I used was the
Agora list, which Greg said might be too controversial to use.

He also suggested that a Venn diagram might not be the most
effective way of showing the enrichment. He suggested instead
going back to the AD-gene heatmap I had at the very beginning
where I had a short-list of genes shown to be genetically 
associated with AD risk. I showed that a number of them, including
APOE and APP were differentially expressed across the Klotho
genotypes. 

So I'm taking out the enrichment stastics and Venn diagrams 
and replacing them with a heatmap. Greg suggested that we can
even show individual stats in the heatmap by having a colum for
each mouse with delimiters, rather than showing means for each
group.

The following plot shows how the differentially expressed
genes and AD-associated genes interact with each other.

```{r venn_fig}
sig.list <- list("All genes" = rownames(effect.fdr), "DEG" = rownames(mean.sig.expr),
  "AD Genes" = ad.ensembl)
#plotVenn(Vlist = sig.list)

circle.cols <- c("all" = "black", "deg" = "#bf812d", "ad" = "#756bb1")

ind.num <- sapply(sig.list, length)
deg.only <- length(setdiff(sig.list$DEG, sig.list$"AD Genes"))
ad.only <- length(setdiff(sig.list$"AD Genes", sig.list$DEG))
deg.and.ad <- length(intersect(sig.list$"AD Genes", sig.list$DEG))

par(mar = c(0,0,0,0))
plot.new()
plot.window(xlim = c(0, 1), ylim = c(0,1))

inner.center <- 0.45
inner.label.y <- 0.62
deg.center <- 0.4 
ad.center <- 0.7
label.cex <- 1.2

#all.circle <- get_circle(radius = 0.45, center_x = 0.5, center_y = inner.center)
#points(all.circle$x, all.circle$y, col = circle.cols["all"], type = "l", lwd = 3)
draw.rectangle(min.x = 0.12, max.x = 0.92, min.y = 0.15, max.y = 0.85, border = circle.cols["all"], lwd = 3)
deg.circle <- get_circle(radius = 0.25, center_x = 0.4, center_y = inner.center)
points(deg.circle$x, deg.circle$y, col = circle.cols["deg"], type = "l", lwd = 3)
ad.circle <- get_circle(radius = 0.15, center_x = 0.7, center_y = inner.center)
points(ad.circle$x, ad.circle$y, col = circle.cols["ad"], type = "l", lwd = 3)


text(x = 0.52, y = 0.81, labels = paste0("All (", ind.num["All genes"], ")"), 
  col = circle.cols["all"], font = 2, cex = label.cex)
text(x = deg.center, y = inner.label.y, labels = paste0("DEG (", ind.num["DEG"], ")"), 
  col = circle.cols["deg"], font = 2, cex = label.cex)
text(x = ad.center+0.035, y = inner.label.y, labels = paste0("AD (", ind.num["AD Genes"], ")"), 
  col = circle.cols["ad"], font = 2, cex = label.cex)

text(x = deg.center, y = inner.center, labels = deg.only, font = 2, cex = label.cex, col = circle.cols["deg"])
text(x = ad.center+0.02, y = inner.center, labels = ad.only, font = 2, cex = label.cex, col = circle.cols["ad"])
text(x = 0.6, y = inner.center, labels = deg.and.ad, font = 2, cex = label.cex, col = mix_colors(circle.cols["deg"], circle.cols["ad"]))
```


## AD gene overview

Are AD genes enriched among the differentially expressed genes?
Are there more AD genes among the DEG than we would expect by 
chance? We used the hypergeometric distribution to answer this.

We have an urn with $N$ total marbles. Some of these marbles
($K$ red marbles) are differentially expressed. $N-K$ green
marbles are not differentially expressed. If we draw $n$ AD 
genes, what is the probability that exaclty $k$ of them are 
red (i.e. differentially expressed)?

. We want to know
the probability of drawing $k$    

```{r hyper, fig.width = 5, fig.height = 5}
#quartz(width = 5, height = 5)
plot.new()
par(mar = c(1,1,1,1))
plot.window(xlim = c(0, 1), ylim = c(0, 1))
draw.rectangle(0.1, 0.9, 0.1, 0.9)

gene.circle <- get_circle(radius = 0.25, center_x = 0.4, center_y = 0.5)
lower.idx <- which(gene.circle$y <= 0.5)
upper.idx <- which(gene.circle$y >= 0.5)
plot.poly.xy(gene.circle$x[upper.idx], gene.circle$y[upper.idx], 
  gene.circle$x[lower.idx], gene.circle$y[lower.idx], border = NA,
  col = rgb(190/256,174/256,212/256))
points(gene.circle$x, gene.circle$y, type = "l")

ad.circle <- get_circle(radius = 0.25, center_x = 0.6, center_y = 0.5)
lower.idx <- which(ad.circle$y <= 0.5)
upper.idx <- which(ad.circle$y >= 0.5)
plot.poly.xy(ad.circle$x[upper.idx], ad.circle$y[upper.idx], 
  ad.circle$x[lower.idx], ad.circle$y[lower.idx], border = NA,
  col = rgb(127/256,201/256,127/256, alpha = 0.5))
points(ad.circle$x, ad.circle$y, type = "l")

#add labels
text(x = 0.72, y = 0.12, labels = "All genes = N", adj = 0)
text(x = 0.25, y = 0.76, labels = "DEG = K")
text(x = 0.83, y = 0.76, labels = "AD genes = n", adj = 1)
text(x = 0.5, y = 0.5, labels = "AD and DEG = k")
mtext("What is the probability that in our draw of AD genes\n
  we get k that are differentially expressed?",
  side = 1, line = -2.5, padj = 0)
```

```{r ad_enrich}
#from: https://montilab.github.io/BS831/articles/docs/HyperEnrichment.html

all.gene.id <- gsub(".value", "", rownames(full.model.result$p_values))
all.gene.name <- gene.info[match(all.gene.id, gene.info[,"ensembl_gene_id"]), "external_gene_name"]
sig.deg <- rownames(sig.expr)
#ad.orthos

p.of.draw <- phyper(q = length(ad.sig)-1, #number of red marbles in our draw (number of AD genes that are DEG)
  m = length(sig.deg), #number of red marbles (DEG) in the urn
  n = length(all.gene.id) - length(sig.deg), #number of green marbles (not DEG)
  k = length(ad.orthos), #number of drawn marbles (AD genes)
  lower.tail = FALSE #compute P(X > 1 overlap), hence the -1 above.
)
```

The calculated probability of there `r length(ad.sig)` differentially
expressed AD genes without enrichment in the set of all DEGs is 
`r signif(p.of.draw, 2)`, which we correct to `r threshold_p(p.of.draw)`.



## Testing individual factors 
we test everything in one model now, and these variables are confusing me.

We also tested the effects of each factor after adjusting
for the other factors to get effect sizes for each.
This allows us to make volcano plots for each factor.

```{r diff_fun}

test_one_var <- function(expr.mat, covar.mat, covar.names, test.var){

  var.test.file <- file.path(results.dir, paste0(expression.type, "_Effects_of_", test.var, ".RDS"))

  adj.covar <- setdiff(covar.names, test.var)
  no.var.covar <- dummy_covar(covar.mat[,adj.covar,drop=FALSE])
  adj.expr.file <- file.path(processed.data.dir, paste0("Adjusted_", expression.type, "_Expression_for_", test.var, "_test.RDS"))
  if(file.exists(adj.expr.file)){
    adj.expr <- readRDS(adj.expr.file)
  }else{
    adj.expr <- adjust(t(expr.mat), no.var.covar)
    saveRDS(adj.expr, adj.expr.file)
  }

  #match covariates to adjusted expression matrix
  common.ind <- intersect(rownames(adj.expr), rownames(covar.mat))
  expr.idx <- match(common.ind, rownames(adj.expr))
  covar.idx <- match(common.ind, rownames(covar.mat))

  diff.test <- apply(adj.expr[expr.idx,], 2, function(x) lm(x~covar.mat[covar.idx,test.var]))
  test.coef <- t(sapply(diff.test, function(x) coef(x)))
  test.p <- t(sapply(diff.test, function(x) summary(x)$coefficients[,"Pr(>|t|)"]))
  colnames(test.coef) <- colnames(test.p) <- gsub("covar.mat[, test.var]", "", colnames(test.coef), fixed = TRUE)
  test.f <- lapply(diff.test, function(x) summary(x)$fstatistic)
  overall.p <- sapply(test.f, function(x) pf(x[1],x[2],x[3],lower.tail=F))
  test.r2 <- sapply(diff.test, function(x) summary(x)$adj.r.squared)
  result <- list("coef" = test.coef, "coef_p_values" = test.p, "overall_p" = overall.p, "R2" = test.r2)
  saveRDS(result, var.test.file)

  return(result)
}
```

```{r run_diff_test}
all.tests <- vector(mode = "list", length = (length(use.covar)-1))
names(all.tests)  <- use.covar[2:length(use.covar)]

for(i in 2:length(use.covar)){
  result.idx <- which(names(all.tests) == use.covar[i])
  all.tests[[result.idx]] <- test_one_var(expr.mat = scaled.expr, 
    covar.mat = plotting.df, covar.names = use.covar, 
    test.var = use.covar[i])
}
```

### Volcano Plots {.tabset .tabset-fade .tabset-pills}

The following volcano plots show the relationship between
coefficients and p values for each of the variables tested.
Here, we used a p value of `r fdr.thresh` to mark genes that
might be differentially expressed.

Again, we removed the sex-specific genes (those with a 
-log10(p) > `r sex.effect.threshold`).

```{r volcano, results = "asis", fig.width = 8, fig.height = 4}

#pdf("~/Desktop/volcano.pdf", width = 8, height = 4)
#par(mfrow = c(3,2))
de.list <- vector(mode = "list", length = length(all.tests))
names(de.list)  <- names(all.tests)

par(mfrow = c(1,2))
for(i in 1:length(all.tests)){
  cat("####", names(all.tests)[i], "\n")
  p <- all.tests[[i]]$overall_p
  coef.p <- all.tests[[i]]$coef_p_values
  coef.table <- all.tests[[i]]$coef[,2:ncol(all.tests[[i]]$coef),drop=FALSE]

  if(names(all.tests)[i] == "sex_ge"){
    #take out the genes with the largest sex effect. These are all X and Y genes.
    big.sex.effect <- which(-log10(p) > sex.effect.threshold)
    big.sex.genes <- gene.info[match(gsub(".value", "", names(big.sex.effect)), gene.info[,"ensembl_gene_id"]),]
    p <- p[-big.sex.effect]
    coef.p <- coef.p[-big.sex.effect,,drop=FALSE]
    coef.table <- coef.table[-big.sex.effect,,drop=FALSE]
  }

  adj.p <- p.adjust(p, "fdr")
  sig.idx <- which(adj.p <= fdr.thresh)
  de.list[[i]] <- sig.idx
  #length(sig.idx)

  qqunif.plot(p,  
    plot.label = paste("qq plot for", names(all.tests)[i], 
    "\n", length(sig.idx), "significant genes"))

  sig.col <- rep("black", length(adj.p))
  sig.col[sig.idx] <- "red"
  coef.names <- gsub("info[, test.var]", "", colnames(coef.table), fixed = TRUE)
  colnames(coef.table) <- coef.names  

  #par(mfrow = c(2,2))
  for(cv in 1:ncol(coef.table)){  
    plot(coef.table[,cv], -log10(coef.p[,(cv+1)]), xlab = "Coefficient",
      ylab = "-log10(p value)", 
      main = paste(names(all.tests)[i], 
        gsub("covar.mat[covar.idx, test.var]", "", colnames(coef.table)[cv], fixed = TRUE), sep = ", "), 
      col = sig.col, pch = 16, cex = 0.7)
  }

  if(expression.type == "gene"){
    gene.id <- rownames(scaled.expr)[sig.idx]
    transcript.id <- sub.id[match(gene.id, sub.id[,1]),2]
  }
  if(expression.type == "transcript"){
    transcript.id <- rownames(scaled.expr)[sig.idx]
    gene.id <- sub.id[match(transcript.id, sub.id[,1]),2]
  }
  gene.idx <- match(gene.id, gene.info[, "ensembl_gene_id"])
  sig.gene.info <- gene.info[gene.idx,c("external_gene_name", "chromosome_name", "start_position")]
  sig.table <- cbind("geneID" = gene.id, "transcriptID" = transcript.id,
    sig.gene.info,
    "coef" = coef.table[sig.idx,],
    "p" = as.numeric(p[sig.idx]), "p_adj" = as.numeric(adj.p[sig.idx]))

  ordered.table <- sig.table[order(sig.table[,"p_adj"], decreasing = FALSE),,drop=FALSE]
  sig.gene.file <- file.path(results.dir, paste0("Significant_", expression.type, "_", 
    names(all.tests)[i], ".txt"))
  write.table(ordered.table, sig.gene.file, sep = "\t", quote = FALSE, row.names = FALSE)
  cat("\n\n")
}
#dev.off()

#num.de <- sapply(de.list, length)
#write.table(num.de, here("Results", "for_paper", paste0("num_de_", basename(results.dir), ".txt")),
#  quote = FALSE, sep = "\t", col.names = FALSE)
```

## plotting and clustering differentially expressed AD genes

This looks basically identical to the plots of the overall
DEG, so it is not worth doing over in the ms.

These expression patterns fall into two distinct groups:

```{r ad_decomp}
ad.decomp <- plot.decomp(ad.sig.mat, plot.results = FALSE)
ad.cl <- pam(ad.decomp$u, k = 2)$clustering
plot.decomp(ad.sig.mat, cols = ad.cl)
```

```{r ad_sig_gene_figure}
if(subgroup[[1]] == 12){
  #make sure FC up is the first cluster to match the general DEG results
  alt.cl <- ad.cl
  alt.cl[which(ad.cl == 1)] <- 2
  alt.cl[which(ad.cl == 2)] <- 1
  ad.cl <- alt.cl 

  geno.order <- c(3,2,4,5)
  cl.order <- rev(order(ad.cl))
  ad.plot.mat <- ad.sig.mat[cl.order,geno.order]
  scaled.mat <- t(apply(ad.plot.mat, 1, scale))
  dimnames(scaled.mat) <- dimnames(ad.plot.mat)

  png(here("Results", "for_paper", "ad_clusters.png"), width = 7, height = 3,  units = "in", res = 300)
  layout.mat <- matrix(c(1,2), nrow = 1)
  layout(layout.mat, widths = c(1, 0.2))
  par(mar = c(2,4,4,1), xpd = NA)
  imageWithText(t(scaled.mat), 
    show.text = FALSE, use.pheatmap.colors = TRUE,
    col.names = NULL, row.text.shift = 0.07, row.text.adj = 0.5)
  plot.dim <- par("usr")
  plot.height <- plot.dim[4] - plot.dim[3]
  plot.width <- plot.dim[2] - plot.dim[1]
  c1.mid <- mean(which(ad.cl[col.order] == 1))
  c2.mid <- mean(which(ad.cl[col.order] == 2))
  label.y <- (plot.dim[4]+(plot.height*0.08))
  text(x = c1.mid, y = label.y, labels = "Neuron and Synapse")
  text(x = c2.mid, y = label.y, labels = "Mitochondria and Ribosome")

  #add colored labels for the genotypes
  y.vals <- 1:4
  y.shift <- plot.height*0.06
  label.xmin <- plot.dim[1] - (plot.width*0.09)
  label.xmax <- plot.dim[1] + (plot.width*0.03)
  for(i in 1:length(y.vals)){
    draw.rectangle(label.xmin, label.xmax, (y.vals[i] - y.shift), (y.vals[i] + y.shift),
      border = geno.cols[match(rev(colnames(ad.plot.mat))[i], names(geno.cols))], 
      lwd = 3)
  }

  #add scale bar
  par(mar = c(2,4,4,1))
  imageWithTextColorbar(scaled.mat, use.pheatmap.colors = TRUE, cex = 1, bar.lwd = 3)
  dev.off()
}
```


The mean expression of the genes in each group
is shown below. The rows are scaled to accentuate
the cross-genotype differences.

```{r cl1}
cl1.idx <- which(ad.cl == 1)
pheatmap(ad.sig.mat[cl1.idx,ordered.geno], cluster_cols = FALSE, 
  scale = "row", show_rownames = FALSE)
```

The enrichments for this cluster are shown below.

```{r cl1_enrich}
cl1.enrich <- gost(rownames(ad.sig.mat)[cl1.idx], organism = "mmusculus",
  source = c("GO", "KEGG", "CORUM", "REACTOME", "HP"))
plot.enrichment(cl1.enrich, num.terms = 25, plot.label = "Cluster 1 enrichment")
```

The mean expression of the genes in the second group is
shown below.

```{r cl2}
cl2.idx <- which(ad.cl == 2)
pheatmap(ad.sig.mat[cl2.idx,ordered.geno], cluster_cols = FALSE, 
  scale = "row", show_rownames = FALSE)
```

These enrichments match the clusters seen in the differentially 
expressed genes overall, and suggest that the differential expression
caused by the variants is related to AD pathology.

```{r cl2_enrich}
cl2.enrich <- gost(rownames(ad.sig.mat)[cl2.idx], organism = "mmusculus",
  source = c("GO", "KEGG", "CORUM", "REACTOME", "HP"))
plot.enrichment(cl2.enrich, num.terms = 25, plot.label = "Cluster 2 enrichment")
```


```{r wgcna, eval = FALSE}
#I checked WGCNA. Unfortunately, it didn't really yield much.

library(WGCNA)

enableWGCNAThreads()

powers = c(c(1:10), seq(from = 12, to=20, by=2))
sft = pickSoftThreshold(adj_expr, powerVector = powers, verbose = 5)
use.power <- sft$powerEstimate
if(is.na(use.power)){use.power = 6}
use.power

#signed if no negative values
#unsigned if negative values are included
type = "unsigned"
#type = "signed"
net = blockwiseModules(adj_expr, power = use.power, 
    TOMType = type, minModuleSize = 10, 
    reassignThreshold = 0, mergeCutHeight = 0.25, 
    numericLabels = TRUE, pamRespectsDendro = FALSE, 
    saveTOMs = FALSE,verbose = 3)

mergedColors = labels2colors(net$colors)
modules <- net$colors
num.wgcna.modules <- length(unique(modules))

u_mods <- sort(unique(modules))
mod.size <- sapply(u_mods, function(x) length(which(modules == x)))
barplot(mod.size)

mod.genes <- lapply(u_mods, function(x) names(modules)[which(modules == x)])

mean.mod.vals <- non_nested_mean_vals(non_nested_gene_list = mod.genes, 
    expr.mat = adj_expr, group.representation = group.representation)

#Mean module expression is not strongly associated with genotype
mod.result <- plot_mean_bd_k(mean.bd.k.mat = mean.mod.vals, 
    covar.mat = covar.table, plot.results = TRUE, order.by = "PC",
    path.stats = "R2", stat.x = 0.8, stat.y = 0.2, label.margin = 20)

high.idx <- which(mod.result$row.r2 > 0.2)
high.enrich <- gost(mod.genes[high.idx], organism = "mmusculus", sources = "GO")
plot.enrichment(high.enrich, num.terms = 30, max.term.size = 3000)

pdf(paste0("~/Desktop/WGCNA_", type, ".pdf"), width = 10, height = 7)
test <- lapply(1:length(mod.genes), function(x) plot_individual_genes(gene.list = mod.genes[[x]], 
  plot.label = paste0("Module", x), return.full.result = TRUE))
dev.off()

mod.r2 <- sapply(test, function(x) x$overall.r2)
hist(mod.r2)
mod.p <- sapply(test, function(x) x$vioplot_results$model.p)
bonf.thresh <- 0.05/length(mod.p)
hist(-log10(mod.p));abline(v = -log10(bonf.thresh))
sig.mod <- which(mod.p <= bonf.thresh)

high.mod.eig <- which(mod.r2 >= 0.2)
high.mod.enrich <- lapply(mod.genes[high.mod.eig], function(x) gost(x, organism = "mmusculus", sources = "GO"))
names(high.mod.enrich)  <- paste0("Module", high.mod.eig)
plot.enrichment.group(high.mod.enrich, max.term.size = 3000, transformation = "sqrt")


enrich.file <- file.path(results.dir, paste0("WGCNA_Module_Enrichment_", type, ".RDS"))
if(!file.exists(enrich.file)){
    mod.enrich <- lapply(u_mods, function(x) gost(names(which(modules == x)), 
        organism = "mmusculus", sources = c("GO", "KEGG", "REACTOME", "CORUM", "HP")))
    saveRDS(mod.enrich, enrich.file)
}else{
    mod.enrich <- readRDS(enrich.file)
}

pdf("~/Desktop/enrichment.pdf", width = 9, height = 40)
plot.enrichment.group(mod.enrich, max.term.size = 3000, transformation = "sqrt")
dev.off()

wgcna.list <- lapply(u_mods, function(x) names(which(modules == x)))
gene.r2 <- vector(mode = "list", length = length(wgcna.list))
mod.r2 <- rep(NA, length(wgcna.list))

for(i in 1:length(u_mods)){
    print(i)
    ind.genes <- lapply(wgcna.list[[i]], function(x) x)
    wgcna.vals <- non_nested_mean_vals(ind.genes, adj_expr, 
        group.representation = group.representation)
    rownames(wgcna.vals) <- unlist(ind.genes)
    wgcna.result <- plot_mean_bd_k(wgcna.vals, covar.mat = covar.table, plot.results = FALSE)
    gene.r2[[i]] <- wgcna.result$row.r2
    mod.r2[i] <- wgcna.result$overall.r2
}

barplot(sort(mod.r2))

mod.order <- order(mod.r2, decreasing = TRUE)
i = 1
ind.result <- plot_individual_genes(gene.list = names(which(modules == u_mods[mod.order[i]])))

#quartz()
plot.enrichment(mod.enrich[[u_mods[high.pred[i]]]], max.term.size = 3000)


hist(unlist(gene.r2))

sorted.genes <- sort(unlist(gene.r2), decreasing = TRUE)
plot(sorted.genes)


#This gives basically the same results to other analyses.
library(fgsea)
#gsea.enrich <- fgsea::fgsea(kegg.list, sorted.genes, scoreType = "pos")
#gsea.enrich <- fgsea::fgsea(bd.list, sorted.genes, scoreType = "pos")
gsea.enrich <- fgsea::fgsea(unlist(kegg.bd.list, recursive = FALSE), sorted.genes, scoreType = "pos")


norm.es <- as.numeric(as.matrix(gsea.enrich[,"NES"]))
names(norm.es) <- gsea.enrich$pathway

par(mar = c(4,24, 2,2))
barplot(tail(sort(norm.es), 19), las = 2, horiz = TRUE)
```

## Latent Pathway Identification Analysis (LPIA)

Here we use a method developed in Pham et al. 2011
called Latent Pathway Identification Analyais (LPIA).
This is a network-based method that combines expression
data with GO-KEGG intersections into a network in which
an edge exists if there are genes in the GO-KEGG intersection.
The edge is weighted by the product of two terms: (1) the 
Jaccard index of the two terms (in terms of genes), and 
(2) the differential expression of the genes in the overlapping
genes. In Pam et al. they were comparing two groups, so they 
used the median differential expression. We could compare the
FC and VS carriers directly to do this, or we can use a different
term that is based on our linear model with ordered groups.

They then project this bipartite network onto the the KEGG 
terms. The nodes in this network are linked if they share 
at least one biological process. The edge weights are larger
if the two KEGG terms relate more GO terms. We would maybe
use the Biodomain projection? We can try both. 

They use eigenvector centrality to calculate centrality for 
each pathway.

```{r lpia}
#sub.vals
intersect.result = sub.val.result
intersect.jaccard = kegg.bd.jaccard

#generate a network between all Biodomains and KEGG pathways
#that combines the Jaccard index of the pathways and the 
#degree to which the intersection is differentially expressed.

lpia_adj <- function(intersect.result, intersect.jaccard){
    net.adj <- matrix(0, nrow = nrow(intersect.jaccard), ncol = ncol(intersect.jaccard))
    dimnames(net.adj) <- dimnames(intersect.jaccard)
    split.name <- strsplit(names(intersect.result$row.r2), " : ")
    row.kegg <- sapply(split.name, function(x) x[1])
    row.bd <- sapply(split.name, function(x) x[2])

    for(bd in 1:nrow(intersect.jaccard)){
        bd.idx <- which(row.bd == rownames(net.adj)[bd])
        kegg.idx <- match(row.kegg[bd.idx], colnames(net.adj))
        intersect.r2 <- intersect.result$row.r2[bd.idx]
        #edge weight is the R2 times the jaccard indes
        eweight <- intersect.r2*intersect.jaccard[bd,kegg.idx]
        net.adj[bd, kegg.idx] <- eweight
    }
    return(net.adj)
}

#create a biadjacency matrix in which the edge weights 
#are a product of the KEGG-Biodomain Jaccard index and the
#R2 of the effect of genotoype on the intersection (we might
#want to change this to better capture effect size).
lpia_mat <- lpia_adj(sub.val.result, kegg.bd.jaccard)
lpia_mat[which(lpia_mat < 0)] <- 0 #zero out any negative numbers
lpia_net <- graph_from_biadjacency_matrix(lpia_mat)
lpia_net <- delete_vertices(lpia_net, which(degree(lpia_net) == 0))
#plot(lpia_net)

lpia_proj <- bipartite_projection(lpia_net)


bd.proj <- lpia_proj[[1]]
bd.proj <- delete_vertices(bd.proj, which(degree(bd.proj) == 0))

kegg.proj <- lpia_proj[[2]]
kegg.proj <- delete_vertices(kegg.proj, which(degree(kegg.proj) == 0))


#plot(bd.proj)
#plot(kegg.proj)
bd_cent <- eigen_centrality(bd.proj)
kegg_cent <- eigen_centrality(kegg.proj)

barplot(bd_cent$vector, las = 2)
par(mar = c(4,16,4,2))
barplot(tail(sort(kegg_cent$vector), 10), las = 2, horiz = TRUE)
```

## From smaller intersection analysis

## KEGG pathways with altered intersections {.tabset .tabset-fade .tabset-pills}

We then looked at all the KEGG intersections with the 
APP Processing biodomain. Because we've already calculated
all the significant correlations in the APP biodomain, we
can just look for where these fall in the overlapping 
KEGG pathways.

```{r all_intersections}
app.list <- kegg.bd.list$"APP Metabolism"
app.kegg.sig <- vector(mode = "list", length = length(app.list))
names(app.kegg.sig) <- names(app.list)
for(i in 2:length(app.list)){
    overlapping.genes <- app.list[[i]]
    overlapping.names <- mouse.genes[match(overlapping.genes, mouse.genes[,"ensembl_gene_id"]),"external_gene_name"]
    common.names <- intersect(overlapping.names, colnames(corr_result[[1]]))
    sub.net <- corr_result[[1]][common.names, common.names]
    sub.p <- corr_result[[2]][common.names, common.names]
    sig.sub <- which(sub.p <= 0.001, arr.ind = TRUE)
    if(nrow(sig.sub) > 0){
        sig.sub.names <- cbind(colnames(sub.net)[sig.sub[,1]], colnames(sub.net)[sig.sub[,2]])
        sig.sub.result <- cbind(sig.sub.names, sub.net[sig.sub], sub.p[sig.sub])
        colnames(sig.sub.result) <- c("gene1", "gene2", "effect", "p")
        app.kegg.sig[[i]] <- sig.sub.result
    }
}
has.vals <- which(sapply(app.kegg.sig, length) > 0)
sig.app.k.list <- app.kegg.sig[has.vals]
```

There were `r length(app.kegg.sig)` KEGG pathways with genes overlapping
the APP processing biodomain. Of these `r length(sig.app.k.list)` had
gene-gene correlations that were significantly dependent on Klotho genotype.

The bar plot below shows how many gene pairs in each term had altered
correlations across genotypes.

```{r num_alt}
num.pairs <- sapply(sig.app.k.list, nrow)
par(mar = c(4,22,4,2))
barplot_with_num(sort(num.pairs), horiz = TRUE, las = 2, xlab = "Number of Gene Pairs",
    main = "KEGG pathways in APP Metabolism\nwith altered gene-gene correlations")
```

```{r plot_kegg, results = "asis", fig.width = 10, fig.height = 10}
#pdf("~/Desktop/kegg_overlaps.pdf", width = 10, height = 10)
for(i in 1:length(sig.app.k.list)){
    cat("###", names(sig.app.k.list)[i], "\n")
    plot_net(dir.net, layout.mat = net.layout, plot.label = names(sig.app.k.list)[i])
    common.genes <- unique(as.vector(sig.app.k.list[[i]][,1:2]))
    common.idx <- match(common.genes, rownames(net.layout))
    scaled.idx <- net.layout
    scaled.idx[,1] <- scale.between.vals(net.layout[,1], -1, 1)
    scaled.idx[,2] <- scale.between.vals(net.layout[,2], -1, 1)
    points(scaled.idx[common.idx,1], scaled.idx[common.idx,2], cex = 2, col = "red", pch = 16)
    cat("\n\n")
}
#dev.off()
```


## Latent Pathway Identification Analysis (LPIA)

Here we use a method developed in Pham et al. 2011
called Latent Pathway Identification Analyais (LPIA).
This is a network-based method that combines expression
data with GO-KEGG intersections into a network in which
an edge exists if there are genes in the GO-KEGG intersection.
The edge is weighted by the product of two terms: (1) the 
Jaccard index of the two terms (in terms of genes), and 
(2) the differential expression of the genes in the overlapping
genes. In Pam et al. they were comparing two groups, so they 
used the median differential expression. We could compare the
FC and VS carriers directly to do this, or we can use a different
term that is based on our linear model with ordered groups.

They then project this bipartite network onto the the KEGG 
terms. The nodes in this network are linked if they share 
at least one biological process. The edge weights are larger
if the two KEGG terms relate more GO terms. We would maybe
use the Biodomain projection? We can try both. 

They use eigenvector centrality to calculate centrality for 
each pathway.

Below, we generate lists of KEGG pathways and Biodomain 
intersections and calculate Jaccard indices for all 
intersections. 

We take up the rest of the analysis in 1.KEGG_and_Biodomains.Rmd
for historical reasons.

```{r intersect}

intersect_bd_kegg <- function(bd.list, kegg.list){

  intersect.list <- vector(mode = "list", length = length(bd.list))
  names(intersect.list) <- names(bd.list)

  for(bd in 1:length(bd.list)){
    k.intersections <- vector(mode = "list", length = length(kegg.list))
    names(k.intersections) <- names(kegg.list)  
    
    for(k in 1:length(kegg.list)){
      k.intersections[[k]] <- intersect(bd.list[[bd]], kegg.list[[k]])
    }

    has.genes <- which(sapply(k.intersections, length) >= min.term.size)
    intersect.list[[bd]] <- k.intersections[has.genes]
  }
  return(intersect.list)
}

bd_kegg_jaccard <- function(bd.list, kegg.list){

  jacc.mat <- matrix(NA, nrow = length(bd.list), ncol = length(kegg.list))
  rownames(jacc.mat) <- names(bd.list)
  colnames(jacc.mat) <- names(kegg.list)

  for(bd in 1:length(bd.list)){
    k.intersections <- vector(mode = "list", length = length(kegg.list))
    names(k.intersections) <- names(kegg.list)  
    
    for(k in 1:length(kegg.list)){
      jacc.mat[bd,k] <- jaccard.ind(bd.list[[bd]], kegg.list[[k]])
    }

  }
  return(jacc.mat)
}


#save grouped form, so we can do the boxplot thingy that Greg Cary does.
mouse.bd.k.intersections <- intersect_bd_kegg(bd.list = bd.for.gsea, kegg.list = mouse.path.gene.ensembl)
mouse.bd.k.jacc <- bd_kegg_jaccard(bd.list = bd.for.gsea, kegg.list = mouse.path.gene.ensembl)
saveRDS(mouse.bd.k.intersections, file.path(general.processed.data.dir, 
  paste0("Mouse_KEGG_Intersections_for_GSEA_", expression.type, ".RDS")))
saveRDS(mouse.bd.k.jacc, file.path(general.processed.data.dir, 
  paste0("Mouse_KEGG_Intersection_Jaccard_", expression.type, ".RDS")))

human.bd.k.intersections <- intersect_bd_kegg(bd_genes, hum.path.gene.ensembl)
human.bd.k.jacc <- bd_kegg_jaccard(bd_genes, hum.path.gene.ensembl)
saveRDS(human.bd.k.intersections, file.path(general.processed.data.dir, 
  paste0("Human_KEGG_Intersections_for_GSEA_", expression.type, ".RDS")))
saveRDS(human.bd.k.jacc, file.path(general.processed.data.dir, 
  paste0("Human_KEGG_Intersections_Jaccard_", expression.type, ".RDS")))
```


```{r intersect_vals}

mean.bd.k.file <- file.path(results.dir, paste0("KEGG_BD_mean_expression_", group.representation, ".RDS"))

if(!file.exists(mean.bd.k.file)){
  mean.bd.k.vals <- vector(mode = "list", length = length(mouse.path.gene.ensembl))
  names(mean.bd.k.vals) <- names(mouse.path.gene.ensembl)

  #for each kegg pathway
  for(k in 1:length(mouse.path.gene.ensembl)){
    kegg.path <- names(mouse.path.gene.ensembl)[k]
    has.intersection <- which(sapply(lapply(mouse.bd.k.intersections, function(x) which(names(x) == kegg.path)), length) > 0)
    if(length(has.intersection) == 0){next()}

    intersect.mean.vals <- vector(mode = "list", length = length(has.intersection))
    names(intersect.mean.vals) <- names(has.intersection)

    all.val.mat <- matrix(NA, nrow = length(mouse.bd.genes), ncol = 5)
    rownames(all.val.mat) <- names(mouse.bd.genes)
    colnames(all.val.mat) <- genotypes[1:5]

    #pull out all gene values for intersections with biodomains
    for(bd in 1:length(has.intersection)){
      kegg.idx <- sapply(mouse.bd.k.intersections[has.intersection], function(x) which(names(x) == kegg.path))
      
      intersect.ids <- mouse.bd.k.intersections[[has.intersection[bd]]][[kegg.idx[bd]]]
      bd.k.vals <- lapply(intersect.ids, 
        function(x) plot_tx_with_genotype(expr.mat = scaled.expr, covar.table = covar.mat, 
        tx_name = x, tx_label = NA, ylab = "Expression (A.U.)", plot.results = FALSE)[[1]])

      has.vals <- which(sapply(bd.k.vals, length) > 0)
      k.mean <- t(sapply(bd.k.vals[has.vals], function(x) sapply(x[[1]], mean)))
      #pheatmap(k.mean)
      rownames(k.mean) <- intersect.ids[has.vals]
      intersect.mean.vals[[bd]] <- k.mean
    }
    has.vals <- which(sapply(intersect.mean.vals, length) > 0)
    if(group.representation == "means"){
      mean.mat <- t(sapply(intersect.mean.vals[has.vals], colMeans))
    }
    if(group.representation == "eigenegene"){
      mean.mat <- t(sapply(intersect.mean.vals[has.vals], function(x) plot.decomp(x, plot.results = FALSE)$v[,1]))
    }
    all.val.mat[rownames(mean.mat), colnames(mean.mat)] <- mean.mat[rownames(mean.mat), colnames(mean.mat)]
    mean.bd.k.vals[[k]] <- all.val.mat
    #pheatmap(mean.bd.k.vals[[k]], main = names(path.gene.ensembl)[k], cluster_cols = FALSE, cluster_rows = FALSE)
  }
  saveRDS(mean.bd.k.vals, mean.bd.k.file)
}else{
  mean.bd.k.vals <- readRDS(mean.bd.k.file)
}

has.vals <- which(sapply(mean.bd.k.vals, length) > 0)
min.val <- min(unlist(mean.bd.k.vals[has.vals]), na.rm = TRUE)
max.val <- max(unlist(mean.bd.k.vals[has.vals]), na.rm = TRUE)
#custom.dist <- unlist(mean.bd.k.vals[has.vals])
custom.dist <- rnorm(1000, 0, 0.2)
#hist(custom.dist)

kegg.plot.dir <- file.path(results.dir, "KEGG_plots")
if(!file.exists(kegg.plot.dir)){dir.create(kegg.plot.dir)}


pdf(file.path(kegg.plot.dir, "kegg_bd_intersections.pdf"), width = 7, height = 7)
hist(unlist(mean.bd.k.vals), xlab = "Mean Value", main = "Mean Biodomain-Kegg Intersection Expression")

for(i in 1:length(mean.bd.k.vals)){
  if(length(mean.bd.k.vals[[i]]) > 0){
    par(mar = c(4,12,2,2))
    neg.vals <- length(which(mean.bd.k.vals[[i]] < 0))
    pos.vals <- length(which(mean.bd.k.vals[[i]] > 0))
    if(neg.vals > 0 && pos.vals > 0){
      col.scale = c("blue", "brown")
      grad.dir = "ends"
      split.at.vals = TRUE
    }
    if(neg.vals > 0 && pos.vals == 0){
      col.scale = "blue"
      grad.dir = "low"
      split.at.vals = TRUE
    }
    if(neg.vals == 0 && pos.vals > 0){
      col.scale = "brown"
      grad.dir = "high"
      split.at.vals = TRUE
    }

  imageWithText(mat = mean.bd.k.vals[[i]], main = names(mean.bd.k.vals)[i], 
      col.scale = col.scale, grad.dir = grad.dir, col.text.rotation = 0, 
      col.text.adj = 0.5, col.text.shift = 0.06, row.text.shift = 0.15, 
      main.shift = 0.05, color.fun = "linear", split.at.vals = split.at.vals,
      light.dark = "f", global.color.scale = TRUE, global.min = min.val, 
      global.max = max.val)

  }
}
dev.off()

global.min <- min(unlist(mean.bd.k.vals), na.rm = TRUE)
global.max <- max(unlist(mean.bd.k.vals), na.rm = TRUE)
#plot all results for each biodomain on one page
for(bd in 1:length(bd_genes)){
  bd.mat <- t(sapply(mean.bd.k.vals, function(x) if(length(x) > 0){x[bd,]}else{rep(NA, 5)}))
  has.vals <- which(!is.na(rowMeans(bd.mat)))

  #plot.decomp(t(bd.mat[has.vals,]), label.points = TRUE, main = names(bd_genes)[bd])
  #bd.decomp <- plot.decomp(bd.mat[has.vals,], main = names(bd_genes)[bd])
  #plot.decomp(bd.mat[has.vals,], main = names(bd_genes)[bd], label.points = TRUE)

  if(length(has.vals) > 0){
    png(file.path(kegg.plot.dir, paste0("bd_kegg_intersections_", names(bd_genes)[bd], ".png")), 
      width = 7, height = max(c(length(has.vals)/7, 7)), units = "in", res = 200)
    row.order <- hclust(dist(bd.mat[has.vals,]))$order
    par(mar = c(4,16, 2, 2))
    test <- imageWithText(mat = bd.mat[has.vals[row.order],], main = names(bd_genes)[bd], 
        col.scale = c("blue", "brown"), grad.dir = "ends", col.text.rotation = 0, 
        col.text.adj = 0.5, col.text.shift = 0.01, row.text.shift = 0.15, 
        main.shift = 0.01, color.fun = "linear", split.at.vals = TRUE,
        color.dist = NULL, light.dark = "f", global.color.scale = TRUE,
        global.min = global.min, global.max = global.max)
    dev.off()
  }
}
```


We put all the results together to see if there were any combinations
that stood out overall. The plot below shows that the expression across
all KEGG pathway and biodomain intersections separates the genotypes
nicely. Along the first principal component, the VS genotypes are negative,
the WT mice are at 0, and the FC genotypes are positive.

```{r all_kegg, fig.width = 5, fig.height = 5}

big.mat <- Reduce("rbind", mean.bd.k.vals)
group.names <- unlist(lapply(1:length(mean.bd.k.vals), function(x) if(length(mean.bd.k.vals[[x]] > 0)){paste(names(mean.bd.k.vals)[x], rownames(mean.bd.k.vals[[x]]), sep = "--")}))
rownames(big.mat) <- group.names
#pdf("~/Desktop/big.mat.pdf", height = 20, width = 10)
#pheatmap(big.mat, cluster_rows = FALSE, cluster_cols = FALSE)
#dev.off()
not.na <- which(!is.na(rowSums(big.mat)))

sub.mat <- big.mat[not.na,]
process.decomp <- plot.decomp(t(sub.mat), label.points = TRUE, xlim = c(-1, 1), 
  cols = geno.cols[match(names(geno.cols), colnames(sub.mat))], cex = 1.5)
process.vals <- process.decomp$v

pc.order <- order(process.decomp$u[,1])
barplot(process.decomp$u[pc.order,1], 
  col = geno.cols[match(names(geno.cols), colnames(sub.mat))][pc.order], 
  names = colnames(sub.mat)[pc.order], ylab = "PC1")
abline(h = 0)
#process.decomp <- plot.decomp(big.mat[has.vals,], label.points = TRUE)
#plot(process.decomp$v, type = "n")
#text(process.decomp$v[,1], process.decomp$v[,2], labels = colnames(big.mat))
#process.vals <- process.decomp$u

```

The processes don't have any particular clustering as
shown below. But there are some pathways that have 
large negative and positive values in the first PC.

```{r process_pc, fig.width = 8, fig.height = 7}
plot(process.decomp$v, xlab = "PC1", ylab = "PC2", pch = 16)
large.neg <- which(process.decomp$v[,1] < -0.04)
large.pos <- which(process.decomp$v[,1] > 0.05)
text(process.decomp$v[large.pos,1], process.decomp$v[large.pos,2], labels = rownames(sub.mat)[large.pos], cex = 0.5)
text(process.decomp$v[large.neg,1], process.decomp$v[large.neg,2], labels = rownames(sub.mat)[large.neg], cex = 0.5)
```

The plots below show the mean expression for each of these
processes with large PC values. 

```{r processes, fig.width = 10, fig.height = 5}
pc.order <- order(process.vals[,1], decreasing = TRUE)
most.neg <- tail(pc.order, 20)
most.pos <- head(pc.order, 20)
#most.pos <- pc.order[1620:1630] #check middle

#test.mat <- process.decomp$v[most.neg,1,drop=FALSE]
#rownames(test.mat) <-  rownames(sub.mat)[most.neg]
#test.mat[order(test.mat[,1]),]

par(mar = c(4, 24, 2, 2))
layout(matrix(c(1,2), ncol = 2), widths = c(1, 0.5))
imageWithText(sub.mat[rev(most.neg),], col.scale = c("blue", "brown"), split.at.vals = TRUE,
  grad.dir = "ends", col.text.rotation = 0, col.text.adj = 0.5, col.text.shift = 0.05, row.text.cex = 0.7,
  row.text.shift = 0.15, main = "Large Negative PC pathways")
par(mar = c(3.8, 0, 1.8, 2))
barplot(process.vals[most.neg,1], horiz = TRUE)
mtext("PC1", side = 1, line = 2.2)


#test.mat <- process.decomp$v[most.pos,1,drop=FALSE]
#rownames(test.mat) <-  rownames(sub.mat)[most.pos]
#test.mat[order(test.mat[,1]),]

par(mar = c(4, 18, 2, 0))
#row.order <- hclust(dist(sub.mat[most.pos,]))$order
layout(matrix(c(1,2), ncol = 2), widths = c(1, 0.5))
imageWithText(sub.mat[most.pos,], col.scale = c("blue", "brown"), split.at.vals = TRUE,
  grad.dir = "ends", col.text.rotation = 0, col.text.adj = 0.5, col.text.shift = 0.05, row.text.cex = 0.7,
  row.text.shift = 0.15, main = "Large Positive PC pathways")
par(mar = c(3.8, 0, 1.8, 2))
barplot(process.vals[rev(most.pos),1], horiz = TRUE)
mtext("PC1", side = 1, line = 2.2)
```

Can we build a biodomain network that shows us where the 
biggest differences are? The biodomains were designed such

```{r bipartite}
#This is quite derived, but let's see what happens when we collect
#the loadings on PC1 for each kegg-biodomain pair.

umat <- matrix(0, nrow = length(mouse.path.gene.ensembl), ncol = length(bd_genes))
rownames(umat) <- names(mouse.path.gene.ensembl)
colnames(umat) <- names(bd_genes)
split.names <- strsplit(rownames(sub.mat), "--")
split.kegg <- sapply(split.names, function(x) x[1])
split.bd <- sapply(split.names, function(x) x[2])
for(i in 1:nrow(process.vals)){
  umat[split.kegg[i], split.bd[i]] <- process.vals[i,1]
}

#pheatmap(umat)
no.vals <- which(rowSums(umat) == 0)
has.vals <- which(rowSums(umat) != 0)
row.order <- order(rowSums(umat[has.vals,]))
col.order <- order(colSums(umat[has.vals,]))

pdf(file.path(kegg.plot.dir, "KEGG_Umat.pdf"), width = 9, height = 40)
pheatmap(umat[has.vals[row.order],col.order], cluster_cols = TRUE, cluster_rows = TRUE)
dev.off()
```

The following image shows just the rows with strong loadings
for a given biodomain-KEGG intersection.

```{r strong_vals, fig.width = 8, fig.height = 9}

#filter to rows with the largest values
#hist(umat)
lower.thresh <- -0.04
upper.thresh <- 0.04
high.val <- which(apply(umat, 1, function(x) any(x > upper.thresh)))
low.val <- which(apply(umat, 1, function(x) any(x < lower.thresh)))
strong.vals <- union(high.val, low.val)

pheatmap(umat[strong.vals,])

#pdf("~/Desktop/sub_mat.pdf", width = 8, height = 9)
#pheatmap(umat[strong.vals,])
#dev.off()

#net <- graph_from_incidence_matrix(abs(umat[strong.vals,]))
#plot(net)

#pdf("~/Desktop/bip_plot.pdf", width = 10, height = 10)
#plot_bipartite(abs(umat[strong.vals,]))
#dev.off()
```


# excised from 2.Splicing_Variation.Rmd


## Literature search

We searched for the top differentially spliced genes in the literature
along with the terms "splicing", and "brain". The number of abstracts
retrieved for each gene is shown below. 

Synj1 (Synaptojanin1) has the largest number of associated abstracts among 
the top 10 genes. Synj1 is the major synaptic polyphosphoinositide phosphatase, 
and regulates uncoating and remobilization of synaptic vesicles during 
clathrin-mediated endocytosis (pmid: 10535736). It interacts with BIN1 
(pmid: 9195986), and APOE4-induced upregulation of SYNJ1
leads to phospholipid homeostatic dysregulation in the brain, which is associated
with cognitive decline in AD (pmid: 24052255). Downregulation of Synj1 restores 
phospholipid homeostasis and reduces levels of phosphorylated tau in vitro 
(pmid: 28231468). SYNJ1 is also associated with Parkinson's Disease 
(pmid: 23804563)

A recent paper looks at effects of SNPs, including splicing effects,
in this gene and two other endocysis-related genes (pmid: 31592138).
They predicted a number of deleterious SNPs that affected splicing
of SYNJ1. 

```{r pubmed}
library(easyPubMed)
pubmed.dir <- file.path(results.dir, "PubMed")
if(!file.exists(pubmed.dir)){dir.create(pubmed.dir, recursive = TRUE)}

search.term <- lapply(sig.gene.names, function(x) paste(x, "AND splicing AND brain"))
all.pubmed <- lapply(search.term, get_pubmed_ids)
num.papers <- unlist(lapply(all.pubmed, function(x) x$Count))
with.papers <- which(num.papers != 0)
paper.counts <- cbind(sig.gene.names[with.papers], num.papers[with.papers])
number.order <- order(as.numeric(paper.counts[,2]))

barplot_with_num(as.numeric(paper.counts[number.order,2]), 
    names = paper.counts[number.order,1], las = 2, ylab = "Number of Abstracts")

#has.data <- all.pubmed[with.papers]
#all.abstracts <- lapply(has.data, function(x) fetch_pubmed_data(x, format = "abstract"))
#for(i in 1:length(all.abstracts)){
#    write.table(all.abstracts[[i]], file.path(pubmed.dir, 
#        paste0("PubMed_Query_", tissue.names[tx], "_Up_", 
#        rownames(up.table)[with.papers[i]], ".txt")), 
#        sep = "\t", quote = FALSE, row.names = FALSE, col.names = FALSE)
#}

```

## Exon Stats

Here we investigate some basic questions about transcript
length and relative expression in `r ref.geno` and `r alt.geno` 
animals.

```{r sandbox, eval = FALSE}
gene.name = "Synj1"
gene.id <- tx.info[match(gene.name, tx.info[,"external_gene_name"]), "ensembl_gene_id"]
plotProportions(d, gene_id = gene.id, group_variable = "genotype", 
  plot_type = "boxplot1", group_colors = geno.cols[c(1,3)])

gene.idx <- which(padj[,"geneID"] == gene.id)
tx.id <- padj[gene.idx,"txID"]
tx.idx <- match(tx.id, tx.info[,"ensembl_transcript_id"])
tx.len <- tx.info[tx.idx,"transcript_length"]
names(tx.len) <- tx.info[tx.idx,"ensembl_transcript_id"]
canonical.idx <- which(tx.info[tx.idx,"transcript_is_canonical"] == 1)
len.order <- order(tx.len)
bar.col <- rep("gray", length = length(tx.len))
bar.col[canonical.idx] <- "red"

par(mar = c(4, 12, 4, 2))
barplot(tx.len[len.order], col = bar.col[len.order], las = 2, horiz = TRUE,
  main = paste(gene.name, "transcript lengths"), xlab = "bp")

#is the canonical transcript differentially abundant?
sig.tx.idx <- which(padj[,"transcript"] <= alpha)
sig.tx.id <- padj[sig.tx.idx,"txID"]
common.tx <- intersect(sig.tx.id, tx.info[,"ensembl_transcript_id"])
is.canonical <- sapply(common.tx, function(x) 
  tx.info[match(x, tx.info[,"ensembl_transcript_id"]), "transcript_is_canonical"])
barplot_with_num(c("canonical" = length(which(is.canonical == 1)), "non-canonical" = length(which(is.na(is.canonical)))))

#what are the lengths of the differentially expressed transcripts?
tx.len <- sapply(common.tx, function(x) 
  tx.info[match(x, tx.info[,"ensembl_transcript_id"]), "transcript_length"])
#hist(tx.len)


common.idx <- match(common.tx, props[,"feature_id"])
allele.prop <- lapply(allele.names, function(x) rowMeans(as.matrix(props[common.idx,x])))
plot(allele.prop$FC, tx.len)
plot(allele.prop$VS, tx.len)
pt.col <- rep("black", length(tx.len))
pt.col[which(is.canonical == 1)] <- "red"
plot(allele.prop$FC, allele.prop$VS, col = pt.col, pch = 16)
abline(0,1)

allele.diff <- allele.prop$VS - allele.prop$FC
vs.higher <- which(allele.diff > 0)
vs.lower <- which(allele.diff < 0)

#VS has reduced expression of longer transcripts? May indicate
#intron retention in FC?
boxplot(list("up in VS" = tx.len[vs.higher], "down in VS" = tx.len[vs.lower]))

canonical.idx <- which(is.canonical == 1)
non.canonical.idx <- which(is.na(is.canonical))

plot(allele.diff, tx.len, col = pt.col, pch = 16, xlab = "Proportion difference",
  ylab = "Transcript length")
abline(v = 0)
text(x = -0.3, y = 25000, labels = "Lower in VS")
text(x = 0.2, y = 25000, labels = "Higher in VS")

#vs has slightly lower expression of canonical transcripts?
#and slightly higher of non-canonical? I don't think they
#are really different.
canon.list <- list("FC" = list("canonical" = allele.prop$FC[canonical.idx], 
  "non-canonical" = allele.prop$FC[non.canonical.idx]),
  "VS" = list("canonical" = allele.prop$VS[canonical.idx], 
  "non-canonical" = allele.prop$VS[non.canonical.idx]))
plot.grouped.boxes(canon.list, group.cols = geno.cols[c(1,3)],
  ylab = "proportion")

len.list <- list("FC" = list("canonical" = allele.prop$FC[canonical.idx], 
  "non-canonical" = allele.prop$FC[non.canonical.idx]),
  "VS" = list("canonical" = allele.prop$VS[canonical.idx], 
  "non-canonical" = allele.prop$VS[non.canonical.idx]))

```


```{r transcript_comparison}
#make a big comparison table that we can use to ask different quesitons 
#of the significant transcripts

sig.tx.idx <- which(padj[,"transcript"] <= alpha)
sig.tx.id <- padj[sig.tx.idx,"txID"]
tx.idx <- match(sig.tx.id, tx.info[,"ensembl_transcript_id"])
tx.len <- tx.info[tx.idx,"transcript_length"]
tx.canon <- tx.info[tx.idx,"transcript_is_canonical"]

alleles <- c(ref.geno, alt.geno)
props <- proportions(d)
allele.names <- lapply(alleles, 
  function(x) rownames(plotting.df)[which(plotting.df[,"ordered_geno"] == x)])
names(allele.names) <- alleles
tx.prop.idx <- match(sig.tx.id, props[,"feature_id"])
allele.prop <- sapply(allele.names, function(x) rowMeans(as.matrix(props[tx.prop.idx,x])))

#if the difference is greater than 0, the alternative allele is higher
allele.diff <- allele.prop[,alt.geno] - allele.prop[,ref.geno]
alt.higher <- which(allele.diff > 0)
alt.lower <- which(allele.diff < 0)

comp.table <- cbind("txLen" = tx.len, 
  "canonical" = tx.canon, allele.prop, "difference" = allele.diff)
rownames(comp.table) <- sig.tx.id
```

Are the transcripts that are upregulated in `r alt.geno`
longer or shorter than transcripts upregulated in `r alt.geno`?

The transcripts that are up in VS may be slightly shorter 
than those that are down in VS. Maybe this indicates more
intron retention in the FC animals?

```{r tx_len, fig.width = 5, fig.height = 5}
len.list <- list(log10(comp.table[alt.higher,"txLen"]),
  log10(comp.table[alt.lower, "txLen"]))
test <- aov.by.list(len.list)
boxplot(len.list, names = paste(c("Higher in", "Lower in"), alt.geno),
  ylab = "log10 Transcript length", main = paste("p =", signif(test$"Pr(>F)"[1], 2)))
```

Are the transcripts that are up in VS more likely to be
the canonical transcripts?



```{r canonical, fig.width = 4, fig.height = 3}
canonical.table <- matrix(NA, nrow = 2, ncol = 2)
colnames(canonical.table) <- c(c("canonical", "non-canonical"))
rownames(canonical.table) <- paste(c("Higher in", "Lower in"), alt.geno)
canonical.table[1, "canonical"] <- 
  length(intersect(alt.higher, which(comp.table[,"canonical"] == 1)))
canonical.table[1, "non-canonical"] <- 
  length(intersect(alt.higher, which(is.na(comp.table[,"canonical"]))))

canonical.table[2, "canonical"] <- 
  length(intersect(alt.lower, which(comp.table[,"canonical"] == 1)))
canonical.table[2, "non-canonical"] <- 
  length(intersect(alt.lower, which(is.na(comp.table[,"canonical"]))))

test <- fisher.test(canonical.table)
pheatmap(canonical.table, display_numbers = TRUE, cluster_rows = FALSE, 
  cluster_cols = FALSE, main = paste("p =", signif(test$p.value, 2)))
```

The plot below shows the expected counts for these categories.

```{r null, fig.width = 4, fig.height = 3}
#permute to see what the expected numbers are
nperm = 100
null.counts <- vector(mode = "list", length = nperm)
for(p in 1:nperm){
  sample.table <- comp.table
  sample.diff <- sample(comp.table[,"difference"])
  sample.higher <- which(sample.diff > 0)
  sample.lower <- which(sample.diff < 0)

  sample.counts <- matrix(NA, nrow = 2, ncol = 2)
  colnames(sample.counts) <- c(c("canonical", "non-canonical"))
  rownames(sample.counts) <- paste(c("Higher in", "Lower in"), alt.geno)
  sample.counts[1, "canonical"] <- 
    length(intersect(sample.higher, which(comp.table[,"canonical"] == 1)))
  sample.counts[1, "non-canonical"] <- 
    length(intersect(sample.higher, which(is.na(comp.table[,"canonical"]))))

  sample.counts[2, "canonical"] <- 
    length(intersect(sample.lower, which(comp.table[,"canonical"] == 1)))
  sample.counts[2, "non-canonical"] <- 
    length(intersect(sample.lower, which(is.na(comp.table[,"canonical"]))))
  null.counts[[p]] <- sample.counts
}

null.avg <- Reduce("+", null.counts)/nperm
pheatmap(null.avg, display_numbers = TRUE, cluster_rows = FALSE, cluster_cols = FALSE)
```

The plot below shows the direct comparison. I think 
there isn't a real difference in the expression of 
canonical transcripts. There are just so many transcripts
that the p value is significant.

If anything, the expression of non-canonical transcripts
is slightly higher in the VS group. 

```{r count_comp, fig.height = 5, fig.width = 5}
mat.idx <- which(canonical.table > 0, arr.ind = TRUE)
pt.names <- apply(mat.idx, 1, 
  function(x) paste(rownames(canonical.table)[x[1]], colnames(canonical.table)[x[2]]))
plot(log(canonical.table), log(null.avg), xlim = c(5.5, 7.5), ylim = c(6, 7), pch = 16)
abline(0, 1)
text(log(canonical.table), log(null.avg), labels = pt.names)
```

## from 1a.KEGG_and_Biodomains

work about network localzation and more visualizations of enrichments
They weren't particularly enlightening

What about localization in the network? How spread out 
are these nodes? Does that matter?

I selected 5000 random samples of genes using the number of 
genes in the biodomains. I got a distribution of node sizes 
and average node distances for each group. There are a couple 
domains in Proteostasis and Structural Stabilization that have
lower distances relative to their size than the random distribution.
They are in Ribosome and COVID (read virus response?) KEGG pathways.

So Ribosome genes and COVID (virus response genes?) are closer
to each other in this network than we expect by chance. Not 
particularly interesting...

```{r node_clustering, eval = FALSE}

get_list_dist <- function(net, gene.list, plot.results = TRUE){
    edges <- as_edgelist(net) #we need to remove the weights from the network
    flat.net <- graph_from_edgelist(edges, directed = FALSE)
    node.dist <- distances(flat.net)

    gene.name <- lapply(gene.list, function(x) mouse.genes[match(x, mouse.genes[,"ensembl_gene_id"]), "external_gene_name"])
    common.name <- lapply(gene.name, function(x) intersect(V(flat.net)$name, x))
    common.dist <- lapply(common.name, function(x) node.dist[x,x])
    avg.dist <- sapply(common.dist, mean)
    
    if(plot.results){
        par(mfrow = c(1,2), mar = c(4,12,4,2))
        barplot(sort(avg.dist, decreasing = TRUE), las = 2, horiz = TRUE)

        par(mar = c(4,4,4,2))
        n.nodes <- sapply(common.name, length)
        plot(n.nodes, avg.dist)
        #boxplot(lapply(bd.dist, function(x) x[upper.tri(x, diag = FALSE)]), las = 2)
    }
    invisible(common.dist)
}

bd.list.len <- sapply(bd.list, length)
rnd_sample <- lapply(1:5000, function(x) sample(unique(unlist(bd.list)), sample(bd.list.len, 1))) #pull some random samples from the biodomainlist
rnd_cl <- get_list_dist(large.comp, rnd_sample, plot.results = FALSE)
rnd.dist <- sapply(rnd_cl, mean)
rnd.nodes <- sapply(rnd_cl, function(x) if(class(x)[1] == "matrix"){nrow(x)}else{0})
#plot(rnd.nodes, rnd.dist) #the randomly sampled cluster sizes and distances in the large component
#plot.hexbin.as.plot(rnd.nodes, rnd.dist)

pdf("~/Desktop/cl_test.pdf", width = 5, height = 5)
for(bd in 1:length(kegg.bd.list)){
    if(length(kegg.bd.list[[bd]]) > 0){
        test <- get_list_dist(large.comp, kegg.bd.list[[bd]], plot.results = FALSE)
        test.dist <- sapply(test, mean)
        test.nodes <- sapply(test, function(x) if(class(x)[1] == "matrix"){nrow(x)}else{0})
        plot.hexbin.as.plot(rnd.nodes, rnd.dist, main = names(kegg.bd.list)[bd], xlim = c(0, 200))
        points(test.nodes, test.dist, pch = 16)

        #check <- intersect(which(test.nodes > 25), which(test.dist < 4))
        #names(kegg.bd.list[[bd]])[check]
    }
}
dev.off()
```

 
The biodomain-KEGG intersections that were significantly
enriched in the large connected component are the following.
Purple vertices show biodomains, green show Kegg pathways.
Edge color and width reflect the -log10(p) of the enrichment.

```{r enrich_int, fig.width = 7, fig.height = 7}
enrich.p.mat <- matrix(NA, nrow = length(bd.list), ncol = length(kegg.list))
rownames(enrich.p.mat) <- names(bd.list)
colnames(enrich.p.mat) <- names(kegg.list)
for(bd in 1:length(kegg.bd.list)){
    bd.name <- names(kegg.bd.list)[bd]
    if(length(kegg.bd.list[[bd]]) == 0){next()}
    for(k in 1:length(kegg.bd.list[[bd]])){
        kegg.name <- names(kegg.bd.list[[bd]])[k]
        test.set <- kegg.bd.list[[bd]][[k]]
        sub.enrich <- gene_set_enrich(test.set, large.id, test.id)
        enrich.p.mat[bd.name, kegg.name] <- sub.enrich
    }
}
sig.enrich <- which(enrich.p.mat <= 1e-5, arr.ind = TRUE)
int.enrich <- cbind(rownames(enrich.p.mat)[sig.enrich[,1]], colnames(enrich.p.mat)[sig.enrich[,2]])
test <- graph_from_edgelist(int.enrich, directed = FALSE)
ecol <- colors.from.values(-log10(enrich.p.mat[sig.enrich]), col.scale = "blue", grad.dir = "high")
vcol <- rep("#7FC97F", vcount(test)) #kegg color
vcol[which(V(test)$name %in% names(bd.list))] <- "#beaed4" #biodomain color

par(mar = c(0,8,0,6))
plot(test, vertex.color = vcol, edge.color = ecol, edge.width = -log10(enrich.p.mat[sig.enrich])/2)

#imageWithTextColorbar(matrix(-log10(enrich.p.mat[sig.enrich]), ncol = 1), col.scale = "blue", 
#    grad.dir = "high", cex = 1, bar.lwd = 2)
```


### Enrichments Localized {.tabset .tabset-fade .tabset-pills}

The following plots show where these intersections lie in the 
network.

```{r enrich_net, fig.width = 20, fig.height = 20, results = "asis"}
#pdf("~/Desktop/enriched_intersections.pdf", width = 20, height = 20)
for(i in 1:nrow(int.enrich)){
    bd.name <- int.enrich[i,1]
    bd.idx <- which(names(kegg.bd.list) == bd.name)
    kegg.name <- int.enrich[i,2]
    kegg.idx <- which(names(kegg.bd.list[[bd.idx]]) == kegg.name)
    cat("####", bd.name, "-", kegg.name, "\n")

    int.genes <- kegg.bd.list[[bd.idx]][[kegg.idx]]
    int.names <- mouse.genes[match(int.genes, mouse.genes[,"ensembl_gene_id"]),"external_gene_name"]
    common.genes <- intersect(int.names, large.name)

    plot_net(large.comp, layout.mat = comp.layout, vertex.size = 5, edge.width.factor = 5,
    highlight.nodes = common.genes, highlight.width = 0.05, highlight.height = 0.02, 
    highlight.cex = 1.5, plot.label = paste(bd.name, kegg.name, sep = " - "))
    cat("\n\n")
}
#dev.off()
```

## selecting large connected component of network for plotting

```{r lcc}

  #there is an unconnected pair that makes plotting difficult. Just plot the large connected component

  comp <- components(high.net, mode = "weak")
  lcc.id <- which.max(comp$csize)
  lcc.verts <- V(high.net)[comp$membership == lcc.id]
  lcc.net <- induced_subgraph(high.net, lcc.verts)

  #suppress labels for nodes with degree less than 3
  high.names <- V(lcc.net)$name
  high.names[which(degree(lcc.net) < 5)] <- ""
  #put back things connected to APP metabolism
  keep.names <- c("Dopaminergic synapse", "Nicotine addiction", "APP Metabolism")
  for(i in 1:length(keep.names)){
      high.names[which(V(lcc.net)$name == keep.names[i])] <- keep.names[i]
  }
```

## Proteomics

In case we need to do a more careful thresholding for proteins with odd
measurement, I'm keeping this code.


I have done some spot checking, and it looks like the genes that 
have very high variance are cases where two replicates in one 
animal have relatively close expression on one side of the population
mean, and the third replicate has an outlier expression on the 
other side of the population mean. Or, one of the replicates 
is just way outside of the population distribution in general.

If I delve further down the list of proteins with high variance,
I start encountering proteins with multimodal distribution that
isn't explained by any of the variables we measured, like sex,
age, or genotype.

```{r spot_check}
high.var <- which(gene.var > 5, arr.ind = TRUE)
check.idx <- 1

animal.which <- final.mouse.table[high.var[check.idx,2], "animalID"]
gene.which <- rownames(high.var)[check.idx]
animal.idx <- which(final.obs.info[,"animalID"] == animal.which)
gene.idx <- which(rownames(log.gene.mat) == gene.which)
barplot(as.numeric(log.gene.mat[gene.idx,animal.idx]), names = paste("rep", 1:3), 
    ylab = "log2 protein level", main = gene.which)

#check other expression levels of this gene
par(mfrow = c(2,3))
hist(as.numeric(log.gene.mat[gene.idx,]))
plot(as.numeric(log.gene.mat[gene.idx,]), col = as.numeric(as.factor(final.obs.info[,"age_months"])), main = gene.which)

abline(h = as.numeric(log.gene.mat[gene.idx,animal.idx]))
plot(as.numeric(log.gene.mat[gene.idx,]), col = as.numeric(as.factor(final.obs.info[,"genotype"])))
plot(as.numeric(log.gene.mat[gene.idx,]), col = as.numeric(as.factor(final.obs.info[,"sex"])))
plot(as.numeric(log.gene.mat[gene.idx,]), col = as.numeric(as.factor(final.obs.info[,"animalID"])))
rep.id <- sapply(strsplit(rownames(final.obs.info), "_"), function(x) tail(x, 1))
plot(as.numeric(log.gene.mat[gene.idx,]), col = as.numeric(as.factor(rep.id)))
```



## limma

I tried limma, since that is frequently used for proteomics
analyses. It gives me basically identical results to my initial
results, so I'm going to go with the simpler analysis.

```{r set_up_data, eval = FALSE}
library(limma)
library(edgeR)


protein_data <- t(mean.gene.mat[twelve.idx,])
sample_info <- pr.info.table[twelve.idx,]

sex_twelve <- sex[twelve.idx]
geno_twelve <- genotype[twelve.idx]
design <- model.matrix(~0 + sex_twelve + geno_twelve)
fit <- lmFit(protein_data, design)

contrast_matrix <- makeContrasts(
  geno_twelve.L,
  levels = design
)

fit2 <- contrasts.fit(fit, contrast_matrix)
fit2 <- eBayes(fit2)

#qqunif.plot(fit2$p.value)
adj.p <- p.adjust(fit2$p.value, method = "fdr")
geno.effect <- coef(fit2)
sig.idx <- which(adj.p <= 0.1)
up.idx <- which(geno.effect > 0)
down.idx <- which(geno.effect < 0)

sig.up <- intersect(sig.idx, up.idx)
sig.down <- intersect(sig.idx, down.idx)
up.pr <- rownames(geno.effect)[sig.up]
down.pr <- rownames(geno.effect)[sig.down]


#the limma results are basically the same as the lm results 
plotVenn(list("limma" = up.pr, "lm" = pos.pr))
plotVenn(list("limma" = down.pr, "lm" = neg.pr))

#setdiff(pos.pr, up.pr)
#setdiff(up.pr, pos.pr)
#setdiff(neg.pr, down.pr)
#setdiff(down.pr, neg.pr)
```



## Check for sample mix-ups in proteomics

I wanted to do some QC to look for the possibility of sample 
mix-ups. I tried checking abundance of genes on the X and Y.
I've tried a few things that I'm not showing here. Basically,
there is no good way to make sure the mouse labels are correct.

```{r xy, eval = FALSE}

x.gene.table <- gene.info[which(gene.info[,"chromosome_name"] == "X"),]
y.gene.table <- gene.info[which(gene.info[,"chromosome_name"] == "Y"),]

pr.ensembl <- sapply(strsplit(obs_data$pr_info[,"ensembl_id"], ".", fixed = TRUE), function(x) x[1])
x.pr.ensembl <- pr.ensembl[which(pr.ensembl %in% x.gene.table[,"ensembl_gene_id"])]
y.pr.ensembl <- pr.ensembl[which(pr.ensembl %in% y.gene.table[,"ensembl_gene_id"])]

x.pr.pg <- obs_data$pr_info[match(x.pr.ensembl, pr.ensembl), "protein_group"]
common.x <- intersect(x.pr.pg, rownames(obs_data$mean_abundance))
decomp_with_nas(t(obs_data$mean_abundance[common.x,]), cols = as.numeric(as.factor(obs_data$mouse_info[,"sex"])))
#decomp_with_nas(t(obs_data$mean_abundance[common.x,]), cols = as.numeric(as.factor(obs_data$mouse_info[,"age_months"])))
#test <- decomp_with_nas(t(obs_data$mean_abundance[common.x,]), cols = as.numeric(as.factor(obs_data$mouse_info[,"genotype"])))
pheatmap(obs_data$mean_abundance[common.x,], show_rownames = FALSE, show_colnames = FALSE, scale = "row") #proteins in rows

y.pr.pg <- obs_data$pr_info[match(y.pr.ensembl, pr.ensembl), "protein_group"]
common.y <- intersect(y.pr.pg, rownames(obs_data$mean_abundance))
plot(obs_data$mean_abundance[common.y,]~as.factor(obs_data$mouse_info[,"sex"]),
    xlab = "", ylab = "Y chromosome protein abundance")

#The gene on Y should definitely be different in males and females
#but Tim said that the fragments might be mis-mapped to a gene on 
#X. #Ddx genes are scattered all over
ddx.info <- gene.info[grep("Ddx", gene.info[,"external_gene_name"]),]
#barplot(table(ddx.info[,"chromosome_name"]))


factor.name <- "sex"; p.thresh = 1e-5
factor.name <- "age_months"; p.thresh = 1e-20 #I'm convinced that all the age labels are correct
factor.name <- "genotype"; p.thresh = 1e-3 #this doesn't have enough of an effect
#look for any sex effects
fac.test <- apply(obs_data$mean_abundance, 1, function(x) lm(rankZ(x)~as.factor(obs_data$mouse_info[,factor.name])))
fac.p <- sapply(fac.test, function(x) coef(summary(x))[2,"Pr(>|t|)"])
qqunif.plot(fac.p);abline(h = -log10(p.thresh))
sig.idx <- which(fac.p < p.thresh)
length(sig.idx)
sig.info <- obs_data$pr_info[match(names(sig.idx), obs_data$pr_info[,"protein_group"]), ]
gene.info <- gene.info[match(sig.info[,"gene_name"], gene.info[,"external_gene_name"]),]
#cat(gene.info[,"external_gene_name"], sep = "\n")
#pheatmap(obs_data$mean_abundance[sig.idx,], scale = "row")
#could individuals in the wrong cluster be mis-labeled
decomp_with_nas(t(obs_data$mean_abundance[sig.idx,]), cols = as.factor(obs_data$mouse_info[,factor.name]))



plot(rankZ(obs_data$mean_abundance[names(sig.idx)[4],])~as.factor(obs_data$mouse_info[,"sex"]))

gene.name <- "Ddx3x"
gene.name <- "Myh10"
gene.name <- "Set"
gene.name <- "Erbb4"
gene.pg <- obs_data$pr_info[which(obs_data$pr_info[,"gene_name"] == gene.name), "protein_group"]
print(gene.pg)
par(mfrow = c(1,length(gene.pg)))
for(i in 1:length(gene.pg)){
    plot(rankZ(obs_data$mean_abundance[gene.pg[i],])~as.factor(obs_data$mouse_info[,"sex"]))
}
```


## Mediation with R package mediate
I am switching to bmediatR

```{r mediate}
#Here I am attempting to use the R package mediate
#to test whether each transcript mediates the effect
#of genotype on each protein. I'm coding the genotype
#as a dosage. Not great, but good for now.
gd <- c("FC" = 0, "WT" = 1, "VS" = 2) #allele dosage
exp.var <- old.var[colnames(test.tx), "genotype"]
Gdosage <- sapply(exp.var, function(x) gd[which(names(gd) == x)])
geno.design <- model.matrix(~exp.var-1)
rownames(geno.design) <- colnames(test.tx)

#here we test iteratively to a minimum p value of 10-5
#filtering out proteins that have higher p values for each round
min.sim <- 10
max.sim <- 10000

mediation.result.file <- file.path(results.dir, "Mediation.Results.RDS")
if(!file.exists(mediation.result.file)){

    tx.med.mat <- prop.med.mat <- matrix(NA, nrow = nrow(test.tx), ncol = length(pr.p))
    rownames(tx.med.mat) <- rownames(prop.med.mat) <- rownames(test.tx)
    colnames(tx.med.mat) <- colnames(prop.med.mat) <- names(pr.p)
    
    for(i in 1:nrow(test.tx)){
        this.sim <- min.sim
        strong.sig <- TRUE
        test.further <- sig.idx

        while(this.sim <= max.sim && strong.sig){
            print(paste("Transcript", i, ":", this.sim, "simulations"))
            one.tx <- test.tx[i,]
            out.var <- t(aligned_pr[names(test.further),colnames(test.tx),drop=FALSE])

            all.df <- lapply(1:ncol(out.var), 
                function(x) data.frame("P" = out.var[,x], "T" = one.tx, "G" = exp.var, "Gdosage" = Gdosage))    
            #make sure all individuals with missing protein data are removed. Otherwise we can't compare the models
            for(j in 1:length(all.df)){
                all.df[[j]] <- all.df[[j]][which(!is.na(all.df[[j]]$P)),]
            }
            

            med.fit <- lapply(all.df, function(x) lm(T ~ Gdosage, data = x))
            out.fit <- lapply(all.df, function(x) lm(P ~ Gdosage + T, data = x))

            med.out <- lapply(1:length(med.fit), 
                function(x) suppress_messages(mediate(med.fit[[x]], out.fit[[x]], 
                treat = "Gdosage", mediator = "T",
                boot = TRUE,sims = this.sim)))
            names(med.out) <- names(all.df)

            med.p <- sapply(med.out, function(x) if(length(x) > 1){summary(x)$n1.p}else{NA})
            tx.med.mat[i,names(med.p)] <- med.p
            prop.med <- sapply(med.out, function(x) if(length(x) > 1){summary(x)$n1}else{NA})
            prop.med.mat[i,names(prop.med)] <- prop.med

            #find proteins worth testing further
            test.further <- which(med.p == 0)
            strong.sig <- as.logical(length(test.further))
            this.sim <- this.sim*10
            cat(paste("\t", length(test.further), "proteins to test further\n"))
        }
    }

    has.vals <- which(!is.na(colSums(prop.med.mat)))
    med.result <- list("proportion_mediated" = prop.med.mat[,has.vals], "p" = tx.med.mat[,has.vals])
    saveRDS(med.result, mediation.result.file)
}else{
    med.result <- readRDS(mediation.result.file)
}

med.p.mat <- med.result$p
med.prop.mat <- med.result$proportion_mediated
rownames(med.p.mat) <- rownames(med.prop.mat) <- gene.info[match(rownames(med.p.mat), gene.info[,"ensembl_gene_id"]), "external_gene_name"]
colnames(med.p.mat) <- colnames(med.prop.mat) <- obs_data$pr_info[match(colnames(med.p.mat), obs_data$pr_info[,id.col]),gene.col]


pheatmap(-log10(med.p.mat+(1/max.sim)))
plot(med.prop.mat, -log10(med.p.mat+(1/max.sim)))

#take only the most significant proteins
med.prot.idx <- which(med.p.mat <= 1e-3, arr.ind = TRUE)
pr.tx.pairs <- med.prot.idx
pr.tx.pairs[,1] <- rownames(med.p.mat)[med.prot.idx[,1]]
pr.tx.pairs[,2] <- colnames(med.p.mat)[med.prot.idx[,2]]

edge.med <- abs(med.prop.mat[med.prot.idx])
```

```{r forward}

pr.tx.net <- plot_mediation_network(pr.tx.pairs, 
    edge.weights = edge.med, main = "Forward Mediation Network")

in_deg <- degree(pr.tx.net, mode = "in")
out_deg <- degree(pr.tx.net, mode = "out")

par(mfrow = c(1,2))
barplot(sort(in_deg[which(in_deg > 0)]), las = 2, ylab = "Number of Mediators")
barplot(sort(out_deg[which(out_deg > 0)]), las = 2, ylab = "Number of Targets")
#plot_gene("Dzank1", obs_data, data.type = "scaled", stat.y = 0.2)


pdf("~/Desktop/test_med.pdf", width = 10, height = 8)
for(i in 1:nrow(pr.tx.pairs)){
    plot_genotype_mediation(aligned_data = test_scaled, sample_data = old_data, 
        rna_data = rna_data, transcript.gene = pr.tx.pairs[i,1], peptide.gene = pr.tx.pairs[i,2], 
        autoplace.text = TRUE)
}
dev.off()

u_med_prot <- unique(pr.tx.pairs[,2])

plot_multi_gene(gene.names = u_med_prot, sample_data = old_data, 
    test_factor = "eigengene", pr_order_by = "clust", mouse_order_by = "clust",
    plot.label = "Mediated by glutamatergic transcripts", stat.y = 0.2)

if(!offline){
    enrich <- gost(u_med_prot, organism = "mmusculus")
    enrich.mat <- plot.enrichment(enrich, num.terms = 25)
}

#cat(u_med_prot, sep = "\n")
#cat(sort(u_med_prot), sep = ", ")
```


```{r reverse_mediation}
reverse.mediation.result.file <- file.path(results.dir, "Reverse_Mediation.Results.RDS")
if(!file.exists(reverse.mediation.result.file)){

    tx.med.mat <- prop.med.mat <- matrix(NA, ncol = nrow(test.tx), nrow = length(pr.p))
    colnames(tx.med.mat) <- colnames(prop.med.mat) <- rownames(test.tx)
    rownames(tx.med.mat) <- rownames(prop.med.mat) <- names(pr.p)
    
    for(i in 1:nrow(test.tx)){
        this.sim <- 10
        strong.sig <- TRUE
        test.further <- sig.idx

        while(this.sim <= max.sim && strong.sig){
            print(paste("Transcript", i, ":", this.sim, "simulations"))
            one.tx <- test.tx[i,]
            out.var <- t(aligned_pr[names(test.further),colnames(test.tx),drop=FALSE])

            all.df <- lapply(1:ncol(out.var), 
                function(x) data.frame("P" = out.var[,x], "T" = one.tx, "G" = exp.var, "Gdosage" = Gdosage))    
            #make sure all individuals with missing protein data are removed. Otherwise we can't compare the models
            for(j in 1:length(all.df)){
                all.df[[j]] <- all.df[[j]][which(!is.na(all.df[[j]]$P)),]
            }
            names(all.df) <- colnames(out.var)

            #fit the reverse mediation model
            med.fit <- lapply(all.df, function(x) lm(P ~ Gdosage, data = x))
            out.fit <- lapply(all.df, function(x) lm(T ~ Gdosage + P, data = x))

            #start with 100 sims.
            med.out <- lapply(1:length(med.fit), 
                function(x) suppress_messages(mediate(med.fit[[x]], out.fit[[x]], 
                treat = "Gdosage", mediator = "P",
                boot = TRUE,sims = this.sim)))
            names(med.out) <- names(all.df)

            med.p <- sapply(med.out, function(x) if(length(x) > 1){summary(x)$n1.p}else{NA})
            tx.med.mat[names(med.p),i] <- med.p
            prop.med <- sapply(med.out, function(x) if(length(x) > 1){summary(x)$n1}else{NA})
            prop.med.mat[names(prop.med),i] <- prop.med

            #find proteins worth testing further
            test.further <- which(med.p == 0)
            strong.sig <- as.logical(length(test.further))
            this.sim <- this.sim*10
            cat(paste("\t", length(test.further), "proteins to test further\n"))
        }
    }

    has.vals <- which(!is.na(rowSums(prop.med.mat)))
    reverse.med.result <- list("proportion_mediated" = prop.med.mat[has.vals,], "p" = tx.med.mat[has.vals,])
    saveRDS(reverse.med.result, reverse.mediation.result.file)
}else{
    reverse.med.result <- readRDS(reverse.mediation.result.file)
}


reverse.med.p.mat <- reverse.med.result$p
reverse.med.prop.mat <- reverse.med.result$proportion_mediated
colnames(reverse.med.p.mat) <- colnames(reverse.med.prop.mat) <- gene.info[match(colnames(reverse.med.p.mat), gene.info[,"ensembl_gene_id"]), "external_gene_name"]
rownames(reverse.med.p.mat) <- rownames(reverse.med.prop.mat) <- obs_data$pr_info[match(rownames(reverse.med.p.mat), obs_data$pr_info[,id.col]),gene.col]


pheatmap(-log10(reverse.med.p.mat+(1/max.sim)))
plot(reverse.med.prop.mat, -log10(reverse.med.p.mat+(1/max.sim)))

#take only the most significant proteins
reverse.med.prot.idx <- which(reverse.med.p.mat <= 1e-3, arr.ind = TRUE)
reverse.pr.tx.pairs <- reverse.med.prot.idx
reverse.pr.tx.pairs[,1] <- rownames(reverse.med.p.mat)[reverse.med.prot.idx[,1]]
reverse.pr.tx.pairs[,2] <- colnames(reverse.med.p.mat)[reverse.med.prot.idx[,2]]

reverse.edge.med <- abs(reverse.med.prop.mat[reverse.med.prot.idx])

reverse.pr.tx.net <- plot_mediation_network(reverse.pr.tx.pairs, reverse.edge.med, 
    main = "Reverse Mediation Network")


#bi.dir.net <- rbind(pr.tx.pairs, reverse.pr.tx.pairs)
#all.weights <- c(edge.med, reverse.edge.med)
#plot_mediation_network(bi.dir.net, all.weights)

pdf("~/Desktop/reverse_test_med.pdf", width = 10, height = 8)
for(i in 1:nrow(reverse.pr.tx.pairs)){
    plot_genotype_mediation(aligned_data = test_scaled, sample_data = old_data, 
        rna_data = rna_data, transcript.gene = reverse.pr.tx.pairs[i,2], 
        peptide.gene = reverse.pr.tx.pairs[i,1], autoplace.text = TRUE,
        reverse.mediation = TRUE)
}
dev.off()
```


## comparing effects - doesn't need to be done

We can also compare the factors on these levels. These don't look great.
Is is interesting that genotype has the opposite effect on transcripts
and proteins? Does that 

```{r compare_factors}


test_co_effect <- function(sample_data, rna.matrix, prot.matrix, factor.name = "age"){
    
    all.mouse.info <- get_factor_var(sample_data, data.type = "mean")
    sub.mouse.info <- all.mouse.info[match(colnames(rna.matrix), rownames(all.mouse.info)),]

    non.var <- names(which(apply(sub.mouse.info, 2, function(x) length(unique(x))) == 1))
    other.factors <- setdiff(colnames(sub.mouse.info), c(non.var, factor.name))

    fmla <- paste("y ~ ", paste(c(other.factors, factor.name), collapse = " + "))
    
    test.pr.df <- lapply(1:nrow(prot.matrix), 
        function(x) data.frame("y" = rankZ(prot.matrix[x,]), sub.mouse.info))
    test.pr <- lapply(test.pr.df, function(x) lm(fmla, data = x))
    
    test.rna.df <- lapply(1:nrow(rna.matrix), 
        function(x) data.frame("y" = rankZ(rna.matrix[x,]), sub.mouse.info))
    test.rna <- lapply(test.rna.df, function(x) lm(fmla, data = x))

    #x <- test.rna[[1]]
    coef.idx <- intersect(grep("L", names(coef(test.rna[[1]]))), grep(factor.name, names(coef(test.rna[[1]]))))
    test.pr.effect <- sapply(test.pr, function(x) coef(x)[coef.idx])
    test.rna.effect <- sapply(test.rna, function(x) coef(x)[coef.idx])
    
    plot.with.model(rankZ(test.rna.effect), rankZ(test.pr.effect), report = "cor.test", 
        xlab = "Effect on RNA", ylab = "Effect on Protein", 
        main = paste("Effect of", factor.name))
    abline(h = 0, v = 0)

    gene.names <- sapply(rownames(prot.matrix), function(x) as.character(get_prg_info(x, obs_data)[gene.col]))
    result.table <- cbind("rna_effect" = test.rna.effect, "pr_effect" = test.pr.effect)
    rownames(result.table) <- gene.names
    
    invisible(result.table)
}

```

The plots below show the effects of age, sex, and genotype
on abundances of both matrices. Weird.

```{r compare_effect, fig.width = 9, fig.height = 6, eval = FALSE}
#quartz(width = 9, height = 3)
par(mfrow = c(1,3))

unadj_aligned <- align_protein_rna(obs_data, rna_data, 
    data.type = "mean", gene.names = NULL, adjust.for = NULL, 
    align.genes = TRUE)

age.effect <- test_co_effect(obs_data, rna.matrix = unadj_aligned$aligned_transcripts, 
    prot.matrix = unadj_aligned$aligned_proteins, factor.name = "age")

sex.effect <- test_co_effect(obs_data, rna.matrix = unadj_aligned$aligned_transcripts, 
    prot.matrix = unadj_aligned$aligned_proteins, factor.name = "sex")

genotype.effect <- test_co_effect(obs_data, rna.matrix = unadj_aligned$aligned_transcripts, 
    prot.matrix = unadj_aligned$aligned_proteins, factor.name = "genotype")
```

What about just in young and old mice? Negative correlations
across the board. What does that mean?

```{r young_old_effect, fig.width = 8, fig.height = 4}

par(mfrow = c(1,2))
unadj_young <- align_protein_rna(young_data, rna_data,
    data.type = "mean", gene.names = NULL, adjust.for = NULL, align.genes = TRUE)

sex.effect <- test_co_effect(young_data, rna.matrix = unadj_young$aligned_transcripts, 
    prot.matrix = unadj_young$aligned_proteins, factor.name = "sex")

genotype.effect <- test_co_effect(young_data, rna.matrix = unadj_young$aligned_transcripts, 
    prot.matrix = unadj_young$aligned_proteins, factor.name = "genotype")
mtext("Young Mice", side = 3, line = -1.5, outer = TRUE)



par(mfrow = c(1,2))
unadj_old <- align_protein_rna(old_data, rna_data,
    data.type = "mean", gene.names = NULL, adjust.for = NULL, align.genes = TRUE)

sex.effect <- test_co_effect(obs_data, rna.matrix = unadj_old$aligned_transcripts, 
    prot.matrix = unadj_old$aligned_proteins, factor.name = "sex")

genotype.effect <- test_co_effect(obs_data, rna.matrix = unadj_old$aligned_transcripts, 
    prot.matrix = unadj_old$aligned_proteins, factor.name = "genotype")
mtext("Old Mice", side = 3, line = -1.5, outer = TRUE)

```

## pathview didn't really work with this data


```{r pathview, eval = FALSE}
#This didn't really work out. There are not many overlapping 
#genes in each cluster.

kegg.list <- readRDS(file.path(general.processed.data.dir, "Mouse_KEGG_for_GSEA.RDS"))

library(pathview)

kegg.file <- here("Data", "general", "KEGG.Mouse.RDS")
if(!file.exists(kegg.file)){
    all.kegg <- download_KEGG("mmu", "KEGG", "kegg")
    saveRDS(all.kegg, kegg.file)
}else{
    all.kegg <- readRDS(kegg.file)
}

u_path <- gsub(" - Mus musculus (house mouse)", "", all.kegg[[2]][,"to"], fixed = TRUE)

# build GSEA list
path.id <- all.kegg[[2]][,1]
path.idx <- lapply(path.id, function(x) which(all.kegg[[1]][,1] == unlist(x)[1]))
path.gene.id <- lapply(path.idx, function(x) all.kegg[[1]][x,2])
names(path.gene.id) <- u_path

path.gene.ensembl <- lapply(path.gene.id, 
    function(x) rna_data$tx_info[match(x, rna_data$tx_info[,"entrezgene_id"]), "ensembl_gene_id"])


comp_groups <- list("FC" = grep("FC", old.var[common.id,"genotype"]), "VS" = grep("VS", old.var[common.id,"genotype"]))

#int.name <- rownames(sub.vals)[pos.sig[1]]
#int.name <- "Longevity regulating pathway : Endolysosome"
#int.name <- "Alzheimer disease : Mitochondrial Metabolism"

kegg.name <- "Ribosome"
kegg.name <- "Oxidative phosphorylation"

#give each gene a "loading", which is a comparison of expression
#between the FC and VS groups. This is group2 - group1, so if you
#want a positive value to indicate up in VS, put VS second
clust.abund <- peptide_vals(unique(pr.clust[[9]][,gene.col]), old_data, data.type = "scaled", 
    adjust.for = "sex")

expr.groups <- sapply(comp_groups, function(x) rowMeans(clust.abund[,common.id[x]], na.rm = TRUE))
pr_loading <- expr.groups[,2] - expr.groups[,1]
pr_info <- get_prg_info(rownames(clust.abund), obs_data) 
names(pr_loading) <- sapply(strsplit(pr_info[,"ensembl_id"], ".", fixed = TRUE), function(x) x[1])
#hist(pr_loading)
pr_loading <- pr_loading[which(!is.na(names(pr_loading)))]
scaled.vals <- scale.between.vals(pr_loading, -1, 1)

#convert ensembl IDs to entrez IDs for pathview
names(scaled.vals) <- rna_data$tx_info[match(names(scaled.vals), rna_data$tx_info[,"ensembl_gene_id"]),"entrezgene_id"]

#sort(sapply(path.gene.id, function(x) length(intersect(names(scaled.vals), x))))
kegg.name <- names(which.max(sapply(path.gene.id, function(x) length(intersect(names(scaled.vals), x)))))

#hist(scaled.vals)
#plot(pr_loading, scaled.vals);abline(h = 0, v = 0)
path.num <- path.id[which(u_path == kegg.name)]
pv.out <- pathview(gene.data = scaled.vals, 
    pathway.id = path.num, species = "mmu", 
    out.suffix = kegg.name,
    kegg.dir = ".")
```


## A lot of bmediator trials


```{r bmediatr, eval = FALSE}
#this only does transcript-protein pairs

#gd <- c("FC" = 0, "WT" = 1, "VS" = 2) #allele dosage
gd <- 0:(length(ordered.geno)-1)
names(gd) <- ordered.geno
exp.var <- old.var[colnames(test.tx), "genotype"]
Gdosage <- sapply(exp.var, function(x) gd[which(names(gd) == x)])
geno.design <- model.matrix(~exp.var-1)
rownames(geno.design) <- colnames(test.tx)

med.probs <- vector(mode = "list", length = length(sig.idx))
names(med.probs) <- names(sig.idx)

#go by protein because multiple mediators are supported
#the reactive prior is uniform across all models.

for(i in 1:length(sig.idx)){ 
    out.var <- t(aligned_pr[names(sig.idx)[i],colnames(test.tx),drop=FALSE])

    med.test <- bmediatR(y = out.var, M = t(test.tx), 
        X = geno.design, ln_prior_c = "reactive", 
        options_X = list(sum_to_zero = TRUE, 
        center = FALSE, scale = FALSE))
    
    post <- get_posterior(med.test)
    post.mat <- as.matrix(post[,2:ncol(post)])
    rownames(post.mat) <- post[,1]
    med.probs[[i]] <- post.mat
    #pheatmap(post[,2:ncol(post)], cluster_rows = FALSE, cluster_cols = FALSE, display_numbers  = TRUE)
    #plot_posterior_bar(med.test, mediator_id = rownames(test.tx)[9])
}
```

```{r mediate_all}
#This chunk tests all pairs including transcript-transcript and protein-protein

gd <- c("FC" = 0, "WT" = 1, "VS" = 2) #allele dosage
exp.var <- old.var[colnames(test.tx), "genotype"]
Gdosage <- sapply(exp.var, function(x) gd[which(names(gd) == x)])
geno.design <- model.matrix(~exp.var-1)
rownames(geno.design) <- colnames(test.tx)

med.probs <- vector(mode = "list", length = length(sig.idx))
names(med.probs) <- names(sig.idx)

#test all possible protein and transcript pairs, even within class
pr.for.med <- aligned_pr[names(sig.idx),]
tx.for.med <- test.tx
all.for.med <- t(rbind(pr.for.med, tx.for.med))

all.med.probs <- vector(mode = "list", length = ncol(all.for.med))
names(all.med.probs) <- colnames(all.for.med)
for(i in 1:ncol(all.for.med)){ 
    output <- all.for.med[,i,drop=FALSE]
    meds <- all.for.med[,-i] #everything but the output
    
    med.test <- bmediatR(y = output, M = meds, 
        X = geno.design, ln_prior_c = "reactive", 
        options_X = list(sum_to_zero = TRUE, 
        center = FALSE, scale = FALSE))
    
    post <- get_posterior(med.test)
    post.mat <- as.matrix(post[,2:ncol(post)])
    rownames(post.mat) <- post[,1]
    all.med.probs[[i]] <- post.mat
    #pheatmap(post[,2:ncol(post)], cluster_rows = FALSE, cluster_cols = FALSE, display_numbers  = TRUE)
    #plot_posterior_bar(med.test, mediator_id = rownames(test.tx)[9])
}

```

```{r entropy_check, eval = FALSE}
check.ent <- matrix(NA, ncol = length(med.probs), nrow = nrow(med.probs[[1]]))
rownames(check.ent) <- gene.info[match(rownames(med.probs[[1]]), gene.info[,"ensembl_gene_id"]),"external_gene_name"]
for(i in 1:length(med.probs)){
    check.ent[,i] <- apply(med.probs[[i]], 1, function(x) DescTools::Entropy(x))
}
pheatmap(check.ent)
boxplot(t(check.ent), las  = 2)
hist(check.ent)

#which(check.ent > 2.5, arr.ind = TRUE)
```

```{r med_thresh}
prob.thresh = 0.51
```

Sometimes the partial forward and partial reactive 
models are pretty much evenly split, so we want to have
a relatively high threshold for the posterior models.
I'm going with `r prob.thresh` for now, although I may raise it
later. We also filtered for models that were either 
partial or complete forward or reverse mediation. 


```{r thresh_med}
med.probs <- all.med.probs
interesting.models <- c("partial med", "complete med", 
    "complete med (react)", "partial med (react)")

filtered_med <- vector(mode = "list", length = length(med.probs))
for(i in 1:length(med.probs)){
    max.probs <- apply(med.probs[[i]], 1, max)
    pass.thresh <- which(max.probs > prob.thresh)
    max.model <- apply(med.probs[[i]], 1, which.max)
    target.id <- rep(names(med.probs)[i], length(max.model))
    target.name <- get_mixed_id(ids = target.id, obs_data, rna_data)
    med.id <- rownames(med.probs[[i]])
    med.name <- get_mixed_id(med.id, obs_data, rna_data)
    med.table <- cbind(med.id[pass.thresh], target.id[pass.thresh],
        med.name[pass.thresh], target.name[pass.thresh], max.probs[pass.thresh],
        colnames(med.probs[[i]])[max.model[pass.thresh]])
    is_interesting <- which(med.table[,6] %in% interesting.models)
    filtered_med[[i]] <- med.table[is_interesting,]
}

filtered.results <- Reduce("rbind", filtered_med)
no.name <- which(filtered.results[,3] == "")
filtered.results[no.name,3] <- filtered.results[no.name,1]
no.name <- which(filtered.results[,4] == "")
filtered.results[no.name,4] <- filtered.results[no.name,2]
colnames(filtered.results) <- c("mediator_id", "target_id", 
    "mediator_name", "target_name", "posterior_prob", "best_model")

#We don't need pairs and their reverse, so just keep one of each
u_edges <- unique(t(apply(filtered.results[,c("mediator_name", "target_name")], 1, sort)))
unique_results <- vector(mode = "list", length = nrow(u_edges))
for(i in 1:nrow(u_edges)){
    gene1 <- u_edges[i,1]
    gene2 <- u_edges[i,2]
    line1 <- intersect(which(filtered.results[,c("mediator_name")] == gene1), which(filtered.results[,c("target_name")] == gene2))
    line2 <- intersect(which(filtered.results[,c("mediator_name")] == gene2), which(filtered.results[,c("target_name")] == gene1))
    #there might be multiple peptides assigned to a gene. 
    #pick the lines with forward mediation
    all_lines <- c(line1, line2)
    is_react <- grep("react", filtered.results[all_lines, "best_model"])
    not_react <- setdiff(1:length(all_lines), is_react)
    unique_results[[i]] <- filtered.results[all_lines[not_react],,drop=FALSE]
}
result_table <- Reduce("rbind", unique_results)

u_prot_tx <- unique(as.vector(result_table[,1:2]))
#sort(u_prot_tx)
```

There were `r length(u_prot_tx)` mediating elements in this
set of proteins and transcripts


```{r check_prob_ent, eval = FALSE}
#I've checked filtering by entropy and by maximum individual probability
#and I think the individual probability is better. Everything with a max.prob
#over 0.51 looks good, but even with low entropy levels we get some ambiguous
#model selection. I'm leaving this here so I don't forget I did this.
plot(max.prob, check.ent);abline(v = 0.51)
spot.check <- idx_to_row_col(intersect(which(max.prob < 0.51), which(check.ent < 1.5)), n.row = nrow(max.prob)) #low entropy but also low max prob
spot.check <- idx_to_row_col(intersect(which(max.prob > 0.51), which(check.ent > 2)), n.row = nrow(max.prob)) #high entrop but pretty high max prob
spot.check <- which(max.prob > 0.51, arr.ind = TRUE)
i = 5
barplot(med.probs[[spot.check[i,2]]][spot.check[i,1],])
```

There were `r nrow(filtered.results)` mediation relationships.

We ordered the mediation interactions based on whether the 
best mediation model was forward or reverse to make a directed
mediation network.



```{r order_med}
ordered.bmed <- result_table[order(as.numeric(result_table[,"posterior_prob"]), decreasing = TRUE),]

#there shouldn't be any reactive interactions left, but check anyway
react.idx <- grep("react", ordered.bmed[,"best_model"])
edge_id <- ordered.bmed[,c("mediator_id", "target_id")]
edge_name <- ordered.bmed[,c("mediator_name", "target_name")]

if(length(react.idx) > 0){
    #if there are any switch the direction for the reactive edges.
    edge_id[react.idx,] <- ordered.bmed[react.idx,c("target_id", "mediator_id")]
    edge_name[react.idx,] <- ordered.bmed[react.idx,c("target_name", "mediator_name")]
}


#also get the correlations for these edges.
edge.cor <- rep(NA, nrow(edge_id))
for(i in 1:nrow(ordered.bmed)){
    gene1 <- get_any_abund(gene.id = edge_id[i,1], test_aligned)
    gene2 <- get_any_abund(gene.id = edge_id[i,2], test_aligned)
    edge.cor[i] <- cor(gene1, gene2, use = "pairwise.complete.obs")
}
#hist(edge.cor)

#plot all edges
plot.idx <- 1:nrow(edge_id)

#filter to edges with transcripts
#plot.idx <- unique(c(grep("ENSMUS", edge_id[,1]), grep("ENSMUS", edge_id[,2])))

#filter to one step away from transcripts
#one.step <- unique(as.vector(edge_id[plot.idx,]))
#plot.idx <- unique(c(which(edge_id[,1] %in% one.step), which(edge_id[,2] %in% one.step)))

#filter to only genes connected to a test gene
#check.gene <- "Grin2b"
#plot.idx <- unique(c(which(edge_name[,1] == check.gene), which(edge_name[,2] == check.gene)))

#filter to only positively or negatively correlated edges
plot.idx <- 1:nrow(edge_id) #all
#plot.idx <- which(edge.cor > 0) #positive edges
#plot.idx <- which(edge.cor < 0) #negative edges

pdf("~/Desktop/Network.pdf", width = 10, height = 10)
bmed.net <- plot_mediation_network(edge.list = edge_id[plot.idx,], 
    sample_data = old_data, rna_data = rna_data,
    mediation.prob = as.numeric(ordered.bmed[,"posterior_prob"])[plot.idx],
    abund.cor = edge.cor[plot.idx], max.edge.width = 5, label.cex = 1,
    max.vertex.size = 10, vertex.label.dist = 0.5)
dev.off()

comm <- cluster_walktrap(bmed.net, weights = E(bmed.net)$mediation_prob, steps = 1000)$membership

#comm <- cluster_fast_greedy(as_undirected(bmed.net))$membership

pdf("~/Desktop/community_colors.pdf", width = 10, height = 10)
plot(bmed.net, vertex.color = comm, vertex.size = 5)
legend("topright", fill = categorical_pal(max(comm))[1:max(comm)], legend = paste("Cluster", 1:max(comm)))
dev.off()

comm.names <- lapply(1:max(comm), function(x) setdiff(V(bmed.net)$name[which(comm == x)], test.genes)) #without transcript names
comm.names <- lapply(1:max(comm), function(x) V(bmed.net)$name[which(comm == x)]) #with transcript names
#names(comm.names) <- paste("cluster", 1:length(comm.names))
#lapply(comm.names, function(x) which(x %in% test.genes))
comm.enrich <- lapply(comm.names, function(x) gost(x, organism = "mmusculus", evcodes = TRUE))
plot.enrichment.group(comm.enrich, max.term.size = 3000, transformation = function(x) log10(x+1))
#plot.enrichment(comm.enrich[[1]], max.term.size = 3000, num.terms = 15)

stop()
```

```{r community_testing, eval = FALSE}
#I'm playing around with this fibrinogen cluster to see what we can 
#get from it.

blood.genes <- comm.names[[1]]
pep.idx <- c(grep("Q", blood.genes), which(blood.genes == "LC")) #remove unnamed peptides and LC. That is some kind of Igk
blood.genes <- blood.genes[-pep.idx]

all.pr.stats <- all.tr.stats <- vector(mode = "list", length = length(blood.genes))
names(all.pr.stats) <- names(all.tr.stats) <- blood.genes
pdf("~/Desktop/blood_proteins_and_transcripts.pdf", width = 10, height = 7)
for(i in 1:length(blood.genes)){
    all.pr.stats[[i]] <- plot_pr_abund(blood.genes[i], obs_data, plot.label = "Protein")
    all.tr.stats[[i]] <- plot_rna_abund(blood.genes[i], rna_data, plot.label = "Transcript")
}
dev.off()


cbind(sapply(all.pr.stats, length), sapply(all.tr.stats, length))
not_in_br <- which(is.na(sapply(all.tr.stats, function(x) x[[1]][1])))
in_br <- setdiff(1:length(blood.genes), not_in_br) #blood.genes[in_br]
check_idx <- in_br
check_idx <- not_in_br #not transcribed in brain
check_idx <- 1:length(blood.genes)

check_tr <- sapply(all.tr.stats[check_idx], function(x) x[[1]][[1]])
colnames(check_tr) <- blood.genes[check_idx]
pheatmap(cor(check_tr))

check_pr <- sapply(all.pr.stats[check_idx], function(x) x[[1]][[1]])
colnames(check_pr) <- blood.genes[check_idx]
pheatmap(cor(check_pr, use = "pairwise.complete.obs"))

png("~/Desktop/blood_proteins.png", width = 12, height = 7, units = "in", res = 300)
check <- plot_multi_gene(gene.names = blood.genes[check_idx], sample_data = obs_data, 
    data.type = "scaled", test_factor = "mean", mouse_order_by = "genotype", 
    adjust.for = NULL, pr_order_by = "cluster", show_r2 = "genotype", 
    stat.x = 0.67)
dev.off()


group1 <- c("Hba", "Hbbt1", "Ca1", "Alas2", "Gypa") #blood component markers, down in VS
group2 <- c("Alb", "Fga", "Fgb", "Fgg", "Kng1", "Hrg", "Apoa4") #coagulation markers, down in VS
group3 <- c("Pdgfrb", "Cspg4", "Rgs5", "Mcam", "Des", "Acta2"); #pericyte markers, no difference
group4 <- c("Cldn5", "Pecam1", "Cdh5", "Emcn", "Kdr", "Flt1", "Eng", "Robo4", "Esam", "Ly6c1", "Ly6ca", "Mfsd2a", "Slc2a1") #endothelial markers; no difference
group5 <- c("Col4a1", "Col4a2", "Lama2", "Lama4", "Lamb1", "Lam1", "Nid2", "Nid1", "Hspg2"); #basmemt membrane proteins; trend up in VS
group6 <- c("Slc2a1", "Mfsd2a", "Abcb1a", "Abcb1b", "Abcg2"); #BBB-specific transporters, no difference
group7 <- c("Aqp4", "Gfap", "Dag1", "Dtna", "Snta1"); #Astrocyte markers, no difference
group8 <- c("Ppbp", "Pf4", "Itga2b", "Itgb2", "Gp1ba", "Gp9", "Trem11", "Vwf", "F13a1"); #platelet-associated markers, trend down in VS
group9 <- c("Apoe", "C3", "C1qa", "C1qb", "C1qc", "Lgals3", "Gfap", "Vim"); #response genes; no difference
injury_response <- read.csv(here("Data", "injury_response2.csv"))
ortho.table <- get_orthologs(human.id = injury_response[,2], id.type = "external_gene_name")
group10 <- unlist(ortho.table[,"mouse_external_gene_name"])

plot_multi_gene(gene.names = group10, sample_data = obs_data, 
    data.type = "scaled", test_factor = "mean", mouse_order_by = "genotype", 
    adjust.for = NULL, pr_order_by = "cluster", show_r2 = "genotype", 
    stat.x = 0.67)

group1_vals <- peptide_vals(c(group1, group2), sample_data = old_data, data.type = "mean", adjust.for = "sex")
group2_vals <- peptide_vals(group2, sample_data = obs_data, data.type = "mean", adjust.for = "sex")
group1_score <- colMeans(group1_vals, na.rm = TRUE)
group2_score <- colMeans(group2_vals, na.rm = TRUE)

model <- lm(group1_score~mouse_var[,"genotype"])
pred <- predict(model)
plot(pred, group1_score)
summary(model)
hist(resid(model))

pr.blood.model <- apply(aligned_pr, 1, function(x) lm(x~group1_score[colnames(aligned_pr)]+old.var[colnames(aligned_pr),"genotype"]))
pr.blood.effect <- sapply(pr.blood.model, function(x) coef(summary(x))[3,1]) #effect of genotype on the protein
pr.blood.p <- sapply(pr.blood.model, function(x) coef(summary(x))[3,4]) #p value of genotype on the protein

blood.sig.idx <- which(p.adjust(pr.blood.p, "fdr") < 0.1) #filter to proteins with some genetic effect
length(blood.sig.idx)

check_data <- get_data(obs_data, data.type = "scaled", adjust.for = "sex")
test_decomp <- decomp_with_nas(t(check_data), plot.results = FALSE, pc = 20)
barplot(abs(apply(test_decomp$u, 2, function(x) cor(x, group1_score))), names = 1:20)
plot(test_decomp$u[,1], group1_score)
plot.with.model(group1_score, group2_score)



mouse_var <- get_factor_var(old_data, data.type = "scaled")

model <- lm(pep_score~mouse_var[,"genotype"]*mouse_var[,"age"])
anova(model)
boxplot(pep_score~mouse_var[,"genotype"]*mouse_var[,"age"])

```


```{r categories, eval = FALSE}
have.cat.genes <- sapply(strsplit(rownames(check), ":"), function(x) x[1])
filtered.cats <- test_cats[match(have.cat.genes, test_cats[,1]),]

gene1.id <- obs_data$pr_info[which(obs_data$pr_info[,gene.col] == "Fgb"),id.col]
gene1.abund <- get_any_abund(gene1.id, test_aligned)

u_cat <- unique(filtered.cats[,"Category"])
cat_genes <- lapply(u_cat, function(x) filtered.cats[which(filtered.cats[,"Category"] == x),1])
names(cat_genes) <- u_cat
cat.cor <- vector(mode = "list", length = nrow(filtered.cats))
pdf("~/Desktop/test.genes.pdf")
for(i in 1:nrow(filtered.cats)){
    print(i)
    gene.name <- filtered.cats[i,1]
    gene2.id <- obs_data$pr_info[which(obs_data$pr_info[,gene.col] == gene.name),id.col]
    plot_pr_abund(gene.name, old_data, data.type = "scaled")
    gene2.abund <- sapply(gene2.id, function(x) get_any_abund(x, test_aligned))
    par(mfrow = c(1,ncol(gene2.abund)))
    all.gene.cor <- rep(NA, ncol(gene2.abund))
    for(g in 1:ncol(gene2.abund)){
        if(!all(is.na(gene2.abund[,g]))){
            all.gene.cor[g] <- plot.with.model(gene1.abund, gene2.abund[,g], main = filtered.cats[i,2])[1]
        }
    }
    cat.cor[[i]] <- all.gene.cor
}
dev.off()

grouped.cor <- lapply(u_cat, function(x) cat.cor[which(filtered.cats[,2] == x)])
names(grouped.cor) <- u_cat
check.cor <- unlist(grouped.cor, recursive = FALSE)
par(mar = c(4,12, 4, 2))
boxplot(check.cor, las = 2, horizontal = TRUE, cex.axis = 0.5)

indeg <- strength(bmed.net, mode = "in", weights = E(bmed.net)$correlation) #weighted in degree
outdeg <- strength(bmed.net, mode = "out", weights = E(bmed.net)$correlation) #weighted out degree

indeg <- strength(bmed.net, mode = "in", weights = E(bmed.net)$mediation_prob) #weighted in degree
outdeg <- strength(bmed.net, mode = "out", weights = E(bmed.net)$mediation_prob) #weighted out degree

barplot(sort(indeg), las = 2)
barplot(sort(outdeg), las = 2)
label.in <- which(indeg > 8)
label.out <- which(outdeg > 10)

#genes with high in degree have genotype effects through other genes
#genes with high out degree are core genes, and affecting others 
plot(jitter(indeg), jitter(outdeg), ylim = c(0, 30), xlab = "In Degree", ylab = "Out Degree",
    pch = 16, cex = 0.7)
text(indeg[c(label.in, label.out)], outdeg[c(label.in, label.out)], 
    labels = c(names(label.in), names(label.out)), pos = 3, cex = 0.7)


#pdf("~/Desktop/bmediatr_test_med.pdf", width = 10, height = 8)
#for(i in 1:nrow(filtered.results)){
#    #quartz(width = 10, height = 7)
#    plot_genotype_mediation(aligned_data = test_scaled, sample_data = old_data, 
#        rna_data = rna_data, transcript.gene = filtered.results[i,1], 
#        peptide.gene = filtered.results[i,2], autoplace.text = TRUE,
#        reverse.mediation = FALSE)
#}
#dev.off()

med.list <- list("forward" = V(pr.tx.net)$name, "reverse" = V(reverse.pr.tx.net)$name,
    "bmediatR" = V(bmed.net)$name)
plotVenn(med.list)
common.med <- Reduce("intersect", med.list)

just.bmed <- setdiff(med.list$bmediatR, union(med.list$forward, med.list$reverse))
cat(just.bmed, sep = "\n")

just.med <- setdiff(union(med.list$forward, med.list$reverse), med.list$bmediatR)
cat(just.med, sep = "\n")
stop()
```


```{r med_enrich}
med.enrich <- gost(u_prot_med, organism = "mmusculus")
enrich.mat <- plot.enrichment(med.enrich, num.terms = 25, max.term.size = 3000)
cat(u_prot_med, sep = "\n")
```

```{r test_reverse}
tx.name = "Grin2b"; pr.name <- "Setdb1"
tx.name = "Gria3"; pr.name <- "Cpeb3"
tx.name = "Cacna1b"; pr.name <- "Dzank1"
tx.name <- "Chrna7"; pr.name = "Wdr5" #partial reactive
tx.name = "Cacna1b"; pr.name = "Afg3l2" #complete med
tx.name = "Cacna1b"; pr.name = "Tnpo1" #complete med

i = 1
tx.name <- filtered.results[i,1]; pr.name <- filtered.results[i,2]

pdf(paste0("~/Desktop/", tx.name, "_and_", pr.name, ".pdf"), width = 10, height = 7)
#quartz(width = 10, height = 7)
forward.result <- plot_genotype_mediation(aligned_data = test_scaled, sample_data = old_data, 
        rna_data = rna_data, source.gene = tx.name, source.gene.type = "transcript",
        target.gene = pr.name, target.gene.type = "peptide", 
        autoplace.text = TRUE, reverse.mediation = FALSE)

#quartz(width = 10, height = 7)
plot_genotype_mediation(aligned_data = test_scaled, sample_data = old_data, 
        rna_data = rna_data, transcript.gene = tx.name, 
        peptide.gene = pr.name, autoplace.text = TRUE, reverse.mediation = TRUE)
dev.off()


individual_bmediatR(tx.name = tx.name, pr.name = pr.name, aligned_data = test_scaled, 
    sample_data = old_data, rna_data = rna_data, plot.results = TRUE)


```


## Older stuff I was playing with


So we want to compare the genotype effects in just the 
`r compare.age.text`-month old animals. However, we also want to be able 
to look at the same stats for the four-month old animals. The following 
sections look at a single age that is defined in the parameters section at 
the top of the workflow. For this run, the age is `r compare.age` months.

We fit linear models to protein abundance in just the animals of the 
specified age. We then compared p values and effect sizes 
across all common genes. We also noted which genes (if any) have significant
effects of genotype in both the proteome and the transcriptome in the 
`r compare.age`
month animals.

The volcano plot for these tests is below. We have reduced power 
to detect differences in this data set compared to the transcriptomics,
likely because of missing data. 

```{r age_thresh}
raised.fdr.thresh <- 0.1
```


```{r age_test}
young.mice <- obs_data$mouse_info[which(obs_data$mouse_info[,"age_months"] == "4"),1]
old.mice <- obs_data$mouse_info[which(obs_data$mouse_info[,"age_months"] == "12"),1]

if(compare.age == 12){
    age_data <- remove_mice(obs_data, young.mice)
}else{
    age_data <- remove_mice(obs_data, old.mice)
}

age_genotype_effects <- test_model(age_data, factor.name = "genotype")
age_genotype_fdr <- apply(age_genotype_effects$p, 2, function(x) p.adjust(x, "fdr"))

age.sig <- which(age_genotype_fdr[,1] <= raised.fdr.thresh)
```

At the same FDR threshold as in the transcriptomics 
(`r raised.fdr.thresh`), we get `r length(age.sig)` 
significantly different proteins by genotype in the 
`r compare.age`-month old animals. These are shown in red 
on the volcano plots below.

```{r age_months, fig.width = 7, fig.height = 7}

pt.col <- rep("black", nrow(age_genotype_effects$effect))
sig.idx <- age.sig
pt.col[sig.idx] <- "red"
par(mfrow = c(1,1))
plot(age_genotype_effects$effect[,1], -log10(age_genotype_fdr[,1]), 
    xlab = paste("Effect of genotype at", compare.age, "months"), ylab = "-log10(p value)", 
    col = pt.col, pch = 16)

if(length(sig.idx) > 0){
    text(age_genotype_effects$effect[sig.idx,1], -log10(age_genotype_fdr[sig.idx,1]), 
    labels = names(age_genotype_effects$effect[sig.idx,1]), pos = 4, cex = 0.5)
}

```

At `r compare.age` months, the upregulated proteins are enriched 
for mitochondrial function. This tracks with the transcriptomic 
results in which mitochondrial metabolism genes overall were up in VS.

The downregulated proteins are enriched for protein 
localization and blood coagulation. This doesn't have 
a clear analog in the transcriptomic data set. Greg thinks
the "blood coagulation" might be a marker of something else,
like ECM, since we take all the blood out of the brains before
testing.

```{r test_sig_age, fig.width = 15, fig.height = 7}
up.sig <- rownames(age_genotype_effects$effect)[intersect(sig.idx, which(age_genotype_effects$effect[,1] > 0))]
down.sig <- rownames(age_genotype_effects$effect)[intersect(sig.idx, which(age_genotype_effects$effect[,1] < 0))]


enrich.file <- file.path(results.dir, paste0("pos_neg_enrich_at_", compare.age, "_months.RDS"))
if(file.exists(enrich.file)){
    enrich <- readRDS(enrich.file)    
}
if(!file.exists(enrich.file) && !offline){
    enrich <- lapply(list("Positive" = up.sig, "Negative" = down.sig), function(x) gost(x, organism = "mmusculus", source = "GO"))
    saveRDS(enrich, enrich.file)
}

#quartz(width = 15, height = 7)
par(mfrow = c(1,2))
for(i in 1:length(enrich)){
    plot.enrichment(enrich[[i]], max.term.size = 3000, num.terms = 25, plot.label = paste(names(enrich)[i], "effects of genotype"))
}
#dev.off()
```

The code below plots out gene-GO term plots all together
as well as split by GO compartment (biological process,
molecular function, and cellular compartment).

Looking at all GO classificatoins together, we see that
the proteins upregulated in VS divided into two large clusters:
one related to mitochondria, and the other related to small
molecule metabolism. These clusters are highly connected,
suggesting that this small molecule metabolism might
be related to mitochondrial activity.

The proteins downregulated in VS cluster aroung three major
and two minor collections of GO terms. The three major clusters
are response to wounding (blood coagulation), protein metabolism,
and protein transport. The two minor clusters are enzyme activity,
and cholesterol/triglyceride stuff (connected to Apo genes).

These are pretty cool, actually, but we can't put these network
plots in a paper.


```{r go_part}
max.pval = 1e-3 #make this low to reduce some of the noise
enrich.file <- file.path(results.dir, paste0("GO_part_enrich_at_", compare.age, "_months.RDS"))

if(file.exists(enrich.file)){
    enrich.list <- readRDS(enrich.file)
    up.enrich <- enrich.list$up
    down.enrich <- enrich.list$down
    all.enrich <- enrich.list$all
}
if(!file.exists(enrich.file) && !offline){
    up.enrich <- gost(up.sig, organism = "mmusculus", evcodes = TRUE, sources = "GO")
    down.enrich <- gost(down.sig, organism = "mmusculus", evcodes = TRUE, sources = "GO")
    all.enrich <- gost(c(up.sig, down.sig), organism = "mmusculus", evcodes = TRUE, sources = "GO")
    enrich.list <- list("up" = up.enrich, "down" = down.enrich, "all" = all.enrich)
    saveRDS(enrich.list, enrich.file)
}

all.var <- ls() #check to see if we have the enrichment data
got.enrich <- length(which(all.var == "all.enrich")) > 0

if(got.enrich){

    go.parts <- c("GO:CC", "GO:MF", "GO:BP")


    pdf(file.path(results.dir, paste0("GO_gene_net_at_", compare.age, "_months.pdf")), width = 30, height = 20)
        up.net <- go_gene_net(enrichment.result = up.enrich, max.term.size = 3000,
            plot.results = TRUE, min.vertex.size = 2, max.vertex.size = 8, 
            max.pval = max.pval)
        mtext("Up in VS", side = 3, outer = TRUE, line = -2.5)    
        
        #up.mat <- go_gene_net(up.enrich, max.term.size = 3000, max.pval = max.pval, plot.as.heatmap = TRUE)
        #up.decomp <- plot.decomp(up.mat$biadjacency_matrix, label.points = TRUE)
        #plot.decomp(t(up.mat$biadjacency_matrix), label.points = TRUE)
        #png("~/Desktop/up_in_vs_adj.png", width = 11, height = 4, units = "in", res = 300)
        #par(mar = c(4,15,4,2))
        #imageWithText(up.mat$biadjacency_matrix[order(up.decomp$u[,1]), order(up.decomp$v[,1])], row.text.shift = 0.03, col.text.shift = 0.05, show.text = FALSE, col.scale = "blue", main = "Up in VS")
        #dev.off()

        down.net <- go_gene_net(enrichment.result = down.enrich, plot.results = TRUE, 
            max.pval = max.pval, max.term.size = 3000, min.vertex.size = 2, max.vertex.size = 8)
        mtext("Down in VS", side = 3, outer = TRUE, line = -2.5) 

        #down.mat <- go_gene_net(down.enrich, max.term.size = 3000, max.pval = max.pval, plot.as.heatmap = TRUE)
        #down.decomp <- plot.decomp(down.mat$biadjacency_matrix, label.points = TRUE)
        #plot.decomp(t(down.mat$biadjacency_matrix), label.points = TRUE)
        #row.order <- hclust(dist(down.mat$biadjacency_matrix))$order
        #col.order <- hclust(dist(t(down.mat$biadjacency_matrix)))$order
        #png("~/Desktop/down_in_vs_adj.png", width = 16, height = 6, units = "in", res = 300)
        #par(mar = c(4,22,4,2))
        #imageWithText(down.mat$biadjacency_matrix[row.order, col.order], row.text.shift = 0.03, col.text.shift = 0.05, show.text = FALSE, col.scale = "blue", main = "Down in VS")
        #dev.off()
    dev.off()

    for(i in 1:length(go.parts)){
        pdf(file.path(results.dir, paste0(gsub(":", "_", go.parts[i]), "_", compare.age, "_months_net.pdf")), 
            width = 15, height = 10)
        up.net <- go_gene_net(enrichment.result = up.enrich, max.term.size = 3000, 
            filter.source = go.parts[i], max.pval = max.pval,
            plot.results = TRUE, min.vertex.size = 2, max.vertex.size = 8)
            mtext("Up in VS", side = 3, outer = TRUE, line = -2.5)    
        
        down.net <- go_gene_net(enrichment.result = down.enrich, plot.results = TRUE, 
            filter.source = go.parts[i], max.pval = max.pval,
            max.term.size = 3000, min.vertex.size = 2, max.vertex.size = 8)
            mtext("Down in VS", side = 3, outer = TRUE, line = -2.5) 
        dev.off()
    }
}
```


Did we see differential expression of any of these genes in the
transcriptomic data? We test this below.

```{r transcriptome_results}
#get gene information for the proteins
protein.info.file <- file.path(processed.data.dir, "protein_info.tsv")
if(!file.exists(protein.info.file)){
    protein.info <- gconvert(colnames(centered.gene.mat), organism = "mmusculus")
    write.table(protein.info[,2:6], protein.info.file, sep = "\t", quote = FALSE, row.names = FALSE)
}else{
    protein.info <- read.delim(protein.info.file)
}
sig.tx <- read.delim(here("Results", paste0("age_batch_", compare.age, "_gene"), "DEG_clusters_carriers.txt"))
```

```{r matching}
pos.pr <- obs_data$pr_info[which(up.sig %in% obs_data$pr_info[,id.col]),gene.col]
neg.pr <- obs_data$pr_info[which(down.sig %in% obs_data$pr_info[,gene.col]),gene.col]

common.pos <- intersect(pos.pr, sig.tx[,"geneName"])
common.neg <- intersect(neg.pr, sig.tx[,"geneName"])

pos.pr.idx <- match(common.pos, rownames(geno.result.table))
neg.pr.idx <- match(common.neg, rownames(geno.result.table))

pos.tr.idx <- match(common.pos, sig.tx[,"geneName"])
neg.tr.idx <- match(common.neg, sig.tx[,"geneName"])

all.pos.table <- cbind(sig.tx[pos.tr.idx,1:5], geno.result.table[pos.pr.idx,,drop=FALSE])
all.neg.table <- cbind(sig.tx[neg.tr.idx,1:5], geno.result.table[neg.pr.idx,,drop=FALSE])
colnames(all.pos.table)[3:5] <- colnames(all.neg.table)[3:5] <- paste0("transcript_", colnames(all.pos.table)[3:5])
colnames(all.pos.table)[6:8] <- colnames(all.neg.table)[6:8] <- paste0("protein_", colnames(all.pos.table)[6:8])

all.common.table <- rbind(all.pos.table, all.neg.table)
write.table(all.common.table, file.path(results.dir, paste0("Common_Effects_", compare.age, "_months.txt")),
    quote = FALSE, sep = "\t", row.names = FALSE)
```

The scatter plot below shows that there is no relationship 
between the effect of genotype on a gene's transcript and its
protein abundance, even for significantly affected proteins. 

If we are looking at 4-month olds, there won't be many
dots. This is because Klotho genotype has basically no
effect on any transcripts in the 4-month olds, so it's
hard to compare. In any case, even with the few genes 
shown, there is no relationship between the effects
of genotype on transcripts and those on proteins. 

```{r compare, fig.width = 5, fig.height = 5}
plot(all.common.table[,"transcript_genotype_beta"], 
    all.common.table[,"protein_Effect"], xlab = "Transcript Beta",
    ylab = "Protein Beta")

text(all.common.table[,"transcript_genotype_beta"], 
    all.common.table[,"protein_Effect"], labels = all.common.table[,"geneName"],
    pos = 4, cex = 0.7)
#abline(0,1)
abline(v = 0, h = 0)
```

Below we expanded the plot to include all transcripts with 
significant effects of genotype. It looks very similar to the
above plot.

```{r common}
#use ensembl IDs
common.gene <- intersect(rownames(sig.tx), protein.info[,"target"])
```

There were `r length(common.gene)` positively identifiable common 
genes between the two data sets.

Below we've plotted the effect size of genotype on each common 
transcript vs. the effect size of gentype on each common protein.
They are uncorrelated to a hilarious degree. Greg thinks that this
is probably grounds to dismiss the data. He says they usually see
a correlation of 0.15 to 0.3 for genotype effects. Bummer.

```{r align_genes, fig.width = 5, fig.height = 5}
tx.idx <- match(common.gene, rownames(sig.tx))
pr.name <- protein.info[match(common.gene, protein.info[,"target"]), "input"]
pr.idx <- match(pr.name, rownames(geno.result.table))

tx.effect <- sig.tx[tx.idx,"genotype_beta"]

combined.result.table <- cbind("Name" = pr.name, "ID" = common.gene, 
    "Transcript Effect" = tx.effect, 
    "Protein Effect" = geno.result.table[pr.idx, "Effect"])

plot.with.model(as.numeric(combined.result.table[,"Transcript Effect"]), 
    as.numeric(combined.result.table[,"Protein Effect"]),
    xlab = "Effect of genotype on transcript", ylab = "Effect of genotype on protein",
    report = "cor.test")

#combined.result.table[which(as.numeric(combined.result.table[,"Protein Effect"]) > 0.4),]

```


## Biodomains and KEGG pathways

We looked at biodomains and KEGG pathways just to round this out.
There were no overall effects on biodomain or KEGG lists

```{r plot_list}
plot_list <- function(gene.list, pr.mat, list.label = "BD", plot.results = TRUE){

    gene.stats <- vector(mode = "list", length = length(gene.list))
    names(gene.stats) <- names(gene.list)
    
    if(plot.results){
        pdf(file.path(results.dir, paste0(list.label, ".pdf")), width = 9, height = 6)
    }
    for(i in 1:length(gene.list)){
        gene.names <- gene.info[match(gene.list[[i]], gene.info[,"ensembl_gene_id"]),"external_gene_name"]
        gene.stats[[i]] <- plot_pr_abund(gene.names, pr.mat, plot.label = names(gene.list)[i], 
            plot.label.cex = 2, plot.results = plot.results)
    }
    if(plot.results){
        dev.off()
    }

    return(gene.stats)
}

plot_nested_list <- function(nested.gene.list, pr.mat, list.label = "KEGG_BD", plot.results = TRUE){

    gene.stats <- vector(mode = "list", length = length(nested.gene.list))
    names(gene.stats) <- names(nested.gene.list)
    
    if(plot.results){
        pdf(file.path(results.dir, paste0(list.label, ".pdf")), width = 9, height = 6)
    }
    for(i in 1:length(nested.gene.list)){
        if(length(nested.gene.list[[i]]) == 0){next()}
        sub.stats <- vector(mode = "list", length = length(nested.gene.list[[i]]))
        names(sub.stats) <- names(nested.gene.list[[i]])
        for(j in 1:length(nested.gene.list[[i]])){
            plot.label <- paste(names(nested.gene.list)[i], "AND\n", names(nested.gene.list[[i]])[j])
            gene.names <- gene.info[match(nested.gene.list[[i]][[j]], gene.info[,"ensembl_gene_id"]),"external_gene_name"]
            group.result <- plot_pr_abund(gene.names, pr.mat, plot.label = plot.label, plot.label.cex = 1, plot.results = plot.results)
            sub.stats[[j]] <- group.result
        }
        gene.stats[[i]] <- sub.stats
    }
    if(plot.results){
        dev.off()
    }

    return(gene.stats)
}

get_list_stat <- function(list.result, 
    stat.label = c("sex", "age", "genotype", "geno_four", paste0("geno_", compare.age.text)), 
    stat.type = c("r2", "model.p", "linear.effect.size", "linear.effect.p")){


    stat.p <- rep(NA, length(list.result))
    names(stat.p) <- names(list.result)
    for(i in 1:length(list.result)){
        stat.mat <- list.result[[i]]$stats
        if(length(stat.mat) > 1){
            stat.p[i] <- as.numeric(stat.mat[stat.label, stat.type])
        }
    }
    return(stat.p)
}

get_nested_stat <- function(nested.list.result, 
    stat.label = c("sex", "age", "genotype", "geno_four", paste0("geno_", compare.age.text)), 
    stat.type = c("r2", "model.p", "linear.effect.size", "linear.effect.p")){

    stat.label <- stat.label[1]
    stat.type <- stat.type[1]

    stat.p <- vector(mode = "list", length(nested.list.result))
    names(stat.p) <- names(nested.list.result)
    
    for(i in 1:length(nested.list.result)){
        sub.stat.p <- rep(NA, length(nested.list.result[[i]]))
        names(sub.stat.p) <- names(nested.list.result[[i]])
        for(j in 1:length(nested.list.result[[i]])){
            if(length(nested.list.result[[i]][[j]]) > 1){
                stat.mat <- nested.list.result[[i]][[j]]$stats
            }else{
                stat.mat <- NA
            }
            if(length(stat.mat) > 1){
                sub.stat.p[j] <- as.numeric(stat.mat[stat.label, stat.type])
            }
        }
        stat.p[[i]] <- sub.stat.p
    }
    return(stat.p)
}
```

```{r gene_lists}
bd.list <- readRDS(file.path(general.processed.data.dir, "Mouse_Biodomains_for_GSEA.RDS"))
sub.bd.list <- readRDS(file.path(general.processed.data.dir, "Mouse_Subdomains_for_GSEA.RDS"))
bd.go.list <- readRDS(file.path(general.processed.data.dir, "Mouse_Biodomains_sub_GO_for_GSEA.RDS"))
kegg.list <- readRDS(file.path(general.processed.data.dir, "Mouse_KEGG_for_GSEA.RDS"))
kegg.bd.list <- readRDS(file.path(general.processed.data.dir, "Mouse_KEGG_Intersections_for_GSEA.RDS"))
```

There were no significantly differentially abundant biodomains at `r compare.age.text` months.

```{r bd, fig.width = 5, fig.height = 5}
bd.stats <- plot_list(gene.list = bd.list, pr.mat = centered.gene.mat, list.label = paste0("Biodomains_", compare.age, "_months"), plot.results = FALSE)
n.bd.genes <- sapply(bd.stats, function(x) if(length(x) > 0){ncol(x$expr_mat)}else{NA})
age.bd.p <- get_list_stat(bd.stats, stat.label = paste0("geno_", compare.age.text), stat.type = "model.p")
qqunif.plot(age.bd.p,  plot.label = "Biodomain p values")
```

There were no significantly differentially abundant biodomains at `r compare.age.text` months.

```{r k, fig.width = 5, fig.height = 5}
kegg.stats <- plot_list(bd.list, centered.gene.mat, "KEGG", plot.results = FALSE)
n.kegg.genes <- sapply(kegg.stats, function(x) if(length(x) > 0){ncol(x$expr_mat)}else{NA})
age.kegg <- get_list_stat(kegg.stats, stat.label = paste0("geno_", compare.age.text), stat.type = "model.p")
qqunif.plot(age.kegg, plot.label = "KEGG p values")
```

There were no significantly differentially abundant subdomains at `r compare.age.text` months.

```{r sbd, fig.width = 5, fig.height = 5}
sub.bd.stats <- plot_nested_list(nested.gene.list = sub.bd.list, centered.gene.mat, "Subdomains", plot.results = FALSE)
n.sbd.genes <- lapply(sub.bd.stats, function(x) sapply(x, function(y) if(length(y) > 0){ncol(y$expr_mat)}else{NA}))
pass.thresh <- which(unlist(n.sbd.genes) >= 10)
age.sub.bd <- get_nested_stat(nested.list.result = sub.bd.stats, stat.label = paste0("geno_", compare.age.text), stat.type = "model.p")
qqunif.plot(unlist(age.sub.bd)[pass.thresh],  plot.label = "Subdomain p values")
```

There were no significantly differentially abundant 
KEGG-biodomain intersections.


```{r bdk}
bd.kegg.stats <- plot_nested_list(nested.gene.list = kegg.bd.list, 
    centered.gene.mat, list.label = "KEGG-Biodomains", plot.results = FALSE)
n.bdk.genes <- lapply(bd.kegg.stats, function(x) sapply(x, function(y) if(length(y) > 1){ncol(y$expr_mat)}else{NA}))
#hist(unlist(n.bdk.genes), breaks = 100)
pass.thresh <- unlist(lapply(n.bdk.genes, function(x) which(x >= 10)))
age.kegg.bd.p <- unlist(get_nested_stat(nested.list.result = bd.kegg.stats, 
    stat.label = paste0("geno_", compare.age.text), stat.type = "model.p"))
age.kegg.bd.effect <- unlist(get_nested_stat(nested.list.result = bd.kegg.stats, 
    stat.label = paste0("geno_", compare.age.text), stat.type = "linear.effect.size"))

qqunif.plot(age.kegg.bd.p[names(pass.thresh)],  plot.label = "KEGG-Biodomain p values")
```

The volcano plot is shown below. Significant intersections (FDR < 0.1)
are highlighted.

```{r bdk_volcano, fig.width = 5, fig.height = 5}
stat.table <- cbind(age.kegg.bd.effect[names(pass.thresh)], 
    age.kegg.bd.p[names(pass.thresh)])
colnames(stat.table) <- c("effect", "p")

adj.p <- p.adjust(stat.table[,"p"], method = "fdr")
#plot(-log10(stat.table[,"p"]), -log10(adj.p))
sig.idx <- which(adj.p <= 0.1)
#stat.table[sig.idx,]
sig.col <- rep("black", length(adj.p))
sig.col[sig.idx] <- "red"

plot(stat.table[,"effect"], -log10(stat.table[,"p"]), 
    col = sig.col, pch = 16, xlab = "Effect Size", 
    ylab = "-log10(p)", main = "KEGG-Biodomain Intersections")
```

## Biodomain-Kegg examples {.tabset .tabset-fade .tabset-pills}

If there are significant effects, we plot the significant ones below.
We show a maximum of 5.

```{r example_bdk, fig.width = 9, fig.height = 6, results = "asis"}
max.show <- 5
show.num <- min(c(max.show, length(sig.idx)))
if(show.num > 0){
    sig.int <- rownames(stat.table)[sig.idx]
    split.int <- strsplit(sig.int, ".", fixed = TRUE)
    sig.table <- Reduce("rbind", split.int)
    if(show.num == 1){
        sig.table <- matrix(sig.table, nrow = 1)
    }
    colnames(sig.table) <- c("Biodomain", "KEGG")

    #plot out the significant KEGG-Biodomain intersections
    #pdf(file.path(results.dir, "KEGG-Biodomain_intersections.pdf"), width = 9, height = 6)
    
    for(i in 1:show.num){
        cat("###", paste(sig.table[i,], collapse = " AND "), "\n")
        bd.idx <- which(names(kegg.bd.list) == sig.table[i,1])
        k.idx <- which(names(kegg.bd.list[[bd.idx]]) == sig.table[i,2])
        gene.names <- gene.info[match(kegg.bd.list[[bd.idx]][[k.idx]], gene.info[,"ensembl_gene_id"]), "external_gene_name"]
        plot_pr_abund(gene.names, centered.gene.mat, autoplace.text = TRUE, 
        plot.label = paste(sig.table[i,], collapse = "\nAND\n"))
        cat("\n\n")
    }
    #dev.off()
}
```


Mike wants a figure comparing the genotype effects on 
the short list of AD genes that I had before from Christian 
Behl's book.

He thinks showing these side-by-side will make the paper fly 
high. I think it will look like a mess. Let's see, shall we?


```{r ad_short_list}
#read in transcript results
tx.ad.short <- as.matrix(read.delim(here("Results", paste0("age_batch_", compare.age, "_gene"), "AD_short_list_means.txt"), comment.char = "#"))

#find the common gene names and IDs between proteins and transcripts
common.ad <- intersect(rownames(tx.ad.short), colnames(centered.gene.mat))
common.id <- gene.info[match(common.ad, gene.info[,"external_gene_name"]), "ensembl_gene_id"]

#get the transcript stats for the age-specific animals
tx.beta <- tx.ad.short[,"effect"]
tx.p <- tx.ad.short[,"p"]

#find the significant genes with an FDR <= 0.1
sig.tx <- names(tx.p)[which(tx.p <= 0.01)]

#adjust the protein data for sex
ad.pr <- adjust(centered.gene.mat[,common.ad], dummy_covar(data.frame("sex" = sex, "age" = age)), retain.intercept = FALSE)
age.ad.pr <- ad.pr[age.idx,]
#calculate means across genotypes for the genes of interest in the age-spcific animals
ad.pr.mean <- sapply(levels(genotype), function(x) colMeans(age.ad.pr[which(genotype[age.idx] == x),], na.rm = TRUE))
#attach the effect sizes and p values from the linear models with the age-specific animals
pr.effect <-  age.geno.effect[common.ad]
pr.p <- age.geno.p[1,common.ad]
ad.pr.mean <- cbind(ad.pr.mean, "effect" = pr.effect, "p" = pr.p)


#align the transcript and protein means based on gene names
aligned.mats <- get.xz(tx.ad.short, ad.pr.mean)
#pull out the names of genes that made it through the alignment
final.names <- rownames(aligned.mats[[1]]) 
#also get the IDs
final.id <- gene.info[match(final.names, gene.info[,"external_gene_name"]), "ensembl_gene_id"]

effect.mat <- sapply(aligned.mats, function(x) x[,"effect"])
colnames(effect.mat) <- c("Transcript", "Protein")
p.mat <- sapply(aligned.mats, function(x) x[,"p"])

sig.idx <- apply(p.mat, 2, function(x) which(x <= 0.01))
sig.tx.names <- lapply(aligned.mats, rownames)
for(i in 1:length(sig.tx.names)){
    sig.flag = rep("", length(sig.tx.names[[i]]))
    sig.flag[sig.idx[[i]]] <- "**"
    sig.tx.names[[i]] <- paste0(sig.flag, sig.tx.names[[i]], sig.flag)
}

row.order <- order(plot.decomp(aligned.mats[[1]], plot.results = FALSE)$u[,1])
#row.order <- order(effect.mat[1,])
#row.order <- order(aligned.mats[[1]][,3] - aligned.mats[[2]][,1], decreasing = TRUE)
#row.order <- order(aligned.mats[[1]][,1])

global.min <- min(unlist(lapply(aligned.mats, function(x) x[,1:3])))
global.max <- max(unlist(lapply(aligned.mats, function(x) x[,1:3])))

png(file.path(results.dir, "AD_comparison.png"), width = 8, height = 6, units = "in", res = 300)
layout.mat <- matrix(1:4, nrow = 1)
layout(layout.mat, widths = c(0.5, 0.7,0.7,1))

par(mar = c(24, 4, 4, 4))
imageWithTextColorbar(Reduce("rbind", lapply(aligned.mats, function(x) x[,1:3])), 
    split.at.vals = TRUE, col.scale = list(heatmap.neg.col, heatmap.pos.col), 
    grad.dir = "ends", cex = 1, round.nearest = 0.1)
mtext("Mean Expression/Abundance (A.U.)", side = 2, cex = 0.7, line = 2.7)

par(mar = c(4,6,4,0))
imageWithText(aligned.mats[[1]][row.order,1:3], show.text = FALSE, split.at.vals = TRUE,
    col.scale = c(heatmap.neg.col, heatmap.pos.col), grad.dir = "ends",
    col.text.shift = 0.04, row.text.shift = 0.3, col.text.rotation = 0,
    col.text.adj = 0.5, global.color.scale = TRUE, global.min = global.min,
    global.max = global.max, row.names = sig.tx.names[[1]])
mtext("Transcript\nAbundance", side = 3, line = 0)

imageWithText(aligned.mats[[2]][row.order,1:3], show.text = FALSE, split.at.vals = TRUE,
    col.scale = c(heatmap.neg.col, heatmap.pos.col), grad.dir = "ends",
    col.text.shift = 0.05, row.text.shift = 0.3, col.text.rotation = 0, 
    col.text.adj = 0.5, global.color.scale = TRUE, global.min = global.min,
    global.max = global.max, row.names = sig.tx.names[[2]])
mtext("Protein\nAbundance", side = 3, line = 0)

par(mar = c(4,6,4,1))
barplot(t(effect.mat[rev(final.names),]), beside = TRUE, las = 2, horiz = TRUE, 
    main = "Effect size of genotype", col = c("gray20", "gray80"),
    xlab = bquote(beta))
plot.dim <- par("usr")
segments(x0 = 0, y0 = 0, y1 = plot.dim[4])
legend("bottomright", fill = c("gray20", "gray80"), legend = colnames(effect.mat), cex = 0.8)

dev.off()

#for the paper
global.min <- min(sapply(aligned.mats, function(x) min(x[,1:3])))
global.max <- max(sapply(aligned.mats, function(x) max(x[,1:3])))
row.order <- order(sapply(rownames(p.mat), function(x) which(rownames(tx.ad.short) == x)), decreasing = FALSE)
sig.text <- rep("", nrow(p.mat))
sig.text[which(p.mat[,2] <= 0.01)] <- "*"
png(here("Results", "for_paper", "AD-related_proteins.png"), width = 8, height = 3, units = "in", res = 300)
par(mar = c(6,4,4,2), bg = NA)
imageWithText(t(aligned.mats[[2]][row.order,1:3]), show.text = FALSE, split.at.vals = TRUE,
    col.scale = c(heatmap.neg.col, heatmap.pos.col), grad.dir = "ends",
    col.text.shift = 0.25, row.text.shift = 0.06, col.text.rotation = 90,
    col.text.adj = 1, row.text.adj = 0.5, global.color.scale = TRUE, 
    global.min = global.min, global.max = global.max, 
    col.names = str_to_upper(rownames(aligned.mats[[2]])[row.order]))
plot.dim <- par("usr")
for(i in 1:length(geno.cols)){
    draw.rectangle(min.x = plot.dim[1]-0.5, max.x = 0.4, min.y = length(geno.cols)-i+0.6, max.y = length(geno.cols)-i+1.4,
        border.col = geno.cols[i], lwd = 3)
}
#text(x = 1:length(short.ad.p), y = rep(plot.dim[4]+0.1, length(short.ad.p)), labels = sig.text[row.order])
dev.off()

#plot_pr_abund("App", centered.gene.mat, autoplace.text = TRUE)
#plot.with.model(as.numeric(aligned.mats[[1]]), as.numeric(aligned.mats[[2]]))
```

## Overall correlations

We wanted to look at whether overall higher transcript 
levels led to higher protein levels.

The plots below show that overall there is no correlation
between transcripts and protein levels. 

Greg said that this is expected in the following statement: 
"I think mass spec is highly variable in how well it detects 
certain peptides, and while RNA seq is better it still has PCR 
biases in amplification that works better for some genes than 
others. In other data sets, we've sometimes seen correlation coeffs 
of 0.15-0.30 for genotype effects (or case vs control), but almost 
no correlation for overall abundance."

```{r cor_test, fig.width = 10, fig.height = 5}
#use tx data that has been adjusted for technical sources of variation, but not biological sources
tx.dir <- here("Results", paste0("age_batch_", compare.age, "_gene"))
all.tx <- readRDS(file.path(tx.dir, "processed_data", "Batch_Adjusted_Expr.RDS"))
tx.covar <- read.csv(file.path(tx.dir, "processed_data", "covariates_carriers.csv"))
#adjust out all known sources of variation. We just want to test whether the transcript
#level is associated with protein level, not whether sex or genotype affects 
#either one. We don't need to adjust for age, because we are only looking in
#one age group at a time.
tx.adj.mat <- cbind(dummy_covar(data.frame("sex" = as.factor(tx.covar[,"sex_ge"]), 
    "geno" = as.factor(tx.covar[,"ordered_geno"]))),
    as.numeric(tx.covar[,"scaled_Ide_expression"]))
#adj.tx <- adjust(all.tx, tx.adj.mat, retain.intercept = FALSE) #adjust and mean-center

#don't adjust
adj.tx <- all.tx
#boxplot(adj.tx[,1:10])

#also adjust proteins for everything. Ide abundance has already been regressed
#out of centered.gene.mat
pr.adj.mat <- dummy_covar(data.frame("sex" =  mouse.info[,"sex"],
    "genotype" = mouse.info[,"genotype"]))

#using age.idx doesn't really matter because the transcript data are for a single age
#adj.pr <- adjust(centered.gene.mat[age.idx,], pr.adj.mat[age.idx,], retain.intercept = FALSE)

#don't adjust
adj.pr <- centered.gene.mat
#adj.pr <- mean.gene.mat #unfiltered for sex-biased genes
#boxplot(adj.pr[,1:10])

pr.id <- gene.info[match(colnames(adj.pr), gene.info[,"external_gene_name"]),"ensembl_gene_id"]
colnames(adj.pr) <- pr.id

common.id <- intersect(pr.id, colnames(adj.tx))
common.ind <- intersect(rownames(adj.pr), rownames(adj.tx))
aligned.pr <- adj.pr[common.ind, common.id]
aligned.tx <- adj.tx[common.ind, common.id]

f.id <- intersect(rownames(tx.covar)[which(tx.covar[,"sex_ge"] == "Female")], common.ind)
m.id <- intersect(rownames(tx.covar)[which(tx.covar[,"sex_ge"] == "Male")], common.ind)
test.cor <- sapply(1:ncol(aligned.pr), function(x) cor(aligned.pr[,x], aligned.tx[,x], 
    use = "pairwise.complete.obs"))
f.test.cor <- sapply(1:ncol(aligned.pr), function(x) cor(aligned.pr[f.id,x], aligned.tx[f.id,x], 
    use = "pairwise.complete.obs"))
m.test.cor <- sapply(1:ncol(aligned.pr), function(x) cor(aligned.pr[m.id,x], aligned.tx[m.id,x], 
    use = "pairwise.complete.obs"))

par(mfrow = c(1,2))
plot(density(test.cor), xlab = "Correlations", main = "Transcript-Protein Correlations", lwd = 3)
points(density(f.test.cor), col = factor.cols["F"], type = "l", lwd = 3)
points(density(m.test.cor), col = factor.cols["M"], type = "l", lwd = 3)
legend("topright", legend = c("all", "M", "F"), lty = 1, lwd = 3, 
    col = c("black", factor.cols[c("M", "F")]))
#abline(v = mean(test.cor), col = "black", lwd = 3)
#abline(v = mean(f.test.cor), col = factor.cols["F"], lwd = 3)
#abline(v = mean(m.test.cor), col = factor.cols["M"], lwd = 3)

abline(v = 0, col = "red")
boxplot(list("all" = test.cor, "F" = f.test.cor, "M" = m.test.cor), 
    ylab = "Correlation", main = "Transcript-Protein Correlations")
abline(h = 0)
mtext(paste(compare.age, "months"), side = 3, outer = TRUE, line = -1.5, font = 2, cex = 1.5)
```

```{r more_comparison, eval = FALSE}
n.pc <- 10
tx.decomp <- plot.decomp(aligned.tx, pc = n.pc, plot.results = FALSE)
tx.pc <- tx.decomp$u
imp.pr <- impute.knn(aligned.pr)$data
pr.decomp <- plot.decomp(imp.pr, pc = n.pc, plot.results = FALSE)
pr.pc <- pr.decomp$u

pc.pairs <- pair.matrix(1:n.pc)
tx.to.pr <- pr.to.tx <- rep(NA, nrow(pc.pairs))
for(i in 1:nrow(pc.pairs)){
    pc1 <- pc.pairs[i,1]
    pc2 <- pc.pairs[i,2]
    tx.to.pr[i] <- cor(tx.pc[,pc1], pr.pc[,pc2])
    pr.to.tx[i] <- cor(pr.pc[,pc1], tx.pc[,pc2])
}
final.result <- cbind(pc.pairs, tx.to.pr, pr.to.tx)
hist(c(tx.to.pr, pr.to.tx), breaks = 15)
boxplot(list(tx.to.pr, pr.to.tx))
top1 <- which.max(abs(tx.to.pr))
top2 <- which.max(abs(pr.to.tx))

final.result[top1,]
final.result[top2,]

tx.pc1 <- final.result[top1,1]
pr.pc1 <- final.result[top1,2]
par(mfrow = c(1,2))
plot.with.model(tx.pc[,tx.pc1], pr.pc[,pr.pc1], report = "cor.test",
    col = as.numeric(covar.table[rownames(aligned.tx),"sex_ge"]),
    xlab = paste("Transcript PC", tx.pc1), ylab = paste("Protein PC", pr.pc1))
plot.with.model(tx.pc[,tx.pc1], pr.pc[,pr.pc1], report = "cor.test",
    col = as.numeric(covar.table[rownames(aligned.tx),"ordered_geno"]),
    xlab = paste("Transcript PC", tx.pc1), ylab = paste("Protein PC", pr.pc1))


tx.pc2 <- final.result[top2,2]
pr.pc2 <- final.result[top2,1]
par(mfrow = c(1,2))
plot.with.model(tx.pc[,tx.pc2], pr.pc[,pr.pc2], report = "cor.test",
    col = as.numeric(covar.table[rownames(aligned.tx),"sex_ge"]))
plot.with.model(tx.pc[,tx.pc2], pr.pc[,pr.pc2], report = "cor.test",
    col = as.numeric(covar.table[rownames(aligned.tx),"ordered_geno"]))

par(mfrow = c(1,2))
pr.models <- apply(pr.pc, 2, function(x) lm(x~covar.table[rownames(aligned.tx),"ordered_geno"]))
pr.effect <- sapply(pr.models, function(x) coef(x)[2])
barplot(pr.effect)

tx.models <- apply(tx.pc, 2, function(x) lm(x~covar.table[rownames(aligned.tx),"ordered_geno"]))
tx.effect <- sapply(tx.models, function(x) coef(x)[2])
barplot(tx.effect)

```

We looked at the enrichments of genes with strong positive
and strong negative correlations between transcript and protein.

The high-correlation pairs are enriched for organelle-associated
genes, and translation genes.

The low-correlation pairs are enriched for synaptic genes. 
This enrichment was predicted by pmid 30664243, citing the 
large distance between the cell body and synaptic regions
of neurons. 

The enrichment tables are shown below.

```{r enrich_cor, fig.widtdh = 7, fig.height = 7}
#check.cor <- m.test.cor
check.cor <- test.cor

high.thresh <- get.percentile(check.cor, 99)
low.thresh <- get.percentile(check.cor, 1)
hist(check.cor, breaks = 100)
abline(v = c(high.thresh, low.thresh))
high.cor <- colnames(aligned.pr)[which(check.cor >= high.thresh)]
low.cor <- colnames(aligned.pr)[which(check.cor <= low.thresh)]

if(!offline){
    high.enrich <- gost(high.cor, organism = "mmusculus", source = "GO")
    low.enrich <- gost(low.cor, organism = "mmusculus", source = "GO")

    plot.enrichment(high.enrich, plot.label = "High Transcript-Protein Correlations",
        num.terms = 25)
    plot.enrichment(low.enrich, plot.label = "Low Transcript-Protein Correlations",
        num.terms = 25)
}

```

For genes that are very highly correlated with sex, we do see
positive correlations between transcript and protein. However,
once you get down below 40\% variance explained by sex, that
correlation disappears.

```{r high_cor_example, fig.width = 5, fig.height = 5}
sex.col <- as.numeric(as.factor(tx.covar[common.ind,"sex_ge"]))
gene.name <- names(sort(factor.effect.mat[,"sex"], decreasing = TRUE))[2]
#gene.name <- "Tppp"
#get_ortho(gene.name)
gene.id <- gene.info[which(gene.info[,"external_gene_name"] == gene.name), "ensembl_gene_id"]

#pdf("~/Desktop/sex_correlation.pdf", height = 5, width = 5)
plot.with.model(aligned.tx[common.ind,gene.id], aligned.pr[common.ind,gene.id], 
    report = "cor.test", main = gene.name, xlab = "Transript level", 
    ylab = "Protein level", col = sex.col)
#dev.off()
stop()
```

```{r sex_specific_QC, eval = FALSE}
#There are two Y chr genes expressed in the brain: Eif2s3y and Ddx3y
#here we check for sex-specific expression
gene.name <- "Eif2s3y"
gene.name <- "Eif2s3x"
gene.name = "Ddx3y"
gene.name = "Ddx3x"

gene.id <- gene.info[which(gene.info[,"external_gene_name"] == gene.name), "ensembl_gene_id"]

par(mfrow = c(1,2))
boxplot(aligned.tx[common.ind,gene.id]~covar.table[common.ind,"sex_ge"], 
    xlab = "", ylab = "Transcript Abundance", main = "Transcript")
boxplot(aligned.pr[common.ind,gene.id]~covar.table[common.ind,"sex_ge"], 
    xlab = "", ylab = "Protein Abundance", main = "Protein")
mtext(gene.name, side = 3, outer = TRUE, font = 2, cex = 1.2, line = -1.5)
```

## Correlation quality control

We are a little suspicious that our protein data might not
be very accurate, so we are conducting some tests. Greg said
that complement transcripts and proteins have been very highly
correlated in their past data sets. Here we look to see if this
is true here. Nope.


```{r check_individ, fig.width = 12, fig.height = 9}

complement.idx <- grep("complement ", gene.features[,"gene_desc"], ignore.case = TRUE)
check.name <- gene.features[complement.idx,gene.col]
comp.cor <- matrix(NA, nrow = length(check.name), ncol = 2)
colnames(comp.cor) <- c("r", "p")
rownames(comp.cor) <- check.name
#pdf("~/Desktop/complement_report.pdf", width = 12, height = 9)
par(mfrow = c(3,4))
for(i in 1:length(check.name)){
    gene.name <- check.name[i]
    gene.id <- gene.info[which(gene.info[,"external_gene_name"] == gene.name), "ensembl_gene_id"]
    if(is.na(gene.id) || length(gene.id) == 0){next()}
    in_data <- Reduce("intersect", list(gene.id, colnames(aligned.tx), colnames(aligned.pr)))
    if(length(in_data) == 0){next()}

    comp.cor[i,] <- plot.with.model(aligned.tx[,gene.id], aligned.pr[,gene.id], report = "cor.test", 
        main = gene.name, xlab = "Transript level", ylab = "Protein level")
    }
boxplot(comp.cor[,"r"], ylab = "Correlation (r)", main = "All correlatoins")
stripchart(comp.cor[,"r"], vertical = TRUE, pch = 16, method = "jitter", 
    add = TRUE, cex = 2)
abline(h = 0)

#dev.off()
```

## Comparison to Matrisome Module 42

Previous proteomic work in the lab has shown that the protein
signatures in the mouse models match the human condition better
than the transcriptomics. That is, proteins that are more abundant
in AD patients relative to controls are also more abundant in 
mouse models with introduced risk variants. 

Greg said that they see the largest signal in a module of genes
related to the "matrisome" module from pmid: 35115731. He sent
me the genes in this module along with some stats for each.

As a measure of quality control we wanted to check how the
Klotho proteomics stack up to that, assuming the VS allele is 
the protective allele. We have set up our models so that the 
order of the genotypes goes FC < WT < VS, so proteins with 
positive effects are up in the *protective* situation, not the risk 
situation. So we would like to see that the proteins in this module
are less abundant in the VS animals, i.e. have negative effects in
the linear model.

The figure below shows the overall effect for the whole module.
It is not different overall across genotypes in the 12-month old
animals.


```{r orthos, fig.width = 9, fig.height = 6.5}
get_ortho <- function(gene.name, target = "mouse"){
    gene.idx <- unlist(apply(orthos, 2, function(x) which(x == gene.name)))
    if(length(gene.idx) == 0){
        return(paste("Couldn't find", gene.name))
    }else{
        gene.line <- orthos[gene.idx,]
    }

    if(target == "mouse"){
        result <- gene.line[,c("Mouse.Ortholog.Name", "Mouse.Ortholog.Ensembl")]
    }else{
        result <- gene.line[,c("Human.Gene.Name", "Human.Ensembl")]
    }
    names(result) <- c("name", "id")
    return(return(result))
}

#get_ortho("PLOD3")

mod42 <- read_excel(here("Data", "general", "TreatAD_scores_M42_module.xlsx")) #matrisome module
mod42.genes <- mod42$GeneName
mod42.mouse <- t(sapply(mod42.genes, get_ortho))

#pdf("~/Desktop/Matrisome.pdf", width = 9, height = 6.5)
mod.result <- plot_pr_abund(mod42.mouse[,"name"], centered.gene.mat, 
    autoplace.text = TRUE, plot.label = "Module 42", plot.label.cex = 2)
```

The plot below shows the results for the individual proteins in
the module. This is a volcano plot showing the effect size of 
genotype at 12 months on each protein against the -log10(p) of
the effect. Proteins with -log10(p) greater than 1 are labeled.

None of these effects are significant, however, the general
trend of the effect is in the right direction. All of these
proteins were more abundant in human AD patients. Here, they
are less abundant in VS animals, which is consistent with their
label of being protective. A few of the proteins, including Flt1
and Bdh2 were more highly abundant in the VS animals. This is 
not consistent with the human data.

```{r mod_ind, fig.width = 5, fig.height = 5}
included.proteins <- colnames(mod.result$expr_mat)
mod.gene.results <- vector(mode = "list", length = length(included.proteins))
for(i in 1:length(included.proteins)){
    mod.gene.results[[i]] <- plot_pr_abund(included.proteins[i], centered.gene.mat, 
        plot.results = FALSE)
}
#dev.off()

twelve.month.effect <- sapply(mod.gene.results, function(x) as.numeric(x$stats["geno_twelve", "linear.effect.size"]))
twelve.month.p <- sapply(mod.gene.results, function(x) as.numeric(x$stats["geno_twelve", "linear.effect.p"]))
#pdf("~/Desktop/Module42_Effect.pdf")
plot(twelve.month.effect, -log10(twelve.month.p), xlim = c(-0.6, 0.6),
    main = "Twelve-month effects on Module 42 proteins")
pretty.big <- which(-log10(twelve.month.p) > 1)
text(twelve.month.effect[pretty.big], -log10(twelve.month.p)[pretty.big], 
    labels = included.proteins[pretty.big], pos = 4)
#dev.off()
```